---
title: "Statistics"
subtitle: "Chapter 12: Analysis of Variance"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format:
  revealjs:
    html-math-method: katex
smaller: true
theme: serif
footer: "Elementary Statistics: A Step by Step Approach"
logo: "logo.svg"
editor: source
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch12-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
```

# Chapter 12: Analysis of Variance {background-color="#2c4a6e"}

## Chapter Overview

::: {.callout-note}
## Why ANOVA?
When comparing **three or more group means**, using multiple t-tests is problematic:

- Each additional comparison inflates the Type I error rate
- With $k$ groups, you need $\binom{k}{2}$ pairwise tests
- The **F test** compares all means simultaneously in a single test
:::

**Topics covered:**

- **Section 12-1:** One-Way Analysis of Variance (ANOVA)
- **Section 12-2:** Post-hoc Tests — Scheffé and Tukey
- **Section 12-3:** Two-Way Analysis of Variance

---

## Historical Note

::: {.callout-note}
## Origin of ANOVA
The methods of analysis of variance were developed by **R. A. Fisher** in the early 1920s.

ANOVA uses **variances** to make inferences about **means** — hence the name "Analysis of Variance."
:::

---

# Section 12-1: One-Way Analysis of Variance {background-color="#2c4a6e"}

## Why Not Multiple t-Tests?

::: {.callout-important}
## Three Reasons to Avoid Multiple t-Tests
1. **Ignores other groups:** Each pairwise comparison ignores the remaining groups
2. **Inflated Type I error:** More comparisons = higher chance of false significance
3. **Explosion of tests:** 3 means → 3 tests; 5 means → 10 tests; 10 means → 45 tests
:::

The **F test** overcomes all three issues by comparing all means at once.

---

## Assumptions for One-Way ANOVA

::: {.callout-important}
## Assumptions
1. Each population must be **normally or approximately normally distributed**
2. All samples must be **independent** of one another
3. The **variances of all populations** must be equal (homogeneity of variance)
:::

---

## The Logic of ANOVA

::: {.callout-note}
## Two Variance Estimates
| Estimate | What it measures |
|----------|-----------------|
| **Between-group variance** ($s_B^2$) | Variation among sample means |
| **Within-group variance** ($s_W^2$) | Variation within each group (pooled) |

- If $H_0$ is **true**: $s_B^2 \approx s_W^2$, so $F \approx 1$
- If $H_0$ is **false**: $s_B^2 \gg s_W^2$, so $F \gg 1$
:::

$$F = \frac{s_B^2}{s_W^2}$$

The F test for comparing means is always **right-tailed**.

---

## Hypotheses and Degrees of Freedom

::: {.callout-note}
## ANOVA Hypotheses
$$H_0: \mu_1 = \mu_2 = \cdots = \mu_k$$
$$H_1: \text{At least one mean differs from the others}$$
:::

::: {.callout-note}
## Degrees of Freedom
$$d.f.N. = k - 1 \qquad d.f.D. = N - k$$

where $k$ = number of groups and $N = n_1 + n_2 + \cdots + n_k$
:::

---

## ANOVA Computational Formulas

::: {.callout-note}
## Step-by-Step Procedure
**Step 1:** Find the mean and variance of each sample: $(\bar{X}_1, s_1^2), \ldots, (\bar{X}_k, s_k^2)$

**Step 2:** Grand mean: $\bar{X}_{GM} = \dfrac{\sum X}{N}$

**Step 3:** Between-group variance: $s_B^2 = \dfrac{\sum n_i(\bar{X}_i - \bar{X}_{GM})^2}{k-1}$

**Step 4:** Within-group variance: $s_W^2 = \dfrac{\sum (n_i - 1)s_i^2}{\sum (n_i - 1)}$

**Step 5:** $F = \dfrac{s_B^2}{s_W^2}$
:::

---

## ANOVA Summary Table

::: {.callout-note}
## Standard Summary Table Format
| Source | SS | d.f. | MS | F |
|--------|-----|------|-----|---|
| Between (Treatment) | SSB | $k-1$ | MSB = SSB/$(k-1)$ | MSB/MSW |
| Within (Error) | SSW | $N-k$ | MSW = SSW/$(N-k)$ | |
| Total | SST | $N-1$ | | |

Where:
- $SSB = \sum n_i(\bar{X}_i - \bar{X}_{GM})^2$
- $SSW = \sum (n_i - 1)s_i^2$
:::

---

## Example 12-1: Lowering Blood Pressure (Part 1)

::: {.callout-tip}
## Problem
A researcher tests three techniques to reduce blood pressure: **medication**, **exercise**, and **diet**. After 4 weeks, blood pressure reductions are recorded. At $\alpha = 0.05$, test the claim that there is no difference among the means.

| Medication | Exercise | Diet |
|-----------|---------|------|
| 10 | 6 | 5 |
| 12 | 8 | 9 |
| 9 | 3 | 12 |
| 15 | 0 | 8 |
| 13 | 2 | 4 |
| $\bar{X}_1=11.8$ | $\bar{X}_2=3.8$ | $\bar{X}_3=7.6$ |
| $s_1^2=5.7$ | $s_2^2=10.2$ | $s_3^2=10.3$ |
:::

---

## Example 12-1: Mathematical Solution

::: {.callout-tip}
## Step-by-Step Solution

**Step 1:** $H_0: \mu_1 = \mu_2 = \mu_3$ (claim); $H_1:$ At least one mean differs

**Step 2:** $k=3$, $N=15$; $d.f.N.=2$, $d.f.D.=12$ → Critical value: **3.89**

**Step 3a:** Means and variances given above

**Step 3b:** Grand mean: $\bar{X}_{GM} = \dfrac{116}{15} = 7.73$

**Step 3c:** Between-group variance:
$$s_B^2 = \frac{5(11.8-7.73)^2 + 5(3.8-7.73)^2 + 5(7.6-7.73)^2}{3-1} = \frac{160.13}{2} = 80.07$$

**Step 3d:** Within-group variance:
$$s_W^2 = \frac{4(5.7) + 4(10.2) + 4(10.3)}{4+4+4} = \frac{104.80}{12} = 8.73$$

**Step 3e:** $F = \dfrac{80.07}{8.73} = 9.17$

**Step 4:** $9.17 > 3.89$ → **Reject** $H_0$

**Step 5:** There is sufficient evidence that at least one mean differs from the others.
:::

---

## Example 12-1: ANOVA Summary Table

| Source | SS | d.f. | MS | F |
|--------|-----|------|-----|---|
| Between | 160.13 | 2 | 80.07 | **9.17** |
| Within (Error) | 104.80 | 12 | 8.73 | |
| **Total** | **264.93** | **14** | | |

Critical value at $\alpha=0.05$: **3.89**

Since $F = 9.17 > 3.89$, we **reject** $H_0$.

---

## Example 12-1: R Solution

```{r ex12-1, echo=T, warning=FALSE}
medication <- c(10, 12, 9, 15, 13)
exercise   <- c(6, 8, 3, 0, 2)
diet       <- c(5, 9, 12, 8, 4)

bp_data <- data.frame(
  reduction = c(medication, exercise, diet),
  group = rep(c("Medication","Exercise","Diet"), each = 5)
)

result <- aov(reduction ~ group, data = bp_data)
summary(result)
```

---

## Example 12-1: Visualization

```{r ex12-1-plot, echo=T, warning=FALSE}
ggplot(bp_data, aes(x = group, y = reduction, fill = group)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, size = 2) +
  labs(title = "Blood Pressure Reduction by Treatment",
       x = "Treatment Group", y = "Reduction (mmHg)") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## Example 12-2: Employees at Toll Road Interchanges (Part 1)

::: {.callout-tip}
## Problem
A state official wants to determine if there is a significant difference in the number of employees at interchanges across three Pennsylvania toll roads. At $\alpha = 0.05$, test the claim.

| PA Turnpike | Greensburg Bypass | Beaver Valley |
|------------|-------------------|---------------|
| 7 | 10 | 1 |
| 14 | 1 | 12 |
| 32 | 1 | 1 |
| 19 | 0 | 9 |
| 10 | 11 | 1 |
| 11 | 1 | 11 |
| $\bar{X}_1=15.5$ | $\bar{X}_2=4.0$ | $\bar{X}_3=5.8$ |
| $s_1^2=81.9$ | $s_2^2=25.6$ | $s_3^2=29.0$ |
:::

---

## Example 12-2: Mathematical Solution

::: {.callout-tip}
## Solution

**Step 1:** $H_0: \mu_1 = \mu_2 = \mu_3$; $H_1:$ At least one mean differs (claim)

**Step 2:** $k=3$, $N=18$; $d.f.N.=2$, $d.f.D.=15$ → Critical value: **3.68**

**Step 3b:** $\bar{X}_{GM} = \dfrac{152}{18} = 8.4$

**Step 3c:** $s_B^2 = \dfrac{6(15.5-8.4)^2 + 6(4.0-8.4)^2 + 6(5.8-8.4)^2}{2} = \dfrac{459.18}{2} = 229.59$

**Step 3d:** $s_W^2 = \dfrac{5(81.9) + 5(25.6) + 5(29.0)}{15} = \dfrac{682.5}{15} = 45.5$

**Step 3e:** $F = \dfrac{229.59}{45.5} = 5.05$

**Step 4:** $5.05 > 3.68$ → **Reject** $H_0$

**Step 5:** Sufficient evidence that at least one interchange mean differs from the others.
:::

---

## Example 12-2: ANOVA Summary Table

| Source | SS | d.f. | MS | F |
|--------|-----|------|-----|---|
| Between | 459.18 | 2 | 229.59 | **5.05** |
| Within (Error) | 682.50 | 15 | 45.50 | |
| **Total** | **1141.68** | **17** | | |

**P-value:** $0.01 < P < 0.025$ (from F-tables); calculator: $P = 0.021$

Since $F = 5.05 > 3.68$, we **reject** $H_0$.

---

## Example 12-2: R Solution

```{r ex12-2, echo=T, warning=FALSE}
pa_turnpike  <- c(7, 14, 32, 19, 10, 11)
greensburg   <- c(10, 1, 1, 0, 11, 1)
beaver       <- c(1, 12, 1, 9, 1, 11)

emp_data <- data.frame(
  employees = c(pa_turnpike, greensburg, beaver),
  road = rep(c("PA Turnpike","Greensburg","Beaver Valley"), each = 6)
)

result2 <- aov(employees ~ road, data = emp_data)
summary(result2)
```

---

## Example 12-2: Visualization

```{r ex12-2-plot, echo=T, warning=FALSE}
ggplot(emp_data, aes(x = road, y = employees, fill = road)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, size = 2) +
  labs(title = "Number of Employees at Toll Road Interchanges",
       x = "Toll Road", y = "Number of Employees") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# Section 12-2: The Scheffé Test and the Tukey Test {background-color="#2c4a6e"}

## Post-Hoc Tests: Overview

::: {.callout-note}
## Why Post-Hoc Tests?
When the F test rejects $H_0$, we know **at least one mean differs**, but not **which pair(s)** differ.

Post-hoc tests make all pairwise comparisons after ANOVA:

| Test | Best Use |
|------|---------|
| **Scheffé** | Unequal sample sizes; most conservative |
| **Tukey** | Equal sample sizes; more powerful for pairwise comparisons |

**General rule:** Use Tukey when $n_i = n_j$; use Scheffé when sizes differ.
:::

---

## The Scheffé Test

::: {.callout-note}
## Scheffé Test Formula
$$F_S = \frac{(\bar{X}_i - \bar{X}_j)^2}{s_W^2\left(\dfrac{1}{n_i} + \dfrac{1}{n_j}\right)}$$

**Critical value for Scheffé:**
$$F' = (k-1) \times C.V._{\text{ANOVA}}$$

Reject $H_0$ for that pair if $F_S > F'$
:::

---

## Example 12-3: Scheffé Test for Blood Pressure Data

::: {.callout-tip}
## Problem
Using the Scheffé test, determine which pairs of means in Example 12-1 differ significantly at $\alpha = 0.05$.

Recall: $\bar{X}_1=11.8$, $\bar{X}_2=3.8$, $\bar{X}_3=7.6$, $s_W^2=8.73$, $n_i=5$ for all groups.
:::

---

## Example 12-3: Mathematical Solution

::: {.callout-tip}
## Scheffé Test Calculations

**Critical value:** $F' = (k-1)(C.V.) = (3-1)(3.89) = 7.78$

**a.** $\bar{X}_1$ vs $\bar{X}_2$:
$$F_S = \frac{(11.8 - 3.8)^2}{8.73(1/5 + 1/5)} = \frac{64}{3.492} = 18.33 \quad \checkmark > 7.78$$

**b.** $\bar{X}_2$ vs $\bar{X}_3$:
$$F_S = \frac{(3.8 - 7.6)^2}{8.73(1/5 + 1/5)} = \frac{14.44}{3.492} = 4.14 \quad \text{Not significant}$$

**c.** $\bar{X}_1$ vs $\bar{X}_3$:
$$F_S = \frac{(11.8 - 7.6)^2}{8.73(1/5 + 1/5)} = \frac{17.64}{3.492} = 5.05 \quad \text{Not significant}$$

**Conclusion:** Only **Medication vs. Exercise** ($\bar{X}_1$ vs $\bar{X}_2$) is significantly different.
:::

---

## Example 12-3: R Solution

```{r ex12-3, echo=T, warning=FALSE}
# Scheffé test using agricolae package (or manual calculation)
# Manual Scheffe test
scheffe_fs <- function(x1_bar, x2_bar, sw2, n1, n2) {
  (x1_bar - x2_bar)^2 / (sw2 * (1/n1 + 1/n2))
}

sw2 <- 8.73
cv_scheffe <- (3 - 1) * 3.89  # F' = 7.78

cat("Critical value F' =", cv_scheffe, "\n\n")

pairs <- list(
  c("Medication vs Exercise", 11.8, 3.8),
  c("Exercise vs Diet",       3.8,  7.6),
  c("Medication vs Diet",     11.8, 7.6)
)

for (p in pairs) {
  fs <- scheffe_fs(as.numeric(p[2]), as.numeric(p[3]), sw2, 5, 5)
  sig <- ifelse(fs > cv_scheffe, "SIGNIFICANT", "not significant")
  cat(p[1], ": Fs =", round(fs, 2), "->", sig, "\n")
}
```

---

## The Tukey Test

::: {.callout-note}
## Tukey Test Formula
Used when **all sample sizes are equal**:

$$q = \frac{\bar{X}_i - \bar{X}_j}{\sqrt{s_W^2 / n}}$$

Compare $|q|$ to the critical value from **Tukey's table** (Table N):

- Rows: $\nu$ = degrees of freedom for $s_W^2$ (= $N - k$)
- Columns: $k$ = number of groups

Reject $H_0$ for that pair if $|q| >$ critical value.
:::

---

## Example 12-4: Tukey Test for Blood Pressure Data

::: {.callout-tip}
## Problem
Using the Tukey test, determine which pairs of means in Example 12-1 differ significantly at $\alpha = 0.05$.

$\bar{X}_1=11.8$, $\bar{X}_2=3.8$, $\bar{X}_3=7.6$, $s_W^2=8.73$, $n=5$
:::

---

## Example 12-4: Mathematical Solution

::: {.callout-tip}
## Tukey Test Calculations

$\sqrt{s_W^2/n} = \sqrt{8.73/5} = \sqrt{1.746} = 1.321$

**Critical value:** $k=3$, $d.f.=12$, $\alpha=0.05$ → $q_{crit} = 3.77$ (from Tukey table)

**a.** $\bar{X}_1$ vs $\bar{X}_2$:
$$q = \frac{11.8 - 3.8}{1.321} = \frac{8}{1.321} = 6.06 \quad \checkmark > 3.77$$

**b.** $\bar{X}_1$ vs $\bar{X}_3$:
$$q = \frac{11.8 - 7.6}{1.321} = \frac{4.2}{1.321} = 3.18 \quad \text{Not significant}$$

**c.** $\bar{X}_2$ vs $\bar{X}_3$:
$$q = \frac{|3.8 - 7.6|}{1.321} = \frac{3.8}{1.321} = 2.88 \quad \text{Not significant}$$

**Conclusion:** Only **Medication vs. Exercise** is significantly different — agrees with Scheffé.
:::

---

## Example 12-4: R Solution

```{r ex12-4, echo=T, warning=FALSE}
# Tukey HSD test using R's built-in function
medication <- c(10, 12, 9, 15, 13)
exercise   <- c(6, 8, 3, 0, 2)
diet       <- c(5, 9, 12, 8, 4)

bp_data <- data.frame(
  reduction = c(medication, exercise, diet),
  group = factor(rep(c("Medication","Exercise","Diet"), each = 5))
)

bp_aov <- aov(reduction ~ group, data = bp_data)
TukeyHSD(bp_aov)
```

---

## Example 12-4: Visualization

```{r ex12-4-plot, echo=T, warning=FALSE}
tukey_result <- TukeyHSD(bp_aov)
plot(tukey_result, las = 1)
```

---

## Scheffé vs. Tukey: Comparison

::: {.callout-note}
## When to Use Each Test

| Feature | Scheffé | Tukey |
|---------|---------|-------|
| Sample sizes | Unequal or equal | Must be equal |
| Power | Lower (conservative) | Higher for pairwise |
| Types of comparisons | Pairwise and combinations | Pairwise only |
| Recommendation | Unequal $n_i$ | Equal $n_i$ |

**Key insight:** When the F test rejects $H_0$, post-hoc tests may occasionally find no significant pairs — the difference may lie in a combination of means rather than any single pair.
:::

---

# Section 12-3: Two-Way Analysis of Variance {background-color="#2c4a6e"}

## Introduction to Two-Way ANOVA

::: {.callout-note}
## Extension of One-Way ANOVA
**One-way ANOVA:** One independent variable (factor)

**Two-way ANOVA:** Two independent variables (factors)

Benefits of two-way ANOVA:
- Tests **two factors simultaneously** in a single experiment
- Tests the **interaction effect** between the two factors
- More efficient than two separate one-way ANOVAs
:::

---

## Two-Way ANOVA: Design Structure

::: {.callout-note}
## Factorial Design Terminology
- **Factor A** has $a$ levels
- **Factor B** has $b$ levels
- **$n$** subjects per cell (must be equal)
- **Total subjects** = $abn$

A **2×2 design** has 2 levels for each factor → 4 treatment groups.

Example: Plant food type (A1, A2) × Soil type (I, II) → 4 groups
:::

---

## Two-Way ANOVA: Three Hypotheses

::: {.callout-note}
## Hypotheses
**1. Interaction effect:**
$$H_0: \text{No interaction between factors A and B}$$
$$H_1: \text{There is an interaction between A and B}$$

**2. Main effect of Factor A:**
$$H_0: \text{No difference in means across levels of A}$$
$$H_1: \text{At least one mean of A differs}$$

**3. Main effect of Factor B:**
$$H_0: \text{No difference in means across levels of B}$$
$$H_1: \text{At least one mean of B differs}$$
:::

---

## Two-Way ANOVA: Assumptions

::: {.callout-important}
## Assumptions
1. Populations must be **normally or approximately normally distributed**
2. All samples must be **independent**
3. **Population variances** must be equal across groups
4. **Group sizes must be equal** ($n$ per cell is the same for all cells)
:::

---

## Two-Way ANOVA: Summary Table

::: {.callout-note}
## General ANOVA Table
| Source | SS | d.f. | MS | F |
|--------|-----|------|-----|---|
| Factor A | SSA | $a-1$ | MSA | $F_A = MSA/MSW$ |
| Factor B | SSB | $b-1$ | MSB | $F_B = MSB/MSW$ |
| Interaction A×B | SS(A×B) | $(a-1)(b-1)$ | MS(A×B) | $F_{A \times B} = MS(A \times B)/MSW$ |
| Within (Error) | SSW | $ab(n-1)$ | MSW | |
| Total | SST | $abn-1$ | | |
:::

---

## Example 12-5: Gasoline Consumption (Part 1)

::: {.callout-tip}
## Problem
A researcher investigates whether **type of gasoline** (regular vs. high-octane) and **type of automobile** (two-wheel vs. four-wheel drive) affect fuel economy (mpg). Two cars are tested in each of the four combinations.

|  | Two-wheel-drive | Four-wheel-drive |
|--|----------------|-----------------|
| **Regular** | 26.7, 25.2 | 28.6, 29.3 |
| **High-octane** | 32.3, 32.8 | 26.1, 24.2 |

Test at $\alpha = 0.05$.
:::

---

## Example 12-5: Setting Up the Test

::: {.callout-tip}
## Step 1: Hypotheses

**Interaction (A×B):**
$H_0$: No interaction between gasoline type and automobile type
$H_1$: There is an interaction effect

**Factor A (Gasoline):**
$H_0$: No difference in means for gasoline types
$H_1$: Means differ by gasoline type

**Factor B (Automobile):**
$H_0$: No difference in means for automobile types
$H_1$: Means differ by automobile type

**Step 2:** $a=2$, $b=2$, $n=2$; all F critical values use $d.f.N.=1$, $d.f.D.=ab(n-1)=4$ → $F_{crit} = 7.71$
:::

---

## Example 12-5: Completed ANOVA Table

::: {.callout-tip}
## Step 3: Computing Mean Squares and F Values

| Source | SS | d.f. | MS | F |
|--------|-----|------|-----|---|
| Gasoline A | 3.920 | 1 | 3.920 | 4.752 |
| Automobile B | 9.680 | 1 | 9.680 | **11.733** |
| Interaction A×B | 54.080 | 1 | 54.080 | **65.552** |
| Within (Error) | 3.300 | 4 | 0.825 | |
| **Total** | **70.980** | **7** | | |

Critical value for all tests: **7.71**
:::

---

## Example 12-5: Decisions and Interpretation

::: {.callout-tip}
## Steps 4 & 5

**Gasoline A:** $F_A = 4.752 < 7.71$ → **Fail to reject** $H_0$

**Automobile B:** $F_B = 11.733 > 7.71$ → **Reject** $H_0$

**Interaction A×B:** $F_{A \times B} = 65.552 > 7.71$ → **Reject** $H_0$

**Important:** Since the interaction is **significant**, the main effect of automobile type should **not** be interpreted independently — it depends on which gasoline type is used.

**Conclusion:** The combination of gasoline type and automobile type significantly affects fuel economy.
:::

---

## Example 12-5: Cell Means and Interaction Plot

Cell means for visualization:

|  | Two-wheel | Four-wheel |
|--|-----------|-----------|
| **Regular** | 25.95 | 28.95 |
| **High-octane** | 32.55 | 25.15 |

```{r ex12-5-plot, echo=T, warning=FALSE}
gas_data <- data.frame(
  mpg  = c(26.7, 25.2, 28.6, 29.3, 32.3, 32.8, 26.1, 24.2),
  gas  = rep(c("Regular","Regular","Regular","Regular",
               "High-octane","High-octane","High-octane","High-octane")),
  auto = rep(c("Two-wheel","Two-wheel","Four-wheel","Four-wheel"), 2)
)

cell_means <- gas_data %>%
  group_by(gas, auto) %>%
  summarise(mean_mpg = mean(mpg), .groups = "drop")

ggplot(cell_means, aes(x = gas, y = mean_mpg, color = auto, group = auto)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Interaction Plot: Gasoline × Automobile Type",
       x = "Gasoline Type", y = "Mean MPG", color = "Auto Type") +
  theme_minimal()
```

---

## Example 12-5: R Solution

```{r ex12-5-aov, echo=T, warning=FALSE}
gas_data2 <- data.frame(
  mpg  = c(26.7, 25.2, 28.6, 29.3, 32.3, 32.8, 26.1, 24.2),
  gas  = factor(rep(c("Regular","High-octane"), each = 4)),
  auto = factor(rep(c("Two-wheel","Two-wheel","Four-wheel","Four-wheel"), 2))
)

result5 <- aov(mpg ~ gas * auto, data = gas_data2)
summary(result5)
```

---

## Interpreting Interaction Effects

::: {.callout-note}
## Types of Interactions
| Graph Pattern | Interaction Type | Interpretation |
|--------------|-----------------|----------------|
| Lines are **parallel** | No interaction | Interpret main effects independently |
| Lines **cross each other** | **Disordinal** interaction | Do NOT interpret main effects independently |
| Lines diverge but **do not cross** | Ordinal interaction | Main effects may still be interpreted |

**Rule:** When the interaction F test is significant:
- Check the interaction plot
- If lines cross → **disordinal** interaction → interpret only with interaction in mind
- If lines don't cross → **ordinal** interaction → main effects can be interpreted
:::

---

## Chapter 12 Summary

::: {.callout-note}
## Key Takeaways

**Section 12-1 — One-Way ANOVA:**
- F test compares 3+ means simultaneously
- $F = s_B^2 / s_W^2$; reject $H_0$ if $F > F_{crit}$ (right-tailed)
- Degrees of freedom: $d.f.N. = k-1$, $d.f.D. = N-k$

**Section 12-2 — Post-Hoc Tests:**
- Scheffé: flexible, works with unequal $n$; more conservative
- Tukey: more powerful for pairwise comparisons with equal $n$

**Section 12-3 — Two-Way ANOVA:**
- Tests two factors and their interaction simultaneously
- If interaction is significant, do not interpret main effects independently
- Interaction plots reveal the nature of the relationship
:::

## Acknowledgements

-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2011). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.

