---
title: "Statistics"
subtitle: "Chapter 14: Sampling and Simulation"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch14-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
set.seed(42)
```

## Overview
::: {.callout-note}
## Why Sampling and Simulation?
When populations are too large to study completely, researchers rely on **samples** to draw valid inferences. And when real-world experiments are too dangerous, expensive, or time-consuming, **simulation** provides a safe, cost-effective alternative.

Random numbers are the engine behind both techniques.
:::

**Topics covered:**

- **14-1:** Common Sampling Techniques
- **14-2:** Surveys and Questionnaire Design
- **14-3:** Simulation Techniques and the Monte Carlo Method

# Section 14-1: Common Sampling Techniques {background-color="#2c4a6e"}

## Why Use Samples?

::: {.callout-note}
## Three Reasons to Sample
1. **Saves time and money** — studying an entire population is often impractical
2. **Enables otherwise impossible research** — e.g., testing cable strength to destruction would leave nothing to sell; blood cholesterol analysis cannot use every drop
3. **Allows more depth** — smaller samples enable more detailed interviews and richer data collection

**Key requirement:** A sample must be **random** — every member of the population must have an equal chance of selection. Random samples are **unbiased**; non-random samples are **biased**.
:::

## Four Basic Sampling Methods

::: {.callout-note}
## Overview
| Method | Core Idea |
|--------|-----------|
| **Random** | Use random numbers to select subjects |
| **Systematic** | Select every $k$th element after a random start |
| **Stratified** | Divide into homogeneous subgroups; sample each |
| **Cluster** | Select an intact, pre-existing group as the sample |
:::

## Random Sampling

::: {.callout-note}
## Random Sampling
Each element of the population is numbered, and random numbers are used to select the sample.

**Common mistakes to avoid:**
- Interviewing "people on the street" (coverage bias)
- Phone-in polls (self-selection bias)
- Mail surveys (non-response bias)

**Correct method:** Use a random number table or computer-generated random numbers.
:::

## Example 14-1 {#example-14-1}

:::{.callout-warning}
## TV Show Interviews — Random Sample

A researcher producing a TV show can interview only **10 governors** from the 50 states. Select a random sample of 10 states using a random number table.
:::

## Example 14-1

:::{.callout-tip}
## Solution

**Mathematical solution**

**Step 1:** Number the states 01–50 alphabetically (01 = Alabama, 02 = Alaska, …, 50 = Wyoming)

**Step 2:** Use the random number table. Start at a chosen point and read two-digit numbers:

Selected numbers: **27, 43, 34, 06, 13, 29, 01, 39, 23, 35**

(Discard numbers > 50 and duplicates; sample without replacement)

**Step 3:** Match to states:

| # | State | | # | State |
|---|-------|-|---|-------|
| 27 | Nebraska | | 29 | New Hampshire |
| 43 | Texas | | 01 | Alabama |
| 34 | North Dakota | | 39 | Rhode Island |
| 06 | Colorado | | 23 | Minnesota |
| 13 | Illinois | | 35 | Ohio |

:::

## Example 14-1

:::{.callout-tip}
## Solution

**R implementation**

```{r e14-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
states <- c("Alabama","Alaska","Arizona","Arkansas","California",
            "Colorado","Connecticut","Delaware","Florida","Georgia",
            "Hawaii","Idaho","Illinois","Indiana","Iowa",
            "Kansas","Kentucky","Louisiana","Maine","Maryland",
            "Massachusetts","Michigan","Minnesota","Mississippi","Missouri",
            "Montana","Nebraska","Nevada","New Hampshire","New Jersey",
            "New Mexico","New York","North Carolina","North Dakota","Ohio",
            "Oklahoma","Oregon","Pennsylvania","Rhode Island","South Carolina",
            "South Dakota","Tennessee","Texas","Utah","Vermont",
            "Virginia","Washington","West Virginia","Wisconsin","Wyoming")
```
:::
---

```{r e14-1-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(123)
sample_indices <- sample(1:50, size = 10, replace = FALSE)
cat("Selected state numbers:", sample_indices, "\n")
cat("Selected states:\n")
for (i in sort(sample_indices)) {
  cat(sprintf("  %02d. %s\n", i, states[i]))
}
```

## Systematic Sampling

::: {.callout-note}
## Systematic Sampling
1. Number all population elements
2. Determine interval $k = N/n$ (population size ÷ sample size)
3. Select a random starting point from 1 to $k$
4. Select every $k$th element thereafter

**Advantage:** Simple to execute; no full list of random numbers needed.

**Caution:** Avoid using systematic sampling on lists with a periodic pattern (e.g., alternating male/female) that matches the interval $k$.
:::

## Example 14-2 {#example-14-2}

:::{.callout-warning}
## TV Show Interviews — Systematic Sample

Using the same 50-state population, select a **systematic sample** of 10 states.
:::

## Example 14-2

:::{.callout-tip}
## Solution

**Mathematical solution**

**Step 1:** Number states 01–50

**Step 2:** $k = 50 / 10 = 5$ → select every 5th state

**Step 3:** Random start: **4** (selected from digits 1–5)

**Step 4:** Selected every 5th from 4:
$$4, 9, 14, 19, 24, 29, 34, 39, 44, 49$$

| # | State | | # | State |
|---|-------|-|---|-------|
| 04 | Arkansas | | 29 | New Hampshire |
| 09 | Florida | | 34 | North Dakota |
| 14 | Indiana | | 39 | Rhode Island |
| 19 | Maine | | 44 | Utah |
| 24 | Mississippi | | 49 | Wisconsin |

:::

## Example 14-2

:::{.callout-tip}
## Solution

**R implementation**

```{r e14-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
N <- 50; n <- 10
k <- N / n  # interval = 5

set.seed(7)
start <- sample(1:k, 1)
cat("Random start:", start, "\n")

sys_indices <- seq(start, N, by = k)
cat("Selected state numbers:", sys_indices, "\n\n")
cat("Selected states:\n")
for (i in sys_indices) {
  cat(sprintf("  %02d. %s\n", i, states[i]))
}
```
:::

## Stratified Sampling

::: {.callout-note}
## Stratified Sampling
The population is divided into **strata** (subgroups) based on shared characteristics (e.g., gender, grade level, region). A random sample is then drawn **from each stratum**.

**Advantage:** Guarantees representation of every important subgroup.

**Disadvantages:**
- Requires extensive effort when many strata exist
- Difficult if grouping criteria are complex or ambiguous

**Proportional stratification:** Sample size from each stratum is proportional to that stratum's share of the total population.
:::

## Example 14-3 {#example-14-3}

:::{.callout-warning}
## Stratified Sample of Students

From a class of 20 students (10 male, 10 female; mixed freshmen/sophomores), select a **stratified sample of 8** based on gender and grade level.

**Four strata:**

| Group 1 | Male Freshmen (5) | Group 3 | Female Freshmen (5) |
|---------|-------------------|---------|---------------------|
| Group 2 | Male Sophomores (5) | Group 4 | Female Sophomores (5) |
:::

## Example 14-3

:::{.callout-tip}
## Solution

**Mathematical solution**

**Step 1:** Divide into 4 strata (Male/Female $\times$ Freshman/Sophomore)

**Step 2:** Each stratum has 5 members. Select $8/4 = 2$ from each.

**Step 3:** Using random numbers, select 2 from each group:

| Group | Selected |
|-------|---------|
| Male Freshmen | Peterson, John; Meloski, Gary |
| Male Sophomores | Brown, Danny; Lyons, Larry |
| Female Freshmen | Bear, Theresa; Collins, Carolyn |
| Female Sophomores | Toms, Debbie; Unger, Roberta |

:::

## Example 14-3

:::{.callout-tip}
## Solution

**R implementation**

```{r e14-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
students <- data.frame(
  name   = c("Ald P","Brown D","Bear T","Carson S","Collins C",
             "Davis W","Hogan M","Jones L","Lutz H","Lyons L",
             "Martin J","Meloski G","Oeler G","Peters M","Peterson J",
             "Smith N","Thomas J","Toms D","Unger R","Zibert M"),
  gender = c("M","M","F","F","F","M","M","F","M","M",
             "F","M","M","F","M","F","M","F","F","F"),
  grade  = c("Fr","So","Fr","Fr","Fr","Fr","Fr","So","So","So",
             "Fr","Fr","So","So","Fr","Fr","So","So","So","So")
)
```

----

```{r e14-3-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# Stratify and sample 2 from each stratum
set.seed(42)
stratified <- students %>%
  group_by(gender, grade) %>%
  slice_sample(n = 2) %>%
  arrange(gender, grade)

print(stratified)
```
:::

## Cluster Sampling

::: {.callout-note}
## Cluster Sampling
A **naturally occurring group** (cluster) is selected as the sample. All members of the cluster may be used, or a random subset is taken.

**Examples:** An entire school classroom, all residents of a city block, all voters in an electoral district.

**Advantages:** Reduces cost; simplifies fieldwork; convenient.

**Disadvantage:** Members within a cluster tend to be more **homogeneous** than the general population — the sample may lack diversity.
:::

## Sampling Methods: Summary Comparison

::: {.callout-note}
## At a Glance

| Method | Best When | Watch Out For |
|--------|-----------|---------------|
| Random | General purpose | Time-consuming with large $N$ |
| Systematic | Pre-numbered list exists | Periodic patterns in list |
| Stratified | Subgroup representation matters | Many strata = lots of effort |
| Cluster | Cost/logistics drive decisions | Within-cluster homogeneity |

**Other methods:** sequence sampling (quality control), double sampling (two-stage eligibility), multistage sampling (combination approach for very large populations).
:::

# Section 14-2: Surveys and Questionnaire Design {background-color="#2c4a6e"}

## Types of Surveys

::: {.callout-note}
## Two Survey Formats

**Interviewer-administered:** A person poses questions face-to-face, by phone, or in a mall.

**Self-administered:** Respondents complete the survey independently — by mail, email, or in a group setting.

**Important:** How a question is worded can dramatically change the responses. Researchers must design questions carefully to avoid leading respondents toward a predetermined answer.
:::

## Five Common Questionnaire Flaws

::: {.callout-important}
## Avoid These Mistakes
1. **Biased questions** — Wording steers the respondent toward a specific answer
   *e.g.,* "Are you going to vote for Jones even though polls show he will lose?"
2. **Confusing words** — Ambiguous terms lead to misinterpretation
   *e.g.,* "Do you think people live longer on a diet?" (which diet?)
3. **Double-barreled questions** — One question asks two things at once
   *e.g.,* "Do you favor a special tax to fund national health care?"
4. **Double negatives** — Two negatives create confusion
   *e.g.,* "Do you feel it is not appropriate to have areas where people cannot smoke?"
5. **Improper question order** — Earlier questions prime responses to later ones
:::

## Additional Survey Biases

::: {.callout-note}
## Other Sources of Bias
- **Social desirability bias:** Respondents answer what they think the interviewer wants to hear
- **Acquiescence bias:** Some people agree with any statement to avoid conflict
- **Confidentiality vs. anonymity:** Respondents behave differently when they believe their identity is known
- **Timing and location:** Surveys about airline safety yield different results right after a crash
- **Open vs. closed questions:**
  - *Open-ended:* "List three retirement activities" — rich but hard to summarize
  - *Closed-ended:* Choose from given options — easy to tally but limits response
:::

---

# Section 14-3: Simulation Techniques and the Monte Carlo Method {background-color="#2c4a6e"}

## What Is Simulation?

::: {.callout-note}
## Simulation Defined
A **simulation technique** uses a probability experiment to replicate a real-life situation that is too costly, dangerous, or time-consuming to study directly.

**Historical roots:** Chess was invented as a simulation of warfare. Modern mathematical simulation dates to the 1940s when physicists **John Von Neumann** and **Stanislaw Ulam** used it to study neutron behavior in nuclear reactor design.

**Modern tools:** Computers generate random numbers, tally outcomes, and compute probabilities millions of times per second.
:::

## The Monte Carlo Method

::: {.callout-note}
## Six Steps of the Monte Carlo Method
1. **List** all possible outcomes of the experiment
2. **Assign probabilities** to each outcome
3. **Map** outcomes to random number ranges
4. **Draw** random numbers and run the experiment
5. **Repeat** and tally results
6. **Compute statistics** and draw conclusions

**Key insight:** Results approach the theoretical values as the number of repetitions increases (Law of Large Numbers).
:::

## Setting Up Random Number Correspondences

::: {.callout-note}
## Examples of Outcome-to-Number Mapping

| Experiment | Outcome | Random Digits |
|------------|---------|--------------|
| Coin toss (50/50) | Heads | 1,3,5,7,9 |
| Coin toss | Tails | 0,2,4,6,8 |
| Die roll (1/6 each) | 1–6 | 1,2,3,4,5,6 (ignore 7,8,9,0) |
| Spinner (4 values, 1/4 each) | 1 | 1–2; 2→3–4; 3→5–6; 4→7–8 (ignore 9,0) |
| Snoring (20% chance) | Snores | 1,2; Non-snore→3–9,0 |
:::

## Example 14-4 {#example-14-4}

:::{.callout-warning}
## Snoring Simulation

According to the CDC, approximately **20%** of people snore while sleeping. Simulate a sample of **20 people** and identify how many snore.
:::

## Example 14-4

:::{.callout-tip}
## Solution

**Mathematical solution**

Since $20\% = 1/5$, one in five people snores.

**Mapping:** Digits 1, 2 → snores; digits 0, 3–9 → does not snore

**Sample random digits:**
$$7\ 2\ 7\ 4\ 8\ 2\ 2\ 3\ 6\ 1\ 6\ 0\ 4\ 6\ 1\ 5\ 5\ 3\ 8\ 9$$

Snorers (digits 1 or 2): positions 2, 6, 7, 10, 15 → **5 snorers out of 20**

Proportion: $5/20 = 0.25$, close to the theoretical 20%.

:::

## Example 14-4

:::{.callout-tip}
## Solution

**R implementation**

```{r e14-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(101)
n_people <- 20
p_snore  <- 0.20

# Simulate with R's built-in Bernoulli sampler
snores <- rbinom(n_people, size = 1, prob = p_snore)

cat("Simulation results (1 = snores, 0 = does not):\n")
cat(snores, "\n\n")
cat("Number who snore:", sum(snores), "out of", n_people, "\n")
cat("Proportion:", round(mean(snores), 3),
    "  (Theoretical: 0.20)\n")
```
:::

## Example 14-5 {#example-14-5}

:::{.callout-warning}
## Tennis Game Simulation

Bill is **twice as good** as Mike at tennis. Simulate a **5-game match** between them.

Since $P(\text{Bill wins}) = 2/3$ and $P(\text{Mike wins}) = 1/3$:

**Mapping:** Digits 1–6 → Bill wins; digits 7–9 → Mike wins; digit 0 → ignore

**Random number used:** 86314

| Digit | 8 | 6 | 3 | 1 | 4 |
|-------|---|---|---|---|---|
| Winner | Mike | Bill | Bill | Bill | Bill |

**Result:** Bill wins 4 games, Mike wins 1.
:::

## Example 14-5

:::{.callout-tip}
## Solution

**R implementation**

```{r e14-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(99)
n_games  <- 5
p_bill   <- 2/3

winners <- sample(c("Bill","Mike"),
                  size = n_games,
                  replace = TRUE,
                  prob = c(p_bill, 1 - p_bill))

cat("Game-by-game results:\n")
for (i in seq_along(winners)) cat("  Game", i, ":", winners[i], "\n")

cat("\nBill wins:", sum(winners == "Bill"), "games\n")
cat("Mike wins:", sum(winners == "Mike"), "games\n")
```
:::

## Example 14-6 {#example-14-6}

:::{.callout-warning}
## Rolling a Die Until a 6 Appears

A die is rolled until a **6** appears. Using simulation, find the **average number of rolls** needed. Repeat the experiment **20 times**.
:::

## Example 14-6

:::{.callout-tip}
## Solution

**Mathematical solution**

**Mapping:** Digits 1–6 → die face; digits 7, 8, 9, 0 → ignore

**Example:** Random block 857236 → ignore 8, ignore 5, ignore 7, roll 2, roll 3, **roll 6** → **4 rolls**

After 20 simulated trials, total rolls = **96**:
$$\bar{X} = \frac{96}{20} = 4.8 \text{ rolls}$$

**Theoretical value** (expected value of geometric distribution):
$$E(X) = \frac{1}{p} = \frac{1}{1/6} = 6$$

More repetitions would push the simulation closer to 6.

:::

## Example 14-6

:::{.callout-tip}
## Solution

**R implementation**

```{r e14-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(55)
n_trials <- 20

rolls_needed <- replicate(n_trials, {
  count <- 0
  repeat {
    roll <- sample(1:6, 1)
    count <- count + 1
    if (roll == 6) break
  }
  count
})

cat("Rolls per trial:", rolls_needed, "\n")
cat("Total rolls:", sum(rolls_needed), "\n")
cat("Simulated average:", round(mean(rolls_needed), 2), "\n")
cat("Theoretical average (1/p = 6):", 6, "\n")
```
:::

## Example 14-6

:::{.callout-tip}
## Solution

**Visualization**

```{r e14-6-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(55)
many_trials <- replicate(1000, {
  count <- 0
  repeat {
    roll <- sample(1:6, 1)
    count <- count + 1
    if (roll == 6) break
  }
  count
})
```
:::
---

```{r e14-6-3, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
ggplot(data.frame(rolls = many_trials), aes(x = rolls)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(many_trials), color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 6, color = "darkgreen", linetype = "dashed", linewidth = 1) +
  labs(title = "Rolls Until First 6 (1000 trials)",
       subtitle = "Red = simulated mean | Green = theoretical mean (6)",
       x = "Number of Rolls", y = "Frequency") +
  theme_minimal()
```

## Example 14-7 {#example-14-7}

:::{.callout-warning}
## Selecting a Key

A person has **4 keys**, only one of which opens a lock. She tries them at random (without replacement) until the correct one is found. Find the **average number of tries** needed. Repeat **25 times**.
:::

## Example 14-7

:::{.callout-tip}
## Solution

**Mathematical solution**

**Setup:** Keys numbered 1–4; key 2 is the correct one.

**Mapping:** Use digits 1–4 only (ignore 5–9, 0). Count digits until 2 appears.

**Example:** 1432 → tries key 1, then 4, then 3, then **2** → **4 tries**

After 25 trials: $\sum X = 54$

$$\bar{X} = \frac{54}{25} = 2.16 \text{ tries}$$

**Theoretical average:**
$$E(X) = \frac{n+1}{2} = \frac{4+1}{2} = 2.5$$

:::

## Example 14-7

:::{.callout-tip}
## Solution

**R implementation**

```{r e14-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(77)
n_trials   <- 25
correct_key <- 2

tries_needed <- replicate(n_trials, {
  keys <- sample(1:4, 4, replace = FALSE)  # random order
  which(keys == correct_key)               # position of correct key
})

cat("Tries per trial:", tries_needed, "\n")
cat("Total tries:", sum(tries_needed), "\n")
cat("Simulated average:", round(mean(tries_needed), 3), "\n")
cat("Theoretical average ((n+1)/2 = 2.5):", 2.5, "\n")
```
:::

## Example 14-8 {#example-14-8}

:::{.callout-warning}
## Selecting a Monetary Bill

A box contains **5 one-dollar bills**, **3 five-dollar bills**, and **2 ten-dollar bills**. A person randomly selects one bill. Find the **expected value** of the selected bill. Perform the experiment **25 times**.
:::

## Example 14-8

:::{.callout-tip}
## Solution

**Mathematical solution**

**Probabilities:** $P(\$1) = 5/10 = 0.5$, $P(\$5) = 3/10 = 0.3$, $P(\$10) = 2/10 = 0.2$

**Mapping:** Digits 1–5 → $\$1$; digits 6–8 → $\$5$; digits 9, 0 → $\$10$

**25 simulated draws:** Total = $\$116$

$$\bar{X} = \frac{116}{25} = \$4.64$$

**Theoretical expected value:**
$$E(X) = 0.5(1) + 0.3(5) + 0.2(10) = 0.5 + 1.5 + 2.0 = \$4.00$$

The simulation is close to the theoretical value; more repetitions bring the results even closer.

:::

## Example 14-8

:::{.callout-tip}
## Solution

**R implementation**

```{r e14-8-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(33)
n_trials <- 25

bills    <- c(1, 5, 10)
probs    <- c(0.5, 0.3, 0.2)

draws <- sample(bills, size = n_trials, replace = TRUE, prob = probs)

cat("Bill values drawn:\n")
cat(draws, "\n\n")
cat("Total:", sum(draws), "\n")
cat("Simulated average: $", round(mean(draws), 2), "\n")
cat("Theoretical E(X): $", sum(bills * probs), "\n")
```
:::

## Example 14-8

:::{.callout-tip}
## Solution

**Visualization**

```{r e14-8-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(33)
many_draws <- sample(bills, size = 1000, replace = TRUE, prob = probs)

ggplot(data.frame(value = many_draws), aes(x = factor(value))) +
  geom_bar(fill = c("gold","steelblue","darkgreen"), alpha = 0.85) +
  geom_hline(yintercept = 1000 * probs, linetype = "dashed",
             color = c("gold3","steelblue4","darkgreen"), linewidth = 0.8) +
  labs(title = "Bill Selection Simulation (1000 trials)",
       x = "Bill Value ($)", y = "Frequency") +
  theme_minimal()
```
:::

## Monte Carlo Method: Summary

::: {.callout-note}
## Summary

| Step | Action |
|------|--------|
| 1 | List all possible outcomes |
| 2 | Assign probabilities |
| 3 | Map outcomes to random number ranges |
| 4 | Draw random numbers, run experiment |
| 5 | Repeat and tally |
| 6 | Compute statistics, draw conclusions |

**Remember:** Simulation approximates — it does not replace — exact probability calculations. The Law of Large Numbers guarantees convergence with enough repetitions.

**Modern use:** NASA flight simulators, video game physics, financial risk modeling, drug trial design, climate modeling.
:::

## Key Takeaways


**Section 14-1 — Sampling:**
- Samples save time, money, and enable otherwise impossible research
- Four methods: Random, Systematic, Stratified, Cluster
- Unbiased samples require every subject to have an equal selection probability

**Section 14-2 — Surveys:**
- Question wording, order, and format all affect results
- Avoid: biased questions, confusing words, double-barreled questions, double negatives, improper ordering

**Section 14-3 — Simulation:**
- Monte Carlo method: map outcomes to random number ranges, repeat, tally
- Simulated averages approach theoretical values as $n \to \infty$
- R's `sample()`, `rbinom()`, and `replicate()` make simulation straightforward

## Acknowledgement
-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2011). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
