---
title: "Statistics"
subtitle: "Chapter 7: Confidence Intervals and Sample Size"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch7-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview
**Sections covered:**

- **7-1** Confidence Intervals for the Mean When σ Is Known
- **7-2** Confidence Intervals for the Mean When σ Is Unknown
- **7-3** Confidence Intervals and Sample Size for Proportions
- **7-4** Confidence Intervals for Variances and Standard Deviations

---

**Estimation** is the process of using sample data to approximate the value of a population parameter. Because collecting data from an entire population is usually impractical, we rely on sample statistics — estimators — to infer what the population looks like.

This chapter introduces two types of estimates: **point estimates** (a single value) and **interval estimates** (a range of plausible values). It also addresses a critical practical question: *how large a sample do we need?*

# Section 7-1: Confidence Intervals for the Mean When σ Is Known {background-color="#2c4a6e"}

## Point Estimates vs. Interval Estimates

::: {.callout-note}
## Point Estimate
A **point estimate** is a single numerical value used to estimate a population parameter. The best point estimate of the population mean μ is the sample mean $\bar{X}$.
:::

::: {.callout-note}
## Interval Estimate
An **interval estimate** is a range of values used to estimate a parameter. It acknowledges that the sample mean will differ from μ due to sampling variability.
:::

A point estimate alone gives no sense of accuracy. An interval estimate — paired with a **confidence level** — tells us how reliable the estimate is.

## Properties of a Good Estimator

::: {.callout-important}
## Three Properties of a Good Estimator
1. **Unbiased**: The expected value of the estimator equals the true parameter
2. **Consistent**: As sample size grows, the estimator converges to the true value
3. **Relatively Efficient**: Among all unbiased estimators, it has the smallest variance
:::

The sample mean $\bar{X}$ satisfies all three properties — it is the preferred estimator of μ.

## Confidence Level and Confidence Interval

::: {.callout-note}
## Confidence Level
The **confidence level** is the probability that the interval estimate will contain the true population parameter, assuming the estimation process is repeated many times.
:::

::: {.callout-note}
## Confidence Interval
A **confidence interval** is a specific interval estimate of a parameter, determined from sample data and a chosen confidence level. Common levels: **90%, 95%, 99%**.
:::

Higher confidence → wider interval. There is always a trade-off between precision and certainty.

## Formula: Confidence Interval for μ (σ Known)

::: {.callout-note}
## CI Formula (σ Known)
$$\bar{X} - z_{\alpha/2}\left(\frac{\sigma}{\sqrt{n}}\right) < \mu < \bar{X} + z_{\alpha/2}\left(\frac{\sigma}{\sqrt{n}}\right)$$

| Confidence Level | $z_{\alpha/2}$ |
|-----------------|----------------|
| 90% | 1.65 |
| 95% | 1.96 |
| 99% | 2.58 |

The term $E = z_{\alpha/2}(\sigma/\sqrt{n})$ is the **margin of error**.
:::

**Assumptions:** 

(1) Random sample; 

(2) $n \geq 30$, OR population is normally distributed if n < 30.

----

**Find the z value in R**

```{r, echo=TRUE}
round(qnorm((1-.9) / 2), 2)
round(qnorm((1-.95) / 2), 2)
round(qnorm((1-.99) / 2), 2)
```

## Example 7-1 {#example-7-1}

:::{.callout-warning}
## Days It Takes to Sell an Aveo

A researcher wants to estimate the average number of days it takes to sell a Chevrolet Aveo. A sample of 50 cars had a mean time of 54 days on the dealer's lot. Assume the population standard deviation is 6.0 days. Find the best point estimate of the population mean and the **95% confidence interval**.
:::

## Example 7-1

:::{.callout-tip}
## Solution

**Solve**

A survey of 30 emergency room patients found an average waiting time of 174.3 minutes. The population standard deviation is known to be 46.5 minutes. Find the best point estimate and the **99% confidence interval** of the true mean waiting time.


**Solution (Mathematical):**

$$174.3 - 2.58\left(\frac{46.5}{\sqrt{30}}\right) < \mu < 174.3 + 2.58\left(\frac{46.5}{\sqrt{30}}\right)$$
$$174.3 - 21.9 < \mu < 174.3 + 21.9$$
$$\mathbf{152.4 < \mu < 196.2}$$
:::
---

```{r e7-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 174.3; sigma <- 46.5; n <- 30; z <- 2.58
E <- z * sigma / sqrt(n)
cat("Point estimate:", xbar, "\n")
cat("99% CI: (", round(xbar - E, 1), ",", round(xbar + E, 1), ")\n")
```

**Interpretation:** We are 99% confident that the true mean ER waiting time is between 152.4 and 196.2 minutes.

## Finding Other $z_{\alpha/2}$ Values

For confidence levels other than 90%, 95%, or 99%, find $z_{\alpha/2}$ as follows:

1. Convert the confidence level to a decimal: e.g., 98% → 0.98
2. Find $\alpha = 1 - 0.98 = 0.02$, so $\alpha/2 = 0.01$
3. The area in each tail is 0.01; look up 1 − 0.01 = 0.9900 in Table E
4. The corresponding z value is **2.33**

```{r e7-1-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# z_alpha/2 for common confidence levels
conf_levels <- c(0.90, 0.95, 0.98, 0.99)
z_vals <- qnorm(1 - (1 - conf_levels)/2)
data.frame(Confidence = paste0(conf_levels*100, "%"),
           z_alpha2 = round(z_vals, 3))
```

## Example 7-3 {#example-7-3}

:::{.callout-warning}
## Credit Union Assets

The following data represent the assets (in millions of dollars) for a sample of 30 credit unions in southwestern Pennsylvania. Find the **90% confidence interval** of the mean. Assume σ = 14.405.

12.23, 16.56, 4.39, 2.89, 1.24, 2.17, 13.19, 9.16, 1.42, 73.25, 1.91, 14.64, 11.59, 6.69, 1.06, 8.74, 3.17, 18.13, 7.92, 4.78, 16.85, 40.22, 2.42, 21.58, 5.01, 1.47, 12.24, 2.27, 12.77, 2.76
:::

## Example 7-3

:::{.callout-tip}
## Solution

**Solve**

$$n = \left(\frac{z_{\alpha/2} \cdot \sigma}{E}\right)^2$$

where E is the desired **margin of error**. Always round **up** to the next whole number.


**What you need to know:**
- The desired confidence level (determines $z_{\alpha/2}$)
- The population standard deviation σ (or an estimate from a prior study)
- The acceptable margin of error E
:::

## Example 7-4 {#example-7-4}

:::{.callout-warning}
A scientist wants to estimate the average depth of a river. She wants to be **99% confident** that the estimate is accurate within **2 feet**. A previous study found σ = 4.33 feet. Find the minimum sample size needed.

**Solution (Mathematical):**

$z_{\alpha/2} = 2.58$, $E = 2$, $\sigma = 4.33$

$$n = \left(\frac{2.58 \times 4.33}{2}\right)^2 = \left(\frac{11.17}{2}\right)^2 = (5.585)^2 = 31.2 \rightarrow \mathbf{n = 32}$$

```{r e7-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
z <- 2.58; sigma <- 4.33; E <- 2
n_raw <- (z * sigma / E)^2
n_raw
ceiling(n_raw)  # Always round UP
```

**Answer:** The scientist needs at least **32 measurements** to achieve the desired precision with 99% confidence.

:::

# Section 7-2: Confidence Intervals for the Mean When σ Is Unknown {background-color="#2c4a6e"}

## The t Distribution

When σ is unknown, the sample standard deviation s is used as a substitute. However, this introduces additional uncertainty — especially for small samples. The appropriate distribution is the **Student's t distribution**.


::: {.callout-note}
## Characteristics of the t Distribution
**Similar to the normal distribution:**
1. Bell-shaped and symmetric about the mean
2. Mean, median, and mode all equal 0
3. The curve never touches the x-axis

**Differences from the normal:**
1. Variance is greater than 1 (heavier tails)
2. It is a **family of curves** indexed by **degrees of freedom** (d.f.)
3. As d.f. increases, the t distribution approaches the standard normal
:::
---

```{r e7-4-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(-4, 4, length.out = 300)
df_plot <- data.frame(
  x = rep(x, 3),
  y = c(dnorm(x), dt(x, df = 5), dt(x, df = 20)),
  dist = rep(c("Normal (z)", "t (df=5)", "t (df=20)"), each = 300)
)
ggplot(df_plot, aes(x, y, color = dist, linetype = dist)) +
  geom_line(linewidth = 1) +
  labs(title = "Comparing Normal and t Distributions",
       x = "Value", y = "Density", color = "Distribution", linetype = "Distribution") +
  theme_minimal()
```

## Degrees of Freedom and the t Formula

::: {.callout-note}
## Degrees of Freedom
For a confidence interval for the mean: **d.f. = n − 1**

The degrees of freedom specify which t curve to use. In R: `qt(p, df = n-1)`
:::

::: {.callout-note}
## CI Formula (σ Unknown)
$$\bar{X} - t_{\alpha/2}\left(\frac{s}{\sqrt{n}}\right) < \mu < \bar{X} + t_{\alpha/2}\left(\frac{s}{\sqrt{n}}\right)$$

where $t_{\alpha/2}$ is from Table F with d.f. = n − 1.
:::

**Assumptions:** (1) Random sample; (2) $n \geq 30$, OR population is approximately normally distributed if n < 30.

## When to Use z vs. t

::: {.callout-important}
## Decision Rule
- **σ is known** → use $z_{\alpha/2}$ with σ in the formula (any n, if normal or $n \geq 30$)
- **σ is unknown** → use $t_{\alpha/2}$ with s in the formula (normal population or $n \geq 30$)

```{r e7-4-3, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
knitr::kable(data.frame(
  Condition = c("σ known, n ≥ 30", "σ known, n < 30 (normal pop.)",
                "σ unknown, n ≥ 30", "σ unknown, n < 30 (normal pop.)"),
  Use = c("z and σ", "z and σ", "t and s", "t and s")
), caption = "When to Use z vs. t")
```
:::

## Example 7-5: Finding $t_{\alpha/2}$ {.smaller}



## Example 7-5 {#example-7-5}


:::{.callout-warning}
For a confidence interval, always use `qt(1 - alpha/2, df = n-1)`.
For a 95% CI: `qt(0.975, df = n-1)`.
:::

## Example 7-5

:::{.callout-tip}
## Solution

**Solve**

Ten randomly selected people reported their nightly sleep duration. The sample mean was 7.1 hours and the standard deviation was 0.78 hour. Find the **95% confidence interval** of the mean. Assume the variable is normally distributed.


**Solution (Mathematical):**

σ unknown, n = 10 < 30 → use t distribution. d.f. = 9, $t_{\alpha/2}$ = 2.262

$$7.1 - 2.262\left(\frac{0.78}{\sqrt{10}}\right) < \mu < 7.1 + 2.262\left(\frac{0.78}{\sqrt{10}}\right)$$
$$7.1 - 0.56 < \mu < 7.1 + 0.56$$
$$\mathbf{6.54 < \mu < 7.66}$$
:::
---

```{r e7-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 7.1; s <- 0.78; n <- 10
t_crit <- qt(0.975, df = n - 1)
E <- t_crit * s / sqrt(n)
cat("t_alpha/2 =", round(t_crit, 3), "\n")
cat("95% CI: (", round(xbar - E, 2), ",", round(xbar + E, 2), ")\n")
```

## Example 7-7 {#example-7-7}

:::{.callout-warning}
## Home Fires Started by Candles

The data below represent the number of home fires started by candles each year over a 7-year period. Find the **99% confidence interval** for the mean number of candle-caused home fires per year.

5460, 5900, 6090, 6310, 7160, 8440, 9930
:::
---

```{r e7-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
fires <- c(5460, 5900, 6090, 6310, 7160, 8440, 9930)
n <- length(fires)
xbar <- mean(fires); s <- sd(fires)
cat("n =", n, "  X-bar =", round(xbar, 1), "  s =", round(s, 1), "\n")

t_crit <- qt(0.995, df = n - 1)   # 99% CI
cat("t_alpha/2 (df=6) =", round(t_crit, 3), "\n")
E <- t_crit * s / sqrt(n)
cat("99% CI: (", round(xbar - E, 1), ",", round(xbar + E, 1), ")\n")
```

**Answer:** $4785.2 < \mu < 9297.6$ fires per year. We are 99% confident that the true mean falls in this range, based on 7 years of data.

# Section 7-3: Confidence Intervals and Sample Size for Proportions {background-color="#2c4a6e"}

## Proportions

::: {.callout-note}
## Notation for Proportions
$$\hat{p} = \frac{X}{n} \quad \text{(sample proportion)} \qquad \hat{q} = \frac{n-X}{n} = 1 - \hat{p}$$

where X = number of units in the sample with the characteristic of interest, n = sample size.

The population proportion is denoted **p** (and q = 1 − p).

The sample proportion $\hat{p}$ is an unbiased, consistent, and relatively efficient estimator of p. It also serves as the best point estimate of p.

**Confidence interval formula:**
$$\hat{p} - z_{\alpha/2}\sqrt{\frac{\hat{p}\hat{q}}{n}} < p < \hat{p} + z_{\alpha/2}\sqrt{\frac{\hat{p}\hat{q}}{n}}$$

**Condition:** $n\hat{p} \geq 5$ and $n\hat{q} \geq 5$
:::

## Example 7-8 {#example-7-8}


:::{.callout-warning}
## Air-Conditioned Households

In a recent survey of 150 households, 54 had central air conditioning. Find $\hat{p}$ and $\hat{q}$, where $\hat{p}$ is the proportion of households with central air conditioning.
:::

## Example 7-8

:::{.callout-tip}
## Solution

**Solve**

**Solution (Mathematical):**

$$\hat{p} = \frac{X}{n} = \frac{54}{150} = 0.36 = 36\%$$
$$\hat{q} = \frac{n - X}{n} = \frac{150 - 54}{150} = \frac{96}{150} = 0.64 = 64\%$$
:::
---

```{r e7-8-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 54; n <- 150
p_hat <- X / n
q_hat <- 1 - p_hat
cat("p-hat =", p_hat, "  q-hat =", q_hat, "\n")
```

This gives the best point estimate: **36%** of households have central air conditioning.

## Example 7-9 {#example-7-9}

:::{.callout-warning}
## Covering College Costs

A survey of 1404 respondents found that 323 students paid for their education using student loans. Find the **90% confidence interval** for the true proportion of students who fund their education through student loans.
:::

## Example 7-9

:::{.callout-tip}
## Solution

**Solve**

**Solution:**

$\hat{p} = 323/1404 = 0.230$, $\hat{q} = 0.770$, $z_{\alpha/2} = 1.65$

$$0.230 \pm 1.65\sqrt{\frac{(0.230)(0.770)}{1404}} = 0.230 \pm 0.019$$
$$\mathbf{0.211 < p < 0.249}$$
:::
---

```{r e7-9-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 323; n <- 1404
p_hat <- X / n; q_hat <- 1 - p_hat
z <- qnorm(0.95)   # 90% CI
E <- z * sqrt(p_hat * q_hat / n)
cat("p-hat =", round(p_hat, 3), "\n")
cat("90% CI: (", round(p_hat - E, 3), ",", round(p_hat + E, 3), ")\n")
```

**Interpretation:** We are 90% confident that between **21.1%** and **24.9%** of students pay for college via student loans.

## Example 7-10 {#example-7-10}

:::{.callout-warning}
## Religious Books

A survey of 1721 people found that 15.9% purchase religious books at a Christian bookstore. Find the **95% confidence interval** for the true proportion of people who purchase religious books there.
:::

## Example 7-10

:::{.callout-tip}
## Solution

**Solve**

$$n = \hat{p}\hat{q}\left(\frac{z_{\alpha/2}}{E}\right)^2$$

- If a prior estimate of $\hat{p}$ is available, use it
- If no prior estimate is available, use $\hat{p} = 0.5$ (maximizes n, giving the most conservative estimate)

Always round **up** to the next whole number.

:::

## Example 7-11 {#example-7-11}
:::{.callout-warning}
A researcher wishes to estimate with **95% confidence** the proportion of people who own a home computer, to within **2%** of the true proportion. A previous study showed that 40% had a computer at home. Find the minimum sample size needed.

**Solution:**

$z_{\alpha/2} = 1.96$, $E = 0.02$, $\hat{p} = 0.40$, $\hat{q} = 0.60$

$$n = (0.40)(0.60)\left(\frac{1.96}{0.02}\right)^2 = (0.24)(9604) = 2304.96 \rightarrow \mathbf{n = 2305}$$

```{r e7-11-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
p_hat <- 0.40; q_hat <- 0.60; z <- 1.96; E <- 0.02
n_raw <- p_hat * q_hat * (z / E)^2
n_raw
ceiling(n_raw)
```

:::

## Example 7-12 {#example-7-12}

:::{.callout-warning}
A researcher wishes to estimate the percentage of M&M's that are brown. He wants to be **95% confident** and accurate within **3%** of the true proportion. No prior estimate of $\hat{p}$ is available. Find the minimum sample size.

**Solution:**

Since no prior $\hat{p}$ is known, use $\hat{p} = \hat{q} = 0.5$

$$n = (0.5)(0.5)\left(\frac{1.96}{0.03}\right)^2 = (0.25)(4268.4) = 1067.1 \rightarrow \mathbf{n = 1068}$$
:::
---

```{r e7-12-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
p_hat <- 0.5; q_hat <- 0.5; z <- 1.96; E <- 0.03
n_raw <- p_hat * q_hat * (z / E)^2
n_raw
ceiling(n_raw)
```

::: {.callout-important}
## Key Rule
When $\hat{p}$ is unknown, always use $\hat{p} = 0.5$ — it produces the **largest possible** sample size, ensuring sufficient precision regardless of the true proportion.
:::

# Section 7-4: Confidence Intervals for Variances and Standard Deviations {background-color="#2c4a6e"}

## The Chi-Square Distribution

Confidence intervals for variances and standard deviations require a different distribution — the **chi-square** (χ²) distribution.

::: {.callout-note}
## Chi-Square Distribution Properties
- Obtained from values of $(n-1)s^2/\sigma^2$ from normal populations
- Cannot be negative; skewed to the right
- A family of curves indexed by **degrees of freedom** (d.f. = n − 1)
- Total area under each curve = 1
- Approaches symmetry as d.f. increases (around 100+)

**Key difference from z and t:** The chi-square distribution is **not symmetric**, so two separate critical values ($\chi^2_{\text{right}}$ and $\chi^2_{\text{left}}$) are needed.
:::

---

```{r e7-12-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(0, 40, length.out = 300)
df_plot <- data.frame(
  x = rep(x, 4),
  y = c(dchisq(x, 1), dchisq(x, 4), dchisq(x, 9), dchisq(x, 15)),
  df = rep(c("df=1","df=4","df=9","df=15"), each = 300)
)
ggplot(df_plot, aes(x, y, color = df)) +
  geom_line(linewidth = 1) +
  coord_cartesian(ylim = c(0, 0.5)) +
  labs(title = "Chi-Square Family of Curves",
       x = expression(chi^2), y = "Density", color = "d.f.") +
  theme_minimal()
```

## Formulas for Variance and Standard Deviation CIs

::: {.callout-note}
## Confidence Interval for a Variance
$$\frac{(n-1)s^2}{\chi^2_{\text{right}}} < \sigma^2 < \frac{(n-1)s^2}{\chi^2_{\text{left}}}$$
d.f. = n − 1
:::

::: {.callout-note}
## Confidence Interval for a Standard Deviation
$$\sqrt{\frac{(n-1)s^2}{\chi^2_{\text{right}}}} < \sigma < \sqrt{\frac{(n-1)s^2}{\chi^2_{\text{left}}}}$$
:::

**Assumptions:** (1) Random sample; (2) Population **must** be normally distributed.

**Finding critical values:** Use `qchisq()` in R. For a 95% CI with d.f. = n − 1:
- $\chi^2_{\text{right}}$: `qchisq(0.975, df = n-1)`
- $\chi^2_{\text{left}}$: `qchisq(0.025, df = n-1)`

## Example 7-13 {#example-7-13}

:::{.callout-warning}
## Finding Chi-Square Critical Values

Find the values $\chi^2_{\text{right}}$ and $\chi^2_{\text{left}}$ for a **90% confidence interval** when n = 25.
:::

## Example 7-13

:::{.callout-tip}
## Solution

**Solve**

**Solution (Mathematical):**

d.f. = 25 − 1 = 24

- $\alpha = 1 - 0.90 = 0.10$, so $\alpha/2 = 0.05$
- $\chi^2_{\text{right}}$: area to right = 0.05 → column 0.05 → **36.415**
- $\chi^2_{\text{left}}$: area to right = 0.95 → column 0.95 → **13.848**
:::
---

```{r e7-13-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 25; df <- n - 1
chi_right <- qchisq(0.95, df = df)   # upper critical value
chi_left  <- qchisq(0.05, df = df)   # lower critical value
cat("chi-sq right =", round(chi_right, 3), "\n")
cat("chi-sq left  =", round(chi_left, 3), "\n")
```

::: {.callout-warning}
## Caution
Note the counterintuitive naming: $\chi^2_{\text{right}}$ is the **larger** value used in the **denominator of the lower bound**, while $\chi^2_{\text{left}}$ is the **smaller** value used in the **denominator of the upper bound**.
:::

## Example 7-14 {#example-7-14}

:::{.callout-warning}
## Nicotine Content

Find the **95% confidence interval** for the variance and standard deviation of the nicotine content of cigarettes manufactured, given that a sample of n = 20 cigarettes has a standard deviation of s = 1.6 milligrams.
:::

## Example 7-14

:::{.callout-tip}
## Solution

**Solve**

**Solution (Mathematical):**

d.f. = 19; 95% CI → $\chi^2_{\text{right}} = 32.852$, $\chi^2_{\text{left}} = 8.907$

$$\frac{(19)(1.6)^2}{32.852} < \sigma^2 < \frac{(19)(1.6)^2}{8.907}$$
$$\frac{48.64}{32.852} < \sigma^2 < \frac{48.64}{8.907} \implies \mathbf{1.5 < \sigma^2 < 5.5}$$

$$\sqrt{1.5} < \sigma < \sqrt{5.5} \implies \mathbf{1.2 < \sigma < 2.3}$$
:::
---

```{r e7-14-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 20; s <- 1.6; df <- n - 1
chi_right <- qchisq(0.975, df = df)
chi_left  <- qchisq(0.025, df = df)
cat("chi-sq right =", round(chi_right, 3), "\n")
cat("chi-sq left  =", round(chi_left, 3), "\n")

var_lower <- (df * s^2) / chi_right
var_upper <- (df * s^2) / chi_left
cat("95% CI for variance: (", round(var_lower, 1), ",", round(var_upper, 1), ")\n")
cat("95% CI for std dev:  (", round(sqrt(var_lower), 1), ",",
    round(sqrt(var_upper), 1), ")\n")
```

## Example 7-15 {#example-7-15}

:::{.callout-warning}
## Cost of Ski Lift Tickets

The data below represent the prices (in dollars) of adult single-day ski lift tickets at a sample of nationwide ski resorts. Find the **90% confidence interval** for the variance and standard deviation. Assume the variable is normally distributed.

59, 54, 53, 52, 51, 39, 49, 46, 49, 48
:::

## Example 7-15

:::{.callout-tip}
## Solution

**R implementation**

----

```{r e7-15-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
tickets <- c(59, 54, 53, 52, 51, 39, 49, 46, 49, 48)
n <- length(tickets)
s2 <- var(tickets); s <- sd(tickets)
df <- n - 1
cat("n =", n, "  s² =", round(s2, 1), "  s =", round(s, 2), "\n")

chi_right <- qchisq(0.95, df = df)   # 90% CI
chi_left  <- qchisq(0.05, df = df)
cat("chi-sq right =", round(chi_right, 3), "\n")
cat("chi-sq left  =", round(chi_left, 3), "\n")

var_lower <- (df * s2) / chi_right
var_upper <- (df * s2) / chi_left
cat("90% CI for variance: (", round(var_lower, 1), ",", round(var_upper, 1), ")\n")
cat("90% CI for std dev:  (", round(sqrt(var_lower), 2), ",",
    round(sqrt(var_upper), 2), ")\n")
```

**Answer:** $15.0 < \sigma^2 < 76.3$; $3.87 < \sigma < 8.73$ dollars.
:::

## Chapter 7 Summary {.smaller}

| Parameter | Formula | R Function |
|-----------|---------|------------|
| Mean (σ known) | $\bar{X} \pm z_{\alpha/2}(\sigma/\sqrt{n})$ | `qnorm(1-alpha/2)` |
| Mean (σ unknown) | $\bar{X} \pm t_{\alpha/2}(s/\sqrt{n})$ | `qt(1-alpha/2, df=n-1)` |
| Sample size (mean) | $n = (z_{\alpha/2} \cdot \sigma / E)^2$ | `ceiling(...)` |
| Proportion | $\hat{p} \pm z_{\alpha/2}\sqrt{\hat{p}\hat{q}/n}$ | `prop.test()` |
| Sample size (proportion) | $n = \hat{p}\hat{q}(z_{\alpha/2}/E)^2$ | `ceiling(...)` |
| Variance | $(n-1)s^2/\chi^2_R < \sigma^2 < (n-1)s^2/\chi^2_L$ | `qchisq(...)` |

## Key Takeaways

1. Use z when σ is known; use t when σ is unknown
2. Wider confidence intervals → more confidence, less precision
3. Larger samples → narrower intervals → more precision
4. When no prior $\hat{p}$ is known for proportions, use $\hat{p} = 0.5$
5. Chi-square CIs for variance **require** the population to be normally distributed
6. Always round sample size calculations **up** to the nearest whole number

## Acknowledgement
-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2011). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
