---
title: "Statistics"
subtitle: "Chapter 7: Confidence Intervals and Sample Size"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
language:
  callout-warning-title: "Problem"
  callout-important-title: "Key point"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch7-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_minimal())
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Chapter 7 is about **estimation** — using a sample statistic to estimate a population parameter, with a stated confidence level and a stated sample size.

| Section | Topics |
|---------|--------|
| 7-1 | Estimation; point estimates; properties of a good estimator; interval estimates; confidence level and interval; critical value $z_{\alpha/2}$; margin of error |
| 7-2 | Confidence interval for the mean, $\sigma$ known; assumptions; rounding rule; minimum sample size |
| 7-3 | The $t$ distribution; degrees of freedom; confidence interval for the mean, $\sigma$ unknown; $z$ versus $t$ |

---

## Overview (continued)

| Section | Topics |
|---------|--------|
| 7-4 | Sample proportion $\hat{p}$ and $\hat{q}$; confidence interval for a proportion; minimum sample size |
| 7-5 | The chi-square distribution; confidence intervals for a variance and a standard deviation |

## Chapter Objectives

:::{.callout-important}
## After completing this chapter, you should be able to

1. Find the confidence interval for the mean when $\sigma$ is known.
2. Determine the minimum sample size for finding a confidence interval for the mean.
3. Find the confidence interval for the mean when $\sigma$ is unknown.
4. Find the confidence interval for a proportion.
5. Determine the minimum sample size for finding a confidence interval for a proportion.
6. Find a confidence interval for a variance and a standard deviation.
:::

# Section 7-1: Confidence Intervals {background-color="#2c4a6e"}

## Estimation

One aspect of inferential statistics is **estimation**: the process of estimating the value of a parameter from information obtained from a sample.

Sample measures (statistics) are used to estimate population measures (parameters) — a sample mean estimates a population mean, a sample proportion estimates a population proportion. Statistics used this way are called **estimators**.

:::{.callout-note}
## Estimation

**Estimation** is the process of estimating the value of a parameter from information obtained from a sample.
:::

:::{.callout-important}
## Common Assumptions for the Procedures in This Chapter

1. The sample must be **randomly selected**.
2. Either the sample size is $n \geq 30$, **or** the population is normally or approximately normally distributed when $n < 30$.

Check the second assumption with the Chapter 6 methods: look at the histogram for a bell shape, check for outliers, and inspect a normal quantile plot.
:::

## Point Estimates

A college president selects a random sample of 100 students and finds their average age to be 22.3 years. Inferring that the average age of *all* students is 22.3 years is a **point estimate**.

:::{.callout-note}
## Point Estimate

A **point estimate** is a specific numerical value estimate of a parameter. The best point estimate of the population mean $\mu$ is the sample mean $\bar{X}$.
:::

:::{.callout-important}
## Three Properties of a Good Estimator

1. The estimator should be an **unbiased estimator** — the expected value (the mean of the estimates obtained from samples of a given size) equals the parameter being estimated.
2. The estimator should be **consistent** — as the sample size increases, the value of the estimator approaches the value of the parameter estimated.
3. The estimator should be a **relatively efficient estimator** — of all the statistics that can be used to estimate a parameter, it has the smallest variance.
:::

## Interval Estimates

A sample mean is almost always somewhat different from the population mean because of sampling error, and there is **no way of knowing** how close a particular point estimate is to $\mu$. For that reason statisticians prefer an interval estimate.

:::{.callout-note}
## Interval Estimate

An **interval estimate** of a parameter is an interval or a range of values used to estimate the parameter. This estimate may or may not contain the value of the parameter being estimated.
:::

An interval estimate for the average age of all students might be $21.9 < \mu < 22.7$, or $22.3 \pm 0.4$ years. An interval estimate for a proportion might be $0.107 < p < 0.113$, or $11\% \pm 0.3\%$.

Confidence intervals may also be written with parentheses: $(21.9,\ 22.7)$ and $(10.7\%,\ 11.3\%)$.

## Confidence Level and Confidence Interval

:::{.callout-note}
## Confidence Level

The **confidence level** of an interval estimate of a parameter is the probability that the interval estimate will contain the parameter, assuming that a large number of samples are selected and that the estimation process on the same parameter is repeated.
:::

:::{.callout-note}
## Confidence Interval

A **confidence interval** is a specific interval estimate of a parameter determined by using data obtained from a sample and by using the specific confidence level of the estimate.
:::

Three common confidence intervals are used: the **90%**, the **95%**, and the **99%** confidence intervals.

---

## Confidence Level and Confidence Interval

A population has a fixed value for the mean or proportion. A confidence interval constructed from one sample either includes that parameter or it does not. If 95% confidence intervals were constructed for **all possible samples** of a specific size, 95% of them would contain the population parameter.

To be more confident — 99% instead of 95% — you must make the interval **wider**. A 99% interval for the mean age of college students might be $21.7 < \mu < 22.9$, or $22.3 \pm 0.6$. Hence a tradeoff occurs.

:::{.callout-warning}
## Caution

The confidence level describes the long-run behaviour of the **procedure**, not one particular interval. Once an interval has been computed, the parameter is either in it or it is not.
:::

## The Critical Value $z_{\alpha/2}$

The Greek letter $\alpha$ represents the total area in **both** tails of the standard normal distribution curve, and $\alpha/2$ represents the area in **each** tail. The value $z_{\alpha/2}$ is called a **critical value**.

The stated confidence level is the percentage equivalent of $1 - \alpha$:

- 95% confidence interval: $\alpha = 0.05$, since $1 - 0.05 = 0.95$
- 99% confidence interval: $\alpha = 0.01$, since $1 - 0.01 = 0.99$

---

## The Critical Value $z_{\alpha/2}$

Table A-4 is the paper route to these critical values; `qnorm()` is the R route. For a 95% confidence interval $z_{\alpha/2} = 1.960$, and for a 99% confidence interval $z_{\alpha/2} = 2.576$.

```{r s7-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# 90%, 95% and 99% confidence
qnorm(c(0.95, 0.975, 0.995))
```

## Margin of Error

:::{.callout-note}
## Margin of Error

The **margin of error**, also called the **maximum error of the estimate**, is the maximum likely difference between the point estimate of a parameter and the actual value of the parameter.
:::

In the age-of-students example, where the confidence interval is $21.9 < \mu < 22.7$ or $22.3 \pm 0.4$, the margin of error is **0.4**. For the proportion interval $10.7\% < p < 11.3\%$, or $11\% \pm 0.3\%$, the margin of error is **0.003**, or 0.3%.

For a specific value, say $\alpha = 0.05$, 95% of the sample means or proportions will fall within this error value on either side of the population mean or proportion.

## 95% Confidence Intervals in R

**Prepare data** — repeated sampling from a population with $\mu = 50$, $\sigma = 10$, $n = 25$.

```{r s7-1-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(2023)
x_bar  <- replicate(40, mean(rnorm(25, 50, 10)))
margin <- qnorm(0.975) * 10 / sqrt(25)
lower  <- x_bar - margin
upper  <- x_bar + margin

sum(lower <= 50 & 50 <= upper)
```

---

## 95% Confidence Intervals in R

**Draw figure**

```{r s7-1-3, echo=T, warning=FALSE, out.width="80%", fig.align='center', fig.height=4}
intervals <- data.frame(sample = 1:40, x_bar, lower, upper)

ggplot(intervals, aes(sample, x_bar)) +
  geom_pointrange(aes(ymin = lower, ymax = upper)) +
  geom_hline(yintercept = 50) +
  labs(title = "95% Confidence Intervals for 40 Sample Means",
       x = "Sample", y = "Sample mean")
```

About 95% of the intervals contain $\mu$; the ones that miss are the price of sampling variability.

# Section 7-2: Confidence Intervals for the Mean When $\sigma$ Is Known {background-color="#2c4a6e"}

## Assumptions and Formula

:::{.callout-important}
## Assumptions for Finding a Confidence Interval for a Mean When $\sigma$ Is Known

1. The sample is a random sample.
2. Either $n \geq 30$ or the population is normally distributed when $n < 30$.
:::

:::{.callout-note}
## Robust

A statistical technique is **robust** when the distribution of the variable can depart somewhat from normality and valid conclusions can still be obtained.
:::

When you estimate a population mean you should use the **sample mean**: the means of samples vary less than medians or modes when many samples are drawn from the same population.

---

## Assumptions and Formula

:::{.callout-important}
## Formula for the Confidence Interval of the Mean When $\sigma$ Is Known

$$\bar{X} - z_{\alpha/2}\left(\frac{\sigma}{\sqrt{n}}\right) < \mu < \bar{X} + z_{\alpha/2}\left(\frac{\sigma}{\sqrt{n}}\right)$$

For a 90% confidence interval $z_{\alpha/2} = 1.645$; for a 95% confidence interval $z_{\alpha/2} = 1.960$; and for a 99% confidence interval $z_{\alpha/2} = 2.576$.

The term $z_{\alpha/2}\left(\sigma/\sqrt{n}\right)$ is the **margin of error**.
:::

:::{.callout-important}
## Rounding Rule for a Confidence Interval for a Mean

When computing a confidence interval for a population mean from **raw data**, round off to **one more decimal place** than the number of decimal places in the original data.

When computing it from a **sample mean and a standard deviation**, round off to the **same number of decimal places** as given for the mean.
:::

## Example 7-1 {#example-7-1}

:::{.callout-warning}
## Air-Freight Transit Time to Osaka

A logistics analyst at a Taipei freight forwarder wants to estimate the mean transit time, in hours, of air-freight shipments from Taoyuan to Osaka. A random sample of 49 shipments had a mean transit time of 34.0 hours. Company records give a population standard deviation of 2.8 hours. Find the best point estimate of the mean and the **95% confidence interval** for the population mean.

*(hypothetical data)*
:::

---

## Example 7-1

:::{.callout-tip}
## Solution

**Step 1** The best point estimate of the mean is **34.0 hours**.

**Step 2** For the 95% confidence interval, use $z_{\alpha/2} = 1.96$.

$$34.0 - 1.96\left(\frac{2.8}{\sqrt{49}}\right) < \mu < 34.0 + 1.96\left(\frac{2.8}{\sqrt{49}}\right)$$
$$34.0 - 0.8 < \mu < 34.0 + 0.8$$
$$\mathbf{33.2 < \mu < 34.8}$$
:::

---

## Example 7-1

:::{.callout-tip}
## Solution

**Compute in R**

```{r e7-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
margin <- qnorm(0.975) * 2.8 / sqrt(49)

margin
34.0 - margin
34.0 + margin
```

One can say with **95% confidence** that the interval between 33.2 and 34.8 hours contains the population mean transit time, based on a sample of 49 shipments.
:::

## Example 7-2 {#example-7-2}

:::{.callout-warning}
## Value of Cross-Border Orders

A cross-border e-commerce team wants to estimate the average value, in US dollars, of the orders its Taiwanese sellers ship overseas. A random sample of 64 orders had a mean value of 84.5 dollars. Assume the population standard deviation is 12.0 dollars. Find the best point estimate of the mean and the **99% confidence interval** of the population mean.

*(hypothetical data)*
:::

---

## Example 7-2

:::{.callout-tip}
## Solution

**Step 1** The best point estimate of the population mean is **84.5 dollars**.

**Step 2** Use $z_{\alpha/2} = 2.576$ for the 99% confidence interval.

$$84.5 - 2.576\left(\frac{12.0}{\sqrt{64}}\right) < \mu < 84.5 + 2.576\left(\frac{12.0}{\sqrt{64}}\right)$$
$$84.5 - 3.9 < \mu < 84.5 + 3.9$$
$$\mathbf{80.6 < \mu < 88.4}$$
:::

---

## Example 7-2

:::{.callout-tip}
## Solution

**Compute in R**

```{r e7-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
margin <- qnorm(0.995) * 12.0 / sqrt(64)

margin
84.5 - margin
84.5 + margin
```

One can be **99% confident** that the mean value of a cross-border order is between 80.6 and 88.4 dollars.
:::

## Finding Other Values of $z_{\alpha/2}$

Confidence levels other than 90%, 95% and 99% are sometimes used. To find $z_{\alpha/2}$ for a **98%** confidence interval:

1. Change 98% to 0.98 and find $\alpha = 1 - 0.98 = 0.02$.
2. Divide by 2: $\alpha/2 = 0.02/2 = 0.01$.
3. Subtract 0.01 from 1.0000 to get 0.9900. That is the area you look up in Table A-4, and `qnorm(0.99)` returns the same critical value, $z_{\alpha/2} = \mathbf{2.326}$.

---

## Finding Other Values of $z_{\alpha/2}$

For confidence intervals, only the **positive** $z$ score is used in the formula.

```{r s7-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# 90%, 94%, 95%, 98% and 99% confidence
qnorm(c(0.95, 0.97, 0.975, 0.99, 0.995))
```

## Example 7-3 {#example-7-3}

:::{.callout-warning}
## Container Dwell Time at Kaohsiung Port

The following data represent the dwell time, in hours, of a random sample of 30 import containers at the Port of Kaohsiung. Assume the population standard deviation is 12.6 hours. Find the **90% confidence interval** of the mean.

```
36.42    28.15    51.70
19.83    44.26    33.08
62.55    25.71    40.19
47.63    30.94    22.38
55.02    38.47    29.66
43.81    26.50    58.24
34.75    49.12    21.07
31.89    45.33    27.46
53.18    37.62    24.91
41.05    32.27    46.88
```

*(hypothetical data)*
:::

---

## Example 7-3

:::{.callout-tip}
## Solution

**Step 1** Find the mean for the data: $\bar{X} = 38.002$ hours. The population standard deviation is 12.6 hours.

**Step 2** Find $\alpha/2$. For a 90% confidence interval, $\alpha = 1 - 0.90 = 0.10$ and $\alpha/2 = 0.10/2 = 0.05$.

**Step 3** Find $z_{\alpha/2}$. The area to the left of the critical value is $1 - 0.05 = 0.9500$; that is the column you look up in Table A-4, and `qnorm(0.95)` returns the same critical value, $z_{\alpha/2} = \mathbf{1.645}$.

**Step 4** Substitute in the formula.

$$38.002 - 1.645\left(\frac{12.6}{\sqrt{30}}\right) < \mu < 38.002 + 1.645\left(\frac{12.6}{\sqrt{30}}\right)$$
$$38.002 - 3.784 < \mu < 38.002 + 3.784$$
$$\mathbf{34.218 < \mu < 41.786}$$
:::

---

## Example 7-3

:::{.callout-tip}
## Solution

**Compute in R**

```{r e7-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
dwell_hours <- c(36.42, 28.15, 51.70, 19.83, 44.26, 33.08, 62.55, 25.71, 40.19, 47.63,
                 30.94, 22.38, 55.02, 38.47, 29.66, 43.81, 26.50, 58.24, 34.75, 49.12,
                 21.07, 31.89, 45.33, 27.46, 53.18, 37.62, 24.91, 41.05, 32.27, 46.88)
x_bar  <- mean(dwell_hours)
margin <- qnorm(0.95) * 12.6 / sqrt(30)

x_bar
margin
x_bar - margin
x_bar + margin
```

One can be **90% confident** that the mean dwell time of all import containers is between 34.218 and 41.786 hours, based on a sample of 30 containers.
:::

## Sample Size for Estimating a Mean

How large a sample is necessary to make an accurate estimate? The answer depends on three things: the **margin of error**, the **population standard deviation**, and the **degree of confidence**.

The formula is derived from the margin-of-error formula $E = z_{\alpha/2}\left(\sigma/\sqrt{n}\right)$, solved for $n$:

$$E\sqrt{n} = z_{\alpha/2}\,\sigma \quad \Longrightarrow \quad \sqrt{n} = \frac{z_{\alpha/2} \cdot \sigma}{E}$$

:::{.callout-important}
## Formula for the Minimum Sample Size Needed for an Interval Estimate of the Population Mean

$$n = \left(\frac{z_{\alpha/2} \cdot \sigma}{E}\right)^2$$

where $E$ is the margin of error. If necessary, **round the answer up** to obtain a whole number: if there is any fraction or decimal portion in the answer, use the next whole number for $n$.
:::

## Example 7-4 {#example-7-4}

:::{.callout-warning}
## Sample Size for Delivery Time

An operations manager at a food-delivery platform in Taichung wishes to estimate the average delivery time per order to within **2 minutes**. She wishes to be **99% confident**, and a previous study puts the standard deviation at 4.5 minutes. How many orders should she sample?

*(hypothetical data)*
:::

---

## Example 7-4

:::{.callout-tip}
## Solution

Since $\alpha = 0.01$ (or $1 - 0.99$), $z_{\alpha/2} = 2.576$ and $E = 2$. Substitute in the formula.

$$n = \left(\frac{z_{\alpha/2} \cdot \sigma}{E}\right)^2 = \left[\frac{(2.576)(4.5)}{2}\right]^2 = 33.59$$

```{r e7-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- (qnorm(0.995) * 4.5 / 2)^2

n
ceiling(n)
```

Round the value up to **34**. To be 99% confident that the estimate is within 2 minutes of the true mean, the manager needs to sample at least 34 orders.
:::

## Notes on Sample Size

:::{.callout-warning}
## Caution

In most cases in statistics we round off; however, when determining sample size we **always round up** to the next whole number.
:::

:::{.callout-important}
## Practical Points

- When the population is large or infinite, or when sampling is done with replacement, the **size of the population is irrelevant** to the sample size calculation.
- The formula requires $\sigma$. When $\sigma$ is unknown, estimate it with the standard deviation $s$ from a previous sample, or by dividing the **range by 4**.
- An interval estimate is often reported in words: "an American family of two spends an average of \$84 per week for groceries, accurate within \$3 with 95% confidence" means $81 < \mu < 87$.
:::

# Section 7-3: Confidence Intervals for the Mean When $\sigma$ Is Unknown {background-color="#2c4a6e"}

## The $t$ Distribution

Usually $\sigma$ is unknown and is estimated by $s$. Critical values **greater** than $z_{\alpha/2}$ are then needed, especially for small $n$; they come from the **Student's $t$ distribution**.

:::{.callout-note}
## Characteristics of the $t$ Distribution

**Similar to the standard normal distribution:**

1. It is bell-shaped.
2. It is symmetric about the mean.
3. The mean, median and mode are equal to 0 and are located at the center of the distribution.
4. The curve approaches but never touches the $x$ axis.

**Different from the standard normal distribution:**

1. The variance is greater than 1.
2. The $t$ distribution is a **family of curves** based on degrees of freedom, which is related to sample size.
3. As the sample size increases, the $t$ distribution approaches the standard normal distribution.
:::

## The $t$ Family of Curves

**Prepare data**

```{r s7-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(-4, 4, by = 0.01)
curves <- data.frame(x, z = dnorm(x), t5 = dt(x, 5), t20 = dt(x, 20))
```

---

## The $t$ Family of Curves

**Draw figure**

```{r s7-3-2, echo=T, warning=FALSE, out.width="80%", fig.align='center', fig.height=3.5}
ggplot(curves, aes(x)) +
  geom_line(aes(y = z)) +
  geom_line(aes(y = t5), linetype = "dashed") +
  geom_line(aes(y = t20), linetype = "dotted") +
  labs(title = "The t Family of Curves", x = "Value", y = "Density")
```

The solid line is the $z$ curve, the dashed line is the $t$ curve for $\text{d.f.} = 5$, and the dotted line is the $t$ curve for $\text{d.f.} = 20$.

## Degrees of Freedom

:::{.callout-note}
## Degrees of Freedom

The **degrees of freedom** are the number of values that are free to vary after a sample statistic has been computed, and they tell the researcher which specific curve to use when a distribution consists of a family of curves.
:::

If the mean of 5 values is 10, then 4 of the 5 values are free to vary; once 4 are chosen the fifth must be a specific number so that the sum is 50. Hence $\text{d.f.} = 5 - 1 = 4$.

The symbol **d.f.** is used for degrees of freedom. For a confidence interval for the mean, $\text{d.f.} = n - 1$. (For some tests later in the book the degrees of freedom are *not* $n - 1$.)

:::{.callout-warning}
## Caution

The values $t_{\alpha/2}$ come from the row labelled **Confidence Intervals** at the top of Table A-5. The rows labelled *One tail* and *Two tails* belong to Chapter 8 and should not be used here. When d.f. is greater than 30 and falls between two printed rows, the table convention is to **round d.f. down** to the nearest row (a conservative choice): d.f. = 68 is read as 65. That convention is a lookup aid; `qt()` takes the exact d.f., and where the two differ these slides report the value R computes.
:::

## Example 7-5 {#example-7-5}

:::{.callout-warning}
## Finding a $t$ Critical Value

Find the $t_{\alpha/2}$ value for a **95% confidence interval** when the sample size is 18.
:::

---

## Example 7-5

:::{.callout-tip}
## Solution

The $\text{d.f.} = 18 - 1$, or 17. Find 17 in the left column of Table A-5 and 95% in the row labelled **Confidence Intervals**; `qt(0.975, df = 17)` returns the same critical value, $t_{\alpha/2} = \mathbf{2.110}$.

```{r e7-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
qt(0.975, df = 17)
```

In R a confidence-interval critical value is always `qt(1 - alpha/2, df = n - 1)`; for a 95% interval that is `qt(0.975, df = n - 1)`.
:::

## Formula for the Mean When $\sigma$ Is Unknown

:::{.callout-important}
## Formula for a Specific Confidence Interval for the Mean When $\sigma$ Is Unknown

$$\bar{X} - t_{\alpha/2}\left(\frac{s}{\sqrt{n}}\right) < \mu < \bar{X} + t_{\alpha/2}\left(\frac{s}{\sqrt{n}}\right)$$

The degrees of freedom are $n - 1$.
:::

:::{.callout-important}
## Assumptions for Finding a Confidence Interval for a Mean When $\sigma$ Is Unknown

1. The sample is a random sample.
2. Either $n \geq 30$ or the population is normally distributed when $n < 30$.
:::

## Example 7-6 {#example-7-6}

:::{.callout-warning}
## Weekday Occupancy of Taipei Hotels

A random sample of 16 business hotels in Taipei had a mean weekday occupancy rate of 72.4%. The standard deviation of the sample was 6.8 percentage points. Assume the variable is normally distributed. Find the **95% confidence interval** of the population mean weekday occupancy rate.

*(hypothetical data)*
:::

---

## Example 7-6

:::{.callout-tip}
## Solution

$\bar{X} = 72.4$, $s = 6.8$, $n = 16$.

Since $\sigma$ is unknown and $s$ must replace it, the $t$ distribution is used. With 15 degrees of freedom, $t_{\alpha/2} = 2.131$.

$$72.4 - 2.131\left(\frac{6.8}{\sqrt{16}}\right) < \mu < 72.4 + 2.131\left(\frac{6.8}{\sqrt{16}}\right)$$
$$72.4 - 3.6 < \mu < 72.4 + 3.6$$
$$\mathbf{68.8 < \mu < 76.0}$$
:::

---

## Example 7-6

:::{.callout-tip}
## Solution

**Compute in R**

```{r e7-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
margin <- qt(0.975, df = 15) * 6.8 / sqrt(16)

margin
72.4 - margin
72.4 + margin
```

One can be **95% confident** that the population mean weekday occupancy rate of Taipei business hotels is between 68.8% and 76.0%.
:::

## Example 7-7 {#example-7-7}

:::{.callout-warning}
## Monthly Revenue of a Night-Market Stall

The data represent the monthly revenue, in thousands of NT dollars, of a night-market beverage stall for a random sample of 7 months. Find the **99% confidence interval** for the mean monthly revenue.

```
268    292    305    318    331    347    384
```

*(hypothetical data)*
:::

---

## Example 7-7

:::{.callout-tip}
## Solution

**Step 1** Find the mean and standard deviation for the data: $\bar{X} = 320.7$ and $s = 38.0$.

**Step 2** Find $t_{\alpha/2}$ for a 99% confidence interval with $\text{d.f.} = 6$. Table A-5 and `qt(0.995, df = 6)` both give **3.707**.

**Step 3** Substitute in the formula and solve.

$$320.7 - 3.707\left(\frac{38.0}{\sqrt{7}}\right) < \mu < 320.7 + 3.707\left(\frac{38.0}{\sqrt{7}}\right)$$
$$320.7 - 53.2 < \mu < 320.7 + 53.2$$
$$\mathbf{267.5 < \mu < 373.9}$$
:::

---

## Example 7-7

:::{.callout-tip}
## Solution

**Compute in R**

```{r e7-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
monthly_revenue <- c(268, 292, 305, 318, 331, 347, 384)
x_bar  <- mean(monthly_revenue)
s      <- sd(monthly_revenue)
margin <- qt(0.995, df = 6) * s / sqrt(7)

x_bar
s
margin
x_bar - margin
x_bar + margin
```

One can be **99% confident** that the mean monthly revenue of the stall is between 267.5 and 373.9 thousand NT dollars, based on a sample of 7 months.
:::

## When to Use $z$ or $t$

:::{.callout-important}
## Decision Rule (Figure 7-8)

- **Is $\sigma$ known?** Yes: use $z_{\alpha/2}$ values and $\sigma$ in the formula.
- **Is $\sigma$ known?** No: use $t_{\alpha/2}$ values and $s$ in the formula.

In either case, if $n < 30$ the variable must be normally distributed.
:::

---

## When to Use $z$ or $t$

| Situation | Use | R |
|--------|---|---------|
| $\sigma$ known, $n \geq 30$ | $z$ and $\sigma$ | `qnorm(1 - alpha/2)` |
| $\sigma$ known, $n < 30$ (normal) | $z$ and $\sigma$ | `qnorm(1 - alpha/2)` |
| $\sigma$ unknown, $n \geq 30$ | $t$ and $s$ | `qt(1 - alpha/2, df = n - 1)` |
| $\sigma$ unknown, $n < 30$ (normal) | $t$ and $s$ | `qt(1 - alpha/2, df = n - 1)` |

# Section 7-4: Confidence Intervals and Sample Size for Proportions {background-color="#2c4a6e"}

## Proportions

A **proportion** represents a part of a whole; it can be expressed as a fraction, a decimal or a percentage. If 12% of the parcels a courier delivers in Taipei go to a convenience-store pickup point, then $12\% = 0.12 = \frac{12}{100} = \frac{3}{25}$, and the probability that a randomly selected parcel goes to a pickup point is 0.12.

:::{.callout-note}
## Symbols Used in Proportion Notation

$$p = \text{population proportion} \qquad \hat{p} = \text{sample proportion (read "p hat")}$$

For a sample proportion,

$$\hat{p} = \frac{X}{n} \qquad \text{and} \qquad \hat{q} = \frac{n - X}{n} \quad \text{or} \quad \hat{q} = 1 - \hat{p}$$

where $X$ is the number of sample units that possess the characteristic of interest and $n$ is the sample size.
:::

When $\hat{p}$ and $\hat{q}$ are in decimal or fraction form, $\hat{p} + \hat{q} = 1$; when they are percentages, $\hat{p} + \hat{q} = 100\%$. The same relations hold for the population: $p + q = 1$.

## Example 7-8 {#example-7-8}

:::{.callout-warning}
## Convenience-Store Pickup

A random sample of 180 online shoppers on a Taiwanese e-commerce platform found that 63 chose convenience-store pickup instead of home delivery. Find $\hat{p}$ and $\hat{q}$, where $\hat{p}$ is the proportion of shoppers who chose convenience-store pickup.

*(hypothetical data)*
:::

---

## Example 7-8

:::{.callout-tip}
## Solution

In this case, $X = 63$ and $n = 180$.

$$\hat{p} = \frac{X}{n} = \frac{63}{180} = 0.35 = 35\%$$
$$\hat{q} = \frac{n - X}{n} = \frac{180 - 63}{180} = \frac{117}{180} = 0.65 = 65\%$$

```{r e7-8-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
63 / 180
(180 - 63) / 180
```

So **65%** of the shoppers chose home delivery instead.
:::

## Confidence Intervals for a Proportion

For a point estimate of $p$, the sample proportion $\hat{p}$ is used; it is unbiased, consistent and relatively efficient. When $n$ is no more than 5% of the population size, the sampling distribution of $\hat{p}$ is approximately normal with mean $p$ and standard deviation $\sqrt{pq/n}$.

:::{.callout-important}
## Formula for a Specific Confidence Interval for a Proportion

$$\hat{p} - z_{\alpha/2}\sqrt{\frac{\hat{p}\hat{q}}{n}} < p < \hat{p} + z_{\alpha/2}\sqrt{\frac{\hat{p}\hat{q}}{n}}$$

when $n\hat{p}$ and $n\hat{q}$ are each greater than or equal to 5. The margin of error is $E = z_{\alpha/2}\sqrt{\hat{p}\hat{q}/n}$.
:::

---

## Confidence Intervals for a Proportion

:::{.callout-important}
## Assumptions for Finding a Confidence Interval for a Population Proportion

1. The sample is a random sample.
2. The conditions for a binomial experiment are satisfied (see Chapter 5).
:::

:::{.callout-important}
## Rounding Rule for a Confidence Interval for a Proportion

Round off to **three decimal places**.
:::

:::{.callout-warning}
## Caution

When a specific percentage is given in a problem, that percentage becomes $\hat{p}$ once it is changed to a decimal. If 12% of the applicants were men, then $\hat{p} = 0.12$.
:::

## Example 7-9 {#example-7-9}

:::{.callout-warning}
## Transfer Experience at Taoyuan Airport

A passenger survey at Taoyuan International Airport sampled 1250 transfer passengers, and 375 of them rated the transfer experience as excellent. Find the **90% confidence interval** of the true proportion of transfer passengers who rate the experience as excellent.

*(hypothetical data)*
:::

---

## Example 7-9

:::{.callout-tip}
## Solution

**Step 1** Determine $\hat{p}$ and $\hat{q}$.

$$\hat{p} = \frac{X}{n} = \frac{375}{1250} = 0.30 \qquad \hat{q} = 1 - \hat{p} = 1.00 - 0.30 = 0.70$$

**Step 2** Determine the critical value. $\alpha = 1 - 0.90 = 0.10$, so $\alpha/2 = 0.05$ and $z_{\alpha/2} = 1.645$.

**Step 3** Substitute in the formula.

$$0.30 - 1.645\sqrt{\frac{(0.30)(0.70)}{1250}} < p < 0.30 + 1.645\sqrt{\frac{(0.30)(0.70)}{1250}}$$
$$0.30 - 0.021 < p < 0.30 + 0.021$$
$$\mathbf{0.279 < p < 0.321} \qquad \text{or} \qquad 27.9\% < p < 32.1\%$$
:::

---

## Example 7-9

:::{.callout-tip}
## Solution

**Compute in R**

```{r e7-9-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
p_hat  <- 375 / 1250
margin <- qnorm(0.95) * sqrt(p_hat * (1 - p_hat) / 1250)

p_hat
margin
p_hat - margin
p_hat + margin
```

You can be **90% confident** that the percentage of transfer passengers who rate the experience as excellent is between 27.9% and 32.1%.
:::

## Example 7-10 {#example-7-10}

:::{.callout-warning}
## Buying from Overseas Sellers

A survey of 2400 online shoppers in Taiwan found that 64% had bought at least one item from an overseas seller in the past year. Find the **90% confidence interval** of the true proportion of shoppers who buy from overseas sellers.

*(hypothetical data)*
:::

---

## Example 7-10

:::{.callout-tip}
## Solution

**Step 1** Determine $\hat{p}$ and $\hat{q}$. Here $\hat{p}$ is already given: 64%, or 0.64. Then $\hat{q} = 1 - 0.64 = 0.36$.

**Step 2** Determine the critical value. $\alpha = 1 - 0.90 = 0.10$, so $\alpha/2 = 0.05$ and $z_{\alpha/2} = 1.645$.

**Step 3** Substitute in the formula.

$$0.64 - 1.645\sqrt{\frac{(0.64)(0.36)}{2400}} < p < 0.64 + 1.645\sqrt{\frac{(0.64)(0.36)}{2400}}$$
$$0.64 - 0.016 < p < 0.64 + 0.016$$
$$\mathbf{0.624 < p < 0.656} \qquad \text{or} \qquad 62.4\% < p < 65.6\%$$
:::

---

## Example 7-10

:::{.callout-tip}
## Solution

**Compute in R**

```{r e7-10-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
margin <- qnorm(0.95) * sqrt(0.64 * 0.36 / 2400)

margin
0.64 - margin
0.64 + margin
```

You can say with **90% confidence** that the true percentage of shoppers who buy from overseas sellers is between 62.4% and 65.6%.
:::

## Sample Size for Proportions

:::{.callout-important}
## Formula for Minimum Sample Size Needed for Interval Estimate of a Population Proportion

$$n = \hat{p}\hat{q}\left(\frac{z_{\alpha/2}}{E}\right)^2$$

If necessary, **round up** to obtain a whole number.
:::

There are two situations to consider:

1. If some approximation of $\hat{p}$ is known (for example, from a previous study), that value is used in the formula.
2. If no approximation of $\hat{p}$ is known, use $\hat{p} = 0.5$. This gives a sample size large enough to guarantee an accurate prediction, because the product $\hat{p}\hat{q}$ is at a maximum when $\hat{p} = \hat{q} = 0.5$.

The disadvantage of the second method is that it can lead to a larger sample size than is necessary.

---

## Sample Size for Proportions

**Prepare data**

```{r s7-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
p_hat <- seq(0.1, 0.9, by = 0.1)

p_hat
1 - p_hat
p_hat * (1 - p_hat)
```

The product $\hat{p}\hat{q}$ is largest at $\hat{p} = \hat{q} = 0.5$. Using that maximum value yields the largest possible value of $n$ for a given margin of error and a given confidence interval.

## Example 7-11 {#example-7-11}

:::{.callout-warning}
## Mobile Payment at Checkout

A market-research team wishes to estimate, with **95% confidence**, the proportion of Taiwanese online shoppers who pay with a mobile wallet at checkout. A previous study shows that 65% of shoppers did so. The team wishes to be accurate within **3%** of the true proportion. Find the minimum sample size necessary.

*(hypothetical data)*

Continued in [Example 7-12](#example-7-12)
:::

---

## Example 7-11

:::{.callout-tip}
## Solution

Since $z_{\alpha/2} = 1.960$, $E = 0.03$, $\hat{p} = 0.65$ and $\hat{q} = 0.35$, then

$$n = \hat{p}\hat{q}\left(\frac{z_{\alpha/2}}{E}\right)^2 = (0.65)(0.35)\left(\frac{1.960}{0.03}\right)^2 = 971.04$$

```{r e7-11-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 0.65 * 0.35 * (qnorm(0.975) / 0.03)^2

n
ceiling(n)
```

Which, when rounded up, is **972** shoppers to interview.
:::

## Example 7-12 {#example-7-12}

:::{.callout-warning}
## No Previous Study

In [Example 7-11](#example-7-11) assume that no previous study was done, so no approximation of $\hat{p}$ is available. Find the minimum sample size necessary to be accurate within **3%** of the true proportion.
:::

---

## Example 7-12

:::{.callout-tip}
## Solution

Here we do not know the values of $\hat{p}$ and $\hat{q}$, so we use $\hat{p} = 0.5$ and $\hat{q} = 0.5$, with $E = 0.03$ and $z_{\alpha/2} = 1.960$.

$$n = \hat{p}\hat{q}\left(\frac{z_{\alpha/2}}{E}\right)^2 = (0.5)(0.5)\left(\frac{1.960}{0.03}\right)^2 = 1067.07$$

```{r e7-12-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 0.5 * 0.5 * (qnorm(0.975) / 0.03)^2

n
ceiling(n)
```

Hence **1068** shoppers must be interviewed when $\hat{p}$ is unknown — 96 more than are needed when $\hat{p}$ is known.
:::

---

## Sample Size for Proportions

:::{.callout-warning}
## Caution

In determining the sample size, the **size of the population is irrelevant**. Only the degree of confidence and the margin of error are necessary to make the determination.
:::

# Section 7-5: Confidence Intervals for Variances and Standard Deviations {background-color="#2c4a6e"}

## The Chi-Square Distribution

The variance and standard deviation of a variable are as important as the mean. When pipes that must fit together are made, the variation of the diameters must stay small; in medicines, the variance of the amount of medication in each pill decides whether patients receive the proper dosage.

These confidence intervals need a new distribution: the **chi-square distribution**, written $\chi^2$ (Greek letter chi, pronounced "ki"). It is obtained from the values of $(n-1)s^2/\sigma^2$ for random samples from a normally distributed population with variance $\sigma^2$.

:::{.callout-note}
## Characteristics of the Chi-Square Distribution

1. All chi-square values are greater than or equal to 0.
2. The chi-square distribution is a **family of curves** based on the degrees of freedom.
3. The area under each chi-square distribution curve is equal to 1.
4. The chi-square distributions are **positively skewed**.

At about 100 degrees of freedom the chi-square distribution becomes somewhat symmetric.
:::

## The Chi-Square Family of Curves

**Prepare data**

```{r s7-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(0, 40, by = 0.1)
curves <- data.frame(x, df1 = dchisq(x, 1), df4 = dchisq(x, 4),
                     df9 = dchisq(x, 9), df15 = dchisq(x, 15))
```

---

## The Chi-Square Family of Curves

**Draw figure**

```{r s7-5-2, echo=T, warning=FALSE, out.width="80%", fig.align='center', fig.height=3.5}
ggplot(curves, aes(x)) +
  geom_line(aes(y = df1)) +
  geom_line(aes(y = df4), linetype = "dashed") +
  geom_line(aes(y = df9), linetype = "dotted") +
  geom_line(aes(y = df15), linetype = "dotdash") +
  coord_cartesian(ylim = c(0, 0.5)) +
  labs(title = "The Chi-Square Family of Curves",
       x = "Chi-square value", y = "Density")
```

The solid line is $\text{d.f.} = 1$, the dashed line $\text{d.f.} = 4$, the dotted line $\text{d.f.} = 9$, and the dot-dash line $\text{d.f.} = 15$.

## Finding the Two Critical Values

Because the chi-square distribution is **not symmetric**, two different values are used in the formulas: $\chi^2_{\text{left}}$ and $\chi^2_{\text{right}}$.

Table A-6 lists degrees of freedom in the left column and the **area to the right** of the critical value across the top row. For a 95% confidence interval:

1. Change 95% to a decimal and subtract from 1: $1 - 0.95 = 0.05$.
2. Divide by 2: $\alpha/2 = 0.05/2 = 0.025$. This column gives $\chi^2_{\text{right}}$.
3. Subtract $\alpha/2$ from 1: $1 - 0.025 = 0.975$. This column gives $\chi^2_{\text{left}}$.
4. Read the row for $\text{d.f.} = n - 1$.

:::{.callout-warning}
## Caution

If the number for the degrees of freedom is not printed in Table A-6, use the **closest lower value** in the table. For example, for $\text{d.f.} = 53$, use $\text{d.f.} = 50$. This is a conservative approach for a paper lookup; `qchisq()` takes the exact d.f., and where the two differ these slides report the value R computes.
:::

## Example 7-13 {#example-7-13}

:::{.callout-warning}
## Finding Chi-Square Critical Values

Find the values for $\chi^2_{\text{right}}$ and $\chi^2_{\text{left}}$ for a **95% confidence interval** when $n = 20$.
:::

---

## Example 7-13

:::{.callout-tip}
## Solution

To find $\chi^2_{\text{right}}$, subtract $1 - 0.95 = 0.05$; then divide 0.05 by 2 to get 0.025.
To find $\chi^2_{\text{left}}$, subtract $1 - 0.025 = 0.975$.

Then use the 0.975 and 0.025 columns of Table A-6 with $\text{d.f.} = n - 1 = 20 - 1 = 19$.

$$\chi^2_{\text{right}} = 32.852 \qquad \chi^2_{\text{left}} = 8.907$$

```{r e7-13-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
qchisq(0.975, df = 19)
qchisq(0.025, df = 19)
```
:::

## Formulas for Variance and Standard Deviation

:::{.callout-important}
## Formula for the Confidence Interval for a Variance

$$\frac{(n-1)s^2}{\chi^2_{\text{right}}} < \sigma^2 < \frac{(n-1)s^2}{\chi^2_{\text{left}}} \qquad \text{d.f.} = n - 1$$
:::

:::{.callout-important}
## Formula for the Confidence Interval for a Standard Deviation

$$\sqrt{\frac{(n-1)s^2}{\chi^2_{\text{right}}}} < \sigma < \sqrt{\frac{(n-1)s^2}{\chi^2_{\text{left}}}} \qquad \text{d.f.} = n - 1$$
:::

Useful estimates for $\sigma^2$ and $\sigma$ are $s^2$ and $s$. If the problem gives the sample standard deviation $s$, be sure to **square it**; if the problem gives the sample variance $s^2$, do **not** square it, since the variance is already in square units.

---

## Formulas for Variance and Standard Deviation

:::{.callout-important}
## Assumptions for Finding a Confidence Interval for a Variance or Standard Deviation

1. The sample is a random sample.
2. The population **must** be normally distributed.
:::

:::{.callout-important}
## Rounding Rule for a Confidence Interval for a Variance or Standard Deviation

When computing from **raw data**, round off to **one more decimal place** than the number of decimal places in the original data.

When computing from a **sample variance or standard deviation**, round off to the **same number of decimal places** as given for the sample variance or standard deviation.
:::

## Example 7-14 {#example-7-14}

:::{.callout-warning}
## Variation in Batch Defect Rates

A quality engineer at an electronics plant sampled 30 production batches and found that the standard deviation of the defect rate is 1.6 defects per thousand units. Find the **95% confidence interval** of the variance and standard deviation of the defect rate. Assume the variable is normally distributed.

*(hypothetical data)*
:::

---

## Example 7-14

:::{.callout-tip}
## Solution

Since $\alpha = 0.05$ and the degrees of freedom is 29, the two critical values are $\chi^2_{\text{left}} = 16.047$ and $\chi^2_{\text{right}} = 45.722$.

$$\frac{(30-1)(1.6)^2}{45.722} < \sigma^2 < \frac{(30-1)(1.6)^2}{16.047}$$
$$\mathbf{1.62 < \sigma^2 < 4.63}$$

For the standard deviation, take the square root of each limit:

$$\mathbf{1.27 < \sigma < 2.15}$$
:::

---

## Example 7-14

:::{.callout-tip}
## Solution

**Compute in R**

```{r e7-14-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
chi_right <- qchisq(0.975, df = 29)
chi_left  <- qchisq(0.025, df = 29)

chi_right
chi_left
(30 - 1) * 1.6^2 / chi_right
(30 - 1) * 1.6^2 / chi_left
sqrt((30 - 1) * 1.6^2 / chi_right)
sqrt((30 - 1) * 1.6^2 / chi_left)
```

One can be **95% confident** that the true variance of the defect rate is between 1.62 and 4.63, and that the true standard deviation is between 1.27 and 2.15 defects per thousand units, based on a sample of 30 batches.
:::

## Example 7-15 {#example-7-15}

:::{.callout-warning}
## Daily Call Volume at a Service Centre

Find the **90% confidence interval** for the variance and standard deviation of the number of inbound calls received per day at a customer-service centre. A random sample of 10 days has been used. Assume the distribution is approximately normal.

```
124   127   130   130   135
140   148   155   158   163
```

*(hypothetical data)*
:::

---

## Example 7-15

:::{.callout-tip}
## Solution

**Step 1** Find the variance for the data: $s^2 = 198$.

**Step 2** Find $\chi^2_{\text{right}}$ and $\chi^2_{\text{left}}$ from Table A-6, using $10 - 1 = 9$ degrees of freedom. With $\alpha/2 = 0.05$ and $1 - \alpha/2 = 0.95$: $\chi^2_{\text{right}} = 16.919$ and $\chi^2_{\text{left}} = 3.325$.

**Step 3** Substitute in the formula.

$$\frac{(10-1)(198)}{16.919} < \sigma^2 < \frac{(10-1)(198)}{3.325}$$
$$\mathbf{105.3 < \sigma^2 < 535.9}$$
$$\sqrt{105.3} < \sigma < \sqrt{535.9} \qquad \Longrightarrow \qquad \mathbf{10.3 < \sigma < 23.1}$$
:::

---

## Example 7-15

:::{.callout-tip}
## Solution

**Compute in R**

```{r e7-15-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
daily_calls <- c(124, 127, 130, 130, 135, 140, 148, 155, 158, 163)
chi_right <- qchisq(0.95, df = 9)
chi_left  <- qchisq(0.05, df = 9)

var(daily_calls)
chi_right
chi_left
(10 - 1) * 198 / chi_right
(10 - 1) * 198 / chi_left
sqrt((10 - 1) * 198 / chi_right)
sqrt((10 - 1) * 198 / chi_left)
```

One can be **90% confident** that the standard deviation of the daily number of inbound calls is between 10.3 and 23.1 calls, based on a random sample of 10 days.
:::

## Important Terms

:::{.callout-important}
## Chapter 7 Vocabulary

assumptions · chi-square distribution · confidence interval · confidence level · consistent estimator · degrees of freedom · estimation · estimator · interval estimate · margin of error · point estimate · proportion · relatively efficient estimator · robust · $t$ distribution · unbiased estimator
:::

## Key Formulas

:::{.callout-important}
## Confidence Intervals for a Mean

Confidence interval of the mean when $\sigma$ is known (when $n \geq 30$, $s$ can be used if $\sigma$ is unknown):

$$\bar{X} - z_{\alpha/2}\left(\frac{\sigma}{\sqrt{n}}\right) < \mu < \bar{X} + z_{\alpha/2}\left(\frac{\sigma}{\sqrt{n}}\right)$$

Confidence interval of the mean when $\sigma$ is unknown, with $\text{d.f.} = n - 1$:

$$\bar{X} - t_{\alpha/2}\left(\frac{s}{\sqrt{n}}\right) < \mu < \bar{X} + t_{\alpha/2}\left(\frac{s}{\sqrt{n}}\right)$$

Sample size for means, where $E$ is the margin of error:

$$n = \left(\frac{z_{\alpha/2} \cdot \sigma}{E}\right)^2$$
:::

---

## Key Formulas

:::{.callout-important}
## Proportions, Variances and Standard Deviations

Confidence interval for a proportion, where $\hat{p} = X/n$ and $\hat{q} = 1 - \hat{p}$:

$$\hat{p} - z_{\alpha/2}\sqrt{\frac{\hat{p}\hat{q}}{n}} < p < \hat{p} + z_{\alpha/2}\sqrt{\frac{\hat{p}\hat{q}}{n}} \qquad n = \hat{p}\hat{q}\left(\frac{z_{\alpha/2}}{E}\right)^2$$

Confidence interval for a variance and for a standard deviation, with $\text{d.f.} = n - 1$:

$$\frac{(n-1)s^2}{\chi^2_{\text{right}}} < \sigma^2 < \frac{(n-1)s^2}{\chi^2_{\text{left}}} \qquad \sqrt{\frac{(n-1)s^2}{\chi^2_{\text{right}}}} < \sigma < \sqrt{\frac{(n-1)s^2}{\chi^2_{\text{left}}}}$$
:::

## Key Takeaways

:::{.callout-important}
- **Estimation** uses a sample statistic to estimate a parameter; a good estimator is **unbiased**, **consistent** and **relatively efficient**, and $\bar{X}$ is the best point estimate of $\mu$
- A **point estimate** is a single value; an **interval estimate** is a range, and its **confidence level** is the long-run probability that such intervals capture the parameter
- The **margin of error** (maximum error of the estimate) is the largest likely gap between the point estimate and the parameter; higher confidence means a **wider** interval
- Use $z_{\alpha/2}$ and $\sigma$ when $\sigma$ is known; use $t_{\alpha/2}$ and $s$ with $\text{d.f.} = n - 1$ when $\sigma$ is unknown — and if $n < 30$ the variable must be normally distributed
- Sample sizes come from $n = \left(z_{\alpha/2}\sigma/E\right)^2$ and $n = \hat{p}\hat{q}\left(z_{\alpha/2}/E\right)^2$, and are **always rounded up**; the population size is irrelevant
- For proportions, $\hat{p} = X/n$ and $\hat{q} = 1 - \hat{p}$, the interval requires $n\hat{p} \geq 5$ and $n\hat{q} \geq 5$, and limits are rounded to **three decimal places**; when $\hat{p}$ is unknown use $\hat{p} = 0.5$
- Confidence intervals for $\sigma^2$ and $\sigma$ use the **positively skewed** chi-square distribution with two critical values, and **require** a normally distributed population
:::

## Acknowledgement

-   **Copyright notice.** These teaching materials follow the organization and terminology of **Bluman, A. G. (2023). Elementary statistics: A step by step approach (11th ed.). McGraw Hill**. All rights in the original work are reserved by its authors and publishers.

-   **Original examples.** Every worked example, data set, and R script in these slides was written for this course. The data are hypothetical unless stated otherwise.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
