---
title: "Master Specification for Auto-Generating Quarto Revealjs Statistics Slides"
subtitle: "A complete, self-contained prompt for AI-assisted chapter slide generation"
author: "Yu-You Liou"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

> **How to use this document.** Paste the whole file to the AI agent and say: *"Generate `chX.qmd` (and `chX-zh.qmd`) for Chapter X following this specification exactly."* The agent MUST read the document end-to-end before producing any slide content, and MUST run the validation checklist in Section 12 before returning a file. This specification is the single source of truth; it supersedes any conflicting habit or default.

---

# Context and Goals

These slides accompany a university-level introductory statistics course based on **Bluman, A. G. (2012). *Elementary Statistics: A Step by Step Approach* (8th ed.). McGraw-Hill**. Each chapter is delivered as a Quarto Revealjs (`.qmd`) presentation that renders to a self-contained `.html` deck.

Design goals, in priority order:

1. **Structural correctness** — the deck must render without error and every callout box must be perfectly balanced (this is the single most common failure; see Section 5).
2. **Pedagogical clarity** — definitions, rules, and worked R examples follow a fixed, predictable layout so students always know where to look.
3. **Consistency across chapters** — `ch2`–`ch5` are the **clean reference standard**. When in doubt, match those files.
4. **Bilingual parity** — every chapter has an English `chX.qmd` and a Traditional-Chinese `chX-zh.qmd` that are structurally identical (see Section 11).

---

# YAML Front Matter (fixed)

Every slide file begins with the YAML header below. Modify **only**: `subtitle` (chapter number and title) and the language-switch link target (`chX-zh.html` in the English file, `chX.html` in the Chinese file). Everything else is fixed and must be reproduced verbatim, including indentation.

````yaml
---
title: "Statistics"
subtitle: "Chapter X: <Chapter Title>"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="chX-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---
````

> In the **Chinese** file, the link text becomes `[English version]` and the target becomes `chX.html`.

Immediately after the YAML header, include the global setup chunk. Add only the libraries the chapter actually uses; `tidyverse` is always present.

````````{r,eval= F}
```{r setup, include=FALSE}
library(tidyverse)
# add only what the chapter needs, e.g.:
# library(ggplot2)
```
````````

> **Indentation warning.** The `format: revealjs:` block must keep its nested indentation. A known broken file (`ch-example.qmd`) lost the indentation under `revealjs:` so the options were ignored — do not replicate that. Verify the nesting matches the block above exactly.

---

# Overall Slide Sequence

Slides must appear in this mandatory order:

```
Title slide (auto-generated from YAML)
  ↓
## Overview                         (theme paragraph + Section/Topics table)
  ↓
# Section X-1: <Name> {background-color="#2c4a6e"}   (divider)
   ## content slides, definitions, rules, examples ...
  ↓
# Section X-2: <Name> {background-color="#2c4a6e"}
   ...
  ↓
(additional sections)
  ↓
## Key Formulas      (only if the chapter has multiple central formulae)
  ↓
## Key Takeaways     (mandatory, penultimate)
  ↓
## Acknowledgement   (mandatory, final, fixed text)
```

Slides are separated by a bare horizontal rule: three hyphens `---` alone on a line. New top-level content opens with a `##` heading; a new section opens with a single `#` divider.

---

# The Overview Slide

The slide right after the title is **Overview**: one short paragraph naming the chapter's central theme, followed by a two-column table summarizing each section.

```markdown
## Overview

Chapter X introduces <one-sentence description of the chapter's central theme>.

| Section | Topics |
|---------|--------|
| X-1     | <key topics of Section X-1> |
| X-2     | <key topics of Section X-2> |
| X-3     | <key topics of Section X-3> |
```

Rules:

- The table has **exactly two columns**: `Section` and `Topics`. No more.
- Keep the theme sentence to one line; bold the key phrase.
- An optional `::: {.callout-note}` titled "Key Vocabulary" listing the chapter's core terms as a bullet list is permitted on the Overview slide (used in later chapters such as `ch8`), but the Section/Topics table is still required.

---

# Callout Boxes — THE CRITICAL RULE

Callouts are the most error-prone element. The following rules are **non-negotiable** and are the #1 thing to verify before returning any file.

## The closing rule

**Every callout MUST be closed with its own `:::` before anything else begins.** A callout has exactly this shape:

```markdown
::: {.callout-xxx}
## <Callout Title>

<content>
:::
```

You may open the next callout, the next `##` slide, or the next `# Section` divider **only after** the current callout's closing `:::` has appeared. Concretely, a closing `:::` is required before each of:

- another `::: {.callout-...}` opener,
- a slide separator `---`,
- a `## ` slide heading,
- a `# Section ...` divider.

## Forbidden patterns (each caused real render breakage)

1. **Missing close** — content runs straight into the next slide heading or the next section with no `:::`. (Most common; especially frequent in hypothesis-testing chapters where multi-step examples flow into `# Section` dividers.)
2. **Extra/stray opener-adjacent `:::`** — an opener immediately followed by a spurious `:::`, e.g.:

   ```markdown
   ::: {.callout-note}
   :::            <- WRONG: creates an empty callout, pushes title/content outside the box
   ## Term
   ```
3. **Heading trapped inside a callout** — the closing `:::` placed *after* a `## ` slide heading, swallowing the next slide into the box. This does not show up as a count imbalance, so it must be checked structurally, not just by counting.
4. **Orphan `:::`** — a lone `:::` with no matching opener (including stray bytes left from edits).

## Style convention

Write the opener as `:::{.callout-xxx}` (no space — this matches the entire existing corpus; a space after `:::` is also legal but do not mix styles within a file) and give the callout a `## ` titled heading on the next line, matching the `ch2`–`ch5` reference. The four callout types and their roles are fixed (Section 6). Do not mix in `###` headings inside callouts; use `## `.

---

# Callout Types and Their Roles

Use exactly these four types for these purposes:

| Type | Purpose |
|------|---------|
| `callout-note` | **Definitions** of terminology and key vocabulary |
| `callout-important` | **Rules, guidelines, formulas, error tables** — formal/methodological content |
| `callout-warning` | **Example problem statements**, and separately a **"Caution"** box for caveats/boundary conditions |
| `callout-tip` | **Solutions** to worked examples (always titled `## Solution`) |

## Definitions — `callout-note`

```markdown
::: {.callout-note}
## Frequency Distribution

A **frequency distribution** is a table that organizes data values (or ranges of values)
along with the count of how often each value or range occurs in the dataset.
:::
```

Bold the defined term; typeset any math in LaTeX (Section 9).

## Rules / Guidelines / Error tables — `callout-important`

```markdown
::: {.callout-important}
## Six Construction Guidelines

1. Use between **5 and 20 classes** — enough to show pattern without excessive detail.
2. The **class width** should be an odd number when possible so midpoints are whole numbers.
3. Classes must be **mutually exclusive** — each value belongs to exactly one class.
4. Classes must be **continuous** — no gaps, even if some classes have zero frequency.
5. Classes must be **exhaustive** — every value falls into some class.
6. Classes should be **equal in width** (except possibly the first or last).

Formula: $\text{Class width} \approx \dfrac{\text{highest} - \text{lowest}}{\text{number of classes}}$
:::
```

Decision tables (e.g., the Type I / Type II error matrix) also use `callout-important` with a Markdown table inside.

---

# Worked Examples

Worked examples are the core of every chapter. They follow a strict question → solution layout with cross-navigation.

## Question slide

- Title carries a named anchor: `## Example X-Y {#example-X-Y}`.
- The problem statement sits in a `callout-warning` box whose `## ` heading is the example's descriptive title.
- Raw data laid out in rows/columns goes inside a **fenced code block with no language tag** (triple backticks) so spacing is preserved.
- If a later example reuses this one's data/results, add a back-link at the bottom of the box.

```markdown
## Example X-Y {#example-X-Y}

::: {.callout-warning}
## <Example Title>

<Problem statement.>

```
<raw data, aligned in columns>
```

Back to [Example X-Z](#example-X-Z)   <- only if cited elsewhere
:::
```

## Solution slide(s)

- Separate question from solution with a bare `---`.
- Repeat `## Example X-Y` **without** re-declaring the anchor.
- Put the solution in a `callout-tip` titled `## Solution`.
- **Prefix each step with a bold descriptor** of what that slide does: `**Prepare data**`, `**Draw figure**`, `**Output figure**`, `**Check conditions**`, `**Step 1**`, etc.
- If the solution spans multiple slides, add another `---`, repeat `## Example X-Y`, and open a fresh `callout-tip`. **Close every `callout-tip` before the next `---`.**

```markdown
---

## Example X-Y

::: {.callout-tip}
## Solution

**Prepare data**

```{r eX-Y-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# R code
```
:::

---

## Example X-Y

::: {.callout-tip}
## Solution

**Output figure**

```{r eX-Y-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# R code
```
:::
```

## R chunk label convention

Labels follow `e<chapter>-<example>-<slide sequence>`, e.g. the 3rd slide of Example 2 in Chapter 2 is `e2-2-3`. **Labels must be globally unique within the file.** Duplicate labels break the render.

## Caution boxes

For a boundary condition, approximation caveat, or methodological warning, add a **separate** `callout-warning` titled `## Caution` (distinct from the problem statement box):

```markdown
::: {.callout-warning}
## Caution

Here $np = 5,\ nq = 5$ — exactly at the minimum threshold; the normal approximation may be less accurate for small $n$.
:::
```

## Hypothesis-testing example pattern (Chapters 8+)

Hypothesis-test examples use a numbered-step solution inside the `callout-tip`. State hypotheses with LaTeX, compute the test statistic in R, then decide:

```markdown
## Example 8-7 {#example-8-7}

::: {.callout-warning}
## z Test for a Mean

A researcher claims the average wind speed in a city is 8 mph. A sample of 32 days
gives $\bar{X} = 8.2$ mph, $\sigma = 0.6$. Use $\alpha = 0.05$.
:::

---

## Example 8-7

::: {.callout-tip}
## Solution

**Step 1.** $H_0:\ \mu = 8$ *(claim)*; $H_1:\ \mu \neq 8$ *(two-tailed)*.

**Step 2 — compute the test value**

```{r e8-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 8.2; mu <- 8; sigma <- 0.6; n <- 32
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)
p_val <- 2 * (1 - pnorm(abs(z)))   # two-tailed P-value
round(p_val, 4)
```

**Step 3.** Since $P = 0.0588 > \alpha = 0.05$, **do not reject $H_0$**.

**Step 4.** There is not enough evidence to reject the claim that the average wind speed is 8 mph.
:::
```

> **Reminder.** The `callout-warning` and `callout-tip` here are on different slides; each is closed by its own `:::` before the `---`. Never let a hypothesis-test example flow into the next `# Section` divider without closing its callout first.

---

# Cross-Example Navigation

When a later example cites an earlier one, make the reference a hyperlink and add a reciprocal back-link in the cited example's question box.

In the later example's statement:

```markdown
Using the blood type data from [Example 2-1](#example-2-1), construct a pie graph.
```

In the earlier (cited) example's `callout-warning`, before its closing `:::`:

```markdown
Back to [Example 2-5](#example-2-5)
```

This bidirectional linking lets readers jump between related examples in both directions.

---

# Mathematical Notation

**All math — symbols, variables, expressions — must be LaTeX. Unicode math characters are prohibited.** This includes the hypothesis-testing symbols that are easy to type as Unicode.

| Prohibited (Unicode) | Required (LaTeX) |
|----------------------|------------------|
| `n ≥ 30` | `$n \geq 30$` |
| `μ`, `σ`, `α`, `β` | `$\mu$`, `$\sigma$`, `$\alpha$`, `$\beta$` |
| `H₀`, `H₁` | `$H_0$`, `$H_1$` |
| `X̄`, `x̄` | `$\bar{X}$`, `$\bar{x}$` |
| `χ²`, `σ²` | `$\chi^2$`, `$\sigma^2$` |
| `≠`, `≤`, `≥`, `×`, `°` | `$\neq$`, `$\leq$`, `$\geq$`, `$\times$`, `^\circ` |
| `%` inside an equation | `$\% = \frac{f}{n}\times 100$` |

Inline math uses `$...$`; display math uses `$$...$$`. Examples:

```markdown
The class width is $\text{Class width} \approx \dfrac{H - L}{k}$.

$$\% = \frac{f}{n} \times 100$$
```

> Some existing chapter files still contain Unicode `H₀`, `α`, `χ²`, etc. Those are legacy and should be converted to LaTeX when touched; new files must be pure LaTeX from the start.

---

# R Code Standards

Default chunk options for every R chunk unless a slide genuinely needs otherwise:

```
echo = T, warning = FALSE, out.width = "80%", fig.align = 'center'
```

Conventions:

- Render tables with `knitr::kable(df, align = "c")`.
- Build figures with `ggplot2`; prefer `theme_void()` / clean themes for charts like pie graphs.
- Keep data vectors aligned in readable rows inside the chunk (mirrors the raw-data block on the question slide).
- When a figure is heavy, split "build the object" and "print the object" across two slides using the `**Draw figure**` / `**Output figure**` step descriptors.
- Every chunk label is unique (Section 7).

---

# Bilingual Workflow (`chX.qmd` + `chX-zh.qmd`)

Each chapter ships in two structurally identical files:

- `chX.qmd` — English. Language link: `[中文版由此去]` → `chX-zh.html`.
- `chX-zh.qmd` — Traditional Chinese. Language link: `[English version]` → `chX.html`.

Rules for the Chinese version:

- Translate prose, slide titles, callout titles, Overview text, Key Takeaways, and example narrative into Traditional Chinese (Taiwan usage).
- **Do not translate**: R code, chunk labels, variable names, anchors (`{#example-X-Y}`), LaTeX, the Acknowledgement block, or any YAML field except the language-link target and `title` (the Chinese file uses `title: "統計學"`).
- The Chinese Key Takeaways slide is titled `## 重點整理` and sits immediately before the (English) Acknowledgement slide.
- Keep the **same slide count, same anchors, same example numbering, and same callout structure** so the two decks stay in lockstep.
- The Acknowledgement slide text stays in English in both versions (it is a fixed copyright notice).

---

# Mandatory Closing Slides

## Key Formulas (conditional)

Include only if the chapter has several central formulae. Place it immediately before Key Takeaways.

```markdown
## Key Formulas

::: {.callout-important}
## Important Formulas

$$\text{Class width} \approx \frac{\text{highest} - \text{lowest}}{\text{number of classes}}$$

$$X_m = \frac{\text{lower limit} + \text{upper limit}}{2}$$

$$\% = \frac{f}{n} \times 100$$
:::
```

## Key Takeaways (mandatory, penultimate)

Concise declarative statements, math in LaTeX:

```markdown
## Key Takeaways

1. <First key insight.>
2. <Second key insight.>
3. The normal approximation to the binomial applies when $np \geq 5$ **and** $nq \geq 5$.
```

## Acknowledgement (mandatory, final, fixed text)

Reproduce verbatim; do not alter:

```markdown
## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2012). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
```

---

# Validation Checklist (run before returning the file)

Before delivering any `.qmd`, verify ALL of the following. The callout checks are the highest priority — they have repeatedly broken real decks.

**Callout integrity**

- [ ] Every `::: {.callout-...}` has exactly one matching closing `:::`.
- [ ] No callout opener is immediately followed by a stray `:::` (no empty callouts).
- [ ] No `## ` slide heading or `# Section` divider is trapped *inside* a callout — every heading/divider sits *after* a closing `:::`.
- [ ] No orphan `:::` (lone closer with no opener) and no stray NULL/control bytes.
- [ ] Total `:::` count is even and balanced.

**Structure**

- [ ] YAML matches Section 2 exactly, including the nested indentation under `revealjs:`.
- [ ] Order is: Title → Overview → Sections → (Key Formulas) → Key Takeaways → Acknowledgement.
- [ ] Overview table has exactly two columns.
- [ ] Section dividers use single `#` with `{background-color="#2c4a6e"}`.
- [ ] Acknowledgement text is verbatim.

**Examples & code**

- [ ] Every example question slide has an anchor `{#example-X-Y}`; solution slides repeat the title without re-declaring the anchor.
- [ ] All R chunk labels follow `e<ch>-<ex>-<seq>` and are globally unique.
- [ ] Cited examples have reciprocal back-links.

**Math**

- [ ] No Unicode math characters anywhere; all math is LaTeX (`$...$` / `$$...$$`).

**Bilingual (if producing both)**

- [ ] `chX.qmd` and `chX-zh.qmd` have identical slide count, anchors, and callout structure; only prose/titles are translated; the language link target is correct in each.

**Reference balance algorithm.** This Python check passes on all current clean chapters (`ch2`–`ch14`) with zero issues; reproduce its logic when self-checking. Note the two subtleties: (a) skip everything inside fenced code blocks so R comments like `# build the figure` are not mistaken for section dividers; (b) the **first** `## ` line right after a callout opener is the callout's *title* and is allowed — only a **second** `## ` inside the same open callout signals a trapped slide heading.

```python
import re, glob
def check(f):
    txt = open(f, encoding="utf-8", errors="replace").read()
    depth = 0; title_seen = 0; in_fence = False; issues = []
    for i, line in enumerate(txt.splitlines(), 1):
        s = line.strip()
        if s.startswith("```"):            # toggle code fence, skip its body
            in_fence = not in_fence; continue
        if in_fence:
            continue
        if re.match(r"^:::\s*\{\.callout", s):
            if depth > 0:
                issues.append((i, "opened a callout before closing the previous one"))
            depth = 1; title_seen = 0
        elif s == ":::":
            if depth == 0:
                issues.append((i, "orphan closing :::"))
            depth = 0
        elif depth == 1:
            if s.startswith("## "):
                title_seen += 1
                if title_seen > 1:
                    issues.append((i, "slide heading trapped inside callout"))
            elif re.match(r"^#\s", s):
                issues.append((i, "section divider trapped inside callout"))
            elif s == "---":
                issues.append((i, "--- slide separator inside an open callout"))
    if depth != 0:
        issues.append(("EOF", "file ends with a callout still open"))
    return issues

for f in sorted(glob.glob("ch*/ch*.qmd")):
    iss = check(f)
    if iss:
        print(f, "->", iss)
```

If this reports anything, fix the file before returning it.

---

# Minimal Complete Skeleton

````````{r,eval= F}
---
title: "Statistics"
subtitle: "Chapter X: <Chapter Title>"
... (full YAML per Section 2) ...
---

```{r setup, include=FALSE}
library(tidyverse)
```
````````

## Overview

Chapter X introduces <central theme>.

| Section | Topics |
|---------|--------|
| X-1 | ... |
| X-2 | ... |

# Section X-1: <Section Title> {background-color="#2c4a6e"}

## <Content Slide Title>

<Prose.>

::: {.callout-note}
## <Term>

<Definition.>
:::

---

## Example X-1 {#example-X-1}

::: {.callout-warning}
## <Example Title>

<Problem statement.>
:::

---

## Example X-1

::: {.callout-tip}
## Solution

**Prepare data**

```{r eX-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# R code
```
:::

# Section X-2: <Section Title> {background-color="#2c4a6e"}

## Key Formulas

::: {.callout-important}
## Important Formulas

$$\text{Formula}$$
:::

## Key Takeaways

1. ...
2. ...

## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2012). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
````

---

# Rendering

Render a single chapter, or all chapters at once:

```bash
quarto render "ch8/ch8.qmd"
quarto render "ch*/ch*.qmd"
```

*End of specification. Generate the `.qmd` slide file(s) in strict accordance with every rule above, and run the Section 12 checklist before returning.*
