---
title: "**Statistics Teaching System: Architecture and Operations Guide**"
author: "Yu-You Liou"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
header-includes:
  - \usepackage{geometry}
  - \geometry{a4paper, margin=2.5cm}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{Statistics Teaching System}
  - \fancyhead[R]{\thepage}
  - \fancyfoot[C]{}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue}
  - \usepackage{xcolor}
  - \definecolor{codebg}{gray}{0.95}
  - \usepackage{listings}
  - \lstset{basicstyle=\ttfamily\small, breaklines=true,
            frame=single, backgroundcolor=\color{codebg}}
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

\newpage

# System Overview

This document describes the complete architecture and operational procedures of the **Statistics Teaching System** (`stat`), designed for the English Taught Program in International Business (ETP) at Shih Chien University. The system supports two sequential courses---*Statistics (1)* and *Statistics (2)*---and is built on a core technology stack of R, Quarto, and R Markdown, integrating website publication, in-class exercises, and automated grading into a unified workflow.

> **Course website:** <https://yyliou.github.io/stat>
> **Instructor:** Yu-You Liou (`d10627008@ntu.edu.tw`)

## Course Structure

```{r course-structure}
df <- data.frame(
  Course = c("Statistics (1)", "Statistics (2)"),
  Code   = c("EIB-21E-01-A1", "EIB-21F-01-A1"),
  Schedule = c("Tuesday, Periods 8-9 (15:10-17:00)",
               "Thursday, Periods 6-7 (13:10-15:00)"),
  Room = c("L309", "L309"),
  Chapters = c("Ch. 1-7", "Ch. 8-14")
)
knitr::kable(df, booktabs = TRUE,
             caption = "Course Identification and Scheduling")
```

Each session is structured as two consecutive hours: the first hour is a **Lecture** covering theoretical foundations, and the second hour is a **Lab** in which students apply concepts through in-class exercises using the R programming language.

## Assessment Structure

```{r assessment}
df <- data.frame(
  Component = c("In-Class Exercises", "Midterm Examination", "Final Examination"),
  Weight    = c("36%", "32%", "32%"),
  Details   = c("12 exercises x 3% each", "", "")
)
knitr::kable(df, booktabs = TRUE,
             caption = "Assessment Weighting and Components")
```

\newpage

# Directory Architecture

## Root Directory Structure

The system root (`stat/`) is organized as follows:

```
stat/
|-- _quarto.yml          # Quarto website global configuration
|-- web.Rproj            # RStudio project file
|-- styles.css           # Custom CSS stylesheet
|-- index.qmd            # Homepage (course overview, examination regulations)
|-- sylla.qmd            # Syllabus landing page (links to PDF syllabi)
|-- slide.qmd            # Slide index page
|-- prob.qmd             # Problem set page (submission requirements)
|-- r.qmd                # R installation and usage guide
|
|-- ch2/ ... ch14/       # Per-chapter lecture slide directories
|   `-- chN.qmd          # Quarto Reveal.js slide source for chapter N
|
|-- stat1-ps/            # Statistics (1) problem sets
|   |-- ex1.qmd          # Exercise 1 source (Reveal.js format)
|   `-- ex1.html         # Compiled HTML for publication
|
|-- stat2-ps/            # Statistics (2) problem sets (same structure)
|
|-- stat1-ex/            # Statistics (1) student submissions
|   |-- ex1/
|   |   `-- ex1-[StudentID].pdf
|   `-- ex2/ ... ex12/
|
|-- stat2-ex/            # Statistics (2) student submissions (same structure)
|
|-- stu1/                # Statistics (1) roster and grading outputs
|   |-- [timestamp].xls  # Official roster exported from university system
|   `-- ex[N]/
|       |-- ex-score.xlsx
|       |-- stat.Rmd
|       `-- stat.pdf
|
|-- stu2/                # Statistics (2) roster and grading outputs (same structure)
|
|-- sylla/               # Syllabus source files
|   |-- stat1.Rmd / stat1.html
|   `-- stat2.Rmd / stat2.html
|
`-- prompt/              # AI-assisted prompt documents
    |-- prompt.Rmd       # Slide generation prompt (instructor use)
    |-- ex-review1.Rmd   # Grading prompt for Statistics (1)
    |-- ex-review2.Rmd   # Grading prompt for Statistics (2)
    `-- stat-system-guide.Rmd   # This document
```

## Design Principles

The directory structure adheres to four organizational principles:

1. **One directory per chapter.** Directories `ch2/` through `ch14/` are self-contained units, facilitating independent version control and selective rendering.
2. **Separation of problem sets and submissions.** `stat1-ps/` (instructor-authored questions) and `stat1-ex/` (student submissions) maintain distinct responsibilities and access patterns.
3. **Co-location of rosters and grading outputs.** Each `stu2/` or `stu1/` directory holds both the official class roster and all automated grading artifacts, enabling single-location lookup.
4. **Centralized prompt management.** All AI-assisted documents are isolated within `prompt/`, decoupled from pedagogical content.

\newpage

# Technology Stack

## Core Tools

```{r tech-stack}
df <- data.frame(
  Tool = c("R", "RStudio Desktop", "Quarto", "R Markdown", "TinyTeX"),
  `Minimum Version` = c(">= 4.2", "Latest stable", ">= 1.3",
                        ">= 2.20", "Current"),
  Purpose = c("Statistical computation and document generation",
              "Integrated development environment",
              "Website and slide publication",
              "Assignment templates and grading reports",
              "LaTeX engine for PDF output"),
  check.names = FALSE
)
knitr::kable(df, booktabs = TRUE,
             caption = "Core Software Components")
```

## Required R Packages

The following packages must be installed prior to system operation:

```r
# Document generation
install.packages(c("rmarkdown", "knitr", "tinytex"))
tinytex::install_tinytex()

# Data manipulation and visualization
install.packages(c("tidyverse", "ggplot2", "patchwork"))

# Spreadsheet I/O for grading reports
install.packages(c("readxl", "openxlsx"))

# Academic integrity (UUID generation)
install.packages("uuid")
```

## Website Publication Architecture

The course website is built as a **Quarto Website** project, governed by `_quarto.yml`. Key configuration decisions are summarized below.

```{r quarto-config}
df <- data.frame(
  Setting = c("output-dir", "Theme", "Navigation bar", "Slide format",
              "Math rendering"),
  Value = c(". (HTML output at repository root for GitHub Pages)",
            "Minty + custom brand CSS",
            "Syllabus -> Slides -> Problem Sets -> R",
            "Reveal.js",
            "KaTeX")
)
knitr::kable(df, booktabs = TRUE,
             caption = "Quarto Website Configuration Summary")
```

\newpage

# Lecture Slide Authoring Workflow

## File Format

Each chapter's slides reside in `chN/chN.qmd` with the following YAML header:

```yaml
---
title: "Statistics 1"
subtitle: "Chapter N: [Title]"
author: "Yu-You Liou"
date: "`r Sys.Date()`"
format:
  revealjs:
    html-math-method: katex
theme: serif
footer: "Elementary Statistics: A Step by Step Approach"
editor: visual
---
```

## Authoring Standards

All slide content must conform to the following conventions:

- Each slide begins with a level-2 heading (`##`).
- Mathematical expressions use standard LaTeX syntax, e.g., `$\bar{x} = \frac{\sum x}{n}$`.
- Definitions and key formulae are enclosed in Quarto callout blocks:

```markdown
::: {.callout-note}
**Definition** ...
:::
```

- Every textbook example provides two parallel solutions: (1) a complete algebraic derivation in LaTeX, and (2) an executable R implementation with displayed output.
- All prose must be original; verbatim reproduction of textbook text is prohibited.

## Generating a New Chapter

```
Step 1. Duplicate the previous chapter directory:
        cp -r ch2/ ch3/

Step 2. Use the prompt in prompt/prompt.Rmd with an AI assistant
        to generate content from the corresponding textbook chapter.

Step 3. Render in RStudio:
        quarto render ch3/ch3.qmd

Step 4. Update the chapter index in slide.qmd.
```

\newpage

# Problem Set Management

## Directory Mapping

```{r ps-mapping}
df <- data.frame(
  Course = c("Statistics (1)", "Statistics (2)"),
  `Problem Set Directory` = c("stat1-ps/", "stat2-ps/"),
  `Submission Directory`  = c("stat1-ex/", "stat2-ex/"),
  check.names = FALSE
)
knitr::kable(df, booktabs = TRUE,
             caption = "Problem Set and Submission Directory Mapping")
```

## Problem Set Format

Problem sets are authored as Reveal.js presentations (`stat1-ps/exN.qmd`), with one question per slide. A representative header:

```yaml
---
title: "Statistics 1"
subtitle: "In-Class Exercise N"
format:
  revealjs:
    html-math-method: katex
---
```

## Submission Requirements

Students must use the official template below and submit to `stat1-ex/exN/` using the exact naming convention `exN-[StudentID].pdf`.

```r
---
title: "Ex. N"
author: "StudentID"
date: "`r Sys.Date()`"
output: pdf_document
---

` ` `{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
` ` `

` ` `{r}
print(uuid::UUIDgenerate())
` ` `
```

> **Every submission must include the UUID generation code block. This serves as the primary mechanism for academic integrity verification.**

\newpage

# Automated Grading System

## Process Overview

The grading pipeline consists of four sequential stages:

```
STAGE 1 | Initialization and Roster Reconciliation
         Read stu[N]/[timestamp].xls -> construct student roster
         Cross-reference stat[N]-ex/exN/ -> flag absent students
         as Missing (Score = 0)

STAGE 2 | Compliance Filtering (sequential; stops at first violation)
         (i)  Submission timestamp > deadline (UTC+8)
              -> Late Submission (Score = 0)
         (ii) File is not PDF or does not match exN-[StudentID].pdf
              -> Incorrect File Name/Format (Score = 0)
         (iii) YAML missing title/author/date, or UUID chunk absent,
               or author field does not match submitting student ID
              -> Incorrect Internal Format (Score = 0)
         (iv) UUID value identical to another submission
              -> Plagiarism (Score = 0; all implicated students)

STAGE 3 | Academic Content Grading
         Total: 100 points, allocated equally across N questions
         Each question scored against stat[N]-ps/exN.qmd answer key
         Incorrect items noted as: [QN] is incorrect

STAGE 4 | Output Generation
         stu[N]/exN/ex-score.xlsx  <- Scoreboard (4 columns)
         stu[N]/exN/stat.Rmd       <- Analytics report source
         stu[N]/exN/stat.pdf       <- Compiled analytics report
```

## Scoreboard Format (`ex-score.xlsx`)

```{r scoreboard-fmt}
df <- data.frame(
  Column = c("StudentID", "Name", "Score", "Note"),
  Type   = c("Character", "Character", "Integer (0-100)", "Character"),
  Description = c("University student identification number",
                  "Student full name as listed in official roster",
                  "Final score awarded for the submission",
                  "Compliance violation code or incorrect question numbers")
)
knitr::kable(df, booktabs = TRUE,
             caption = "Scoreboard Column Specification")
```

## Analytics Report (`stat.Rmd`)

The analytics report reads `ex-score.xlsx` and produces three outputs:

1. **Descriptive statistics table:** minimum, $Q_1$, median, $Q_3$, maximum, and mean of the score distribution.
2. **Boxplot:** a publication-quality visualization of the score distribution rendered via `ggplot2`.
3. **Submission summary table:** counts for Total, Missing, Late Submission, Incorrect Format, and Plagiarism.

## Grading Note Codes

```{r note-codes}
df <- data.frame(
  Code = c("Missing", "Late Submission",
           "Incorrect File Name/Format", "Incorrect Internal Format",
           "Plagiarism", "[QN] is incorrect"),
  Description = c(
    "No file submitted by the deadline",
    "File submitted after the official deadline",
    "File is not PDF or does not follow naming convention exN-[ID].pdf",
    "YAML header incomplete, author ID mismatch, or UUID chunk absent",
    "UUID value matches that of one or more other submissions",
    "Question N answer is incorrect or methodology is flawed"
  ),
  Score = c("0", "0", "0", "0", "0", "Partial deduction")
)
knitr::kable(df, booktabs = TRUE,
             caption = "Grading Note Code Reference")
```

\newpage

# Academic Integrity Mechanisms

## UUID Verification

Every submission must execute and display output from the following R code block:

```r
print(uuid::UUIDgenerate())
```

`uuid::UUIDgenerate()` produces a Version 4 UUID (Universally Unique Identifier) on each invocation, e.g., `6d2b9fda-2d9b-48f3-84bc-f82eabc39b4d`. The grading system compares this value across all submissions for a given exercise. If two or more files share an identical UUID, all implicated students receive a score of zero with the note `Plagiarism`.

**Important implementation notes:**

- The compliance filter is applied sequentially per file. A file that fails Stage 2 check (iii) (Incorrect Internal Format) is not evaluated for plagiarism within its own record; however, its UUID remains in the comparison pool and may trigger a Plagiarism flag in other students' records.
- The `author` field in the YAML header must contain the submitting student's own ID. Use of another student's ID constitutes a format violation independent of the plagiarism check.

## Examination Regulations

```{r exam-regs}
df <- data.frame(
  Regulation = c("Mandatory attendance", "Network restriction",
                 "Authorized resources", "Media prohibition",
                 "Communication prohibition", "Copy-paste disabled"),
  Policy = c(
    "Roll call 5 minutes before start; late arrival or absence results in score = 0",
    "System tracks IP addresses; campus Wi-Fi required; non-campus IP results in score = 0",
    "Open-book; reference materials and AI tools are permitted as auxiliary aids",
    "Photography, video/audio recording, and screenshotting result in score = 0",
    "Interpersonal communication and disruptive behavior result in score = 0",
    "Copy-paste functionality is disabled by the examination system"
  )
)
knitr::kable(df, booktabs = TRUE,
             caption = "Examination Conduct Regulations")
```

## AI Use Policy

Students may use AI tools (e.g., ChatGPT, Gemini) as learning aids. However, students must be capable of explaining and defending any code or analysis they submit. Uncritical submission of AI-generated content without demonstrated understanding constitutes academic dishonesty and is subject to the same consequences as plagiarism.

\newpage

# AI Prompt Management

The `prompt/` directory contains structured prompt documents for AI-assisted course material generation and grading.

```{r prompt-files}
df <- data.frame(
  File = c("prompt.Rmd", "ex-review1.Rmd",
           "ex-review2.Rmd", "stat-system-guide.Rmd"),
  Purpose = c(
    "Slide generation prompt: instructs AI to extract all textbook examples from a given chapter and render them with dual mathematical/R solutions",
    "Automated grading prompt for Statistics (1) exercises",
    "Automated grading prompt for Statistics (2) exercises",
    "This technical document"
  )
)
knitr::kable(df, booktabs = TRUE,
             caption = "Prompt Directory File Inventory")
```

## Slide Generation Prompt (`prompt.Rmd`)

Key directives encoded in the slide generation prompt:

- **Role:** University-level statistics instructor.
- **Exhaustive extraction:** Every example in the target chapter must be incorporated; summarization or omission is not permitted.
- **Dual-solution requirement:** Each example must present (1) a full algebraic derivation in LaTeX and (2) executable R code with displayed output.
- **Paraphrasing mandate:** All prose must be rewritten in original language; verbatim textbook reproduction is prohibited.
- **Callout block usage:** Definitions, key formulae, and example prompts must be enclosed in appropriate callout blocks.
- **Directory replication:** Output must mirror the established file structure (e.g., `ch3/ch3.qmd`).

## Grading Prompts (`ex-review1.Rmd`, `ex-review2.Rmd`)

Key components of the grading prompts:

- Definition of environment variables: `[no]`, timezone (UTC+8), and root directory structure.
- Specification of all four grading stages with precise conditional logic.
- Output format requirements: scoreboard column names, Rmd content sections, and PDF compilation.

\newpage

# Operations Guide

## New Semester Initialization

```
1. Export official class roster from the university administrative system.
   Save to stu1/ (Statistics 1) or stu/ (Statistics 2).

2. Update sylla/stat1.Rmd and sylla/stat2.Rmd with the current
   semester's schedule and policies. Knit to PDF.

3. Verify _quarto.yml navigation links and semester metadata.

4. Rebuild the full website:
   quarto render
```

## Weekly Problem Set Publication

```
1. Create stat1-ps/exN.qmd (or stat2-ps/exN.qmd) in Reveal.js format.
   Each slide corresponds to one question.

2. Render the problem set:
   quarto render stat1-ps/exN.qmd

3. Update the problem set index in prob.qmd.
```

## Weekly Submission Collection and Grading

```
1. Students submit to: stat1-ex/exN/exN-[StudentID].pdf

2. Execute the AI grading prompt (prompt/ex-review1.Rmd) with
   environment variable [no] set to the exercise number.

3. Verify generated scoreboard: stu1/exN/ex-score.xlsx

4. Verify analytics report: stu1/exN/stat.pdf
```

## Generating a New Lecture Chapter

```
1. Duplicate the previous chapter directory:
   cp -r ch2/ ch3/

2. Generate content using prompt/prompt.Rmd with an AI assistant,
   referencing the corresponding textbook chapter.

3. Review, edit, and finalize ch3/ch3.qmd in RStudio.

4. Render:
   quarto render ch3/ch3.qmd

5. Add the chapter link to slide.qmd.
```

## Full-Site Publication (GitHub Pages)

```bash
# From the stat/ root directory
quarto render

# Commit and push to trigger GitHub Pages deployment
git add .
git commit -m "Update: Week N — [brief description]"
git push origin main
```

## Troubleshooting Reference

```{r troubleshooting}
df <- data.frame(
  Symptom = c("PDF output fails",
              "UUID package not found",
              "Quarto render error",
              "Roster file read error",
              "Slide math not rendering"),
  `Probable Cause` = c("TinyTeX not installed or incomplete",
                       "uuid package not installed",
                       "Quarto version outdated",
                       "Legacy .xls format compatibility",
                       "KaTeX configuration missing"),
  Resolution = c("Run: tinytex::install_tinytex()",
                 "Run: install.packages('uuid')",
                 "Update Quarto to latest stable release",
                 "Use readxl package or resave as .xlsx",
                 "Set html-math-method: katex in YAML"),
  check.names = FALSE
)
knitr::kable(df, booktabs = TRUE,
             caption = "Common Issues and Resolutions")
```

\newpage

# Course Chapter Reference

## Statistics (1): Chapter Schedule

```{r sched1}
df <- data.frame(
  Week = c(2,3,4,5,6,7,8,9,10,11,12,13,14,15,16),
  Chapter = c("Ch. 1","Ch. 2","Ch. 3","Ch. 3","Ch. 4","Ch. 4",
              "—","—","Ch. 5","Ch. 5","Ch. 6","Ch. 6",
              "Ch. 7","Ch. 7","—"),
  Topic = c(
    "The Nature of Probability and Statistics",
    "Frequency Distributions and Graphs",
    "Data Description",
    "Data Description (continued)",
    "Probability and Counting Rules",
    "Probability and Counting Rules (continued)",
    "Midterm Review",
    "Midterm Examination",
    "Discrete Probability Distributions",
    "Discrete Probability Distributions (continued)",
    "The Normal Distribution",
    "The Normal Distribution (continued)",
    "Confidence Intervals and Sample Size",
    "Confidence Intervals and Sample Size (continued)",
    "Final Examination"
  ),
  Exercise = c("ex1","ex2","ex3","ex4","ex5","ex6",
               "—","—","ex7","ex8","ex9","ex10",
               "ex11","ex12","—")
)
knitr::kable(df, booktabs = TRUE,
             caption = "Statistics (1) Weekly Schedule and Exercise Mapping")
```

## Statistics (2): Chapter Schedule

```{r sched2}
df <- data.frame(
  Week = c("2-4","5-6","7","8","9","10-11","12-13","14-15","16"),
  Chapter = c("Ch. 8","Ch. 9","Ch. 10","—","—",
              "Ch. 10","Ch. 11","Ch. 12","—"),
  Topic = c(
    "Hypothesis Testing",
    "Testing the Difference Between Two Means, Two Proportions, and Two Variances",
    "Correlation and Regression",
    "Midterm Review",
    "Midterm Examination",
    "Correlation and Regression (continued)",
    "Other Chi-Square Tests",
    "Analysis of Variance",
    "Final Examination"
  ),
  Exercises = c("ex1-3","ex4-5","ex6","—","—",
                "ex7-8","ex9-10","ex11-12","—")
)
knitr::kable(df, booktabs = TRUE,
             caption = "Statistics (2) Weekly Schedule and Exercise Mapping")
```

\newpage

# Appendix

## A. Official Submission Template

```yaml
---
title: "Ex. 1"
author: "A113290001"
date: "2026-03-04"
output: pdf_document
---
```

```r
# This chunk is mandatory in every submission.
print(uuid::UUIDgenerate())

# Exercise solutions follow below.
```

## B. Directory Quick Reference

```{r dir-ref}
df <- data.frame(
  Directory = c("ch2/-ch14/", "stat1-ps/", "stat2-ps/",
                "stat1-ex/exN/", "stat2-ex/exN/",
                "stu1/exN/", "stu/exN/",
                "sylla/", "prompt/"),
  Purpose = c(
    "Reveal.js lecture slide sources, one subdirectory per chapter",
    "Statistics (1) in-class exercise problem sets",
    "Statistics (2) in-class exercise problem sets",
    "Statistics (1) student PDF submissions for exercise N",
    "Statistics (2) student PDF submissions for exercise N",
    "Statistics (1) grading outputs (scoreboard, analytics report)",
    "Statistics (2) grading outputs (scoreboard, analytics report)",
    "Syllabus R Markdown source files",
    "AI-assisted prompt documents for slide generation and grading"
  )
)
knitr::kable(df, booktabs = TRUE,
             caption = "Directory Purpose Quick Reference")
```

## C. References

- Bluman, A. G. (2012). *Elementary Statistics: A Step-by-Step Approach* (8th ed.). McGraw-Hill.
- R Core Team. (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing. <https://www.R-project.org>
- Posit, PBC. (2024). *RStudio Desktop*. <https://posit.co/download/rstudio-desktop>
- Allaire, J. J., et al. (2024). *Quarto*. <https://quarto.org>
- Course website: <https://yyliou.github.io/stat>
