---
title: "Statistics"
subtitle: "Chapter 2: Frequency Distributions and Graphs"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format:
  revealjs:
    html-math-method: katex
smaller: true
theme: serif
footer: "Elementary Statistics: A Step by Step Approach"
logo: "logo.svg"
editor: source
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch2-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
```

## Overview

Chapter 2 introduces methods for **organizing and visualizing data** — moving from raw, unordered observations to structured summaries that reveal patterns.

| Section | Topics |
|---------|--------|
| 2-1 | Frequency distributions: categorical, grouped, ungrouped, cumulative |
| 2-2 | Histograms, frequency polygons, and ogives |
| 2-3 | Bar graphs, Pareto charts, time series, pie graphs, stem and leaf plots |

# Section 2-1: Organizing Data {background-color="#2c4a6e"}

## From Raw Data to Organized Information

When data are first gathered, they exist in **raw form** — an unordered, unprocessed list of values. On their own, raw data reveal very little about the underlying patterns or trends in a dataset.

Organizing data into a **frequency distribution** provides a structured summary that makes patterns visible and analysis manageable.

:::{.callout-note}
## Frequency Distribution

A **frequency distribution** is a table that organizes data values (or ranges of values) along with the count of how often each value or range occurs in the dataset.
:::

---

## Categorical Frequency Distributions

When data consist of **qualitative categories** (non-numeric labels or groups), a **categorical frequency distribution** is used. Each category is listed alongside its frequency and relative frequency (percentage).

**Steps to construct a categorical frequency distribution:**

1. List each distinct category.
2. Tally the occurrences of each category.
3. Record the frequency and compute the percentage: $\% = \frac{f}{n} \times 100$

---

## Example 2-1 {#example-2-1}

:::{.callout-warning}
## Blood Types of Army Inductees
A sample of 25 army inductees was tested for blood type. The results are shown below. Organize the data into a categorical frequency distribution.

```
A  B  B  AB  O
O  O  B  AB  B
B  B  O  A   O
A  O  O  O   AB
AB A  O  B   A
```
Back to [Example 2-5](#example-2-5)
:::

---

## Example 2-1 

:::{.callout-tip}
## Solution
```{r e2-1-1,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
blood <- c("A","B","B","AB","O","O","O","B","AB","B",
           "B","B","O","A","O","A","O","O","O","AB",
           "AB","A","O","B","A")

df <- data.frame(blood) |>
  count(blood) |>
  rename(Class = blood, Frequency = n) |>
  mutate(Percentage = round(Frequency / sum(Frequency) * 100, 1))

knitr::kable(df, align = "c")
```
:::

## Rules for Constructing a Grouped Frequency Distribution

:::{.callout-important}
## Six Construction Guidelines

1. Use between **5 and 20 classes** — enough to show pattern without excessive detail.
2. The **class width** should be an odd number when possible so midpoints are whole numbers.
3. Classes must be **mutually exclusive** — each data value belongs to exactly one class.
4. Classes must be **continuous** — no gaps between classes, even if some classes have zero frequency.
5. Classes must be **exhaustive** — every data value must fit into some class.
6. Classes should be **equal in width** (except possibly the first or last class).

Formula: $\text{Class width} \approx \dfrac{\text{highest value} - \text{lowest value}}{\text{number of classes}}$
:::

## Example 2-2 {#example-2-2}

:::{.callout-warning}
##  Record High Temperatures
The record high temperatures (°F) for each of the 50 U.S. states are listed below. Construct a grouped frequency distribution using 7 classes.

```
112 100 127 120 134 118 105 110 109 112
110 118 117 116 118 122 114 114 105 109
107 112 114 115 118 117 118 122 106 110
116 108 110 121 113 120 119 111 104 111
120 113 120 117 105 110 118 112 114 114
```
:::

---

## Example 2-2

:::{.callout-tip}
## Solution
**Prepare data**

```{r e2-2-1,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
temps <- c(112,100,127,120,134,118,105,110,109,112,
           110,118,117,116,118,122,114,114,105,109,
           107,112,114,115,118,117,118,122,106,110,
           116,108,110,121,113,120,119,111,104,111,
           120,113,120,117,105,110,118,112,114,114)
```

**Draw figure**

```{r e2-2-2,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
# Class width = ceiling((134 - 100) / 7) = 5
breaks <- seq(99.5, 134.5, by = 5)
labels <- c("100-104","105-109","110-114","115-119","120-124","125-129","130-134")

freq_table <- data.frame(temps) |>
  mutate(Class = cut(temps, breaks = breaks, labels = labels)) |>
  count(Class, .drop = FALSE) |>
  rename(Frequency = n) |>
  mutate(
    Boundaries = c("99.5-104.5","104.5-109.5","109.5-114.5",
                   "114.5-119.5","119.5-124.5","124.5-129.5","129.5-134.5"),
    Midpoint   = seq(102, 132, by = 5),
    Cumulative = cumsum(Frequency)
  )
```
:::

---

## Example 2-2

:::{.callout-tip}
## Solution

**Output figure**

```{r e2-2-3,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
knitr::kable(freq_table[, c("Class","Boundaries","Midpoint","Frequency","Cumulative")],
             align = "c")
```
:::

## Ungrouped Frequency Distributions

When the range of values is small and each distinct value is meaningful on its own, an **ungrouped (single-value) frequency distribution** lists each individual data value as its own class.

This approach preserves the exact values rather than collapsing them into intervals.

# Section 2-2: Other Types of Graphs {background-color="#2c4a6e"}

## Beyond the Histogram

In addition to histograms, frequency polygons, and ogives, several other graphical tools are widely used in statistics and everyday data communication:

- **Bar graphs** — for comparing frequencies of categorical data
- **Pareto charts** — bar graphs with bars sorted from highest to lowest frequency
- **Time series graphs** — for tracking a variable's change over time
- **Pie graphs** — for showing each category's share of the whole
- **Stem and leaf plots** — for displaying the actual data values in a graphical form

---

## Example 2-3

::: {.callout-warning}
## Batting Averages
In baseball, a .320 batting average means a player gets a hit 32% of the time. In 100 at-bats, find the probability of getting **26 or fewer** hits.
:::

::: {.callout-tip}
## Solution

**Check:** $np = 32 \geq 5, nq = 68 \geq 5$

**Parameters:** $\mu = 32, \sigma = \sqrt{100 \times 0.32 \times 0.68}$

```{r e2-3-1, echo=T, warning=FALSE}
n <- 100; p <- 0.32; q <- 0.68
mu <- n * p; sigma <- sqrt(n * p * q)
cat("mu =", mu, "  sigma =", round(sigma, 4), "\n")

# P(X <= 26) with continuity correction: P(X < 26.5)
pnorm(26.5, mean = mu, sd = sigma)

# Exact binomial
pbinom(26, n, p)
```

**Answer (normal approx.):** $\approx 0.119$ (11.9%); Exact: $\approx 0.116$
:::

## Example 2-4

::: {.callout-warning}
## Comparing Binomial and Normal Approximation 
A coin is flipped 10 times. Find the probability of getting **exactly 6 heads** using:

a. The binomial distribution

b. The normal approximation
:::

::: {.callout-warning}
## Caution
Here $np = 5, nq = 5$ — these are exactly at the minimum threshold. The normal approximation may be less accurate for small n.
:::

---

::: {.callout-tip}
## Solution
```{r e2-4-1, echo=T, warning=FALSE}
n <- 10; p <- 0.5; q <- 0.5
mu <- n * p; sigma <- sqrt(n * p * q)
cat("mu =", mu, "  sigma =", sigma, "\n")

# a. Exact binomial
dbinom(6, n, p)

# b. Normal approximation with continuity correction
pnorm(6.5, mean = mu, sd = sigma) - pnorm(5.5, mean = mu, sd = sigma)
```

**Comparison:** Exact = **0.2051**; Normal approx. = **0.2034**

The approximation is quite close even for this borderline case.
:::

## Bar Graphs

:::{.callout-note}
## Bar Graph

A **bar graph** represents categorical (qualitative) data using horizontal or vertical bars. The length or height of each bar corresponds to the frequency or value for that category. Unlike a histogram, gaps between bars are acceptable.
:::

## Example 2-5 {#example-2-5}

:::{.callout-warning}
##  Blood Type Pie Graph
Using the blood type data from [Example 2-1](#example-2-1), construct a pie graph.
:::

:::{.callout-tip}
## Solution
```{r e2-5-1,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
blood_pie <- data.frame(
  type = c("A","B","O","AB"),
  freq = c(5, 6, 9, 4)
) |>
  mutate(
    pct    = freq / sum(freq) * 100,
    deg    = pct / 100 * 360,
    label  = paste0(type, "\n", freq, " (", round(pct, 1), "%)")
  )

fig <- ggplot(blood_pie, aes(x = "", y = freq, fill = type)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(type, "\n", round(pct, 1), "%")),
            position = position_stack(vjust = 0.5), size = 4) +
  scale_fill_manual(values = c("#4a90d9","#d9534f","#5cb85c","#f0ad4e")) +
  labs(title = "Blood Type Distribution (n = 25)", fill = "Blood Type") +
  theme_void()
```
:::

---

## Example 2-5

:::{.callout-tip}
## Solution
**Output figure**
```{r e2-5-2,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
fig
```
:::

## Key Formulas

:::{.callout-important}
## Important Formulas

$$\text{Class width} \approx \frac{\text{highest} - \text{lowest}}{\text{number of classes}}$$

$$X_m = \frac{\text{lower limit} + \text{upper limit}}{2} \quad \text{(class midpoint)}$$

$$\% = \frac{f}{n} \times 100 \quad \text{(percentage for each class)}$$

$$\text{Degrees} = \frac{f}{n} \times 360° \quad \text{(pie chart sector)}$$
:::

## Key Takeaways

1. The normal distribution is fully described by μ and σ

2. Convert to z scores to use the standard normal table

3. The CLT guarantees that sample means are approximately normal for $n ≥ 30$

4. Use normal approximation to binomial when $np \geq 5$ **and** $nq \geq 5$

5. Always apply the **correction for continuity** in the normal approximation

## Acknowledgement

-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2011). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
