---
title: "Statistics"
subtitle: "Chapter 2: Frequency Distributions and Graphs"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "Elementary Statistics: A Step-by-Step Approach"
logo: "logo.svg"
editor: source
bibliography: bib.bib
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      out.width = "80%", fig.align = "center")
library(ggplot2)
library(dplyr)
```

# Introduction

## Chapter Overview

**Chapter 2: Frequency Distributions and Graphs**

After collecting data, the analyst's next challenge is to **organise and present** it in a way that reveals meaningful patterns. This chapter introduces two fundamental tools for that task:

- **Frequency distributions** — structured tables that summarise how often each value (or range of values) occurs
- **Statistical graphs** — visual displays that make patterns, trends, and comparisons immediately apparent

::: {.callout-note icon=false}
## Why Organise Data?

Raw data — a collection of numbers or labels as gathered — is difficult to interpret at a glance. Organising data into distributions and graphs allows researchers to:

- Identify the **centre** and **spread** of a data set
- Spot **outliers** or unusual observations
- Communicate findings **efficiently** to a broad audience
:::

## Road Map

| Section | Topic |
|---------|-------|
| 2-1 | Organising Data — frequency distributions |
| 2-2 | Histograms, frequency polygons, and ogives |
| 2-3 | Other graph types: bar, Pareto, time series, pie, stem-and-leaf |

# §2-1 Organising Data

## Raw Data and the Need for Organisation

When data are first collected they appear in **raw** (unsorted) form. Consider 50 state record-high temperatures — a list of 50 numbers tells us very little until we organise them.

::: {.callout-note icon=false}
## Definition: Frequency Distribution

A **frequency distribution** is a table that organises data values into **classes** (categories or intervals) and records the **frequency** — the count — of observations falling in each class.

- **Categorical frequency distribution** — used when data consist of named categories
- **Grouped frequency distribution** — used when data are numerical and span a wide range; values are grouped into equal-width intervals
- **Ungrouped frequency distribution** — used when the range of values is small; each distinct value forms its own class
:::

## Categorical Frequency Distribution

::: {.callout-tip icon=false}
## Example 2-1 — Blood Type Survey

Twenty-five army inductees were blood-typed with the following results:

`A B B AB O A O B B A B B O A O AB AB A B A O B B O A`

Construct a **categorical frequency distribution** for these data.
:::

---

```{r ex2-1}
blood <- c("A","B","B","AB","O","A","O","B","B","A",
           "B","B","O","A","O","AB","AB","A","B","A",
           "O","B","B","O","A")

df <- as.data.frame(table(blood))
colnames(df) <- c("Blood Type", "Frequency")
df$`Relative Frequency` <- round(df$Frequency / sum(df$Frequency), 3)
df$`Percent` <- paste0(round(df$`Relative Frequency` * 100, 1), "%")
knitr::kable(df, align = "c")
```

## Grouped Frequency Distribution — Key Terminology

::: {.callout-note icon=false}
## Definitions: Class, Limits, Boundaries, Width, Midpoint

| Term | Meaning |
|------|---------|
| **Class** | An interval used to group data |
| **Class limits** | Smallest (lower) and largest (upper) values that belong to a class |
| **Class boundaries** | Values halfway between consecutive class limits — eliminate gaps |
| **Class width** | Difference between consecutive lower class limits (or boundaries) |
| **Class midpoint** | Average of the lower and upper class boundaries: $X_m = \dfrac{L + U}{2}$ |

**Choosing the number of classes:** Use between **5 and 20** classes. As a starting rule, use $k$ classes where $2^k \geq n$.

**Class width formula (always round up):**
$$\text{Width} = \frac{\text{Range}}{\text{Number of classes}} = \frac{\max - \min}{k}$$
:::

## Grouped Frequency Distribution — Example

::: {.callout-tip icon=false}
## Example 2-2 — Record High Temperatures

The record high temperatures (°F) for each of the 50 US states are listed below. Construct a **grouped frequency distribution** using 7 classes.

```
112 100 127 120 134 118 105 110 109 112
110 118 117 116 118 122 114 114 105 109
107 112 114 115 118 117 118 122 106 110
116 108 110 121 113 120 119 111 104 111
120 113 120 117 105 110 118 112 114 112
```
:::

```{r}
temps <- c(112,100,127,120,134,118,105,110,109,112,
           110,118,117,116,118,122,114,114,105,109,
           107,112,114,115,118,117,118,122,106,110,
           116,108,110,121,113,120,119,111,104,111,
           120,113,120,117,105,110,118,112,114,112)
```

---

```{r ex2-2}
# Class width = ceiling((134-100)/7) = 5
breaks <- seq(100, 135, by = 5)
labels <- paste0(breaks[-length(breaks)], "–", breaks[-1] - 1)
cut_temps <- cut(temps, breaks = breaks, right = FALSE, labels = labels)
freq_table <- as.data.frame(table(cut_temps))
colnames(freq_table) <- c("Class", "Frequency")
freq_table$`Cumulative Frequency` <- cumsum(freq_table$Frequency)
freq_table$`Relative Frequency` <- round(freq_table$Frequency / length(temps), 3)
knitr::kable(freq_table, align = "c")
```

## Ungrouped Frequency Distribution

::: {.callout-tip icon=false}
## Example 2-3 — Miles per Gallon (SUVs)

The EPA fuel-economy ratings (MPG, city driving) for a sample of sport utility vehicles are:

`12 17 12 14 16 18 16 18 12 16`
`17 15 16 12 15 16 16 12 14 16`

Construct an **ungrouped frequency distribution** with a cumulative frequency column.
:::

---

```{r ex2-3}
mpg <- c(12,17,12,14,16,18,16,18,12,16,
         17,15,16,12,15,16,16,12,14,16)

freq_mpg <- as.data.frame(table(mpg))
colnames(freq_mpg) <- c("MPG", "Frequency")
freq_mpg$MPG <- as.numeric(as.character(freq_mpg$MPG))
freq_mpg$`Cumulative Frequency` <- cumsum(freq_mpg$Frequency)
freq_mpg$`Relative Frequency` <- round(freq_mpg$Frequency / length(mpg), 3)
knitr::kable(freq_mpg, align = "c")
```

## Applying the Concepts 2-1

::: {.callout-tip icon=false}
## ATC 2-1 — Ages of Presidents at Inauguration

The ages (years) at which US presidents were inaugurated are listed below. Construct a grouped frequency distribution with 5 classes and describe what the distribution reveals about presidential inauguration ages.

```
57 61 57 57 58 57 61 54 68 51
49 64 50 48 65 52 56 46 54 49
51 47 55 55 54 42 51 56 55 51
54 51 60 62 43 55 56 61 52 69
64 46 54
```
:::

---

```{r atc2-1}
ages <- c(57,61,57,57,58,57,61,54,68,51,
          49,64,50,48,65,52,56,46,54,49,
          51,47,55,55,54,42,51,56,55,51,
          54,51,60,62,43,55,56,61,52,69,
          64,46,54)

# 5 classes, width = ceiling((69-42)/5) = 6
breaks_age <- seq(42, 72, by = 6)
labels_age  <- paste0(breaks_age[-length(breaks_age)], "–",
                       breaks_age[-1] - 1)
cut_age <- cut(ages, breaks = breaks_age, right = FALSE, labels = labels_age)
age_freq <- as.data.frame(table(cut_age))
colnames(age_freq) <- c("Age Class", "Frequency")
age_freq$`Cumulative Frequency` <- cumsum(age_freq$Frequency)
knitr::kable(age_freq, align = "c")
```

Most presidents were inaugurated between **48 and 59 years** of age, with the distribution showing a slight right skew toward older ages.

# §2-2 Histograms, Frequency Polygons, and Ogives

## Three Key Graphs for Frequency Distributions

::: {.callout-note icon=false}
## Definitions: Histogram, Frequency Polygon, Ogive

**Histogram** — a bar graph where each bar represents one class; the *x*-axis shows class boundaries and the *y*-axis shows frequency (or relative frequency). Bars are adjacent with no gaps.

**Frequency polygon** — a line graph connecting the midpoints of each class at a height equal to the class frequency. The line is anchored to zero at each end by adding an empty class on each side.

**Ogive** (cumulative frequency graph) — a line graph plotting **cumulative frequency** against the **upper class boundary**. Used to determine how many observations fall below a given value.

**Relative frequency versions** replace raw counts with proportions ($f/n$) or percentages on the *y*-axis, enabling fair comparisons across data sets of different sizes.
:::

## Histogram

::: {.callout-tip icon=false}
## Example 2-4 — Histogram of Record High Temperatures

Using the frequency distribution from Example 2-2 (record high temperatures for 50 US states), construct a **histogram**.
:::

---

```{r ex2-4}
temps <- c(112,100,127,120,134,118,105,110,109,112,
           110,118,117,116,118,122,114,114,105,109,
           107,112,114,115,118,117,118,122,106,110,
           116,108,110,121,113,120,119,111,104,111,
           120,113,120,117,105,110,118,112,114,112)

ggplot(data.frame(x = temps), aes(x = x)) +
  geom_histogram(breaks = seq(99.5, 134.5, by = 5),
                 fill = "steelblue", color = "white") +
  scale_x_continuous(breaks = seq(99.5, 134.5, by = 5)) +
  labs(title = "Record High Temperatures for the 50 US States",
       x = "Temperature (°F)", y = "Frequency") +
  theme_classic(base_size = 14)
```

## Frequency Polygon

::: {.callout-tip icon=false}
## Example 2-5 — Frequency Polygon of Record High Temperatures

Using the same data, construct a **frequency polygon**. Plot the midpoint of each class on the *x*-axis and the class frequency on the *y*-axis; connect the points with straight lines and anchor both ends to zero.
:::

---

```{r ex2-5}
midpoints <- seq(102, 132, by = 5)
freqs     <- c(2, 8, 18, 13, 7, 1, 1)

# Add anchor points
mp_full <- c(97, midpoints, 137)
fr_full <- c(0, freqs, 0)

ggplot(data.frame(x = mp_full, y = fr_full), aes(x, y)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_x_continuous(breaks = mp_full) +
  labs(title = "Frequency Polygon — Record High Temperatures",
       x = "Midpoint (°F)", y = "Frequency") +
  theme_classic(base_size = 14)
```

## Ogive (Cumulative Frequency Graph)

::: {.callout-tip icon=false}
## Example 2-6 — Ogive of Record High Temperatures

Construct an **ogive** for the record-high-temperature data. Plot the **upper class boundary** on the *x*-axis and the **cumulative frequency** on the *y*-axis.
:::

---

```{r ex2-6}
upper_bounds <- c(99.5, 104.5, 109.5, 114.5, 119.5, 124.5, 129.5, 134.5)
cum_freq     <- c(0, 2, 10, 28, 41, 48, 49, 50)

ggplot(data.frame(x = upper_bounds, y = cum_freq), aes(x, y)) +
  geom_line(color = "darkred", linewidth = 1) +
  geom_point(color = "darkred", size = 2) +
  scale_x_continuous(breaks = upper_bounds) +
  labs(title = "Ogive — Record High Temperatures",
       x = "Upper Class Boundary (°F)", y = "Cumulative Frequency") +
  theme_classic(base_size = 14)
```

## Relative Frequency Graphs

::: {.callout-tip icon=false}
## Example 2-7 — Relative Frequency Histogram

A researcher recorded the number of miles run per week by a sample of joggers:

| Miles | Frequency |
|-------|-----------|
| 5.5 – 10.5 | 1 |
| 10.5 – 15.5 | 2 |
| 15.5 – 20.5 | 3 |
| 20.5 – 25.5 | 5 |
| 25.5 – 30.5 | 4 |

Construct a **relative frequency histogram** and a **relative frequency ogive**.
:::

```{r}
miles_mid  <- c(8, 13, 18, 23, 28)
miles_freq <- c(1, 2, 3, 5, 4)
n          <- sum(miles_freq)
rel_freq   <- miles_freq / n
cum_rel    <- cumsum(rel_freq)
bounds     <- c(5.5, 10.5, 15.5, 20.5, 25.5, 30.5)
```

---

```{r ex2-7, fig.show='hold', out.width="48%", fig.width=5, fig.height=4}
# Relative frequency histogram
df_hist <- data.frame(
  xmin = bounds[-length(bounds)],
  xmax = bounds[-1],
  y    = rel_freq
)
ggplot(df_hist, aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = y)) +
  geom_rect(fill = "steelblue", color = "white") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Relative Frequency Histogram",
       x = "Miles per Week", y = "Relative Frequency") +
  theme_classic(base_size = 12)
```

---

```{r ex2-7b, fig.show='hold', out.width="48%", fig.width=5, fig.height=4}
# Relative frequency ogive
df_ogive <- data.frame(x = bounds, y = c(0, cum_rel))
ggplot(df_ogive, aes(x, y)) +
  geom_line(color = "darkred", linewidth = 1) +
  geom_point(color = "darkred", size = 2) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Relative Frequency Ogive",
       x = "Upper Boundary (miles)", y = "Cumulative Relative Frequency") +
  theme_classic(base_size = 12)
```

## Distribution Shapes

::: {.callout-note icon=false}
## Common Distribution Shapes

Frequency polygons and histograms reveal the **shape** of a distribution. The most important shapes are:

| Shape | Description |
|-------|-------------|
| **Bell-shaped (symmetric)** | Tapers off equally on both sides; data cluster near the centre |
| **Uniform** | Roughly flat — all classes have similar frequencies |
| **J-shaped** | Frequencies increase or decrease monotonically; resembles the letter J |
| **Reverse-J** | Mirror image of J-shaped |
| **Right-skewed (positive skew)** | Long tail extends to the right; most data are at the lower end |
| **Left-skewed (negative skew)** | Long tail extends to the left; most data are at the upper end |
| **Bimodal** | Two distinct peaks; may indicate two different subgroups |
| **U-shaped** | High frequencies at both extremes, low in the middle |
:::

## Distribution Shapes — Illustrated

```{r shapes, echo=FALSE, fig.width=10, fig.height=6}
set.seed(42)
par(mfrow = c(2, 4), mar = c(3, 2, 2, 1))

# Bell-shaped
x_bell <- rnorm(500)
hist(x_bell, main = "Bell-Shaped", xlab = "", col = "steelblue",
     border = "white", axes = FALSE); axis(1)

# Uniform
x_unif <- runif(500)
hist(x_unif, main = "Uniform", xlab = "", col = "tomato",
     border = "white", axes = FALSE); axis(1)

# J-shaped
x_j <- rexp(500, rate = 0.5)
hist(x_j, main = "J-Shaped", xlab = "", col = "goldenrod",
     border = "white", axes = FALSE); axis(1)

# Reverse-J
x_revj <- max(x_j) - x_j
hist(x_revj, main = "Reverse-J", xlab = "", col = "mediumseagreen",
     border = "white", axes = FALSE); axis(1)

# Right-skewed
x_right <- rgamma(500, shape = 2, rate = 0.5)
hist(x_right, main = "Right-Skewed", xlab = "", col = "orchid",
     border = "white", axes = FALSE); axis(1)

# Left-skewed
x_left <- -rgamma(500, shape = 2, rate = 0.5) + max(rgamma(500,2,0.5))
hist(x_left, main = "Left-Skewed", xlab = "", col = "salmon",
     border = "white", axes = FALSE); axis(1)

# Bimodal
x_bi <- c(rnorm(250, -2), rnorm(250, 2))
hist(x_bi, main = "Bimodal", xlab = "", col = "cornflowerblue",
     border = "white", axes = FALSE); axis(1)

# U-shaped
x_u <- c(rbeta(500, 0.3, 0.3))
hist(x_u, main = "U-Shaped", xlab = "", col = "lightcoral",
     border = "white", axes = FALSE); axis(1)
```

## Applying the Concepts 2-2

::: {.callout-tip icon=false}
## ATC 2-2 — Real Estate Selling Prices

The selling prices (in thousands of dollars) of 30 randomly selected homes in a suburb are listed below. Construct a grouped frequency distribution, a histogram, and a frequency polygon. Describe the shape of the distribution.

```
67  73  98  110  85  92  78  88  105  76
83  91  79  103  88  94  82  96  101  85
77  89  92  100  86  95  84  97  108  90
```
:::

---

```{r atc2-2}
prices <- c(67,73,98,110,85,92,78,88,105,76,
            83,91,79,103,88,94,82,96,101,85,
            77,89,92,100,86,95,84,97,108,90)

# Width = ceiling((110-67)/6) = 8; use 6 classes
breaks_p <- seq(64, 114, by = 8)
cut_p    <- cut(prices, breaks = breaks_p, right = FALSE)
freq_p   <- as.data.frame(table(cut_p))

# Histogram
ggplot(data.frame(x = prices), aes(x = x)) +
  geom_histogram(breaks = breaks_p, fill = "steelblue", color = "white") +
  labs(title = "Home Selling Prices",
       x = "Price ($000)", y = "Frequency") +
  theme_classic(base_size = 14)
```

The distribution is approximately **bell-shaped**, with selling prices clustering around the $84,000–$100,000 range.

# §2-3 Other Types of Graphs

## Overview of Additional Graph Types

::: {.callout-note icon=false}
## When to Use Each Graph Type

| Graph | Best Used When |
|-------|---------------|
| **Bar graph** | Comparing frequencies across distinct categories |
| **Pareto chart** | Bar graph where bars are sorted by frequency (descending); highlights the most important categories |
| **Time series graph** | Data are collected over successive time periods; shows trends |
| **Pie graph** | Showing each category's proportion of a whole |
| **Stem-and-leaf plot** | Displaying the shape of a distribution while retaining exact data values |

**Pie chart formula:**
Degrees for each slice $= \dfrac{f}{n} \times 360°$
:::

## Bar Graph

::: {.callout-tip icon=false}
## Example 2-8 — College Spending Categories

A survey asked college students to identify their single largest spending category. The results were:

| Category | Frequency |
|----------|-----------|
| Housing | 423 |
| Food | 211 |
| Transportation | 176 |
| Clothing | 98 |
| Entertainment | 62 |

Construct a **bar graph** to display these data.
:::

---

```{r ex2-8}
categories <- c("Housing", "Food", "Transportation", "Clothing", "Entertainment")
freq_spend <- c(423, 211, 176, 98, 62)

df_spend <- data.frame(Category = factor(categories, levels = categories),
                       Frequency = freq_spend)

ggplot(df_spend, aes(x = Category, y = Frequency)) +
  geom_col(fill = "steelblue") +
  labs(title = "College Student Spending Categories",
       x = "Category", y = "Number of Students") +
  theme_classic(base_size = 14)
```

## Pareto Chart

::: {.callout-tip icon=false}
## Example 2-9 — Homeless Population by City

The number of homeless individuals counted in selected US cities on a given night were:

| City | Count |
|------|-------|
| New York | 36,000 |
| Los Angeles | 28,000 |
| Seattle | 9,000 |
| San Diego | 5,000 |
| Phoenix | 4,000 |

Construct a **Pareto chart**.
:::

---

```{r ex2-9}
cities  <- c("New York", "Los Angeles", "Seattle", "San Diego", "Phoenix")
homeless <- c(36000, 28000, 9000, 5000, 4000)

df_home <- data.frame(City = cities, Count = homeless) |>
  arrange(desc(Count))
df_home$City <- factor(df_home$City, levels = df_home$City)

ggplot(df_home, aes(x = City, y = Count)) +
  geom_col(fill = "tomato") +
  labs(title = "Pareto Chart — Homeless Population by City",
       x = "City", y = "Number of Homeless Individuals") +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## Time Series Graph

::: {.callout-tip icon=false}
## Example 2-10 — Workplace Homicides Over Time

The number of workplace homicides in the United States for selected years is shown below. Construct a **time series graph** and describe the trend.

| Year | 1995 | 1996 | 1997 | 1998 | 1999 | 2000 | 2001 |
|------|------|------|------|------|------|------|------|
| Homicides | 1024 | 927 | 912 | 714 | 645 | 677 | 639 |
:::

---

```{r ex2-10}
years      <- c(1995, 1996, 1997, 1998, 1999, 2000, 2001)
homicides  <- c(1024, 927, 912, 714, 645, 677, 639)

ggplot(data.frame(Year = years, Homicides = homicides),
       aes(x = Year, y = Homicides)) +
  geom_line(color = "darkred", linewidth = 1) +
  geom_point(color = "darkred", size = 3) +
  scale_x_continuous(breaks = years) +
  labs(title = "US Workplace Homicides (1995–2001)",
       x = "Year", y = "Number of Homicides") +
  theme_classic(base_size = 14)
```

The graph shows a clear **downward trend** — workplace homicides declined substantially from 1995 to 2001.

## Pie Graph — Super Bowl Snacks

::: {.callout-tip icon=false}
## Example 2-11 — Super Bowl Snack Preferences

A survey of 1000 football fans revealed their preferred Super Bowl snack:

| Snack | Frequency |
|-------|-----------|
| Potato chips | 410 |
| Pizza | 330 |
| Pretzels | 130 |
| Chicken wings | 80 |
| Other | 50 |

Construct a **pie graph**.
:::

---

```{r ex2-11}
snacks <- c("Potato chips", "Pizza", "Pretzels", "Chicken wings", "Other")
snack_freq <- c(410, 330, 130, 80, 50)

df_snack <- data.frame(Snack = snacks, Freq = snack_freq)
df_snack$Pct <- df_snack$Freq / sum(df_snack$Freq)
df_snack$Label <- paste0(snacks, "\n", round(df_snack$Pct * 100, 1), "%")

ggplot(df_snack, aes(x = "", y = Freq, fill = Snack)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(title = "Super Bowl Snack Preferences") +
  theme_void(base_size = 13) +
  theme(legend.position = "right")
```

## Pie Graph — Blood Types Revisited

::: {.callout-tip icon=false}
## Example 2-12 — Pie Graph for Blood Type Data

Using the blood type data from Example 2-1 (25 inductees), construct a **pie graph**. Calculate the degree measure of each slice using $\text{Degrees} = (f/n) \times 360°$.
:::

```{r ex2-12}
blood_type <- c("A", "AB", "B", "O")
blood_freq <- c(9, 3, 9, 4)   # from Ex 2-1 table
n_blood    <- sum(blood_freq)
degrees    <- round(blood_freq / n_blood * 360, 1)

df_blood <- data.frame(Type = blood_type, Freq = blood_freq,
                       Degrees = degrees,
                       Pct = round(blood_freq / n_blood * 100, 1))
knitr::kable(df_blood, align = "c",
             col.names = c("Blood Type","Frequency","Degrees","Percent"))
```

---

```{r ex2-12-pie, echo=TRUE}
ggplot(df_blood, aes(x = "", y = Freq, fill = Type)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(title = "Blood Type Distribution (n = 25)") +
  theme_void(base_size = 13)
```

## Misleading Graphs

::: {.callout-note icon=false}
## Caution: Common Ways Graphs Mislead

Graphs can create **false impressions** if the reader is not careful. Watch for:

1. **Truncated y-axis** — starting the vertical axis above zero exaggerates differences between bars or lines
2. **Inconsistent scale** — unequal intervals on an axis distort the visual distance between data points
3. **Three-dimensional effects** — 3-D charts make it difficult to judge the true height of bars or the size of pie slices
4. **Pictographs without area scaling** — doubling a picture's height quadruples its area, overstating a 2× difference

**Rule of thumb:** Always check whether the axis begins at zero and whether the scale is uniform before drawing conclusions from a graph.
:::

## Stem-and-Leaf Plot

::: {.callout-note icon=false}
## Definition: Stem-and-Leaf Plot

A **stem-and-leaf plot** is a method for organising numerical data that retains every individual value. Each data value is split into two parts:

- **Stem** — the leading digit(s), written to the left of a vertical line
- **Leaf** — the trailing digit, written to the right

A stem-and-leaf plot serves as a quick histogram while preserving the exact data, making it easy to find the minimum, maximum, and median at a glance.
:::

---

::: {.callout-tip icon=false}
## Example 2-13 — Cardiogram Waiting Times

The waiting times (minutes) at a cardiac diagnostic clinic for 20 patients were:

`24 40 27 29 40 12 18 22 38 35`
`42 16 45 48 32 32 22 26 51 31`

Construct a **stem-and-leaf plot**.
:::

```{r ex2-13}
waiting <- c(24,40,27,29,40,12,18,22,38,35,
             42,16,45,48,32,32,22,26,51,31)
stem(waiting)
```

## Stem-and-Leaf Plot with Classes

::: {.callout-tip icon=false}
## Example 2-14 — Car Thefts by City

The number of automobile thefts (in thousands) for a sample of US cities is:

`56 48 32 68 90 87 42 59 73 81`
`38 44 66 79 83 62 54 71 91 43`

Construct a **stem-and-leaf plot** using the tens digit as the stem. Then identify the shape of the distribution.
:::

```{r ex2-14}
thefts <- c(56,48,32,68,90,87,42,59,73,81,
            38,44,66,79,83,62,54,71,91,43)
stem(thefts)
```

The data show a roughly **uniform or slightly bell-shaped** distribution spanning the 30s to 90s.

## Back-to-Back Stem-and-Leaf Plot

::: {.callout-tip icon=false}
## Example 2-15 — Atlanta vs Philadelphia Temperatures

Average monthly high temperatures (°F) for Atlanta and Philadelphia are given below. Construct a **back-to-back stem-and-leaf plot** and compare the two cities.

| City | Temperatures |
|------|-------------|
| Atlanta | 52 57 65 73 80 86 88 87 81 71 60 52 |
| Philadelphia | 38 42 52 63 73 83 87 85 79 67 55 41 |
:::

---

```{r ex2-15}
atlanta <- c(52,57,65,73,80,86,88,87,81,71,60,52)
philly  <- c(38,42,52,63,73,83,87,85,79,67,55,41)
aplpack::stem.leaf.backback(atlanta,philly, m = 1, depths=F)
```

Atlanta generally has **higher temperatures** in the cooler months, while both cities are similar in summer.

## Applying the Concepts 2-3

::: {.callout-tip icon=false}
## ATC 2-3 — Leading Causes of Death

The number of deaths (in thousands) in the United States for selected leading causes in a recent year are shown below. Construct both a **Pareto chart** and a **pie graph**. Which display do you find more informative?

| Cause | Deaths (000s) |
|-------|--------------|
| Heart disease | 700 |
| Cancer | 553 |
| Stroke | 163 |
| COPD | 122 |
| Accidents | 101 |
| Diabetes | 71 |
| Influenza/Pneumonia | 63 |
:::

---

```{r atc2-3, fig.show='hold', out.width="48%", fig.width=5, fig.height=4}
causes <- c("Heart disease","Cancer","Stroke","COPD",
            "Accidents","Diabetes","Flu/Pneumonia")
deaths <- c(700, 553, 163, 122, 101, 71, 63)

df_death <- data.frame(Cause = factor(causes, levels = causes[order(deaths, decreasing=TRUE)]),
                       Deaths = deaths[order(deaths, decreasing=TRUE)])

# Pareto chart
ggplot(df_death, aes(x = Cause, y = Deaths)) +
  geom_col(fill = "steelblue") +
  labs(title = "Pareto Chart — Leading Causes of Death",
       x = "", y = "Deaths (000s)") +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

---

```{r atc2-3b, fig.show='hold', out.width="40%", fig.width=5, fig.height=4}
# Pie graph
df_pie <- data.frame(Cause = causes, Deaths = deaths)
ggplot(df_pie, aes(x = "", y = Deaths, fill = Cause)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  labs(title = "Pie Graph — Leading Causes of Death") +
  theme_void(base_size = 9)
```

The **Pareto chart** makes it easier to rank causes and compare magnitudes; the **pie graph** is better for visualising each cause's share of total deaths.

# Summary

## Chapter 2 Key Concepts

**Frequency Distributions**

- A **categorical** distribution tabulates named categories; a **grouped** distribution uses equal-width intervals for numerical data; an **ungrouped** distribution lists each distinct value separately.
- Class width $= \lceil (\max - \min) / k \rceil$; class midpoint $= (L + U) / 2$.
- A **cumulative frequency** column shows how many observations fall at or below each class.

**Graphs for Frequency Distributions**

- **Histogram** — adjacent bars, class boundaries on *x*-axis.
- **Frequency polygon** — connects class midpoints; anchored to zero at both ends.
- **Ogive** — cumulative frequency plotted against upper class boundaries.
- **Relative frequency** versions use proportions rather than raw counts.

---

**Other Graph Types**

- **Bar graph / Pareto chart** — categorical comparisons; Pareto sorts bars in descending order.
- **Time series** — shows trends over time.
- **Pie graph** — proportional slices; each slice's angle $= (f/n) \times 360°$.
- **Stem-and-leaf plot** — retains exact values while revealing distribution shape.

## Looking Ahead

In **Chapter 3** we move from visual summaries to **numerical summaries** — measures of central tendency (mean, median, mode) and measures of variation (range, variance, standard deviation) — which allow us to describe a distribution precisely with a handful of numbers.
