---
title: "Statistics"
subtitle: "Chapter 10: Correlation and Regression"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch10-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
```

## Overview
Researchers often ask: *Is there a relationship between two variables, and if so, how strong is it?*

This chapter addresses four key questions:

1. Are two variables linearly related?
2. If so, how strong is that relationship?
3. What type of relationship exists (positive, negative, linear, curvilinear)?
4. What predictions can be made from the relationship?

**Tools:** Scatter plots, correlation coefficients, regression lines, and prediction intervals.

# Section 10-1: Scatter Plots and Correlation {background-color="#2c4a6e"}

## Key Terms

::: {.callout-note}
- **Independent variable (x):** The predictor variable — can be controlled or manipulated.
- **Dependent variable (y):** The response variable — its value depends on x.
- **Scatter plot:** A graph of ordered pairs (x, y) used to visually assess the relationship between two variables.
- **Positive relationship:** Both variables move in the same direction.
- **Negative relationship:** Variables move in opposite directions.
:::

## Example 10-1 {#example-10-1}

:::{.callout-warning}
## Car Rental Companies — Scatter Plot

Construct a scatter plot for the following car rental company data:

| Company | Cars ($\times$10,000) | Revenue ($ billions) |
|---------|----------------|----------------------|
| A | 63.0 | 7.0 |
| B | 29.0 | 3.9 |
| C | 20.8 | 2.1 |
| D | 19.1 | 2.8 |
| E | 13.4 | 1.4 |
| F | 8.5 | 1.5 |
:::

## Example 10-1

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
cars    <- c(63.0, 29.0, 20.8, 19.1, 13.4, 8.5)
revenue <- c(7.0, 3.9, 2.1, 2.8, 1.4, 1.5)
df <- data.frame(cars, revenue)

ggplot(df, aes(x = cars, y = revenue)) +
  geom_point(size = 3, color = "steelblue") +
  labs(x = "Cars (in 10,000s)", y = "Revenue ($ billions)",
       title = "Car Rental Companies") +
  theme_minimal()
```

The scatter plot suggests a **positive linear relationship** — as fleet size increases, so does revenue.
:::

## Example 10-2 {#example-10-2}

:::{.callout-warning}
## Absences and Final Grades — Scatter Plot

Construct a scatter plot for the number of absences and final grades for seven statistics students:

| Student | Absences x | Final Grade y (%) |
|---------|------------|-------------------|
| A | 6 | 82 |
| B | 2 | 86 |
| C | 15 | 43 |
| D | 9 | 74 |
| E | 12 | 58 |
| F | 5 | 90 |
| G | 8 | 78 |
:::

## Example 10-2

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
absences <- c(6, 2, 15, 9, 12, 5, 8)
grade    <- c(82, 86, 43, 74, 58, 90, 78)
df2 <- data.frame(absences, grade)

ggplot(df2, aes(x = absences, y = grade)) +
  geom_point(size = 3, color = "tomato") +
  labs(x = "Number of Absences", y = "Final Grade (%)",
       title = "Absences vs. Final Grade") +
  theme_minimal()
```

The plot suggests a **negative linear relationship** — more absences correspond to lower grades.
:::

## Example 10-3 {#example-10-3}

:::{.callout-warning}
## Age and Wealth — Scatter Plot

A researcher examines the relationship between age and net worth (in $ billions) for the wealthiest people in the U.S.:

| Person | Age x | Wealth y ($ billions) |
|--------|-------|-----------------------|
| A | 73 | 16 |
| B | 65 | 26 |
| C | 53 | 50 |
| D | 54 | 21.5 |
| E | 79 | 40 |
| F | 69 | 16 |
| G | 61 | 19.6 |
| H | 65 | 19 |
:::

## Example 10-3

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
age   <- c(73, 65, 53, 54, 79, 69, 61, 65)
wealth <- c(16, 26, 50, 21.5, 40, 16, 19.6, 19)
df3 <- data.frame(age, wealth)

ggplot(df3, aes(x = age, y = wealth)) +
  geom_point(size = 3, color = "purple") +
  labs(x = "Age", y = "Wealth ($ billions)",
       title = "Age vs. Net Wealth") +
  theme_minimal()
```

No clear pattern is visible — this suggests **little or no linear relationship** between age and wealth.
:::

## The Pearson Correlation Coefficient

::: {.callout-note}
The **Pearson product moment correlation coefficient** (r) measures the strength and direction of the linear relationship between two variables.

$$r = \frac{n\sum xy - (\sum x)(\sum y)}{\sqrt{[n\sum x^2 - (\sum x)^2][n\sum y^2 - (\sum y)^2]}}$$

- Range: $-1 \leq r \leq +1$
- Close to +1: strong positive relationship
- Close to −1: strong negative relationship
- Close to 0: weak or no linear relationship

**Assumptions:** Random sample; approximately linear relationship; interval/ratio data; bivariate normal distribution.
:::

## Example 10-4 {#example-10-4}

:::{.callout-warning}
## Correlation for Car Rental Companies

Compute the correlation coefficient for the car rental data from [Example 10-1](#example-10-1).
:::

## Example 10-4

:::{.callout-tip}
## Solution

**Mathematical solution**

Build the calculation table:

| Company | x | y | xy | x² | y² |
|---------|-----|------|--------|---------|-------|
| A | 63.0 | 7.0 | 441.00 | 3969.00 | 49.00 |
| B | 29.0 | 3.9 | 113.10 | 841.00 | 15.21 |
| C | 20.8 | 2.1 | 43.68 | 432.64 | 4.41 |
| D | 19.1 | 2.8 | 53.48 | 364.81 | 7.84 |
| E | 13.4 | 1.4 | 18.76 | 179.56 | 1.96 |
| F | 8.5 | 1.5 | 12.75 | 72.25 | 2.25 |
| **Σ** | **153.8** | **18.7** | **682.77** | **5859.26** | **80.67** |

$$r = \frac{6(682.77) - (153.8)(18.7)}{\sqrt{[6(5859.26) - 153.8^2][6(80.67) - 18.7^2]}} \approx \boxed{0.982}$$
:::

## Example 10-4

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- c(63.0, 29.0, 20.8, 19.1, 13.4, 8.5)
y <- c(7.0, 3.9, 2.1, 2.8, 1.4, 1.5)

r <- cor(x, y)
cat("Correlation coefficient r =", round(r, 4), "\n")
cat("Strong positive relationship — r is close to 1.\n")
```
:::

## Example 10-5 {#example-10-5}

:::{.callout-warning}
## Correlation for Absences and Final Grades

Compute the correlation coefficient for the absences and final grade data from [Example 10-2](#example-10-2).
:::

## Example 10-5

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- c(6, 2, 15, 9, 12, 5, 8)
y <- c(82, 86, 43, 74, 58, 90, 78)

r <- cor(x, y)
cat("r =", round(r, 4), "\n")
cat("Strong negative relationship — more absences, lower grade.\n")
```

**Result:** $r \approx -0.944$ → strong negative linear relationship.
:::

## Example 10-6 {#example-10-6}

:::{.callout-warning}
## Correlation for Age and Wealth

Compute the correlation coefficient for the age and wealth data from [Example 10-3](#example-10-3).
:::

## Example 10-6

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- c(73, 65, 53, 54, 79, 69, 61, 65)
y <- c(16, 26, 50, 21.5, 40, 16, 19.6, 19)

r <- cor(x, y)
cat("r =", round(r, 4), "\n")
cat("Very weak negative relationship — r is close to 0.\n")
```

**Result:** $r \approx -0.176$ → very weak relationship, likely due to chance.
:::

## Testing the Significance of r

::: {.callout-note}
To determine whether r reflects a true linear relationship (and not just sampling error), we test:

$$H_0: \rho = 0 \quad H_1: \rho \neq 0$$

**t test for the correlation coefficient:**
$$t = r\sqrt{\frac{n-2}{1-r^2}}, \quad d.f. = n-2$$

Alternatively, use **Table I** (critical values of r) directly — no need to compute t.
:::

## Example 10-7 {#example-10-7}

:::{.callout-warning}
## Test Significance of r = 0.982

Test whether the correlation r = 0.982 from [Example 10-4](#example-10-4) is significant at α = 0.05.
:::

## Example 10-7

:::{.callout-tip}
## Solution

**Solve**

**Step 1:** H₀: ρ = 0 &nbsp; H₁: ρ $\neq 0$

**Step 2:** d.f. = 6 − 2 = 4; CV = ±2.776 (from t-table)

**Step 3:**
$$t = 0.982\sqrt{\frac{6-2}{1-0.982^2}} = 0.982\sqrt{\frac{4}{0.0356}} \approx 10.4$$

**Step 4:** |10.4| > 2.776 → **Reject H₀**

**Step 5:** There is a significant linear relationship between fleet size and revenue.
:::

## Example 10-7

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- c(63.0, 29.0, 20.8, 19.1, 13.4, 8.5)
y <- c(7.0, 3.9, 2.1, 2.8, 1.4, 1.5)

test <- cor.test(x, y, alternative = "two.sided")
cat("r =", round(test$estimate, 4), "\n")
cat("t =", round(test$statistic, 4), "\n")
cat("P-value =", round(test$p.value, 6), "\n")
cat("Decision:", ifelse(test$p.value < 0.05, "Reject H0", "Do not reject H0"), "\n")
```
:::

## Example 10-8 {#example-10-8}

:::{.callout-warning}
## Test Significance Using Table I (r = −0.176)

Using Table I, test the significance of r = −0.176 from [Example 10-6](#example-10-6) at α = 0.01.
:::

## Example 10-8

:::{.callout-tip}
## Solution

**Solve**

A significant correlation between two variables does **not** prove causation. Possible explanations include:

1. **Direct cause-and-effect** (x causes y)
2. **Reverse causation** (y causes x)
3. **Lurking variable** — a third variable causes both x and y
4. **Complex interrelationships** among many variables
5. **Coincidence** — the relationship is due to chance

:::

Always interpret a significant r cautiously and consider alternative explanations.

# Section 10-2: Regression {background-color="#2c4a6e"}

## The Regression Line (Line of Best Fit)

::: {.callout-note}
When r is significant, we can describe the linear relationship with the **regression line**:

$$\hat{y} = a + bx$$

where:
$$b = \frac{n\sum xy - (\sum x)(\sum y)}{n\sum x^2 - (\sum x)^2}$$

$$a = \frac{(\sum y)(\sum x^2) - (\sum x)(\sum xy)}{n\sum x^2 - (\sum x)^2}$$

- $b$ = slope (marginal change in y per one-unit increase in x)
- $a$ = y-intercept
- **Round a and b to three decimal places**
:::
:::

## Example 10-9 {#example-10-9}

:::{.callout-warning}
## Regression Line for Car Rental Data

Find the equation of the regression line for the car rental data from [Example 10-4](#example-10-4).

Known values: n = 6, Σx = 153.8, Σy = 18.7, Σxy = 682.77, Σx² = 5859.26
:::

## Example 10-9

:::{.callout-tip}
## Solution

**Solve**

**Solution:**

$$b = \frac{6(682.77) - (153.8)(18.7)}{6(5859.26) - 153.8^2} = \frac{4096.62 - 2876.06}{35155.56 - 23653.44} = \frac{1220.56}{11502.12} \approx 0.106$$

$$a = \frac{(18.7)(5859.26) - (153.8)(682.77)}{6(5859.26) - 153.8^2} \approx 0.396$$

$$\boxed{\hat{y} = 0.396 + 0.106x}$$
:::

## Example 10-9

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-9-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- c(63.0, 29.0, 20.8, 19.1, 13.4, 8.5)
y <- c(7.0, 3.9, 2.1, 2.8, 1.4, 1.5)

model <- lm(y ~ x)
cat("Intercept a =", round(coef(model)[1], 3), "\n")
cat("Slope b =",     round(coef(model)[2], 3), "\n")

ggplot(data.frame(x, y), aes(x, y)) +
  geom_point(size = 3, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Car Rental: Revenue vs Fleet Size",
       x = "Cars ($\times$10,000)", y = "Revenue ($ billions)") +
  theme_minimal()
```
:::

## Example 10-10 {#example-10-10}

:::{.callout-warning}
## Regression Line for Absences and Grades

Find the equation of the regression line for the absences/grade data from [Example 10-2](#example-10-2).

n = 7, Σx = 57, Σy = 511, Σxy = 3745, Σx² = 579
:::

## Example 10-10

:::{.callout-tip}
## Solution

**Solve**

$$b = \frac{7(3745) - (57)(511)}{7(579) - 57^2} = \frac{26215 - 29127}{4053 - 3249} = \frac{-2912}{804} \approx -3.622$$

$$a = \frac{(511)(579) - (57)(3745)}{7(579) - 57^2} = \frac{295869 - 213465}{804} \approx 102.493$$

$$\boxed{\hat{y} = 102.493 - 3.622x}$$
:::

## Example 10-10

:::{.callout-tip}
## Solution

**R implementation**

Using the regression equation from [Example 10-9](#example-10-9), predict the revenue of a car rental agency with **200,000 automobiles**.

:::

Since x values are in 10,000s: x = 200,000 ÷ 10,000 = **20**

$$\hat{y} = 0.396 + 0.106(20) = 0.396 + 2.120 = \boxed{2.516 \text{ billion dollars}}$$

```{r e10-10-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- c(63.0, 29.0, 20.8, 19.1, 13.4, 8.5)
y <- c(7.0, 3.9, 2.1, 2.8, 1.4, 1.5)
model <- lm(y ~ x)

pred <- predict(model, newdata = data.frame(x = 20))
cat("Predicted revenue for 200,000 cars: $", round(pred, 3), "billion\n")
```

::: {.callout-important}
**Caution:** Only predict within the range of observed x values. Extrapolation beyond the data is unreliable.
:::

# Section 10-3: Coefficient of Determination and Standard Error of the Estimate {background-color="#2c4a6e"}

## Types of Variation

::: {.callout-note}
The **total variation** in y can be decomposed:

$$\underbrace{\sum(y - \bar{y})^2}_{\text{Total variation}} = \underbrace{\sum(\hat{y} - \bar{y})^2}_{\text{Explained variation}} + \underbrace{\sum(y - \hat{y})^2}_{\text{Unexplained variation}}$$

- **Explained variation:** attributed to the linear relationship with x
- **Unexplained variation:** due to chance or other factors (also called **residual** variation)
:::
:::

## Coefficient of Determination

::: {.callout-note}
$$r^2 = \frac{\text{Explained variation}}{\text{Total variation}}$$

Equivalently, $r^2$ is obtained by **squaring the correlation coefficient**.

- $r^2$ expressed as a percentage tells us what proportion of variation in y is explained by x.
- **Coefficient of nondetermination:** $1 - r^2$ (unexplained proportion)

**Example:** If r = 0.982, then r² = 0.964, meaning **96.4%** of the variation in revenue is explained by fleet size.
:::

## Coefficient of Determination — R Examples

```{r e10-10-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
r_vals <- c(0.982, -0.944, -0.176, 0.90, 0.60)
for (r in r_vals) {
  cat(sprintf("r = %6.3f  →  r² = %.4f  (%.1f%% explained)\n",
              r, r^2, r^2 * 100))
}
```

## Standard Error of the Estimate

::: {.callout-note}
The **standard error of the estimate** ($s_{est}$) measures the spread of observed y values around the regression line:

$$s_{est} = \sqrt{\frac{\sum(y - \hat{y})^2}{n-2}}$$

Computational shortcut:

$$s_{est} = \sqrt{\frac{\sum y^2 - a\sum y - b\sum xy}{n-2}}$$

The smaller $s_{est}$, the better the regression line fits the data.
:::

## Example 10-12 {#example-10-12}

:::{.callout-warning}
## Copy Machine Maintenance Costs

A researcher finds a significant relationship between the age of a copy machine (years) and its monthly maintenance cost (dollars). The regression equation is $\hat{y} = 55.57 + 8.13x$. Find the standard error of the estimate.

| Machine | Age x | Cost y |
|---------|-------|--------|
| A | 1 | $62 |
| B | 2 | $78 |
| C | 3 | $70 |
| D | 4 | $90 |
| E | 4 | $93 |
| F | 6 | $103 |
:::

## Example 10-12

:::{.callout-tip}
## Solution

**Mathematical solution**

Compute $\hat{y}$ and residuals for each observation:

| x | y | $\hat{y}$ | $y - \hat{y}$ | $(y - \hat{y})^2$ |
|---|---|-----------|----------------|---------------------|
| 1 | 62 | 63.70 | −1.70 | 2.89 |
| 2 | 78 | 71.83 | 6.17 | 38.07 |
| 3 | 70 | 79.96 | −9.96 | 99.20 |
| 4 | 90 | 88.09 | 1.91 | 3.65 |
| 4 | 93 | 88.09 | 4.91 | 24.11 |
| 6 | 103 | 104.35 | −1.35 | 1.82 |
| | | | **Σ** | **169.74** |

$$s_{est} = \sqrt{\frac{169.74}{6-2}} = \sqrt{42.44} \approx \boxed{6.51}$$
:::

## Example 10-12

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-12-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- c(1, 2, 3, 4, 4, 6)
y <- c(62, 78, 70, 90, 93, 103)

model3 <- lm(y ~ x)
cat("Intercept a =", round(coef(model3)[1], 3), "\n")
cat("Slope b =",     round(coef(model3)[2], 3), "\n")

sest <- summary(model3)$sigma
cat("Standard error of estimate s_est =", round(sest, 4), "\n")
cat("r² =", round(summary(model3)$r.squared, 4), "\n")
```
:::

## Example 10-13 {#example-10-13}

:::{.callout-warning}
## Standard Error via Computational Formula

Using the shortcut formula and the data from [Example 10-12](#example-10-12) (where Σy² = 42,186, Σy = 496, Σxy = 1778, a = 55.57, b = 8.13):
:::

## Example 10-13

:::{.callout-tip}
## Solution

**R implementation**

$$s_{est} = \sqrt{\frac{42186 - 55.57(496) - 8.13(1778)}{6-2}} \approx \boxed{6.48}$$

```{r e10-13-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- c(1, 2, 3, 4, 4, 6)
y <- c(62, 78, 70, 90, 93, 103)
model3 <- lm(y ~ x)
a <- coef(model3)[1]; b <- coef(model3)[2]
sest_shortcut <- sqrt((sum(y^2) - a*sum(y) - b*sum(x*y)) / (length(y) - 2))
cat("s_est (shortcut formula) =", round(sest_shortcut, 4), "\n")
```

*(Small discrepancy due to rounding in a and b.)*
:::

## Prediction Interval

::: {.callout-note}
A **prediction interval** provides a range within which a future individual y value is expected to fall for a given x:

$$\hat{y} \pm t_{\alpha/2} \cdot s_{est} \sqrt{1 + \frac{1}{n} + \frac{n(x - \bar{X})^2}{n\sum x^2 - (\sum x)^2}}$$

with $d.f. = n - 2$
:::

## Example 10-14 {#example-10-14}

:::{.callout-warning}
## 95% Prediction Interval

For the copy machine data, find the 95% prediction interval for the monthly maintenance cost when the machine is **3 years old**.
:::

## Example 10-14

:::{.callout-tip}
## Solution

**Solve**

Known: $\sum x = 20$, $\sum x^2 = 82$, $\bar{X} = 20/6 = 3.3$, $\hat{y}(3) = 79.96$, $s_{est} = 6.48$, $t_{0.025, 4} = 2.776$

$$\text{SE factor} = \sqrt{1 + \frac{1}{6} + \frac{6(3-3.3)^2}{6(82)-20^2}} = \sqrt{1 + 0.167 + \frac{0.54}{92}} \approx 1.08$$

$$79.96 \pm 2.776(6.48)(1.08) = 79.96 \pm 19.43$$

$$\boxed{60.53 < y < 99.39}$$
:::

## Example 10-14

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-14-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- c(1, 2, 3, 4, 4, 6)
y <- c(62, 78, 70, 90, 93, 103)
model3 <- lm(y ~ x)

pred_int <- predict(model3, newdata = data.frame(x = 3),
                    interval = "prediction", level = 0.95)
cat("Predicted y at x=3:", round(pred_int[1], 2), "\n")
cat("95% Prediction Interval: (",
    round(pred_int[2], 2), ",", round(pred_int[3], 2), ")\n")
```

We are 95% confident the actual maintenance cost for a 3-year-old machine falls between **\$60.53 and \$99.39**.
:::

## Residual Plot

```{r e10-14-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- c(1, 2, 3, 4, 4, 6)
y <- c(62, 78, 70, 90, 93, 103)
model3 <- lm(y ~ x)

df_resid <- data.frame(x, residuals = resid(model3))
ggplot(df_resid, aes(x = x, y = residuals)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residual Plot", x = "Age (years)", y = "Residual") +
  theme_minimal()
```

If residuals are randomly scattered around 0, the linear model is appropriate.

# Section 10-4: Multiple Regression (Optional) {background-color="#2c4a6e"}

## Multiple Regression Overview

::: {.callout-note}
**Multiple regression** uses two or more independent variables to predict a single dependent variable:

$$\hat{y} = a + b_1 x_1 + b_2 x_2 + \cdots + b_k x_k$$

- Each $b_i$ is a **partial regression coefficient**: the change in y for a one-unit increase in $x_i$, holding all other x's constant.

**Assumptions:**
1. Normality of y values for each x
2. Equal variances (homoscedasticity)
3. Linearity
4. Non-multicollinearity (independent variables not highly correlated)
5. Independence of observations
:::

## Example 10-13

:::{.callout-tip}
## Solution

A nursing instructor wants to predict a student's state board score (y) from GPA (x₁) and age (x₂). Data for 5 students:

| Student | GPA x₁ | Age x₂ | Score y |
|---------|--------|--------|---------|
| A | 3.2 | 22 | 550 |
| B | 2.7 | 27 | 570 |
| C | 2.5 | 24 | 525 |
| D | 3.4 | 28 | 670 |
| E | 2.2 | 23 | 490 |

Find the multiple regression equation.
:::

## Example 10-13

:::{.callout-tip}
## Solution

**R implementation**

```{r e10-13-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
gpa   <- c(3.2, 2.7, 2.5, 3.4, 2.2)
age   <- c(22, 27, 24, 28, 23)
score <- c(550, 570, 525, 670, 490)

model_mr <- lm(score ~ gpa + age)
summary(model_mr)
```
:::
---

$$\boxed{\hat{y} = -44.81 + 87.64 \cdot \text{GPA} + 14.533 \cdot \text{Age}}$$

For a student with GPA = 3.0 and age = 25: $\hat{y} = -44.81 + 87.64(3.0) + 14.533(25) \approx \mathbf{581}$

## Predicting with the Multiple Regression Equation

```{r e10-13-3, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
new_student <- data.frame(gpa = 3.0, age = 25)
pred_score  <- predict(model_mr, newdata = new_student)
cat("Predicted state board score:", round(pred_score, 1), "\n")
```

**Interpretation of coefficients:**
- For each 1-point increase in GPA (holding age constant), the score increases by about **87.64 points**
- For each 1-year increase in age (holding GPA constant), the score increases by about **14.53 points**

## Multiple Correlation Coefficient R

::: {.callout-note}
The **multiple correlation coefficient R** measures the strength of the relationship between all predictors and the dependent variable:

$$R = \sqrt{\frac{r_{yx_1}^2 + r_{yx_2}^2 - 2r_{yx_1}r_{yx_2}r_{x_1x_2}}{1 - r_{x_1x_2}^2}}$$

- Range: 0 to 1 (never negative)
- $R^2$ = **coefficient of multiple determination** (proportion of variation explained)
- $1 - R^2$ = residual/error variation
:::

## Example 10-15 {#example-10-15}

:::{.callout-warning}
## Multiple Correlation Coefficient

For the state board score data, find R given:
$r_{yx_1} = 0.845$, $r_{yx_2} = 0.791$, $r_{x_1x_2} = 0.371$
:::

## Example 10-15

:::{.callout-tip}
## Solution

**R implementation**

$$R = \sqrt{\frac{0.845^2 + 0.791^2 - 2(0.845)(0.791)(0.371)}{1 - 0.371^2}} = \sqrt{\frac{0.8428}{0.8623}} \approx \boxed{0.989}$$

```{r e10-15-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
ryx1 <- 0.845; ryx2 <- 0.791; rx1x2 <- 0.371
R <- sqrt((ryx1^2 + ryx2^2 - 2*ryx1*ryx2*rx1x2) / (1 - rx1x2^2))
cat("Multiple correlation coefficient R =", round(R, 4), "\n")
cat("Coefficient of multiple determination R² =", round(R^2, 4), "\n")
```
:::

## F Test for Significance of R

::: {.callout-note}
Test $H_0: \rho = 0$ vs $H_1: \rho \neq 0$ using:

$$F = \frac{R^2/k}{(1-R^2)/(n-k-1)}$$

- d.f.N. = k &nbsp;&nbsp; d.f.D. = n − k − 1

where k = number of independent variables, n = number of data groups.
:::

## Example 10-16 {#example-10-16}

:::{.callout-warning}
## F Test for Multiple Correlation

Test the significance of R = 0.989 at α = 0.05 for the state board data (n = 5, k = 2).
:::

## Example 10-16

:::{.callout-tip}
## Solution

**R implementation**

$$F = \frac{0.989^2/2}{(1-0.989^2)/(5-2-1)} = \frac{0.978/2}{0.022/2} = \frac{0.489}{0.011} \approx 44.45$$

Critical value (Table H, d.f.N. = 2, d.f.D. = 2, α = 0.05): **19.16**

44.45 > 19.16 → **Reject H₀** — significant relationship exists.

```{r e10-16-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
R2 <- 0.989^2; n <- 5; k <- 2
F_stat <- (R2/k) / ((1-R2)/(n-k-1))
cv <- qf(0.95, df1 = k, df2 = n-k-1)
cat("F =", round(F_stat, 2), "  Critical value =", round(cv, 2), "\n")
cat("Decision:", ifelse(F_stat > cv, "Reject H0", "Do not reject H0"), "\n")
```
:::

## Example 10-17 {#example-10-17}

:::{.callout-warning}
## Adjusted R²

The **adjusted R²** corrects for the number of predictors and sample size:

$$R^2_{adj} = 1 - \frac{(1-R^2)(n-1)}{n-k-1}$$

It is always $\leq R^2$ and penalizes for adding unnecessary predictors.
:::

## Example 10-17

:::{.callout-tip}
## Solution

**Solve**

Calculate the adjusted R² for R = 0.989, n = 5, k = 2.

:::

$$R^2_{adj} = 1 - \frac{(1-0.989^2)(5-1)}{5-2-1} = 1 - \frac{(0.022)(4)}{2} = 1 - 0.044 = \boxed{0.956}$$

```{r e10-17-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
R2 <- 0.989^2; n <- 5; k <- 2
R2_adj <- 1 - (1-R2)*(n-1)/(n-k-1)
cat("R² =", round(R2, 4), "\n")
cat("Adjusted R² =", round(R2_adj, 4), "\n")
```
:::

## Key Takeaways

| Concept | Formula | Purpose |
|--------|--------------|---------|
| Correlation coefficient | $r = \frac{n\sum xy - \sum x \sum y}{\sqrt{[n\sum x^2 - (\sum x)^2][n\sum y^2 - (\sum y)^2]}}$ | Strength & direction of linear relationship |
| t test for r | $t = r\sqrt{\frac{n-2}{1-r^2}}$, d.f. = n−2 | Test if r is significant |
| Regression line | $\hat{y} = a + bx$ | Predict y from x |
| Coefficient of determination | $r^2$ | % of variation in y explained by x |

---

| Concept | Formula | Purpose |
|--------|--------------|---------|
| Standard error of estimate | $s_{est} = \sqrt{\frac{\sum(y-\hat{y})^2}{n-2}}$ | Accuracy of predictions |
| Prediction interval | $\hat{y} \pm t_{\alpha/2} \cdot s_{est} \cdot \text{SE factor}$ | Range for individual y |
| Multiple R | $R = \sqrt{\frac{r_{yx_1}^2 + r_{yx_2}^2 - 2r_{yx_1}r_{yx_2}r_{x_1x_2}}{1-r_{x_1x_2}^2}}$ | Multiple predictors |

## Acknowledgement
-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2011). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.