---
title: "Statistics"
subtitle: "Chapter 3: Data Description"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
language:
  callout-warning-title: "Problem"
  callout-important-title: "Key point"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch3-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_minimal())
```

## Overview
Chapter 3 shows the statistical methods used to **summarize data** — measures of average, measures of variation, and measures of position.

| Section | Topics |
|---------|--------|
| 3-1 | Mean, median, mode, midrange, weighted mean; distribution shapes |
| 3-2 | Range, variance, standard deviation, coefficient of variation; range rule of thumb; Chebyshev's theorem; empirical rule; linear transformations |
| 3-3 | Standard scores, percentiles, deciles, quartiles, outliers |
| 3-4 | Five-number summary and boxplots |

## Chapter Objectives

:::{.callout-important}
## After completing this chapter, you should be able to

1. Summarize data, using measures of central tendency, such as the mean, median, mode, and midrange.
2. Describe data, using measures of variation, such as the range, variance, and standard deviation.
3. Identify the position of a data value in a data set, using various measures of position, such as percentiles, deciles, and quartiles.
4. Use the techniques of exploratory data analysis, including boxplots and five-number summaries, to discover various aspects of data.
:::

# Section 3-1: Measures of Central Tendency {background-color="#2c4a6e"}

## Statistics and Parameters

Loosely stated, the **average** means the center of the distribution or the most typical case. Measures of average are also called **measures of central tendency** and include the mean, median, mode, and midrange.

:::{.callout-note}
## Statistic and Parameter

- A **statistic** is a characteristic or measure obtained by using the data values from a **sample**.
- A **parameter** is a characteristic or measure obtained by using all the data values from a specific **population**.
:::

:::{.callout-important}
## General Rounding Rule

Rounding should **not** be done until the final answer is calculated. Rounding in intermediate steps tends to increase the difference between that answer and the exact one.
:::

## The Mean

The **mean**, also known as the arithmetic average, is found by adding the values of the data and dividing by the total number of values.

:::{.callout-note}
## The Mean

The **mean** is the sum of the values, divided by the total number of values.

The **sample mean**, denoted by $\bar{X}$, is a statistic:
$$\bar{X} = \frac{X_1 + X_2 + \cdots + X_n}{n} = \frac{\Sigma X}{n}$$

The **population mean**, denoted by $\mu$, is a parameter:
$$\mu = \frac{X_1 + X_2 + \cdots + X_N}{N} = \frac{\Sigma X}{N}$$
:::

## Rounding Rule for the Mean

:::{.callout-important}
## Rounding Rule for the Mean

The mean should be rounded to **one more decimal place** than occurs in the raw data. If the raw data are whole numbers, round the mean to the nearest tenth; if the data are in tenths, round to the nearest hundredth.
:::

Greek letters denote parameters; Roman letters denote statistics. Assume the data are obtained from samples unless otherwise specified.

## Example 3-1 {#example-3-1}

:::{.callout-warning}
## Monthly Revenue of a Coffee Chain

The figures show last month's revenue (in NT\$ million) for the nine branches of a coffee chain. Find the mean.

```
2.8   3.5   4.1   2.6   5.0   3.9   4.4   3.2   4.7
```

*(hypothetical data)*
:::

---

## Example 3-1

:::{.callout-tip}
## Solution

**Solve**

```{r e3-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
coffee_revenue <- c(2.8, 3.5, 4.1, 2.6, 5.0, 3.9, 4.4, 3.2, 4.7)

sum(coffee_revenue)
mean(coffee_revenue)
```

$$\bar{X} = \frac{\Sigma X}{n} = \frac{34.2}{9} = 3.8$$

The mean monthly revenue for the nine branches is **NT\$3.8 million**.
:::

## Example 3-2 {#example-3-2}

:::{.callout-warning}
## Export Value of a Trading Company

The data show the annual export value (in NT\$ million) of a small trading company over a 5-year period. Find the mean.

```
486   512   545   578   614
```

*(hypothetical data)*
:::

---

## Example 3-2

:::{.callout-tip}
## Solution

**Solve**

```{r e3-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
export_value <- c(486, 512, 545, 578, 614)

sum(export_value)
mean(export_value)
```

$$\bar{X} = \frac{\Sigma X}{n} = \frac{2735}{5} = 547$$

The mean annual export value over the 5-year period is **NT\$547 million**.
:::

## The Mean for Grouped Data

The procedure for finding the mean for grouped data assumes that the mean of all the raw data values in each class is equal to the **midpoint** of the class.

:::{.callout-important}
## Procedure — Finding the Mean for Grouped Data

1. Make a table with columns A (class), B (frequency $f$), C (midpoint $X_m$), D ($f \cdot X_m$).
2. Find the midpoint of each class and place it in column C.
3. Multiply the frequency by the midpoint for each class; place the product in column D.
4. Find the sum of column D.
5. Divide the sum of column D by the sum of the frequencies.

$$\bar{X} = \frac{\Sigma f \cdot X_m}{n}$$
:::

## Example 3-3 {#example-3-3}

:::{.callout-warning}
## Revenue of Night Market Stalls

The frequency distribution shows last month's revenue (in NT\$10,000) for 25 stalls in a night market. Find the mean.

```
Class boundaries    Frequency
    9.5-14.5             8
   14.5-19.5             7
   19.5-24.5             5
   24.5-29.5             3
   29.5-34.5             2
             Total      25
```

*(hypothetical data)*

Back to [Example 3-9](#example-3-9)
:::

---

## Example 3-3

:::{.callout-tip}
## Solution

**Prepare data**

```{r e3-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
f  <- c(8, 7, 5, 3, 2)
Xm <- c(12, 17, 22, 27, 32)

f * Xm
```
:::

---

## Example 3-3

:::{.callout-tip}
## Solution

**Solve**

```{r e3-3-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sum(f)
sum(f * Xm)
sum(f * Xm) / sum(f)
```

$$\bar{X} = \frac{\Sigma f \cdot X_m}{n} = \frac{470}{25} = 18.8$$

The mean monthly revenue is **NT\$188,000** per stall.
:::

## The Median

An article reported that the median income for college professors was \$82,974: one-half of all professors surveyed earned more than that, and one-half earned less.

:::{.callout-note}
## Data Array and Median

When the data set is ordered from lowest to highest, it is called a **data array**.

The **median** is the midpoint of the data array. The symbol for the median is $MD$.
:::

:::{.callout-important}
## Procedure — Finding the Median

1. Arrange the data values in ascending order.
2. Determine the number of values in the data set.
3. If $n$ is **odd**, select the middle data value as the median. If $n$ is **even**, find the mean of the two middle values.
:::

## Example 3-4 {#example-3-4}

:::{.callout-warning}
## Passengers at MRT Stations

The data show the average number of daily passengers (in thousands) entering seven MRT stations. Find the median.

```
42   57   38   61   49   55   44
```

*(hypothetical data)*
:::

---

## Example 3-4

:::{.callout-tip}
## Solution

**Solve**

```{r e3-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mrt_passengers <- c(42, 57, 38, 61, 49, 55, 44)

sort(mrt_passengers)
length(mrt_passengers)
median(mrt_passengers)
```

There are an odd number of data values, namely 7, so the middle data value is the median: $MD = 49$.

The median number of daily passengers is **49 thousand**.
:::

## Example 3-5 {#example-3-5}

:::{.callout-warning}
## Container Throughput

The number of containers (in thousands of TEU) handled by a port over eight quarters is as follows. Find the median.

```
146   133   171   119   154   126   162   140
```

*(hypothetical data)*
:::

---

## Example 3-5

:::{.callout-tip}
## Solution

**Solve**

```{r e3-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
containers_teu <- c(146, 133, 171, 119, 154, 126, 162, 140)

sort(containers_teu)
length(containers_teu)
median(containers_teu)
```

The middle two data values are 140 and 146, so

$$MD = \frac{140 + 146}{2} = \frac{286}{2} = 143$$

The median throughput is **143 thousand TEU**.
:::

## The Mode

The third measure of average is the **mode** — sometimes said to be the most typical case.

:::{.callout-note}
## Mode

The value that occurs most often in a data set is called the **mode**.

- **Unimodal** — only one value occurs with the greatest frequency.
- **Bimodal** — two values occur with the same greatest frequency; both are modes.
- **Multimodal** — more than two values occur with the same greatest frequency.
- **No mode** — no data value occurs more than once.
:::

:::{.callout-warning}
## Caution

When no value occurs more than once, do **not** say that the mode is zero. That would be incorrect, because in some data, such as temperature, zero can be an actual value.
:::

## Example 3-6 {#example-3-6}

:::{.callout-warning}
## Food Delivery Times

The delivery times, in minutes, for seven randomly selected food delivery orders are shown. Find the mode.

```
32   28   35   28   41   37   26
```

*(hypothetical data)*
:::

---

## Example 3-6

:::{.callout-tip}
## Solution

**Solve**

```{r e3-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
delivery_time <- c(32, 28, 35, 28, 41, 37, 26)

sort(delivery_time)
table(delivery_time)
```

The mode is **28 minutes**, because it occurs twice, which is more than any other value.
:::

## Example 3-7 {#example-3-7}

:::{.callout-warning}
## Pickers on Duty

The data show the number of pickers on duty at a distribution center over a 15-day period. Find the mode.

```
18   18   18   18   18
21   23   23   23   24
23   25   26   25   23
```

*(hypothetical data)*
:::

---

## Example 3-7

:::{.callout-tip}
## Solution

**Solve**

```{r e3-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
pickers <- c(18, 18, 18, 18, 18,
             21, 23, 23, 23, 24,
             23, 25, 26, 25, 23)

table(pickers)
```

Since the values 18 and 23 both occur 5 times, the modes are **18 and 23**. The data set is said to be **bimodal**.
:::

## Example 3-8 {#example-3-8}

:::{.callout-warning}
## Hotel Ratings

The data show the average customer rating of 8 business hotels on a booking platform. Find the mode.

```
4.3   3.8   4.6   3.5   4.1   3.9   4.8   4.2
```

*(hypothetical data)*
:::

---

## Example 3-8

:::{.callout-tip}
## Solution

**Solve**

```{r e3-8-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
hotel_rating <- c(4.3, 3.8, 4.6, 3.5, 4.1, 3.9, 4.8, 4.2)

table(hotel_rating)
max(table(hotel_rating))
```

Since each value occurs only once, there is **no mode**.
:::

## The Modal Class

:::{.callout-note}
## Modal Class

The mode for grouped data is the **modal class** — the class with the largest frequency.
:::

## Example 3-9 {#example-3-9}

:::{.callout-warning}
## Revenue of Night Market Stalls

Find the modal class for the frequency distribution of night market stall revenue shown in [Example 3-3](#example-3-3).
:::

---

## Example 3-9

:::{.callout-tip}
## Solution

**Solve**

```{r e3-9-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
f
max(f)
```

Since the class 9.5–14.5 has the largest frequency, 8, it is the **modal class**. Sometimes the midpoint of the class is used; in this case it is 12.
:::

## The Mode for Nominal Data

The mode is the only measure of central tendency that can be used in finding the most typical case when the data are **nominal or categorical**.

## Example 3-10 {#example-3-10}

:::{.callout-warning}
## Payment Methods

The data show the payment method chosen by the customers of an online store on a single day. Find the mode.

```
Method            Number
Credit card         214
Mobile payment      168
Cash                 95
Store credit         32
Bank transfer        11
```

*(hypothetical data)*
:::

---

## Example 3-10

:::{.callout-tip}
## Solution

**Solve**

```{r e3-10-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
data.frame(method = c("Credit card", "Mobile payment", "Cash",
                      "Store credit", "Bank transfer"),
           number = c(214, 168, 95, 32, 11))
```

Since the credit card category has the largest frequency, 214, the most typical payment method was the **credit card**.
:::

## Outliers and the Mean

An extremely high or extremely low data value in a data set can have a striking effect on the mean. These extreme values are called **outliers**. A method for identifying outliers is given in Section 3-3.

## Example 3-11 {#example-3-11}

:::{.callout-warning}
## Salaries at a Trading Firm

A small trading firm consists of the owner, the manager, a sales representative, and two clerks, whose monthly salaries in NT dollars are listed here. (Assume that this is the entire population.) Find the mean, median, and mode.

```
Position            Salary
Owner              180,000
Manager             72,000
Sales rep           48,000
Clerk               35,000
Clerk               35,000
```

*(hypothetical data)*
:::

---

## Example 3-11

:::{.callout-tip}
## Solution

**Solve**

```{r e3-11-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
salary <- c(180000, 72000, 48000, 35000, 35000)

mean(salary)
median(salary)
table(salary)
```

$$\mu = \frac{\Sigma X}{N} = \frac{370{,}000}{5} = 74{,}000$$

The mean is **NT\$74,000**, the median is **NT\$48,000**, and the mode is **NT\$35,000**.
:::

---

## Example 3-11

:::{.callout-tip}
## Solution

**Interpretation**

The mean is much higher than the median or the mode, because the extremely high salary of the owner tends to raise the value of the mean. In this and similar situations, the **median** should be used as the measure of central tendency.
:::

## The Midrange

:::{.callout-note}
## Midrange

The **midrange** is defined as the sum of the lowest and highest values in the data set, divided by 2. The symbol $MR$ is used for the midrange.

$$MR = \frac{\text{lowest value} + \text{highest value}}{2}$$
:::

The midrange is a very rough estimate of the average and can be affected by one extremely high or low value.

## Example 3-12 {#example-3-12}

:::{.callout-warning}
## Flight Cancellations

The number of flight cancellations recorded by an airline in each of five months is shown. Find the midrange.

```
6   41   97   112   58
```

*(hypothetical data)*
:::

---

## Example 3-12

:::{.callout-tip}
## Solution

**Solve**

```{r e3-12-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
cancellations <- c(6, 41, 97, 112, 58)

min(cancellations)
max(cancellations)
(min(cancellations) + max(cancellations)) / 2
```

$$MR = \frac{6 + 112}{2} = \frac{118}{2} = 59$$

The midrange for the number of flight cancellations is **59**.
:::

## Example 3-13 {#example-3-13}

:::{.callout-warning}
## Containers Loaded

The number of containers (in thousands) loaded at an export terminal in each of five years is shown. Find the midrange.

```
318   742   906   455   264
```

*(hypothetical data)*
:::

---

## Example 3-13

:::{.callout-tip}
## Solution

**Solve**

```{r e3-13-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
containers_loaded <- c(318, 742, 906, 455, 264)

min(containers_loaded)
max(containers_loaded)
(min(containers_loaded) + max(containers_loaded)) / 2
```

$$MR = \frac{264 + 906}{2} = \frac{1170}{2} = 585$$

The midrange is **585 thousand containers**.
:::

## The Weighted Mean

Sometimes you must find the mean of a data set in which not all values are equally represented.

:::{.callout-note}
## Weighted Mean

Find the **weighted mean** of a variable $X$ by multiplying each value by its corresponding weight and dividing the sum of the products by the sum of the weights.

$$\bar{X} = \frac{w_1 X_1 + w_2 X_2 + \cdots + w_n X_n}{w_1 + w_2 + \cdots + w_n} = \frac{\Sigma wX}{\Sigma w}$$

where $w_1, w_2, \ldots, w_n$ are the weights and $X_1, X_2, \ldots, X_n$ are the values.
:::

## Example 3-14 {#example-3-14}

:::{.callout-warning}
## Grade Point Average

An ETP student received an A in International Trade Practice (4 credits), a B in Business Statistics (3 credits), a C in Managerial Accounting (3 credits), and a B in Business English Presentation (2 credits). Assuming A = 4 grade points, B = 3, C = 2, D = 1, and F = 0, find the student's grade point average.
:::

---

## Example 3-14

:::{.callout-tip}
## Solution

**Solve**

```{r e3-14-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
credits      <- c(4, 3, 3, 2)
grade_points <- c(4, 3, 2, 3)

sum(credits * grade_points) / sum(credits)
```

$$\bar{X} = \frac{\Sigma wX}{\Sigma w} = \frac{4 \cdot 4 + 3 \cdot 3 + 3 \cdot 2 + 2 \cdot 3}{4+3+3+2} = \frac{37}{12} \approx 3.1$$

The grade point average is **3.1**.
:::

## Table 3-1: Summary of Measures of Central Tendency

| Measure | Definition | Symbol(s) |
|---------|------------|-----------|
| Mean | Sum of values, divided by total number of values | $\mu$, $\bar{X}$ |
| Median | Middle point in data set that has been ordered | $MD$ |
| Mode | Most frequent data value | None |
| Midrange | Lowest value plus highest value, divided by 2 | $MR$ |

## Properties and Uses of the Mean and Median

:::{.callout-important}
## The Mean

1. The mean is found by using **all** the values of the data.
2. The mean **varies less** than the median or mode when samples are taken from the same population.
3. The mean is used in computing other statistics, such as the variance.
4. The mean for the data set is **unique** and not necessarily one of the data values.
5. The mean **cannot** be computed for a frequency distribution that has an **open-ended class**.
6. The mean is **affected by outliers** and may not be the appropriate average in these situations.
:::

---

## Properties and Uses of the Mean and Median

:::{.callout-important}
## The Median

1. The median is used to find the center or middle value of a data set.
2. The median is used when it is necessary to find out whether the data values fall into the **upper half or lower half** of the distribution.
3. The median is used for an **open-ended** distribution.
4. The median is **affected less** than the mean by extremely high or extremely low values.
:::

## Properties and Uses of the Mode and Midrange

:::{.callout-important}
## The Mode

1. The mode is used when the **most typical case** is desired.
2. The mode is the **easiest** average to compute.
3. The mode can be used when the data are **nominal or categorical**, such as religious preference, gender, or political affiliation.
4. The mode is **not always unique**. A data set can have more than one mode, or the mode may not exist.
:::

:::{.callout-important}
## The Midrange

1. The midrange is easy to compute.
2. The midrange gives the midpoint.
3. The midrange is affected by extremely high or low values in a data set.
:::

## Distribution Shapes

Frequency distributions can assume many shapes. The three most important shapes are **positively skewed**, **symmetric**, and **negatively skewed** (Figure 3-1).

:::{.callout-note}
## Three Distribution Shapes

- **Positively skewed (right-skewed)** — the majority of the data values fall to the **left** of the mean and cluster at the lower end; the tail is to the right. The **mean is to the right of the median**, and the **mode is to the left of the median**.
- **Symmetric** — the data values are evenly distributed on both sides of the mean. When the distribution is unimodal, the **mean, median, and mode are the same** and lie at the center.
- **Negatively skewed (left-skewed)** — the majority of the data values fall to the **right** of the mean and cluster at the upper end; the tail is to the left. The **mean is to the left of the median**, and the **mode is to the right of the median**.
:::

## Skewness and the Choice of Average

Incomes in the United States are positively skewed: most incomes cluster at the low end, and the few high incomes form a tail on the right. If most students score very high on an examination, the distribution of scores is negatively skewed.

:::{.callout-warning}
## Caution

When a distribution is **extremely skewed**, the value of the mean is pulled toward the tail, so the **median** rather than the mean is the more appropriate measure of central tendency. An extremely skewed distribution can also affect other statistics.
:::

---

## Distribution Shapes in R

**Prepare data**

```{r s3-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(2026)
shapes <- data.frame(
  shape = rep(c("(a) Positively skewed", "(b) Symmetric", "(c) Negatively skewed"),
              each = 4000),
  value = c(rgamma(4000, 2, 0.5), rnorm(4000, 8, 2), 16 - rgamma(4000, 2, 0.5))
)
```

---

## Distribution Shapes in R

**Output figure**

```{r s3-1-2, echo=T, warning=FALSE, out.width="90%", fig.align='center', fig.height=3.5}
ggplot(shapes, aes(value)) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~ shape, scales = "free") +
  labs(title = "Figure 3-1: Types of Distributions", x = NULL, y = "Frequency")
```

In the skewed panels the mean is pulled toward the long tail, while the median stays near the bulk of the data.

# Section 3-2: Measures of Variation {background-color="#2c4a6e"}

## Why Variation Matters

Knowing the average of a data set is not enough to describe it entirely. A shoe store owner who knows that the average size of a man's shoe is size 10 would not be in business very long if she ordered only size 10 shoes.

In addition to knowing the average, you must know how the data values are **dispersed**. The measures that determine the spread of the data values are called **measures of variation**, or **measures of dispersion**: the range, variance, and standard deviation.

## Example 3-15 {#example-3-15}

:::{.callout-warning}
## Comparison of Two Suppliers

A buyer wants to compare two overseas suppliers of packaging film on how long each takes to deliver an order. Six shipments were placed with each supplier. Since only six shipments are involved and no further shipments will be placed under the current contracts, these two groups constitute two small **populations**. The lead times (in days) are shown. Find the mean of each group.

```
Supplier A    Supplier B
     9            14
    21            10
     3            13
    15            11
     6            14
    18            10
```

*(hypothetical data)*

Back to [Example 3-16](#example-3-16), [Example 3-18](#example-3-18), [Example 3-19](#example-3-19)
:::

---

## Example 3-15

:::{.callout-tip}
## Solution

**Solve**

```{r e3-15-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
supplier_a <- c(9, 21, 3, 15, 6, 18)
supplier_b <- c(14, 10, 13, 11, 14, 10)

sum(supplier_a)
mean(supplier_a)
sum(supplier_b)
mean(supplier_b)
```

$$\mu_A = \frac{72}{6} = 12 \text{ days} \qquad \mu_B = \frac{72}{6} = 12 \text{ days}$$

Both means are **12 days**.
:::

---

## Example 3-15

:::{.callout-tip}
## Solution

**Output figure**

```{r e3-15-2, echo=T, warning=FALSE, out.width="70%", fig.align='center', fig.height=3}
lead_times <- data.frame(supplier = rep(c("(a) Supplier A", "(b) Supplier B"), each = 6),
                         days     = c(supplier_a, supplier_b))

ggplot(lead_times, aes(days, supplier)) +
  geom_point() +
  geom_vline(xintercept = 12) +
  labs(title = "Figure 3-2: Examining Data Sets Graphically",
       x = "Lead time (in days)", y = NULL)
```

Even though the means are the same, the **spread** is quite different: supplier B delivers far more consistently.
:::

## The Range

:::{.callout-note}
## Range

The **range** is the highest value minus the lowest value. The symbol $R$ is used for the range.

$$R = \text{highest value} - \text{lowest value}$$
:::

Make sure the range is given as a **single number**.

## Example 3-16 {#example-3-16}

:::{.callout-warning}
## Comparison of Two Suppliers

Find the ranges for the two suppliers in [Example 3-15](#example-3-15).
:::

---

## Example 3-16

:::{.callout-tip}
## Solution

**Solve**

```{r e3-16-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
max(supplier_a) - min(supplier_a)
max(supplier_b) - min(supplier_b)
```

For supplier A, $R = 21 - 3 = 18$ days. For supplier B, $R = 14 - 10 = 4$ days.

For supplier A, 18 days separate the largest data value from the smallest; for supplier B only 4 days do, which is less than one-fourth of supplier A's range.
:::

## Example 3-17 {#example-3-17}

:::{.callout-warning}
## Parcel Volume

The monthly parcel volumes (in ten thousands) handled by a logistics hub for six selected months are shown. Find the range.

```
37   62   19   48   55   41
```

*(hypothetical data)*
:::

---

## Example 3-17

:::{.callout-tip}
## Solution

**Solve**

```{r e3-17-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
parcels <- c(37, 62, 19, 48, 55, 41)

max(parcels) - min(parcels)
```

$$R = 62 - 19 = 43 \text{ (ten thousand parcels)}$$

One extremely high or one extremely low data value can affect the range markedly, so statisticians use the **variance** and **standard deviation** for a more meaningful measure of variability.
:::

## Population Variance and Standard Deviation

Data variation is based on the difference or distance each data value is from the mean; this distance is called a **deviation**, $X - \mu$. The sum of the deviations about the mean is always zero, so the deviations are **squared**.

:::{.callout-note}
## Population Variance and Standard Deviation

The **population variance** is the average of the squares of the distance each value is from the mean:

$$\sigma^2 = \frac{\Sigma(X - \mu)^2}{N}$$

The **population standard deviation** is the square root of the variance:

$$\sigma = \sqrt{\sigma^2} = \sqrt{\frac{\Sigma(X - \mu)^2}{N}}$$

where $X$ = individual value, $\mu$ = population mean, $N$ = population size.
:::

## Procedure and Rounding Rule

:::{.callout-important}
## Procedure — Population Variance and Standard Deviation

1. Find the mean for the data, $\mu = \dfrac{\Sigma X}{N}$.
2. Find the deviation for each data value, $X - \mu$.
3. Square each of the deviations, $(X - \mu)^2$.
4. Find the sum of the squares, $\Sigma(X - \mu)^2$.
5. Divide by $N$ to get the variance.
6. Take the square root of the variance to get the standard deviation.
:::

:::{.callout-important}
## Rounding Rule for the Standard Deviation

The final answer should be rounded to **one more decimal place** than that of the original data.
:::

## Example 3-18 {#example-3-18}

:::{.callout-warning}
## Comparison of Two Suppliers

Find the variance and standard deviation for the lead times of supplier A in [Example 3-15](#example-3-15). The lead times in days were

```
9   21   3   15   6   18
```
:::

---

## Example 3-18

:::{.callout-tip}
## Solution

**Prepare data**

```{r e3-18-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
deviation_a <- supplier_a - mean(supplier_a)

deviation_a
deviation_a^2
sum(deviation_a^2)
```
:::

---

## Example 3-18

:::{.callout-tip}
## Solution

**Solve**

```{r e3-18-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sum(deviation_a^2) / 6
sqrt(sum(deviation_a^2) / 6)
```

$$\sigma^2 = \frac{252}{6} = 42 \qquad \sigma = \sqrt{42} \approx 6.5$$

The variance is **42** and the standard deviation is **6.5 days**.
:::

## Example 3-19 {#example-3-19}

:::{.callout-warning}
## Comparison of Two Suppliers

Find the variance and standard deviation for the lead times of supplier B in [Example 3-15](#example-3-15). The lead times in days were

```
14   10   13   11   14   10
```
:::

---

## Example 3-19

:::{.callout-tip}
## Solution

**Solve**

```{r e3-19-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
deviation_b <- supplier_b - mean(supplier_b)

deviation_b
deviation_b^2
sum(deviation_b^2)
```
:::

---

## Example 3-19

:::{.callout-tip}
## Solution

**Solve**

```{r e3-19-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sum(deviation_b^2) / 6
sqrt(sum(deviation_b^2) / 6)
```

$$\sigma^2 = \frac{18}{6} = 3 \qquad \sigma \approx 1.7$$

With equal means, a larger $\sigma$ means more spread: supplier A ($6.5$) exceeds B ($1.7$).
:::

## Sample Variance and Standard Deviation

Dividing by $n$ usually **underestimates** the population variance, so the sum of squares is divided by $n - 1$, giving a slightly larger value and an **unbiased estimate** of the population variance.

:::{.callout-note}
## Sample Variance and Standard Deviation

$$s^2 = \frac{\Sigma(X - \bar{X})^2}{n - 1} \qquad s = \sqrt{s^2} = \sqrt{\frac{\Sigma(X - \bar{X})^2}{n - 1}}$$

where $X$ = individual value, $\bar{X}$ = sample mean, $n$ = sample size.
:::

## Example 3-20 {#example-3-20}

:::{.callout-warning}
## Export Orders

The numbers of export orders received last month by five randomly selected sales offices are shown. Find the variance and standard deviation for the data.

```
14   50   26   38   32
```

*(hypothetical data)*

Back to [Example 3-21](#example-3-21)
:::

---

## Example 3-20

:::{.callout-tip}
## Solution

**Solve**

```{r e3-20-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
export_orders <- c(14, 50, 26, 38, 32)
deviation <- export_orders - mean(export_orders)

deviation
deviation^2
```
:::

---

## Example 3-20

:::{.callout-tip}
## Solution

**Solve**

```{r e3-20-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sum(deviation^2)
var(export_orders)
sd(export_orders)
```

$$s^2 = \frac{720}{5 - 1} = 180 \qquad s = \sqrt{180} \approx 13.4$$
:::

## Shortcut (Computational) Formulas

:::{.callout-important}
## Shortcut Formulas for $s^2$ and $s$

$$s^2 = \frac{n(\Sigma X^2) - (\Sigma X)^2}{n(n-1)} \qquad s = \sqrt{\frac{n(\Sigma X^2) - (\Sigma X)^2}{n(n-1)}}$$

These are mathematically equivalent to the previous formulas and do not involve using the mean. Note that $\Sigma X^2$ is **not** the same as $(\Sigma X)^2$: $\Sigma X^2$ means square the values first, then sum; $(\Sigma X)^2$ means sum the values first, then square the sum.
:::

## Example 3-21 {#example-3-21}

:::{.callout-warning}
## Export Orders

The numbers of export orders received last month by five randomly selected sales offices are shown. Find the variance and standard deviation for the data, using the shortcut formula.

```
14   50   26   38   32
```

*(hypothetical data)*

Back to [Example 3-20](#example-3-20)
:::

---

## Example 3-21

:::{.callout-tip}
## Solution

**Solve**

```{r e3-21-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sum(export_orders)
sum(export_orders^2)
(5 * 5840 - 160^2) / (5 * 4)
var(export_orders)
```

$$s^2 = \frac{5(5840) - 160^2}{5(4)} = \frac{3600}{20} = 180 \qquad s \approx 13.4$$

The shortcut formula agrees with `var()`, and with [Example 3-20](#example-3-20).
:::

## Variance and Standard Deviation for Grouped Data

:::{.callout-important}
## Shortcut Formula for Grouped Data

$$s^2 = \frac{n(\Sigma f \cdot X_m^2) - (\Sigma f \cdot X_m)^2}{n(n-1)} \qquad s = \sqrt{\frac{n(\Sigma f \cdot X_m^2) - (\Sigma f \cdot X_m)^2}{n(n-1)}}$$

where $X_m$ is the midpoint of each class and $f$ is the frequency of each class.

Be sure to use the **sum of the frequencies** for $n$ — not the number of classes.
:::

## Example 3-22 {#example-3-22}

:::{.callout-warning}
## Picking Times at a Distribution Center

Find the sample variance and the sample standard deviation for the frequency distribution shown. The data represent the time, in minutes, needed to pick each of 20 randomly selected orders.

```
  Class      Frequency   Midpoint
  4.5-9.5         2           7
  9.5-14.5        3          12
 14.5-19.5        5          17
 19.5-24.5        4          22
 24.5-29.5        3          27
 29.5-34.5        2          32
 34.5-39.5        1          37
```

*(hypothetical data)*
:::

---

## Example 3-22

:::{.callout-tip}
## Solution

**Prepare data**

```{r e3-22-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
f  <- c(2, 3, 5, 4, 3, 2, 1)
Xm <- c(7, 12, 17, 22, 27, 32, 37)

f * Xm
f * Xm^2
```
:::

---

## Example 3-22

:::{.callout-tip}
## Solution

**Solve**

```{r e3-22-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sum(f)
sum(f * Xm)
sum(f * Xm^2)
(20 * 9515 - 405^2) / (20 * 19)
sqrt((20 * 9515 - 405^2) / (20 * 19))
```

$$s^2 = \frac{20(9515) - 405^2}{20(19)} = \frac{26{,}275}{380} \approx 69.1 \qquad s \approx 8.3$$
:::

## Table 3-2: Summary of Measures of Variation

| Measure | Definition | Symbol(s) |
|---------|------------|-----------|
| Range | Distance between highest value and lowest value | $R$ |
| Variance | Average of the squares of the distance that each value is from the mean | $\sigma^2$, $s^2$ |
| Standard deviation | Square root of the variance | $\sigma$, $s$ |

## Uses of the Variance and Standard Deviation

:::{.callout-important}
## Four Uses

1. They determine the **spread** of the data, and are useful in comparing two or more data sets to determine which is more variable.
2. They determine the **consistency** of a variable. In the manufacture of nuts and bolts, the variation in diameters must be small or the parts will not fit together.
3. They determine the **number of data values that fall within a specified interval** in a distribution — for example, Chebyshev's theorem shows that at least 75% of the data values fall within 2 standard deviations of the mean.
4. They are used quite often in **inferential statistics**.
:::

## Coefficient of Variation

When two samples have the same units of measure, their standard deviations can be compared directly. To compare the standard deviations of **two different variables**, use the coefficient of variation.

:::{.callout-note}
## Coefficient of Variation

The **coefficient of variation**, denoted by $\text{CVar}$, is the standard deviation divided by the mean. The result is expressed as a percentage.

$$\text{CVar} = \frac{s}{\bar{X}} \cdot 100\% \quad \text{(samples)} \qquad \text{CVar} = \frac{\sigma}{\mu} \cdot 100\% \quad \text{(populations)}$$
:::

## Example 3-23 {#example-3-23}

:::{.callout-warning}
## Store Sales in Two Cities

For the stores of a coffee chain in Taipei, the mean monthly revenue is NT\$2.68 million with a standard deviation of NT\$523,000. For its stores in Kaohsiung, the mean monthly revenue is NT\$1.94 million with a standard deviation of NT\$417,000. Compute the coefficient of variation for the two cities. Which data set is more variable?

*(hypothetical data)*
:::

---

## Example 3-23

:::{.callout-tip}
## Solution

**Solve**

```{r e3-23-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
52.3 / 268 * 100
41.7 / 194 * 100
```

$$\text{Taipei: } \text{CVar} = \frac{52.3}{268} \cdot 100\% = 19.5\% \qquad
\text{Kaohsiung: } \text{CVar} = \frac{41.7}{194} \cdot 100\% = 21.5\%$$

The variation of the Kaohsiung stores is only **slightly larger** than the variation of the Taipei stores. (Both revenue figures are expressed in NT\$10,000.)
:::

## Example 3-24 {#example-3-24}

:::{.callout-warning}
## Hotel Occupancy and Room Rates

For a sample of international hotels in Taipei, the mean occupancy rate is 72.5% with a variance of 27.04, and the mean average daily room rate is NT\$4,850 with a variance of 336,400. Compare the variations of the two data sets.

*(hypothetical data)*
:::

---

## Example 3-24

:::{.callout-tip}
## Solution

**Solve**

```{r e3-24-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sqrt(27.04) / 72.5 * 100
sqrt(336400) / 4850 * 100
```

$$\text{CVar}_{\text{occupancy}} = \frac{\sqrt{27.04}}{72.5} \cdot 100\% = 7.2\% \qquad
\text{CVar}_{\text{rate}} = \frac{\sqrt{336{,}400}}{4850} \cdot 100\% = 12.0\%$$

The variation in the **room rates** is larger than the variation in the occupancy rates.
:::

## Range Rule of Thumb

:::{.callout-important}
## The Range Rule of Thumb

A rough estimate of the standard deviation is

$$s \approx \frac{\text{range}}{4}$$

The rule can also estimate the extreme data values: the smallest data value is approximately $\bar{X} - 2s$ and the largest is approximately $\bar{X} + 2s$.
:::

:::{.callout-warning}
## Caution

The range rule of thumb is **only an approximation** and should be used when the distribution of data values is **unimodal and roughly symmetric**.
:::

---

## The Range Rule of Thumb in R

```{r s3-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
containers_per_hour <- c(16, 13, 22, 10, 18, 14, 19, 16)

mean(containers_per_hour)
sd(containers_per_hour)
(max(containers_per_hour) - min(containers_per_hour)) / 4
mean(containers_per_hour) - 2 * sd(containers_per_hour)
mean(containers_per_hour) + 2 * sd(containers_per_hour)
```

The actual standard deviation is 3.7 and the range is $22 - 10 = 12$, so the range rule of thumb gives $s \approx 3$. The mean is 16, so the estimated smallest and largest values are 8.5 and 23.5 — in the ballpark of the actual 10 and 22.

## Chebyshev's Theorem

:::{.callout-important}
## Chebyshev's Theorem

The proportion of values from a data set that will fall within $k$ standard deviations of the mean will be **at least** $1 - \dfrac{1}{k^2}$, where $k$ is a number greater than 1 ($k$ is not necessarily an integer).

- At least three-fourths, or **75%**, of all data values fall within **2** standard deviations of the mean.
- At least eight-ninths, or **89%**, of all data values fall within **3** standard deviations of the mean.

This theorem can be applied to **any** distribution, regardless of its shape.
:::

## Example 3-25 {#example-3-25}

:::{.callout-warning}
## Rents near Campus

The mean monthly rent for a studio apartment near campus is NT\$12,000, and the standard deviation is NT\$1,500. Find the rent range for which at least 75% of the studios will be listed.

*(hypothetical data)*
:::

---

## Example 3-25

:::{.callout-tip}
## Solution

**Solve**

```{r e3-25-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
(1 - 1 / 2^2) * 100
12000 - 2 * 1500
12000 + 2 * 1500
```

$$12{,}000 + 2(1{,}500) = 15{,}000 \qquad 12{,}000 - 2(1{,}500) = 9{,}000$$

At least 75% of the studios near campus will be listed between **NT\$9,000 and NT\$15,000**.
:::

## Example 3-26 {#example-3-26}

:::{.callout-warning}
## Fuel Cost per Delivery

A survey of delivery riders found that the mean fuel cost per order was NT\$8.00, with a standard deviation of NT\$1.20. Using Chebyshev's theorem, find the minimum percentage of the orders whose fuel cost falls between NT\$5.00 and NT\$11.00.

*(hypothetical data)*
:::

---

## Example 3-26

:::{.callout-tip}
## Solution

**Solve**

```{r e3-26-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
k <- (11.00 - 8.00) / 1.20
k
(1 - 1 / k^2) * 100
```

**Step 1.** $11.00 - 8.00 = 3.00$.  **Step 2.** $k = \dfrac{3.00}{1.20} = 2.5$.

**Step 3.** $1 - \dfrac{1}{2.5^2} = 1 - \dfrac{1}{6.25} = 1 - 0.16 = 0.84$.

Hence, at least **84%** of the orders have a fuel cost between NT\$5.00 and NT\$11.00.
:::

## The Empirical (Normal) Rule

:::{.callout-important}
## The Empirical Rule

When a distribution is **bell-shaped (normal)**:

- Approximately **68%** of the data values will fall within **1** standard deviation of the mean.
- Approximately **95%** of the data values will fall within **2** standard deviations of the mean.
- Approximately **99.7%** of the data values will fall within **3** standard deviations of the mean.

Because the empirical rule requires the distribution to be approximately bell-shaped, its results are **more accurate** than those of Chebyshev's theorem, which applies to all distributions.
:::

---

## The Empirical Rule in R

**Prepare data**

Suppose the monthly revenue of the stores in a convenience store chain is bell-shaped with a mean of NT\$1.8 million and a standard deviation of NT\$200,000 (figures below are in NT\$10,000).

```{r s3-2-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(100, 260, by = 1)
curve <- data.frame(x, density = dnorm(x, 180, 20))
```

---

## The Empirical Rule in R

**Output figure**

```{r s3-2-3, echo=T, warning=FALSE, out.width="65%", fig.align='center', fig.height=3.5}
ggplot(curve, aes(x, density)) +
  geom_line() +
  geom_vline(xintercept = c(120, 140, 160, 200, 220, 240)) +
  labs(title = "Figure 3-4: The Empirical Rule",
       x = "Monthly revenue (NT$10,000)", y = "Density")
```

Approximately 68% of the stores take between 160 and 200, approximately 95% between 140 and 220, and approximately 99.7% between 120 and 240 (NT\$10,000 per month).

## Linear Transformation of Data

Sometimes it is necessary to transform data values into other data values — for example, converting Celsius temperatures collected in Canada into Fahrenheit for a study used in the United States. This change is called the **linear transformation** of the data.

:::{.callout-note}
## Adding a Constant

When a **constant is added to** each data value, the **mean increases by that constant** and the **standard deviation does not change**.
:::

:::{.callout-note}
## Multiplying by a Constant

When each data value is **multiplied by a constant**, the **mean** of the new data set equals the constant times the old mean, and the **standard deviation** of the new data set equals the **absolute value** of the constant times the old standard deviation.
:::

---

## Linear Transformation: Adding a Constant

Suppose a logistics firm employs five part-time packers whose hourly wages are NT\$200, NT\$185, NT\$215, NT\$185, and NT\$215. After a profitable quarter every packer is given a raise of NT\$20 per hour.

```{r s3-2-4, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
hourly_wage <- c(200, 185, 215, 185, 215)
raised_wage <- hourly_wage + 20

mean(hourly_wage)
sd(hourly_wage)
mean(raised_wage)
sd(raised_wage)
```

The mean rises from NT\$200 to NT\$220 — exactly the amount added — while the standard deviation stays at **15**.

## Linear Transformation: Multiplying by a Constant

The same five packers worked 18, 14, 22, 14, and 22 hours last week. Suppose the payroll system records the hours in **minutes** instead, so every value is multiplied by 60.

```{r s3-2-5, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
weekly_hours   <- c(18, 14, 22, 14, 22)
weekly_minutes <- weekly_hours * 60

mean(weekly_hours)
sd(weekly_hours)
mean(weekly_minutes)
sd(weekly_minutes)
```

The mean is multiplied by 60, from 18 hours to 1080 minutes, and the standard deviation is multiplied by 60 as well, from 4 to **240**.

# Section 3-3: Measures of Position {background-color="#2c4a6e"}

## Locating a Value Within a Distribution

In addition to measures of central tendency and measures of variation, there are **measures of position** or location. These measures include **standard scores, percentiles, deciles, and quartiles**. They are used to locate the relative position of a data value in the data set.

If a value is located at the 80th percentile, 80% of the values fall below it and 20% fall above it. The **median** is the value that corresponds to the 50th percentile.

## Standard Scores

A direct comparison of a score of 90 on a music test and 45 on an English exam is impossible, since the exams might not be equivalent in number of questions or value of each question. A comparison of a relative standard can be made using the mean and standard deviation.

:::{.callout-note}
## z Score (Standard Score)

A **z score** or **standard score** for a value is obtained by subtracting the mean from the value and dividing the result by the standard deviation:

$$z = \frac{\text{value} - \text{mean}}{\text{standard deviation}}$$

$$z = \frac{X - \bar{X}}{s} \quad \text{(samples)} \qquad z = \frac{X - \mu}{\sigma} \quad \text{(populations)}$$

The $z$ score represents the number of standard deviations that a data value falls above or below the mean. If $z > 0$ the score is above the mean; if $z = 0$ it equals the mean; if $z < 0$ it is below the mean.
:::

## Example 3-27 {#example-3-27}

:::{.callout-warning}
## Call Handling Times

At a call center's Taipei site the mean handling time is 4.8 minutes with a standard deviation of 0.6 minute; at its Taichung site the mean is 5.4 minutes with a standard deviation of 0.8 minute. One call in Taipei took 6.3 minutes and one call in Taichung took 6.8 minutes. Compare the relative positions of the two calls.

*(hypothetical data)*
:::

---

## Example 3-27

:::{.callout-tip}
## Solution

**Solve**

```{r e3-27-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
(6.3 - 4.8) / 0.6
(6.8 - 5.4) / 0.8
```

$$\text{Taipei: } z = \frac{6.3 - 4.8}{0.6} = 2.50 \qquad
\text{Taichung: } z = \frac{6.8 - 5.4}{0.8} = 1.75$$

Since the $z$ score for the Taipei call is higher, that call is **relatively longer** within its own site than the Taichung call is within its site.
:::

## Example 3-28 {#example-3-28}

:::{.callout-warning}
## Daily Revenue of Convenience Stores

For a convenience store chain, the 24-hour branches have a mean daily revenue of NT\$52,000 with a standard deviation of NT\$5,000, while the campus branches have a mean daily revenue of NT\$38,000 with a standard deviation of NT\$4,000. Find the relative positions of a 24-hour branch that took NT\$45,000 and a campus branch that took NT\$33,000.

*(hypothetical data)*
:::

---

## Example 3-28

:::{.callout-tip}
## Solution

**Solve**

```{r e3-28-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
(45000 - 52000) / 5000
(33000 - 38000) / 4000
```

$$\text{24-hour: } z = \frac{45{,}000 - 52{,}000}{5{,}000} = -1.40 \qquad
\text{Campus: } z = \frac{33{,}000 - 38{,}000}{4{,}000} = -1.25$$

In this case, the **campus** branch stands relatively higher within its own group than the 24-hour branch does within its group.

When all data for a variable are transformed into $z$ scores, the resulting distribution has a mean of 0 and a standard deviation of 1.
:::

## Percentiles

:::{.callout-note}
## Percentiles

**Percentiles** divide the data set into 100 equal groups. They are symbolized by $P_1, P_2, P_3, \ldots, P_{99}$.
:::

:::{.callout-warning}
## Caution

Percentiles are **not** the same as percentages. If a student gets 72 correct answers out of 100, she obtains a percentage score of 72, which says nothing about her position relative to the class. If a raw score of 72 corresponds to the 64th percentile, then she did better than 64% of the students in her class.
:::

## Table 3-3: Percentile Ranks on the Business English Placement Test

| Scaled score | Sec. 1: Listening | Sec. 2: Grammar | Sec. 3: Reading | Total scaled | Percentile rank |
|---|---|---|---|---|---|
| 72 | 99 | 98 | 99 | 720 | 99 |
| 70 | 97 | 95 | 97 | 700 | 96 |
| 68 | 94 | 91 | 93 | 680 | 92 |
| 66 | 89 | 85 | 88 | 660 | 87 |
| 64 | 83 | 78 | 82 | 640 | 80 |
| 62 | 75 | 70 | 74 | 620 | 72 |
| 60 | 66 | 60 | 64 | 600 | 63 |

Percentile ranks published by a business school for the English placement test it gives to incoming students. *(hypothetical data)*

---

## Table 3-3: Percentile Ranks (continued)

| Scaled score | Sec. 1: Listening | Sec. 2: Grammar | Sec. 3: Reading | Total scaled | Percentile rank |
|---|---|---|---|---|---|
| 58 | 56 | 50 | 54 | 580 | 53 |
| 56 | 45 | 40 | 43 | 560 | 42 |
| 54 | 35 | 31 | 33 | 540 | 32 |
| 52 | 26 | 23 | 25 | 520 | 24 |
| 50 | 19 | 17 | 18 | 500 | 17 |
| 48 | 13 | 12 | 13 | 480 | 12 |
| 46 | 9 | 8 | 9 | 460 | 8 |
| 44 | 6 | 5 | 6 | 440 | 5 |

---

## Table 3-3: Percentile Ranks (continued)

| Scaled score | Sec. 1: Listening | Sec. 2: Grammar | Sec. 3: Reading | Total scaled | Percentile rank |
|---|---|---|---|---|---|
| 42 | 4 | 3 | 4 | 420 | 3 |
| 40 | 2 | 2 | 2 | 400 | 2 |
| 38 | 1 | 1 | 1 | 380 | 1 |
| 36 | 1 | 1 | 1 | 360 | 1 |
| 34 | 1 | 1 | 1 | 340 | 1 |
| **Mean** | 58.4 | 59.1 | 58.7 | 588 | |
| **S.D.** | 6.2 | 6.8 | 6.5 | 62 | |

A student with a scaled score of 64 for section 1 has a percentile rank of 83 — that student did better than 83% of the students who took section 1 of the test.

## Example 3-29 {#example-3-29}

:::{.callout-warning}
## Order Values on an E-Commerce Platform

The frequency distribution for the order values (in NT dollars) of 200 randomly selected orders placed on an e-commerce platform is shown here. Construct a percentile graph.

```
Class boundaries      Frequency
   199.5-399.5            18
   399.5-599.5            52
   599.5-799.5            68
   799.5-999.5            38
   999.5-1199.5           16
  1199.5-1399.5            8
               Total     200
```

*(hypothetical data)*
:::

---

## Example 3-29

:::{.callout-tip}
## Solution

**Prepare data**

$$\text{Cumulative \%} = \frac{\text{cumulative frequency}}{n} \cdot 100$$

```{r e3-29-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
f <- c(18, 52, 68, 38, 16, 8)

cumsum(f)
cumsum(f) / 200 * 100
```
:::

---

## Example 3-29

:::{.callout-tip}
## Solution

**Output figure**

```{r e3-29-2, echo=T, warning=FALSE, out.width="60%", fig.align='center', fig.height=3.5}
percentile_points <- data.frame(boundary    = seq(199.5, 1399.5, by = 200),
                                cum_percent = cumsum(c(0, f)) / 200 * 100)

ggplot(percentile_points, aes(boundary, cum_percent)) +
  geom_line() +
  geom_point() +
  labs(title = "Figure 3-6: Percentile Graph for Example 3-29",
       x = "Class boundaries (NT dollars)", y = "Cumulative percentages")
```

Reading from the graph, an order value of NT\$799.50 sits at the 69th percentile, and the 50th percentile falls somewhere between NT\$599.50 and NT\$799.50.
:::

## The Percentile Formula

:::{.callout-important}
## Percentile Formula

The percentile corresponding to a given value $X$ is computed by using

$$\text{Percentile} = \frac{(\text{number of values below } X) + 0.5}{\text{total number of values}} \cdot 100$$
:::

## Example 3-30 {#example-3-30}

:::{.callout-warning}
## Product Returns

The number of product returns processed each day by a fulfillment center over a 10-day period is shown. Find the percentile rank of 18.

```
14   23   9   31   18   27   12   21   35   25
```

*(hypothetical data)*

Back to [Example 3-31](#example-3-31), [Example 3-32](#example-3-32), [Example 3-33](#example-3-33)
:::

---

## Example 3-30

:::{.callout-tip}
## Solution

**Solve**

```{r e3-30-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
returns <- c(14, 23, 9, 31, 18, 27, 12, 21, 35, 25)

sort(returns)
sum(returns < 18)
(sum(returns < 18) + 0.5) / 10 * 100
```

Since there are 3 numbers below the value of 18,

$$\text{Percentile} = \frac{3 + 0.5}{10} \cdot 100 = 35\text{th percentile}$$

Hence, the value of 18 is higher than **35%** of the data values.
:::

## Example 3-31 {#example-3-31}

:::{.callout-warning}
## Product Returns

Using the data in [Example 3-30](#example-3-30), find the percentile rank of 31.
:::

---

## Example 3-31

:::{.callout-tip}
## Solution

**Solve**

```{r e3-31-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sum(returns < 31)
(sum(returns < 31) + 0.5) / 10 * 100
```

There are 8 values below 31; thus

$$\text{Percentile} = \frac{8 + 0.5}{10} \cdot 100 = 85\text{th percentile}$$

Therefore, the data value 31 is higher than **85%** of the values in the data set.
:::

## Finding a Value for a Given Percentile

:::{.callout-important}
## Procedure — Finding a Data Value Corresponding to a Given Percentile

1. Arrange the data in order from lowest to highest.
2. Substitute into the formula $c = \dfrac{n \cdot p}{100}$, where $n$ = total number of values and $p$ = percentile.
3. **If $c$ is not a whole number**, round up to the next whole number and count over to that value from the lowest value.
   **If $c$ is a whole number**, use the value halfway between the $c$th and $(c+1)$st values.
:::

## Example 3-32 {#example-3-32}

:::{.callout-warning}
## Product Returns

Using the data in [Example 3-30](#example-3-30), find the value corresponding to the 65th percentile.
:::

---

## Example 3-32

:::{.callout-tip}
## Solution

**Solve**

```{r e3-32-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sort(returns)
10 * 65 / 100
sort(returns)[7]
```

$$c = \frac{10 \cdot 65}{100} = 6.5$$

Since $c$ is not a whole number, round it up to 7. Starting at the lowest value and counting over to the 7th value gives **25**. Hence, the value of 25 corresponds to the 65th percentile.
:::

## Example 3-33 {#example-3-33}

:::{.callout-warning}
## Product Returns

Using the data in [Example 3-30](#example-3-30), find the data value corresponding to the 30th percentile.
:::

---

## Example 3-33

:::{.callout-tip}
## Solution

**Solve**

```{r e3-33-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
10 * 30 / 100
(sort(returns)[3] + sort(returns)[4]) / 2
```

$$c = \frac{10 \cdot 30}{100} = 3$$

Since $c$ is a whole number, use the value halfway between the 3rd and 4th values. The halfway value between 14 and 18 is **16**, which corresponds to the 30th percentile.
:::

## Quartiles and Deciles

:::{.callout-note}
## Quartiles and Deciles

**Quartiles** divide the distribution into four equal groups, denoted by $Q_1$, $Q_2$, $Q_3$.
**Deciles** divide the distribution into 10 groups, denoted by $D_1, D_2, \ldots, D_9$.

- $Q_1 = P_{25}$, $Q_2 = P_{50}$, $Q_3 = P_{75}$
- $D_1 = P_{10}$, $D_2 = P_{20}$, ..., $D_9 = P_{90}$
- The median is the same as $P_{50}$ or $Q_2$ or $D_5$.
:::

:::{.callout-important}
## Procedure — Finding $Q_1$, $Q_2$, and $Q_3$

1. Arrange the data in order from lowest to highest.
2. Find the median of the data values. This is $Q_2$.
3. Find the median of the data values that fall **below** $Q_2$. This is $Q_1$.
4. Find the median of the data values that fall **above** $Q_2$. This is $Q_3$.
:::

## Example 3-34 {#example-3-34}

:::{.callout-warning}
## Deliveries per Rider

The numbers of orders delivered in one hour by 9 riders on a food delivery platform are shown. Find $Q_1$, $Q_2$, and $Q_3$.

```
22   27   19   25   31   24   27   22   26
```

*(hypothetical data)*

Back to [Example 3-35](#example-3-35)
:::

---

## Example 3-34

:::{.callout-tip}
## Solution

**Solve**

```{r e3-34-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
deliveries <- c(22, 27, 19, 25, 31, 24, 27, 22, 26)
sorted <- sort(deliveries)

sorted
median(sorted)
median(sorted[1:4])
median(sorted[6:9])
quantile(deliveries, c(0.25, 0.50, 0.75))
```

$$Q_2 = 25 \qquad Q_1 = \frac{22 + 22}{2} = 22 \qquad Q_3 = \frac{27 + 27}{2} = 27$$

Hence, $Q_1 = 22$, $Q_2 = 25$, and $Q_3 = 27$. The built-in `quantile()` returns the same three values, so the rest of the chapter simply uses `quantile()`.
:::

## The Interquartile Range

:::{.callout-note}
## Interquartile Range

The **interquartile range (IQR)** is the difference between the third and first quartiles.

$$\text{IQR} = Q_3 - Q_1$$
:::

Quartiles can be used as a rough measure of variability: the IQR is the range of the **middle 50%** of the data values. Like the standard deviation, the more variable the data set is, the larger the IQR.

## Example 3-35 {#example-3-35}

:::{.callout-warning}
## Deliveries per Rider

Find the interquartile range for the data in [Example 3-34](#example-3-34).
:::

---

## Example 3-35

:::{.callout-tip}
## Solution

**Solve**

```{r e3-35-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
quantile(deliveries, c(0.25, 0.75))
IQR(deliveries)
```

$$\text{IQR} = Q_3 - Q_1 = 27 - 22 = 5$$

The interquartile range is **5 orders per hour**.
:::

## Table 3-4: Summary of Position Measures

| Measure | Definition | Symbol |
|---------|------------|--------|
| Standard score, or $z$ score | Standard deviations a value lies above or below the mean | $z$ |
| Percentile | Position of a value in hundredths of the distribution | $P_n$ |
| Decile | Position of a value in tenths of the distribution | $D_n$ |
| Quartile | Position of a value in fourths of the distribution | $Q_n$ |

## Outliers

:::{.callout-note}
## Outlier

An **outlier** is an extremely high or an extremely low data value when compared with the rest of the data values.
:::

:::{.callout-note}
## Resistant and Nonresistant Statistics

The mean and standard deviation are affected by outliers, so they are called **nonresistant statistics**. The median and interquartile range are less affected by outliers, so they are called **resistant statistics**.
:::

## Identifying Outliers

:::{.callout-important}
## Procedure for Identifying Outliers

1. Arrange the data in order from lowest to highest and find $Q_1$ and $Q_3$.
2. Find the interquartile range: $\text{IQR} = Q_3 - Q_1$.
3. Multiply the IQR by 1.5.
4. Subtract the value obtained in step 3 from $Q_1$ and add it to $Q_3$.
5. Check the data set for any value smaller than $Q_1 - 1.5(\text{IQR})$ or larger than $Q_3 + 1.5(\text{IQR})$.
:::

:::{.callout-warning}
## Caution

There are **no hard-and-fast rules** on what to do with outliers, nor complete agreement among statisticians on ways to identify them. When a distribution is normal or bell-shaped, data values beyond 3 standard deviations of the mean can be considered suspected outliers.
:::

## Example 3-36 {#example-3-36}

:::{.callout-warning}
## Complaint Tickets

The numbers of complaint tickets received by a customer service team on 8 randomly selected days are shown. Check the set for outliers.

```
8   4   18   12   57   8   14   18
```

*(hypothetical data)*
:::

---

## Example 3-36

:::{.callout-tip}
## Solution

**Solve**

```{r e3-36-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
tickets <- c(8, 4, 18, 12, 57, 8, 14, 18)
Q1 <- quantile(tickets, 0.25)
Q3 <- quantile(tickets, 0.75)

Q1
Q3
IQR(tickets)
Q1 - 1.5 * IQR(tickets)
Q3 + 1.5 * IQR(tickets)
```

Here $Q_1 = 8$, $Q_3 = 18$, $\text{IQR} = 10$ and $1.5(10) = 15$, so the fences are $-7$ and $33$. The data value **57** is greater than 33; hence, it can be considered an outlier.
:::

# Section 3-4: Exploratory Data Analysis {background-color="#2c4a6e"}

## Exploratory Data Analysis

The purpose of **traditional analysis** is to confirm various conjectures about the nature of the data. The purpose of **exploratory data analysis** is to examine data to find out what information can be discovered about the data, such as the center and the spread.

:::{.callout-note}
## Exploratory Data Analysis (EDA)

In **exploratory data analysis (EDA)**, data can be organized using a stem and leaf plot. The measure of central tendency used in EDA is the **median**; the measure of variation is the **interquartile range** $Q_3 - Q_1$; and the data are represented graphically using a **boxplot** (sometimes called a box and whisker plot).

EDA was developed by **John Tukey** and presented in his book *Exploratory Data Analysis* (Addison-Wesley, 1977).
:::

## The Five-Number Summary and Boxplots

:::{.callout-note}
## Five-Number Summary

A boxplot involves five specific values, called the **five-number summary** of the data set:

1. The lowest value of the data set (minimum)
2. $Q_1$
3. The median
4. $Q_3$
5. The highest value of the data set (maximum)
:::

:::{.callout-note}
## Boxplot

A **boxplot** is a graph of a data set obtained by drawing a horizontal line from the minimum data value to $Q_1$, drawing a horizontal line from $Q_3$ to the maximum data value, and drawing a box whose vertical sides pass through $Q_1$ and $Q_3$ with a vertical line inside the box passing through the median or $Q_2$.
:::

## Constructing a Boxplot

:::{.callout-important}
## Procedure — Constructing a Boxplot

1. Find the five-number summary for the data.
2. Draw a horizontal axis and place the scale on the axis. The scale should start on or below the minimum data value and end on or above the maximum data value.
3. Locate the lowest data value, $Q_1$, the median, $Q_3$, and the highest data value; draw a box whose vertical sides go through $Q_1$ and $Q_3$; draw a vertical line through the median; draw a line from the minimum data value to the left side of the box and from the maximum data value to the right side of the box.
:::

## Example 3-37 {#example-3-37}

:::{.callout-warning}
## Revenue of a Beverage Stand

The daily revenue, in NT\$1,000, of a beverage stand on 10 randomly selected days is shown. Construct a boxplot for the data.

```
14   9   26   6   16   12   8   20   16   9
```

*(hypothetical data)*
:::

---

## Example 3-37

:::{.callout-tip}
## Solution

**Prepare data**

```{r e3-37-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
daily_revenue <- c(14, 9, 26, 6, 16, 12, 8, 20, 16, 9)

sort(daily_revenue)
quantile(daily_revenue)
```

The five-number summary is minimum 6, $Q_1 = 9$, median 13, $Q_3 = 16$, maximum 26.
:::

---

## Example 3-37

:::{.callout-tip}
## Solution

**Output figure**

```{r e3-37-2, echo=T, warning=FALSE, out.width="65%", fig.align='center', fig.height=2.6}
ggplot(data.frame(daily_revenue), aes(daily_revenue)) +
  geom_boxplot() +
  labs(title = "Figure 3-7: Boxplot for Example 3-37",
       x = "Daily revenue (NT$1,000)")
```

The right whisker is much longer than the left, so the distribution is **positively skewed**, with a high value of 26 and a low value of 6 thousand NT dollars.
:::

## Information Obtained from a Boxplot

:::{.callout-important}
## Reading a Boxplot

1. **The median inside the box**
   a. If the median is near the **center** of the box, the distribution is approximately **symmetric**.
   b. If the median falls to the **left** of the center, the distribution is **positively skewed**.
   c. If the median falls to the **right** of the center, the distribution is **negatively skewed**.
2. **The lines (whiskers)**
   a. If the lines are about the **same length**, the distribution is approximately **symmetric**.
   b. If the **right** line is larger, the distribution is **positively skewed**.
   c. If the **left** line is larger, the distribution is **negatively skewed**.
:::

## Example 3-38 {#example-3-38}

:::{.callout-warning}
## Customs Clearance Times at Two Ports

The data shown are the customs clearance times, in hours, for a sample of export shipments at two ports. Compare the distributions by using boxplots.

```
      Port A                    Port B
 22   26   20   18       32   56   24   16
 24   28   20   26       48   64   56   24
```

*(hypothetical data)*
:::

---

## Example 3-38

:::{.callout-tip}
## Solution

**Prepare data**

```{r e3-38-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
port_a <- c(22, 26, 20, 18, 24, 28, 20, 26)
port_b <- c(32, 56, 24, 16, 48, 64, 56, 24)

quantile(port_a, c(0.25, 0.50, 0.75))
quantile(port_b, c(0.25, 0.50, 0.75))
```

For port A, $Q_1 = 20$, $MD = 23$, $Q_3 = 26$; for port B, $Q_1 = 24$, $MD = 40$, $Q_3 = 56$.
:::

---

## Example 3-38

:::{.callout-tip}
## Solution

**Output figure**

```{r e3-38-2, echo=T, warning=FALSE, out.width="65%", fig.align='center', fig.height=3}
clearance <- data.frame(port  = rep(c("Port A", "Port B"), each = 8),
                        hours = c(port_a, port_b))

ggplot(clearance, aes(hours, port)) +
  geom_boxplot() +
  labs(title = "Figure 3-8: Boxplots for Example 3-38",
       x = "Customs clearance time (hours)", y = NULL)
```

The **median** clearance time at port B is much higher, and both the **interquartile range** and the **range** at port B are much larger than those at port A.
:::

## Hinges versus Quartiles

:::{.callout-warning}
## Caution

In exploratory data analysis, **hinges** are used instead of quartiles to construct boxplots. When the data set consists of an **even** number of values, hinges are the same as quartiles; hinges for a data set with an **odd** number of values differ somewhat from quartiles. Since most calculators and computer programs use quartiles, quartiles are used in this textbook.

A **modified boxplot** can be drawn to check for outliers: the whiskers extend only to the highest and lowest values within $1.5(\text{IQR})$ of the box.
:::

## Table 3-5: Traditional versus EDA Techniques

| Traditional | Exploratory data analysis |
|-------------|---------------------------|
| Frequency distribution | Stem and leaf plot |
| Histogram | Boxplot |
| Mean | Median |
| Standard deviation | Interquartile range |

## Important Terms

:::{.callout-important}
## Chapter 3 Vocabulary

bimodal · boxplot · Chebyshev's theorem · coefficient of variation · data array · decile · empirical rule · exploratory data analysis (EDA) · five-number summary · interquartile range (IQR) · mean · median · midrange · modal class · mode · multimodal · negatively skewed or left-skewed distribution · nonresistant statistic · outlier · parameter · percentile · population mean · population standard deviation · population variance · positively skewed or right-skewed distribution · quartile · range · range rule of thumb · resistant statistic · sample mean · standard deviation · statistic · symmetric distribution · unimodal · variance · weighted mean · z score or standard score
:::

## Key Formulas

:::{.callout-important}
## Central Tendency

$$\bar{X} = \frac{\Sigma X}{n} \qquad \mu = \frac{\Sigma X}{N} \qquad \bar{X} = \frac{\Sigma f \cdot X_m}{n}$$

$$\bar{X} = \frac{\Sigma wX}{\Sigma w} \qquad MR = \frac{\text{lowest value} + \text{highest value}}{2}$$
:::

:::{.callout-important}
## Variation

$$R = \text{highest value} - \text{lowest value} \qquad \sigma^2 = \frac{\Sigma(X-\mu)^2}{N} \qquad \sigma = \sqrt{\frac{\Sigma(X-\mu)^2}{N}}$$

$$s^2 = \frac{n(\Sigma X^2) - (\Sigma X)^2}{n(n-1)} \qquad s^2 = \frac{n(\Sigma f \cdot X_m^2) - (\Sigma f \cdot X_m)^2}{n(n-1)}$$

$$\text{CVar} = \frac{s}{\bar{X}} \cdot 100 \qquad s \approx \frac{\text{range}}{4} \qquad 1 - \frac{1}{k^2}$$
:::

---

## Key Formulas

:::{.callout-important}
## Position

$$z = \frac{X - \bar{X}}{s} \qquad z = \frac{X - \mu}{\sigma}$$

$$\text{Cumulative \%} = \frac{\text{cumulative frequency}}{n} \cdot 100$$

$$\text{Percentile} = \frac{(\text{number of values below } X) + 0.5}{\text{total number of values}} \cdot 100$$

$$c = \frac{n \cdot p}{100} \qquad \text{IQR} = Q_3 - Q_1$$
:::

## Key Takeaways

:::{.callout-important}
- **Measures of central tendency** — the mean, median, mode, and midrange; the weighted mean is used when values are not equally represented
- **The mean uses every value** but is pulled toward outliers; the **median** is preferred for skewed or open-ended data; the **mode** is the only average for nominal data
- **Distribution shapes** — positively skewed, symmetric, negatively skewed, with the mean pulled toward the tail
- **Measures of variation** — the range, variance $\sigma^2$ or $s^2$, and standard deviation $\sigma$ or $s$; divide by $n-1$ for an unbiased sample estimate
- **Coefficient of variation** compares variability across variables measured in different units
- **Chebyshev's theorem** applies to any distribution: at least $1 - 1/k^2$ of the values lie within $k$ standard deviations; the **empirical rule** (68-95-99.7%) applies to bell-shaped data
- **Linear transformation** — adding a constant shifts the mean and leaves the standard deviation unchanged; multiplying by a constant multiplies both
- **Measures of position** — $z$ scores, percentiles, deciles and quartiles, with $Q_1 = P_{25}$, $Q_2 = P_{50} = MD$, $Q_3 = P_{75}$
- **Outliers** lie beyond $Q_1 - 1.5(\text{IQR})$ or $Q_3 + 1.5(\text{IQR})$; the median and IQR are **resistant**, the mean and standard deviation are **nonresistant**
- **Exploratory data analysis** uses the five-number summary and the boxplot to reveal center, spread, and skewness
:::

## Acknowledgement

-   **Copyright notice.** These teaching materials follow the organization and terminology of **Bluman, A. G. (2023). Elementary statistics: A step by step approach (11th ed.). McGraw Hill**. All rights in the original work are reserved by its authors and publishers.

-   **Original examples.** Every worked example, data set, and R script in these slides was written for this course. The data are hypothetical unless stated otherwise.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
