---
title: "Statistics"
subtitle: "Chapter 3: Data Description"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "Elementary Statistics: A Step by Step Approach"
logo: "logo.svg"
editor: source
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
---

```{r setup, include=FALSE}
library(tidyverse)
```

# Section 3-1: Measures of Central Tendency {background-color="#2c4a6e"}

## What Is a Measure of Central Tendency?

A **measure of central tendency** is a single numerical value that summarizes the center or "typical" value of a dataset. Rather than examining every individual observation, a central tendency statistic provides a concise snapshot of where data tend to cluster.

The four most widely used measures are:

- **Mean** ŌĆö arithmetic average of all values
- **Median** ŌĆö middle value when data are ordered
- **Mode** ŌĆö most frequently occurring value(s)
- **Midrange** ŌĆö midpoint between the highest and lowest values

---

:::{.callout-note}
## Key Notation

| Symbol | Meaning |
|--------|---------|
| $\bar{X}$ | Sample mean |
| $\mu$ | Population mean |
| $n$ | Sample size |
| $N$ | Population size |
:::

---

## The Mean

The **mean** is computed by summing all data values and dividing by the number of observations. It is the most commonly reported measure of center.

:::{.callout-note}
## Formulas for the Mean

**Sample mean:**
$$\bar{X} = \frac{\Sigma X}{n}$$

**Population mean:**
$$\mu = \frac{\Sigma X}{N}$$

where $\Sigma X$ is the sum of all data values and $n$ (or $N$) is the count of values.
:::

:::{.callout-important}
## Rounding Rule
Round the mean to **one more decimal place** than appears in the original data.
:::

---

## Example 3-1: Days Off per Year

:::{.callout-tip}
## Example 3-1
A survey of six countries found the following average number of days per year that workers receive as paid leave. Find the mean.

**Days:** 13, 15, 26, 36, 37, 42
:::

---

## Example 3-1: Solution

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
days_off <- c(13, 15, 26, 36, 37, 42)

mean_val <- mean(days_off)
cat("Sum of values:", sum(days_off), "\n")
cat("Number of values:", length(days_off), "\n")
cat("Mean:", mean_val, "\n")
```

The mean number of paid days off is approximately **`r round(mean(c(13,15,26,36,37,42)),1)` days**.

---

## Example 3-2: Hospital Infections

:::{.callout-tip}
## Example 3-2
The number of hospital-acquired infections reported at a sample of hospitals over one year is listed below. Determine the mean.

**Values:** 62, 58, 71, 70, 69, 58, 51, 64, 66, 80
:::

---

## Example 3-2: Solution

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
infections <- c(62, 58, 71, 70, 69, 58, 51, 64, 66, 80)
cat("Mean number of infections:", mean(infections), "\n")
cat("Sum:", sum(infections), "  n:", length(infections), "\n")
```

The mean is **`r mean(c(62,58,71,70,69,58,51,64,66,80))`** hospital-acquired infections.

---

## Mean for Grouped Data

When raw data are unavailable and only a frequency distribution is given, use the **midpoint** of each class to approximate the mean.

:::{.callout-note}
## Formula ŌĆö Mean for Grouped Data

$$\bar{X} = \frac{\Sigma f \cdot X_m}{n}$$

where $f$ = class frequency, $X_m$ = class midpoint, $n = \Sigma f$.
:::

---

## Example 3-3: Miles Run per Week (Grouped Mean)

:::{.callout-tip}
## Example 3-3
Twenty runners recorded the number of miles they ran during one week. Using the frequency distribution below, find the mean.

| Class | Frequency |
|-------|-----------|
| 5.5ŌĆō10.5 | 1 |
| 10.5ŌĆō15.5 | 2 |
| 15.5ŌĆō20.5 | 3 |
| 20.5ŌĆō25.5 | 5 |
| 25.5ŌĆō30.5 | 4 |
| 30.5ŌĆō35.5 | 3 |
| 35.5ŌĆō40.5 | 2 |
:::

---

## Example 3-3: Solution

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
midpoints <- c(8, 13, 18, 23, 28, 33, 38)
freq      <- c(1,  2,  3,  5,  4,  3,  2)

df_grouped <- data.frame(midpoint = midpoints, f = freq) |>
  mutate(fXm = f * midpoint)

knitr::kable(df_grouped, col.names = c("Midpoint (Xm)", "Frequency (f)", "f ┬Ę Xm"),
             caption = "Grouped Mean Calculation")
```

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
n    <- sum(freq)
xbar <- sum(df_grouped$fXm) / n
cat("\nn =", n, "  Sum(f┬ĘXm) =", sum(df_grouped$fXm),
    "  Mean =", round(xbar, 1), "miles\n")
```

---

## The Median

The **median** is the value that sits at the exact center of an ordered dataset, dividing it into two equal halves. It is resistant to the influence of extreme values.

:::{.callout-note}
## Finding the Median

1. Arrange all data values from **lowest to highest**.
2. If $n$ is **odd**: the median is the middle value ŌĆö position $\frac{n+1}{2}$.
3. If $n$ is **even**: the median is the **average of the two middle values**.
:::

---

## Example 3-4: Hotel Rooms

:::{.callout-tip}
## Example 3-4
The number of rooms in the seven hotels in a downtown district are listed below. Find the median.

**Rooms:** 713, 300, 618, 595, 311, 401, 292
:::

---

## Example 3-4: Solution

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
rooms <- c(713, 300, 618, 595, 311, 401, 292)
sorted <- sort(rooms)
cat("Sorted:", sorted, "\n")
cat("Median position:", (length(rooms)+1)/2, "\n")
cat("Median:", median(rooms), "\n")
```

With 7 values (odd), the median is the **4th value** when sorted: **`r median(c(713,300,618,595,311,401,292))`** rooms.

---

## Example 3-5: National Park Costs (Even n)

:::{.callout-tip}
## Example 3-5
The entrance fees (in dollars) for a random sample of national parks are: 25, 10, 8, 15, 5, 12, 14, 3. Find the median.
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
fees <- c(25, 10, 8, 15, 5, 12, 14, 3)
sorted_fees <- sort(fees)
cat("Sorted:", sorted_fees, "\n")
cat("n =", length(fees), "(even) ŌåÆ average of positions",
    length(fees)/2, "and", length(fees)/2 + 1, "\n")
cat("Median:", median(fees), "\n")
```

With 8 values (even), the median is the average of the 4th and 5th sorted values: $\frac{10+12}{2}$ = **$11**.

---

## Example 3-6 & 3-7: Medians for Larger Datasets

:::{.callout-tip}
## Example 3-6 & 3-7 ŌĆö Tornadoes & Asthma Cases
1. The number of tornadoes reported in 12 U.S. states during a recent year:
188, 967, 990, 1133, 630, 604, 294, 584, 276, 250, 451, 810

2. Hospital emergency visits for asthma across 10 cities:
345, 243, 111, 144, 280, 300, 213, 201, 403, 201
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
tornadoes <- c(188, 967, 990, 1133, 630, 604, 294, 584, 276, 250, 451, 810)
asthma    <- c(345, 243, 111, 144, 280, 300, 213, 201, 403, 201)

cat("Tornadoes ŌĆö sorted:", paste(sort(tornadoes), collapse=", "), "\n")
cat("Tornadoes median:", median(tornadoes), "\n\n")
cat("Asthma ŌĆö sorted:", paste(sort(asthma), collapse=", "), "\n")
cat("Asthma median:", median(asthma), "\n")
```

---

## Example 3-8: Magazines (Finding the Median from a Distribution)

:::{.callout-tip}
## Example 3-8
Six randomly selected bookstores reported selling this many magazines per week: 3, 7, 8, 3, 1, 3. Find the median.
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
mags <- c(3, 7, 8, 3, 1, 3)
cat("Sorted:", sort(mags), "\n")
cat("Median:", median(mags), "\n")
```

The median is **`r median(c(3,7,8,3,1,3))`** magazines per week.

---

## The Mode

The **mode** is the value that appears most frequently in a dataset. Unlike the mean and median, the mode can be used with both quantitative and qualitative data.

:::{.callout-note}
## Key Properties of the Mode

- A dataset may have **no mode** (all values appear once), **one mode** (unimodal), **two modes** (bimodal), or **more** (multimodal).
- For grouped data, the **modal class** is the class with the highest frequency.
- Mode is the **only** appropriate measure of center for nominal (categorical) data.
:::

---

## Example 3-9: NFL Signing Bonuses

:::{.callout-tip}
## Example 3-9
The signing bonuses (in millions of dollars) for a sample of NFL players are listed. Find the mode.

**Values:** 18, 14, 10, 10, 10, 7, 7, 3, 2, 2, 2, 8
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
bonuses <- c(18, 14, 10, 10, 10, 7, 7, 3, 2, 2, 2, 8)

mode_tbl <- sort(table(bonuses), decreasing = TRUE)
print(mode_tbl)
cat("Mode: $", names(mode_tbl)[1], "million\n")
```

Both 2 and 10 appear 3 times ŌåÆ dataset is **bimodal**: **$2M and $10M**.

---

## Example 3-10 & 3-11: No Mode vs. Bimodal

:::{.callout-tip}
## Example 3-10 & 3-11 ŌĆö No Mode & Bimodal
1. Bank branches in 5 cities: 17, 23, 25, 3, 42 ŌĆö each value appears exactly once ŌåÆ **no mode**.

2. Licensed nuclear reactors in the U.S. over 15 years:
104, 104, 104, 104, 104, 107, 109, 109, 109, 110, 109, 111, 112, 111, 109
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
reactors <- c(104,104,104,104,104,107,109,109,109,110,109,111,112,111,109)
mode_tbl <- sort(table(reactors), decreasing = TRUE)
print(mode_tbl)
```

**104** appears 5 times and **109** appears 6 times ŌåÆ bimodal: **104 and 109**.

---

## Example 3-12: Modal Class for Grouped Data

:::{.callout-tip}
## Example 3-12
Find the modal class for the miles-run-per-week frequency distribution (Example 3-3).
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
classes <- c("5.5-10.5","10.5-15.5","15.5-20.5","20.5-25.5",
             "25.5-30.5","30.5-35.5","35.5-40.5")
freq <- c(1,2,3,5,4,3,2)
df_mode <- data.frame(Class=classes, Frequency=freq)
knitr::kable(df_mode, caption="Miles Run per Week")
cat("\nModal class:", classes[which.max(freq)], "(frequency =", max(freq), ")\n")
```

The **modal class** is **20.5ŌĆō25.5** miles, with the highest frequency of 5.

---

## Example 3-13: Mode for Qualitative Data

:::{.callout-tip}
## Example 3-13
A survey recorded boat registrations (in thousands) for five Pennsylvania counties:
Westmoreland: 11,008 | Butler: 9,147 | Allegheny: 5,765 | Beaver: 3,342 | Lawrence: 1,478
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
county <- c("Westmoreland","Butler","Allegheny","Beaver","Lawrence")
boats  <- c(11008, 9147, 5765, 3342, 1478)
df_boats <- data.frame(County=county, Registrations=boats)
knitr::kable(df_boats, caption="Boat Registrations")
cat("\nMode (highest registrations):", county[which.max(boats)],
    "with", max(boats), "registrations\n")
```

---

## Example 3-14: Effect of an Outlier on the Mean

:::{.callout-tip}
## Example 3-14
Six employees' salaries at a small firm: $5,000 | $6,000 | $9,000 | $12,000 | $15,000 | $45,000

Which measure best represents this dataset?
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
salaries <- c(5000, 6000, 9000, 12000, 15000, 45000)
cat("Mean:  $", mean(salaries), "\n")
cat("Median: $", median(salaries), "\n")
# Mode
mode_s <- as.numeric(names(sort(table(salaries), decreasing=TRUE))[1])
cat("Mode:  No mode (all values unique)\n")
```

---

The **mean ($15,333)** is inflated by the $45,000 outlier. The **median ($10,500)** better represents the typical salary.

:::{.callout-warning}
## When Outliers Are Present
Use the **median** when a dataset contains extreme values, as the mean is strongly pulled in the direction of the outlier.
:::

---

## The Midrange

The **midrange** is the arithmetic mean of the maximum and minimum data values. It is easy to calculate but highly sensitive to extreme values.

:::{.callout-note}
## Formula ŌĆö Midrange

$$MR = \frac{\text{lowest value} + \text{highest value}}{2}$$
:::

---

## Example 3-15 & 3-16: Midrange

:::{.callout-tip}
## Example 3-15 & 3-16: ŌĆö Water-Line Breaks & NFL Signing Bonuses
1. The number of water-line breaks in a city over 9 years: 2, 3, 6, 8, 4, 1, 3, 5, 7. Find the midrange.

2. Using the data from Example 3-9 (min = $2M, max = $18M), find the midrange.
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
water_breaks <- c(2, 3, 6, 8, 4, 1, 3, 5, 7)
nfl_bonuses  <- c(18, 14, 10, 10, 10, 7, 7, 3, 2, 2, 2, 8)

midrange <- function(x) (min(x) + max(x)) / 2

cat("Water-line breaks midrange:", midrange(water_breaks), "breaks\n")
cat("NFL bonuses midrange: $", midrange(nfl_bonuses), "million\n")
```

---

## The Weighted Mean

When different values carry different levels of importance (weight), use the **weighted mean** to compute an accurate average.

:::{.callout-note}
## Formula ŌĆö Weighted Mean

$$\bar{X}_w = \frac{\Sigma w \cdot X}{\Sigma w}$$

where $w$ = weight and $X$ = value.
:::

---

## Example 3-17: GPA Calculation

:::{.callout-tip}
## Example 3-17
A student earned the following grades. Compute the grade-point average (GPA), using credit hours as weights.

| Course | Grade | Credit Hours |
|--------|-------|-------------|
| English | A (4) | 3 |
| Math | C (2) | 4 |
| History | B (3) | 3 |
| Art | D (1) | 2 |
:::

---

## Example 3-17: Solution

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
courses <- c("English", "Math", "History", "Art")
grades  <- c(4, 2, 3, 1)         # A=4, B=3, C=2, D=1
credits <- c(3, 4, 3, 2)

df_gpa <- data.frame(Course=courses, Grade=grades, Credits=credits) |>
  mutate(w_X = Grade * Credits)
knitr::kable(df_gpa, col.names=c("Course","Grade Points","Credits","w ├Ś X"),
             caption="GPA Calculation")

gpa <- sum(df_gpa$w_X) / sum(credits)
cat("\nWeighted Mean (GPA):", round(gpa, 2), "\n")
```

The student's GPA is **`r round(sum(c(4,2,3,1)*c(3,4,3,2))/sum(c(3,4,3,2)),2)`**.

---

## Choosing the Right Measure of Center

:::{.callout-important}
## Which Measure to Use?

| Situation | Best Measure |
|-----------|-------------|
| Most typical value desired | Mode |
| Distribution is open-ended (no max/min) | Median |
| Extreme values are present | Median |
| Data are categorical (nominal) | Mode |
| Further computations are needed | Mean |
| Data must be split into two equal groups | Median |

The **mean** is generally preferred when data are approximately symmetric with no extreme outliers, since it uses every data value in its computation.
:::

# Section 3-2: Measures of Variation {background-color="#2c4a6e"}

## Why Variation Matters

Two datasets can have identical means yet differ dramatically in their spread. **Measures of variation** quantify how spread out or clustered data values are around the center.

:::{.callout-note}
## Three Key Measures of Variation

| Measure | Symbol(s) | Description |
|---------|-----------|-------------|
| Range | $R$ | Distance between highest and lowest |
| Variance | $\sigma^2$, $s^2$ | Average squared distance from mean |
| Standard Deviation | $\sigma$, $s$ | Square root of variance |
:::

---

## Example 3-18: Why the Mean Alone Is Insufficient

:::{.callout-tip}
## Example 3-18
A testing laboratory evaluates two brands of outdoor paint to see how long (in months) each lasts before fading. Six cans of each brand were tested.

**Brand A:** 10, 60, 50, 30, 40, 20 ŌĆö Mean = 35 months
**Brand B:** 35, 45, 30, 35, 40, 25 ŌĆö Mean = 35 months
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
A <- c(10, 60, 50, 30, 40, 20)
B <- c(35, 45, 30, 35, 40, 25)
cat("Brand A mean:", mean(A), "| Brand B mean:", mean(B), "\n")
```

----

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
data.frame(Brand = rep(c("A","B"), each=6),
           Months = c(A, B)) |>
  ggplot(aes(x = Brand, y = Months, color = Brand)) +
  geom_jitter(size=3, width=0.1) +
  geom_hline(yintercept=35, linetype="dashed") +
  labs(title="Paint Brand Longevity: Same Mean, Different Spread",
       y="Months before fading") +
  theme_bw() + theme(legend.position="none")
```

---

## The Range

The **range** is the simplest measure of spread ŌĆö the difference between the largest and smallest values.

:::{.callout-note}
## Formula ŌĆö Range

$$R = \text{highest value} - \text{lowest value}$$
:::

:::{.callout-warning}
## Limitation of the Range
A single extreme value can drastically inflate the range, making it a misleading indicator of typical variation.
:::

---

## Example 3-19: Ranges for Paint Brands

:::{.callout-tip}
## Example 3-19
Find the range for Brand A and Brand B from Example 3-18.
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
A <- c(10, 60, 50, 30, 40, 20)
B <- c(35, 45, 30, 35, 40, 25)

cat("Brand A range:", max(A) - min(A), "months\n")
cat("Brand B range:", max(B) - min(B), "months\n")
cat("\nBrand B is more consistent: range is less than half of Brand A's range.\n")
```

---

## Example 3-20: Employee Salaries Range

:::{.callout-tip}
## Example 3-20
Find the range for the following salaries at XYZ Manufacturing Co.:

Owner: $100,000 | Manager: $40,000 | Sales rep: $30,000 | Workers: $25,000, $15,000, $18,000
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
salaries <- c(100000, 40000, 30000, 25000, 15000, 18000)
cat("Minimum salary: $", min(salaries), "\n")
cat("Maximum salary: $", max(salaries), "\n")
cat("Range: $", max(salaries) - min(salaries), "\n")
```

The range is **$85,000** ŌĆö heavily influenced by the owner's salary.

---

## Population Variance and Standard Deviation

The **variance** measures the average squared deviation of each value from the mean. The **standard deviation** is its square root, expressed in the original units.

:::{.callout-note}
## Population Formulas

$$\sigma^2 = \frac{\Sigma(X - \mu)^2}{N} \qquad \sigma = \sqrt{\frac{\Sigma(X - \mu)^2}{N}}$$

**Step-by-step procedure:**

1. Find the population mean $\mu$.
2. Subtract $\mu$ from each value: $(X - \mu)$.
3. Square each difference: $(X - \mu)^2$.
4. Sum all squared differences.
5. Divide by $N$ to get variance; take the square root for $\sigma$.
:::

---

## Example 3-21: Variance & SD ŌĆö Paint Brand A

:::{.callout-tip}
## Example 3-21
Find the variance and standard deviation for Brand A paint data: 10, 60, 50, 30, 40, 20
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
A <- c(10, 60, 50, 30, 40, 20)
mu <- mean(A)
df_A <- data.frame(X = A) |>
  mutate(`X - mu` = X - mu,
         `(X-mu)^2` = (X - mu)^2)
knitr::kable(df_A, caption=paste("Brand A: mu =", mu))
```

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
sigma2 <- sum(df_A$`(X-mu)^2`) / length(A)
sigma  <- sqrt(sigma2)
cat("\nVariance (sigma^2):", round(sigma2, 1),
    "\nStandard deviation (sigma):", round(sigma, 1), "months\n")
```

---

## Example 3-22: Variance & SD ŌĆö Paint Brand B

:::{.callout-tip}
## Example 3-22
Find the variance and standard deviation for Brand B: 35, 45, 30, 35, 40, 25
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
B <- c(35, 45, 30, 35, 40, 25)
mu_B <- mean(B)
df_B <- data.frame(X = B) |>
  mutate(`X - mu` = X - mu_B,
         `(X-mu)^2` = (X - mu_B)^2)
knitr::kable(df_B, caption=paste("Brand B: mu =", mu_B))
```

----

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
sigma2_B <- sum(df_B$`(X-mu)^2`) / length(B)
sigma_B  <- sqrt(sigma2_B)
cat("\nVariance (sigma^2):", round(sigma2_B, 1),
    "\nStandard deviation (sigma):", round(sigma_B, 1), "months\n")
cat("\nBrand A sigma:", round(sqrt(1750/6), 1),
    "vs Brand B sigma:", round(sigma_B, 1),
    "ŌåÆ Brand B is more consistent.\n")
```

---

## Sample Variance and Standard Deviation

When data come from a **sample** (not the entire population), divide by $n - 1$ rather than $n$ to obtain an **unbiased** estimate of the population variance.

:::{.callout-note}
## Sample Formulas

$$s^2 = \frac{\Sigma(X - \bar{X})^2}{n-1} \qquad s = \sqrt{\frac{\Sigma(X - \bar{X})^2}{n-1}}$$

**Shortcut (computational) formulas:**

$$s^2 = \frac{n(\Sigma X^2) - (\Sigma X)^2}{n(n-1)} \qquad s = \sqrt{\frac{n(\Sigma X^2) - (\Sigma X)^2}{n(n-1)}}$$
:::

:::{.callout-important}
## Why $n-1$?
Dividing by $n-1$ corrects the tendency of sample variance to underestimate population variance. The value $n-1$ is called the **degrees of freedom**.
:::

---

## Example 3-23: European Auto Sales (Shortcut Formula)

:::{.callout-tip}
## Example 3-23
Find the sample variance and standard deviation for European auto sales (in billions of dollars) over 6 years: 11.2, 11.9, 12.0, 12.8, 13.4, 14.3
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
auto <- c(11.2, 11.9, 12.0, 12.8, 13.4, 14.3)
n    <- length(auto)

sum_X  <- sum(auto)
sum_X2 <- sum(auto^2)

s2 <- (n * sum_X2 - sum_X^2) / (n * (n - 1))
s  <- sqrt(s2)

cat("n =", n, "\n")
cat("Sum(X) =", sum_X, "  Sum(X^2) =", sum_X2, "\n")
cat("s^2 =", round(s2, 3), "\n")
cat("s =", round(s, 2), "billion dollars\n")

# Verify with built-in function
cat("Verified with var():", round(var(auto), 3), "\n")
```

---

## Example 3-24: Standard Deviation for Grouped Data

:::{.callout-tip}
## Example 3-24
Find the variance and standard deviation for the miles-run-per-week grouped data (Example 3-3), representing 20 runners.
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
Xm   <- c(8, 13, 18, 23, 28, 33, 38)
f    <- c(1,  2,  3,  5,  4,  3,  2)
n    <- sum(f)

df_g <- data.frame(Xm, f) |>
  mutate(fXm = f * Xm, fXm2 = f * Xm^2)
knitr::kable(df_g, col.names=c("Xm","f","f┬ĘXm","f┬ĘXm┬▓"),
             caption="Grouped Variance Table")
```

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
s2 <- (n * sum(df_g$fXm2) - sum(df_g$fXm)^2) / (n * (n - 1))
s  <- sqrt(s2)
cat("\nn =", n, "  Sum(f┬ĘXm) =", sum(df_g$fXm),
    "  Sum(f┬ĘXm^2) =", sum(df_g$fXm2))
cat("\nVariance s^2 =", round(s2, 1), "  Std Dev s =", round(s, 1), "miles\n")
```

## Coefficient of Variation

The **coefficient of variation (CVar)** expresses the standard deviation as a percentage of the mean, enabling comparison of variability across datasets with different units or different scales.

:::{.callout-note}
## Formula ŌĆö Coefficient of Variation

$$\text{CVar} = \frac{s}{\bar{X}} \times 100\% \quad \text{(samples)} \qquad \text{CVar} = \frac{\sigma}{\mu} \times 100\% \quad \text{(populations)}$$
:::

---

## Example 3-25: Sales vs. Commissions

:::{.callout-tip}
## Example 3-25
The mean number of car sales per 3-month period is 87 with a standard deviation of 5. The mean commission amount is $5,225 with a standard deviation of $773. Which is more variable?
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
cvar_sales <- (5 / 87) * 100
cvar_comm  <- (773 / 5225) * 100

cat("CVar (sales):", round(cvar_sales, 1), "%\n")
cat("CVar (commissions):", round(cvar_comm, 1), "%\n")
cat("\nConclusion: Commissions are more variable (larger CVar).\n")
```

---

## Example 3-26: Pages vs. Advertisements

:::{.callout-tip}
## Example 3-26
For women's fitness magazines: mean pages = 132, variance = 23; mean ads = 182, variance = 62. Which is more variable?
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
cvar_pages <- (sqrt(23) / 132) * 100
cvar_ads   <- (sqrt(62) / 182) * 100

cat("CVar (pages):", round(cvar_pages, 1), "%\n")
cat("CVar (ads):", round(cvar_ads, 1), "%\n")
cat("\nConclusion: Advertisements are more variable.\n")
```

---

## Range Rule of Thumb

The **range rule of thumb** provides a quick approximation of the standard deviation using only the range.

:::{.callout-note}
## Range Rule of Thumb

$$s \approx \frac{\text{range}}{4}$$

Conversely, most data values fall within the interval:
$$[\bar{X} - 2s, \; \bar{X} + 2s]$$
:::

This approximation works best for unimodal, roughly symmetric distributions.

---

## Chebyshev's Theorem

**Chebyshev's theorem** provides a guaranteed lower bound on the proportion of data values that fall within $k$ standard deviations of the mean ŌĆö regardless of the distribution's shape.

:::{.callout-note}
## Chebyshev's Theorem

The proportion of values from any data set that will fall within $k$ standard deviations of the mean is at least:

$$1 - \frac{1}{k^2}, \quad k > 1$$

| $k$ | Minimum % within $k$ SDs |
|-----|--------------------------|
| 2 | At least 75% |
| 3 | At least 88.89% |
| 2.5 | At least 84% |
:::

---

## Example 3-27: Home Prices (Chebyshev)

:::{.callout-tip}
## Example 3-27
The mean price of homes in a neighborhood is $50,000 and the standard deviation is $10,000. Find the price range in which at least 75% of homes will sell.
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
mu  <- 50000
sd  <- 10000
k   <- 2   # gives at least 75%

lower <- mu - k * sd
upper <- mu + k * sd
pct   <- (1 - 1/k^2) * 100

cat("k =", k, "ŌåÆ at least", pct, "% of values\n")
cat("Price range: $", lower, "to $", upper, "\n")
```

At least **75%** of homes will sell between **$30,000 and $70,000**.

---

## Example 3-28: Travel Allowances (Chebyshev ŌĆö Finding k)

:::{.callout-tip}
## Example 3-28
Executive travel allowances: mean = $0.25/mile, SD = $0.02/mile. Find the minimum percentage of values between $0.20 and $0.30.
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
mu  <- 0.25
sd  <- 0.02

# Distance from mean to boundary
diff <- 0.30 - mu
k    <- diff / sd
pct  <- (1 - 1/k^2) * 100

cat("Distance from mean to boundary: $", diff, "\n")
cat("k = distance / SD =", k, "\n")
cat("Minimum percentage: 1 - 1/k^2 =", pct, "%\n")
```

At least **84%** of travel allowances fall between $0.20 and $0.30 per mile.

---

## The Empirical (Normal) Rule

When a distribution is **bell-shaped (normal)**, tighter guarantees apply:

:::{.callout-note}
## The Empirical Rule

For bell-shaped distributions:

| Interval | Approximate % of data |
|----------|----------------------|
| $\bar{X} \pm 1s$ | Ōēł 68% |
| $\bar{X} \pm 2s$ | Ōēł 95% |
| $\bar{X} \pm 3s$ | Ōēł 99.7% |
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
mu <- 480; sigma <- 90
tibble(x = seq(mu - 4*sigma, mu + 4*sigma, length.out = 500),
       y = dnorm(x, mu, sigma)) |>
  ggplot(aes(x, y)) +
  geom_line() +
  geom_area(aes(fill = "99.7%"), alpha=0.2) +
  geom_vline(xintercept = c(mu-sigma, mu+sigma), color="blue", linetype=2) +
  geom_vline(xintercept = c(mu-2*sigma, mu+2*sigma), color="red", linetype=2) +
  geom_vline(xintercept = c(mu-3*sigma, mu+3*sigma), color="darkgreen", linetype=2) +
  annotate("text", x=mu, y=0.003, label="68%", color="blue") +
  annotate("text", x=mu+2.5*sigma, y=0.001, label="95%", color="red") +
  labs(title="Empirical Rule (mu=480, sigma=90)",
       x="Score", y="Density") +
  theme_bw() + theme(legend.position="none")
```

# Section 3-3: Measures of Position {background-color="#2c4a6e"}

## Locating a Value Within a Distribution

**Measures of position** describe where a particular value stands in relation to the rest of the dataset. They answer the question: *How does this value compare to others?*

Key position measures include:

- **Standard score (z-score)** ŌĆö distance from mean in standard deviation units
- **Percentiles** ŌĆö divide data into 100 equal groups
- **Deciles** ŌĆö divide data into 10 equal groups
- **Quartiles** ŌĆö divide data into 4 equal groups

---

## Standard Scores (z-Scores)

A **z-score** (standard score) converts a raw value into the number of standard deviations it lies above or below the mean, enabling fair comparison across different distributions.

:::{.callout-note}
## z-Score Formulas

$$z = \frac{X - \bar{X}}{s} \quad \text{(sample)} \qquad z = \frac{X - \mu}{\sigma} \quad \text{(population)}$$

- $z > 0$: value is **above** the mean
- $z = 0$: value **equals** the mean
- $z < 0$: value is **below** the mean
:::

---

## Example 3-29: Comparing Test Performance

:::{.callout-tip}
## Example 3-29
A student scored 65 on a calculus exam ($\bar{X}=50$, $s=10$) and 30 on a history exam ($\bar{X}=25$, $s=5$). On which test did she perform relatively better?
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
z_calc <- (65 - 50) / 10
z_hist <- (30 - 25) / 5

cat("Calculus z-score:", z_calc, "\n")
cat("History z-score:", z_hist, "\n")
cat("\nBoth z-scores are equal (1.5) ŌåÆ same relative performance.\n")
```

Both scores are 1.5 standard deviations above their respective means ŌĆö **equal relative positions**.

---

## Example 3-30: Which Score Is Relatively Higher?

:::{.callout-tip}
## Example 3-30

| Test | Score (X) | Mean ($\bar{X}$) | SD ($s$) |
|------|-----------|-----------------|----------|
| A | 38 | 40 | 5 |
| B | 94 | 100 | 10 |

Find each z-score and identify the relatively higher score.
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
z_A <- (38 - 40) / 5
z_B <- (94 - 100) / 10

cat("Test A z-score:", z_A, "\n")
cat("Test B z-score:", z_B, "\n")
cat("\nConclusion: Test A score is relatively higher (z = -0.4 > -0.6)\n")
```

The Test A score has a **higher relative position** (z = ŌłÆ0.4 vs z = ŌłÆ0.6).

---

## Percentiles

**Percentiles** locate a data value's position in hundredths. A value at the $p$th percentile exceeds $p$% of all values in the distribution.

:::{.callout-note}
## Percentile Formula

$$\text{Percentile rank of } X = \frac{(\text{number of values below } X) + 0.5}{\text{total number of values}} \times 100$$

To find the **value** corresponding to a given percentile $p$:

$$c = \frac{n \cdot p}{100}$$

- If $c$ is **not** a whole number: round **up** and use that position.
- If $c$ **is** a whole number: average the $c$th and $(c+1)$th values.
:::

---

## Example 3-31: Percentile Graph (Systolic Blood Pressure)

:::{.callout-tip}
## Example 3-31
Construct a percentile graph for the systolic blood pressure readings of 200 college students.

| Class Boundaries | Frequency |
|-----------------|-----------|
| 89.5ŌĆō104.5 | 24 |
| 104.5ŌĆō119.5 | 62 |
| 119.5ŌĆō134.5 | 72 |
| 134.5ŌĆō149.5 | 26 |
| 149.5ŌĆō164.5 | 12 |
| 164.5ŌĆō179.5 | 4 |
:::

---

## Example 3-31: Solution

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
upper <- c(104.5, 119.5, 134.5, 149.5, 164.5, 179.5)
freq  <- c(24, 62, 72, 26, 12, 4)
cum_f <- cumsum(freq)
cum_pct <- (cum_f / sum(freq)) * 100

df_bp <- data.frame(upper, cum_f, cum_pct)
knitr::kable(df_bp, col.names=c("Upper Boundary","Cumulative Freq","Cumulative %"),
             caption="Systolic Blood Pressure ŌĆö Percentile Table")
```

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
ggplot(df_bp, aes(x=upper, y=cum_pct)) +
  geom_line(color="#2c4a6e", linewidth=1) +
  geom_point(size=3, color="#2c4a6e") +
  labs(title="Percentile Graph ŌĆö Systolic Blood Pressure",
       x="Blood Pressure (mm Hg)", y="Cumulative Percentage") +
  theme_bw()
```

---

## Examples 3-32 & 3-33: Finding Percentile Rank

:::{.callout-tip}
## Examples 3-32 & 3-33
A teacher gives a 20-point test to 10 students. Scores: 18, 15, 12, 6, 8, 2, 3, 5, 20, 10

Find the percentile rank of a score of **12** and a score of **6**.
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
scores <- c(18, 15, 12, 6, 8, 2, 3, 5, 20, 10)
n <- length(scores)

prank <- function(x, data) {
  below <- sum(data < x)
  ((below + 0.5) / length(data)) * 100
}

cat("Sorted scores:", paste(sort(scores), collapse=", "), "\n\n")
cat("Percentile rank of 12:", prank(12, scores), "\n")
cat("Percentile rank of 6:", prank(6, scores), "\n")
```

Score 12 is at the **65th percentile** (did better than 65% of class). Score 6 is at the **35th percentile**.

---

## Examples 3-34 & 3-35: Value at a Given Percentile

:::{.callout-tip}
## Example 3-34 & 3-35
1. Using the test scores from 3-32, find the value corresponding to the **25th percentile**.

2. Find the value corresponding to the **60th percentile**.
:::

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
scores_sorted <- sort(c(18, 15, 12, 6, 8, 2, 3, 5, 20, 10))
n <- length(scores_sorted)
cat("Sorted scores:", scores_sorted, "\n\n")

# 25th percentile
c25 <- n * 25 / 100
cat("c for 25th percentile:", c25, "(not whole ŌåÆ round up to 3)\n")
cat("Value at position 3:", scores_sorted[3], "\n\n")

# 60th percentile
c60 <- n * 60 / 100
cat("c for 60th percentile:", c60, "(whole number ŌåÆ average positions 6 and 7)\n")
cat("Average of positions 6 and 7:", (scores_sorted[6] + scores_sorted[7])/2, "\n")
```

---

## Quartiles and Deciles

**Quartiles** divide the dataset into four equal groups ($Q_1$, $Q_2$, $Q_3$), while **deciles** divide it into ten groups ($D_1, \ldots, D_9$).

:::{.callout-note}
## Key Relationships

| Position Measure | Percentile Equivalent |
|-----------------|----------------------|
| $Q_1$ (1st quartile) | $P_{25}$ |
| $Q_2$ (2nd quartile = median) | $P_{50}$ |
| $Q_3$ (3rd quartile) | $P_{75}$ |
| $D_1$ (1st decile) | $P_{10}$ |
| $D_5$ (5th decile) | $P_{50}$ |

**Interquartile Range (IQR):** $\text{IQR} = Q_3 - Q_1$

The IQR measures the spread of the middle 50% of the data.
:::

---

## Finding Quartiles ŌĆö Procedure

:::{.callout-important}
## Step-by-Step Procedure for Q1, Q2, Q3

1. Arrange data from lowest to highest.
2. Find the median ($Q_2$) of the entire dataset.
3. Find the median of the lower half (below $Q_2$) ŌåÆ this is $Q_1$.
4. Find the median of the upper half (above $Q_2$) ŌåÆ this is $Q_3$.
:::

---

## Example 3-36: Finding Quartiles

:::{.callout-tip}
## Example 3-36
Find $Q_1$, $Q_2$, and $Q_3$ for the data set: 15, 13, 6, 5, 12, 50, 22, 18
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
x <- c(15, 13, 6, 5, 12, 50, 22, 18)
xs <- sort(x)
cat("Sorted data:", xs, "\n\n")
```

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
Q2 <- median(xs)
lower <- xs[xs < Q2]
upper <- xs[xs > Q2]

Q1 <- median(lower)
Q3 <- median(upper)
IQR_val <- Q3 - Q1

cat("Q2 (median):", Q2, "\n")
cat("Lower half:", lower, "ŌåÆ Q1 =", Q1, "\n")
cat("Upper half:", upper, "ŌåÆ Q3 =", Q3, "\n")
cat("IQR:", IQR_val, "\n")
```

$Q_1 = 9$, $Q_2 = 14$, $Q_3 = 20$, $\text{IQR} = 11$.

---

## Outliers

An **outlier** is an extremely high or low data value that stands apart from the rest of the distribution. Outliers can distort the mean and standard deviation significantly.

:::{.callout-important}
## Procedure for Identifying Outliers

1. Arrange data and find $Q_1$ and $Q_3$.
2. Compute $\text{IQR} = Q_3 - Q_1$.
3. Multiply: $1.5 \times \text{IQR}$.
4. Compute fences: lower fence $= Q_1 - 1.5(\text{IQR})$; upper fence $= Q_3 + 1.5(\text{IQR})$.
5. Any value **outside** these fences is a suspected outlier.
:::

---

## Example 3-37: Checking for Outliers

:::{.callout-tip}
## Example 3-37
Check the following dataset for outliers: 5, 6, 12, 13, 15, 18, 22, 50

(This is the sorted version of Example 3-36's data.)
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
x <- c(5, 6, 12, 13, 15, 18, 22, 50)

Q1 <- 9; Q3 <- 20  # from Example 3-36
IQR_val <- Q3 - Q1

lower_fence <- Q1 - 1.5 * IQR_val
upper_fence <- Q3 + 1.5 * IQR_val
```

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
cat("Q1:", Q1, "  Q3:", Q3, "  IQR:", IQR_val, "\n")
cat("1.5 ├Ś IQR:", 1.5 * IQR_val, "\n")
cat("Lower fence:", lower_fence, "\n")
cat("Upper fence:", upper_fence, "\n\n")

outliers <- x[x < lower_fence | x > upper_fence]
cat("Values outside fences:", outliers, "\n")
cat("Conclusion: 50 is an outlier.\n")
```

# Section 3-4: Exploratory Data Analysis {background-color="#2c4a6e"}

## Exploratory Data Analysis (EDA)

**Exploratory data analysis (EDA)**, introduced by John Tukey, uses graphical and resistant statistics to uncover patterns, distribution shape, and potential outliers without assuming a particular model.

:::{.callout-note}
## Traditional vs. EDA Approaches

| Traditional | Exploratory Data Analysis |
|-------------|--------------------------|
| Frequency distribution | Stem-and-leaf plot |
| Histogram | Boxplot |
| Mean | Median |
| Standard deviation | Interquartile range (IQR) |

EDA statistics (median, IQR) are **resistant** ŌĆö less influenced by extreme values ŌĆö while mean and standard deviation are **nonresistant**.
:::

---

## The Five-Number Summary

A **five-number summary** captures the full spread and shape of a distribution using five key values:

:::{.callout-note}
## Five-Number Summary

1. **Minimum** ŌĆö lowest data value
2. **$Q_1$** ŌĆö first quartile (25th percentile)
3. **Median ($Q_2$)** ŌĆö middle value (50th percentile)
4. **$Q_3$** ŌĆö third quartile (75th percentile)
5. **Maximum** ŌĆö highest data value
:::

These five values form the basis of the **boxplot** (box-and-whisker plot).

---

## Constructing a Boxplot

:::{.callout-important}
## Steps for Building a Boxplot

1. Find the five-number summary.
2. Draw a horizontal axis scaled to include the min and max.
3. Draw a box from $Q_1$ to $Q_3$ with a vertical line at the median.
4. Extend **whiskers** from the box to the minimum and maximum values.
:::

**Reading the Shape from a Boxplot:**

- Median near center of box ŌåÆ approximately **symmetric**
- Median left of center ŌåÆ **positively skewed** (right-skewed)
- Median right of center ŌåÆ **negatively skewed** (left-skewed)
- Longer right whisker ŌåÆ **positively skewed**
- Longer left whisker ŌåÆ **negatively skewed**

---

## Example 3-38: Meteorites Found

:::{.callout-tip}
## Example 3-38
The number of meteorites found in 10 U.S. states: 89, 47, 164, 296, 30, 215, 138, 78, 48, 39. Construct a boxplot and describe the distribution.
:::

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
meteorites <- c(89, 47, 164, 296, 30, 215, 138, 78, 48, 39)
xs <- sort(meteorites)
cat("Sorted:", xs, "\n")

five_num <- fivenum(meteorites)
names(five_num) <- c("Min","Q1","Median","Q3","Max")
print(five_num)
```

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
boxplot(meteorites, horizontal=TRUE,
        main="Meteorites Found in 10 U.S. States",
        xlab="Number of Meteorites",
        col="#8db4d6")
```

The distribution is **positively skewed** ŌĆö the right whisker extends much farther than the left.

---

## Example 3-38: Using ggplot2

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
meteorites <- c(89, 47, 164, 296, 30, 215, 138, 78, 48, 39)

ggplot(data.frame(x = meteorites), aes(x = "", y = x)) +
  geom_boxplot(fill="#8db4d6", color="#2c4a6e", width=0.4) +
  coord_flip() +
  labs(title="Boxplot: Meteorites Found in 10 U.S. States",
       y="Number of Meteorites", x="") +
  theme_bw()
```

---

## Example 3-39: Comparing Two Distributions

:::{.callout-tip}
## Example 3-39
A dietitian compares the sodium content (mg) of real cheese vs. cheese substitute. Compare the two distributions using boxplots.

**Real cheese:** 310, 420, 45, 40, 220, 240, 180, 90
**Cheese substitute:** 270, 180, 250, 290, 130, 260, 340, 310
:::

---

## Example 3-39: Solution

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
real  <- c(310, 420, 45, 40, 220, 240, 180, 90)
sub   <- c(270, 180, 250, 290, 130, 260, 340, 310)

cat("Real cheese five-number summary:\n")
print(fivenum(real))
cat("\nCheese substitute five-number summary:\n")
print(fivenum(sub))

df_cheese <- data.frame(
  sodium = c(real, sub),
  type   = rep(c("Real Cheese","Cheese Substitute"), each=8)
)
```

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
ggplot(df_cheese, aes(x=type, y=sodium, fill=type)) +
  geom_boxplot(alpha=0.7) +
  coord_flip() +
  scale_fill_manual(values=c("#2c4a6e","#8db4d6")) +
  labs(title="Sodium Content: Real Cheese vs. Cheese Substitute",
       y="Sodium (mg)", x="") +
  theme_bw() + theme(legend.position="none")
```

---

## Example 3-39: Interpretation

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
real <- c(310, 420, 45, 40, 220, 240, 180, 90)
sub  <- c(270, 180, 250, 290, 130, 260, 340, 310)

summary_df <- data.frame(
  Measure = c("Minimum","Q1","Median","Q3","Maximum","IQR"),
  `Real Cheese`    = c(min(real), quantile(real,0.25), median(real),
                       quantile(real,0.75), max(real),
                       IQR(real)),
  `Cheese Sub`     = c(min(sub), quantile(sub,0.25), median(sub),
                       quantile(sub,0.75), max(sub),
                       IQR(sub))
)
```

---

```{r,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
knitr::kable(summary_df, caption="Five-Number Summary Comparison")
```

:::{.callout-note}
**Interpretation:** The cheese substitute has a **higher median** sodium content (265 vs 200 mg) and **less variability** (IQR = 85 vs 207 mg). Real cheese data are more spread out and slightly positively skewed.
:::

---

## Summary: Chapter 3 at a Glance

:::{.callout-note}
## Section 3-1: Measures of Central Tendency

**Mean** $\bar{X} = \frac{\Sigma X}{n}$ ŌĆö uses all values; sensitive to outliers

**Median** ŌĆö middle value; resistant to outliers

**Mode** ŌĆö most frequent; can be used with categorical data

**Midrange** $= \frac{\min + \max}{2}$ ŌĆö easily computed; sensitive to extremes

**Weighted Mean** $= \frac{\Sigma wX}{\Sigma w}$ ŌĆö accounts for different weights
:::

:::{.callout-note}
## Section 3-2: Measures of Variation

**Range** $R = \max - \min$

**Variance** $s^2 = \frac{n(\Sigma X^2)-(\Sigma X)^2}{n(n-1)}$

**Standard Deviation** $s = \sqrt{s^2}$

**CVar** $= \frac{s}{\bar{X}} \times 100\%$ ŌĆö compares variability across datasets

**Chebyshev's Theorem** ŌĆö at least $1-\frac{1}{k^2}$ within $k$ SDs (any distribution)

**Empirical Rule** ŌĆö 68/95/99.7% within 1/2/3 SDs (bell-shaped only)
:::

---

## Summary: Chapter 3 at a Glance (cont.)

:::{.callout-note}
## Section 3-3: Measures of Position

**z-score** $= \frac{X - \bar{X}}{s}$ ŌĆö standard deviations from the mean

**Percentile rank** $= \frac{(\text{# below }X)+0.5}{n} \times 100$

**Quartiles**: $Q_1 = P_{25}$, $Q_2 = P_{50}$, $Q_3 = P_{75}$

**IQR** $= Q_3 - Q_1$ ŌĆö spread of the middle 50%

**Outlier fences**: $Q_1 - 1.5(\text{IQR})$ and $Q_3 + 1.5(\text{IQR})$
:::

:::{.callout-note}
## Section 3-4: Exploratory Data Analysis

**Five-number summary**: Min, $Q_1$, Median, $Q_3$, Max

**Boxplot**: visual display of five-number summary revealing center, spread, and skew

**EDA statistics** (median, IQR) are **resistant** to outliers

**Traditional statistics** (mean, SD) are **nonresistant**
:::

---

## Acknowledgements

-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2011). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
