---
title: "Statistics"
subtitle: "Chapter 11: Other Chi-Square Tests"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch11-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
```

## Overview
The **chi-square distribution** has several important applications beyond testing a single variance:

1. **Goodness-of-fit test** — Does a distribution follow a hypothesized pattern?
2. **Test of independence** — Are two categorical variables related?
3. **Test of homogeneity** — Are proportions equal across several populations?

All three tests use the same chi-square statistic and are **always right-tailed**.

# Section 11-1: Test for Goodness of Fit {background-color="#2c4a6e"}

## The Goodness-of-Fit Test

::: {.callout-note}
The **chi-square goodness-of-fit test** compares observed frequencies with expected frequencies to determine whether a distribution follows a specific pattern.

$$\chi^2 = \sum \frac{(O - E)^2}{E}$$

- $O$ = observed frequency
- $E$ = expected frequency
- **d.f.** = number of categories − 1

**Assumptions:**
1. Data are from a random sample
2. Expected frequency for each category $\geq 5$

The test is always **right-tailed** (small χ² → good fit; large χ² → poor fit).
:::

## Example 11-1 {#example-11-1}

:::{.callout-warning}
## Fruit Soda Flavor Preference

A market analyst surveys 100 people about their preference among five flavors of a new fruit soda. If there is no preference, each flavor should be equally popular. The results are:

| Flavor | Cherry | Strawberry | Orange | Lime | Grape |
|--------|--------|------------|--------|------|-------|
| Observed | 32 | 28 | 16 | 14 | 10 |
| Expected | 20 | 20 | 20 | 20 | 20 |

At α = 0.05, is there enough evidence to reject the claim that consumers show no preference?
:::

## Example 11-1

:::{.callout-tip}
## Solution

**Mathematical solution**

**Step 1:** H₀: Consumers show no preference (claim) &nbsp;&nbsp; H₁: Consumers show a preference

**Step 2:** d.f. = 5 − 1 = 4; CV = 9.488

**Step 3:**
$$\chi^2 = \frac{(32-20)^2}{20} + \frac{(28-20)^2}{20} + \frac{(16-20)^2}{20} + \frac{(14-20)^2}{20} + \frac{(10-20)^2}{20}$$
$$= \frac{144}{20} + \frac{64}{20} + \frac{16}{20} + \frac{36}{20} + \frac{100}{20} = 7.2 + 3.2 + 0.8 + 1.8 + 5.0 = \boxed{18.0}$$

**Step 4:** 18.0 > 9.488 → **Reject H₀**

**Step 5:** There is sufficient evidence to conclude that consumers show a preference for certain flavors.
:::

## Example 11-1

:::{.callout-tip}
## Solution

**R implementation**

```{r e11-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
observed <- c(32, 28, 16, 14, 10)
expected <- c(20, 20, 20, 20, 20)

chi2 <- sum((observed - expected)^2 / expected)
df   <- length(observed) - 1
cv   <- qchisq(0.95, df)
p_val <- pchisq(chi2, df, lower.tail = FALSE)

cat("chi-square =", round(chi2, 4), "\n")
cat("Critical value =", round(cv, 3), "\n")
cat("P-value =", round(p_val, 6), "\n")
cat("Decision:", ifelse(chi2 > cv, "Reject H0", "Do not reject H0"), "\n")
```
:::
---

```{r e11-1-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# Visualize
df_plot <- data.frame(
  Flavor = c("Cherry","Strawberry","Orange","Lime","Grape"),
  Observed = observed, Expected = expected)
df_long <- pivot_longer(df_plot, -Flavor, names_to = "Type", values_to = "Count")
ggplot(df_long, aes(x = Flavor, y = Count, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() + labs(title = "Observed vs. Expected Frequencies")
```

## Example 11-2 {#example-11-2}

:::{.callout-warning}
## Retired Senior Executives Return to Work

A survey found that after returning to work, retired executives were distributed as: 38% employed by another organization, 32% self-employed, 23% freelancing/consulting, 7% formed their own companies. A local researcher surveyed 300 executives and found: 122 working for another company, 85 self-employed, 76 freelancing/consulting, and 17 formed their own companies. At α = 0.10, test the claim that these proportions are the same.
:::

## Example 11-2

:::{.callout-tip}
## Solution

**Mathematical solution**

**Step 1:** H₀: Distribution is 38%/32%/23%/7% (claim) &nbsp;&nbsp; H₁: Distribution differs

**Step 2:** d.f. = 4 − 1 = 3; CV = 6.251

**Step 3:** Expected frequencies: 0.3$8 \times 3$00 = 114; 0.3$2 \times 3$00 = 96; 0.2$3 \times 3$00 = 69; 0.0$7 \times 3$00 = 21

$$\chi^2 = \frac{(122-114)^2}{114} + \frac{(85-96)^2}{96} + \frac{(76-69)^2}{69} + \frac{(17-21)^2}{21}$$
$$= 0.561 + 1.260 + 0.710 + 0.762 = \boxed{3.293}$$

**Step 4:** 3.293 < 6.251 → **Do not reject H₀**

**Step 5:** There is not enough evidence to conclude that the local distribution differs from the national survey.
:::

## Example 11-2

:::{.callout-tip}
## Solution

**R implementation**

```{r e11-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
observed <- c(122, 85, 76, 17)
props    <- c(0.38, 0.32, 0.23, 0.07)
expected <- props * 300

chi2  <- sum((observed - expected)^2 / expected)
df    <- length(observed) - 1
cv    <- qchisq(0.90, df)
p_val <- pchisq(chi2, df, lower.tail = FALSE)

cat("Expected:", expected, "\n")
cat("chi-square =", round(chi2, 4), "\n")
cat("Critical value =", round(cv, 3), "\n")
cat("P-value =", round(p_val, 4), "\n")
cat("Decision:", ifelse(chi2 > cv, "Reject H0", "Do not reject H0"), "\n")
```
:::

## Example 11-3 {#example-11-3}

:::{.callout-warning}
## Firearm Deaths

A researcher read that firearm-related deaths for people aged 1–18 are distributed as: 74% accidental, 16% homicides, 10% suicides. In her district, there were 68 accidental deaths, 27 homicides, and 5 suicides (total = 100) during the past year. At α = 0.10, test the claim.
:::

## Example 11-3

:::{.callout-tip}
## Solution

**Mathematical solution**

**Step 1:** H₀: Deaths are 74% accidental, 16% homicides, 10% suicides (claim) &nbsp;&nbsp; H₁: Distribution differs

**Step 2:** d.f. = 3 − 1 = 2; CV = 4.605

**Step 3:** Expected: 74, 16, 10

$$\chi^2 = \frac{(68-74)^2}{74} + \frac{(27-16)^2}{16} + \frac{(5-10)^2}{10} = 0.486 + 7.5625 + 2.5 = \boxed{10.549}$$

**Step 4:** 10.549 > 4.605 → **Reject H₀**

**Step 5:** There is enough evidence to reject the claim that the distribution in this district matches the national pattern.
:::

## Example 11-3

:::{.callout-tip}
## Solution

**R implementation**

```{r e11-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
observed <- c(68, 27, 5)
expected <- c(74, 16, 10)

chi2  <- sum((observed - expected)^2 / expected)
df    <- length(observed) - 1
cv    <- qchisq(0.90, df)
p_val <- pchisq(chi2, df, lower.tail = FALSE)

cat("chi-square =", round(chi2, 4), "\n")
cat("Critical value =", round(cv, 3), "\n")
cat("P-value =", round(p_val, 4), "\n")
cat("Decision:", ifelse(chi2 > cv, "Reject H0", "Do not reject H0"), "\n")
```
:::

## Example 11-4 {#example-11-4}

:::{.callout-warning}
## Test of Normality (Optional)

Use the chi-square goodness-of-fit test to determine if the following frequency distribution is normally distributed. Use α = 0.05.

| Boundaries | Frequency |
|------------|-----------|
| 89.5–104.5 | 24 |
| 104.5–119.5 | 62 |
| 119.5–134.5 | 72 |
| 134.5–149.5 | 26 |
| 149.5–164.5 | 12 |
| 164.5–179.5 | 4 |
| **Total** | **200** |
:::

## Example 11-4

:::{.callout-tip}
## Solution

**Solve**

A **contingency table** ($R \times C$ table) cross-classifies data from two categorical variables.

**Two chi-square tests use contingency tables:**

1. **Test of independence** — Are two variables related? (single sample, observe both variables)
2. **Test of homogeneity** — Are proportions equal across populations? (multiple samples, sample sizes set in advance)

**Both use the same formulas:**

$$E = \frac{\text{(row sum)} \times \text{(column sum)}}{\text{grand total}}, \quad \chi^2 = \sum \frac{(O-E)^2}{E}$$

$$d.f. = (R-1)(C-1)$$

**Assumption:** All expected cell frequencies $\geq 5$.

:::

## Test of Independence — Overview

When testing whether two variables are **independent**:

- **H₀:** Variable A is independent of Variable B
- **H₁:** Variable A is dependent on Variable B (claim)

Rejecting H₀ means the variables are **related** — but it does not specify the direction of the relationship.

The test is always **right-tailed**.

## Example 11-5 {#example-11-5}

:::{.callout-warning}
## Hospitals and Infections

A researcher examines whether the type of hospital infection is related to which hospital it occurred in. Data for 582 total infections across 3 hospitals:

| Hospital | Surgical Site | Pneumonia | Bloodstream | Total |
|----------|--------------|-----------|-------------|-------|
| A | 41 | 27 | 51 | 119 |
| B | 36 | 3 | 40 | 79 |
| C | 169 | 106 | 109 | 384 |
| **Total** | **246** | **136** | **200** | **582** |

At α = 0.05, test the claim that infection type depends on hospital.
:::

## Example 11-5

:::{.callout-tip}
## Solution

**Mathematical solution**

**Step 1:** H₀: Infection type is independent of hospital &nbsp;&nbsp; H₁: Infection type depends on hospital (claim)

**Step 2:** d.f. = (3−1)(3−1) = 4; CV = 9.488

**Step 3:** Expected values (E = row$\times$col/grand):
$$E_{A,\text{Surg}} = \frac{119 \times 246}{582} = 50.30, \quad E_{B,\text{Pneu}} = \frac{79 \times 136}{582} = 18.46, \quad \ldots$$

$$\chi^2 = \frac{(41-50.30)^2}{50.30} + \frac{(27-27.81)^2}{27.81} + \frac{(51-40.89)^2}{40.89} + \cdots \approx \boxed{30.698}$$

**Step 4:** 30.698 > 9.488 → **Reject H₀**

**Step 5:** There is sufficient evidence that infection type is related to the hospital where it occurred.
:::

## Example 11-5

:::{.callout-tip}
## Solution

**R implementation**

```{r e11-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
M <- matrix(c(41, 27, 51,
              36,  3, 40,
             169, 106, 109),
            nrow = 3, byrow = TRUE,
            dimnames = list(c("A","B","C"),
                            c("Surgical","Pneumonia","Bloodstream")))

result <- chisq.test(M)
cat("chi-square =", round(result$statistic, 4), "\n")
cat("d.f. =", result$parameter, "\n")
cat("P-value =", round(result$p.value, 6), "\n")
cat("Decision:", ifelse(result$p.value < 0.05, "Reject H0", "Do not reject H0"), "\n")

cat("\nExpected frequencies:\n")
round(result$expected, 2)
```
:::

## Example 11-6 {#example-11-6}

:::{.callout-warning}
## Alcohol Consumption and Gender

A researcher examines whether alcohol consumption level is related to gender. A sample of 68 people yielded:

| Gender | Low | Moderate | High | Total |
|--------|-----|----------|------|-------|
| Male | 10 | 9 | 8 | 27 |
| Female | 13 | 16 | 12 | 41 |
| **Total** | **23** | **25** | **20** | **68** |

At α = 0.10, test whether alcohol consumption depends on gender.
:::

## Example 11-6

:::{.callout-tip}
## Solution

**Mathematical solution**

**Step 1:** H₀: Alcohol consumption is independent of gender &nbsp;&nbsp; H₁: Consumption depends on gender (claim)

**Step 2:** d.f. = (2−1)(3−1) = 2; CV = 4.605

**Step 3:** Expected values:
$$E_{\text{Male, Low}} = \frac{27 \times 23}{68} = 9.13, \quad E_{\text{Male, Mod}} = \frac{27 \times 25}{68} = 9.93, \quad \ldots$$

$$\chi^2 = \frac{(10-9.13)^2}{9.13} + \frac{(9-9.93)^2}{9.93} + \frac{(8-7.94)^2}{7.94} + \frac{(13-13.87)^2}{13.87} + \cdots \approx \boxed{0.283}$$

**Step 4:** 0.283 < 4.605 → **Do not reject H₀**

**Step 5:** There is not enough evidence to conclude that alcohol consumption is related to gender.
:::

## Example 11-6

:::{.callout-tip}
## Solution

**R implementation**

```{r e11-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
M2 <- matrix(c(10, 9, 8,
               13, 16, 12),
             nrow = 2, byrow = TRUE,
             dimnames = list(c("Male","Female"),
                             c("Low","Moderate","High")))

result2 <- chisq.test(M2)
cat("chi-square =", round(result2$statistic, 4), "\n")
cat("d.f. =", result2$parameter, "\n")
cat("P-value =", round(result2$p.value, 4), "\n")
cat("Decision:", ifelse(result2$p.value < 0.10, "Reject H0", "Do not reject H0"), "\n")

cat("\nExpected frequencies:\n")
round(result2$expected, 2)
```
:::

## Test of Homogeneity of Proportions

::: {.callout-note}
The **homogeneity test** determines whether proportions are equal across multiple independently sampled populations.

- Sample sizes are fixed **in advance** (by the researcher)
- H₀: p₁ = p₂ = ... = pₖ
- H₁: At least one proportion differs

**Computational procedure is identical** to the independence test — same formula for E and χ².

**Key distinction:** Independence test → one sample, two variables; Homogeneity test → multiple samples, one variable.
:::

## Example 11-7 {#example-11-7}

:::{.callout-warning}
## Money and Happiness

A psychologist selects 100 people from each of four income groups and asks whether they are "very happy." Results:

| | <\$30K | \$30K–\$74.9K | \$75K–\$99.9K | ≥\$100K | Total |
|--|--------|--------------|--------------|---------|-------|
| Yes | 24 | 33 | 38 | 49 | 144 |
| No | 76 | 67 | 62 | 51 | 256 |
| **Total** | **100** | **100** | **100** | **100** | **400** |

At α = 0.05, test the claim that happiness proportions are the same across income groups.
:::

## Example 11-7

:::{.callout-tip}
## Solution

**Mathematical solution**

**Step 1:** H₀: p₁ = p₂ = p₃ = p₄ (claim) &nbsp;&nbsp; H₁: At least one proportion differs

**Step 2:** d.f. = (2−1)(4−1) = 3; CV = 7.815

**Step 3:** Since each column total = 100 and "Yes" total = 144:
$$E_{\text{Yes, col}} = \frac{144 \times 100}{400} = 36, \quad E_{\text{No, col}} = \frac{256 \times 100}{400} = 64 \quad \text{(for each column)}$$

$$\chi^2 = \frac{(24-36)^2}{36} + \frac{(33-36)^2}{36} + \frac{(38-36)^2}{36} + \frac{(49-36)^2}{36}$$
$$+ \frac{(76-64)^2}{64} + \frac{(67-64)^2}{64} + \frac{(62-64)^2}{64} + \frac{(51-64)^2}{64}$$
$$= 4 + 0.25 + 0.111 + 4.694 + 2.25 + 0.141 + 0.063 + 2.641 = \boxed{14.149}$$
:::
---

**Step 4:** 14.149 > 7.815 → **Reject H₀**

**Step 5:** There is sufficient evidence that happiness proportions differ across income groups.

## Example 11-7

:::{.callout-tip}
## Solution

**R implementation**

```{r e11-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
M3 <- matrix(c(24, 33, 38, 49,
               76, 67, 62, 51),
             nrow = 2, byrow = TRUE,
             dimnames = list(c("Yes","No"),
                             c("<30K","30-74.9K","75-99.9K",">=100K")))

result3 <- chisq.test(M3)
cat("chi-square =", round(result3$statistic, 4), "\n")
cat("d.f. =", result3$parameter, "\n")
cat("P-value =", round(result3$p.value, 6), "\n")
cat("Decision:", ifelse(result3$p.value < 0.05, "Reject H0", "Do not reject H0"), "\n")

cat("\nExpected frequencies:\n")
round(result3$expected, 2)
```
:::

## Key Takeaways

| Test | When to Use | d.f. | Formula |
|------|------------|------|---------|
| Goodness of Fit | Single sample vs. hypothesized distribution | k − 1 | $\chi^2 = \sum\frac{(O-E)^2}{E}$, $E$ given by proportions |
| Independence | Single sample; two categorical variables | (R−1)(C−1) | $E = \frac{\text{row} \times \text{col}}{\text{total}}$ |
| Homogeneity | Multiple samples; one categorical variable | (R−1)(C−1) | Same as independence |

::: {.callout-important}
All chi-square tests in this chapter are **right-tailed**. Expected frequencies in each cell must be **$\geq 5$**. If not, combine categories.
:::

## Acknowledgement
-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2011). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
