---
title: "Statistics"
subtitle: "Chapter 6: The Normal Distribution"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
language:
  callout-warning-title: "Problem"
  callout-important-title: "Key point"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch6-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview
**Sections covered:**

- **6-1** Normal Distributions
- **6-2** Applications of the Normal Distribution
- **6-3** The Central Limit Theorem
- **6-4** The Normal Approximation to the Binomial Distribution

The **normal distribution** is the most important and widely used continuous probability distribution in statistics. Its familiar **bell-shaped curve** appears throughout natural phenomena, measurement errors, and statistical inference.

# Section 6-1: Normal Distributions {background-color="#2c4a6e"}

## Properties of the Normal Distribution

:::{.callout-note}
## Definition
A **normal distribution** is a continuous, symmetric, bell-shaped distribution of a variable. The curve is determined entirely by the **mean** μ and **standard deviation** σ.
:::

**Key properties:**

1. The distribution is **bell-shaped** and symmetric about the mean
2. The mean, median, and mode are all **equal**
3. The curve is continuous — it never touches the x-axis
4. The total area under the curve equals **1**
5. Approximately **68%** of values lie within 1σ of μ; **95%** within 2σ; **99.7%** within 3σ

---

```{r normal-shape,echo=T,warning=FALSE,out.width = "80%",fig.align='center'}
x <- seq(-4, 4, length.out = 300)
df <- data.frame(x = x, y = dnorm(x))
ggplot(df, aes(x, y)) +
  geom_line(color = "#2c4a6e", linewidth = 1.2) +
  geom_area(fill = "#2c4a6e", alpha = 0.15) +
  labs(title = "Standard Normal Distribution",
       x = "z", y = "Density") +
  theme_minimal()
```

## The Standard Normal Distribution

:::{.callout-note}
## Definition
The **standard normal distribution** is a normal distribution with mean **μ = 0** and standard deviation **σ = 1**. Values on this distribution are called **z values** or **z scores**.
:::

The **z score** converts any normal variable X to the standard normal:

$$z = \frac{X - \mu}{\sigma}$$

- The z score represents the number of standard deviations X is from the mean
- Area under the curve = probability

## Finding Areas Under the Standard Normal Curve

Areas are found using:

- **Table E** (standard normal table) in textbook appendix
- **R**: `pnorm(z)` gives P(Z < z)

:::{.callout-important}
## Key R Functions
- `pnorm(z)` — area to the **left** of z
- `1 - pnorm(z)` — area to the **right** of z
- `pnorm(z2) - pnorm(z1)` — area **between** z1 and z2
- `qnorm(p)` — find z given left-tail area p
:::

## Example 6-1 {#example-6-1}

:::{.callout-warning}
## Area to the Left of z

Find the area to the left of z = 2.06.
:::

## Example 6-1

:::{.callout-tip}
## Solution

**R implementation**

**Mathematical solution**

- Look up z = 2.06 in Table E
- Area = **0.9803**
- Interpretation: 98.03% of values fall below z = 2.06

```{r e6-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# Area to the LEFT of z = 2.06
pnorm(2.06)
```
:::
---

```{r e6-1-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(-4, 4, length.out = 300)
df <- data.frame(x = x, y = dnorm(x))
ggplot(df, aes(x, y)) +
  geom_line(color = "#2c4a6e", linewidth = 1) +
  geom_area(data = filter(df, x <= 2.06),
            aes(x, y), fill = "#2c4a6e", alpha = 0.4) +
  annotate("text", x = 0, y = 0.15, label = "0.9803", size = 5) +
  geom_vline(xintercept = 2.06, linetype = "dashed", color = "red") +
  labs(title = "P(Z < 2.06) = 0.9803", x = "z", y = "Density") +
  theme_minimal()
```

## Example 6-2 {#example-6-2}

:::{.callout-warning}
## Area to the Right of z

Find the area to the right of z = −1.19.
:::

## Example 6-2

:::{.callout-tip}
## Solution

**R implementation**

**Mathematical solution**

- Area to the left of z = −1.19: Table E gives 0.1170
- Area to the right = 1 − 0.1170 = **0.8830**

```{r e6-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# Area to the RIGHT of z = -1.19
1 - pnorm(-1.19)
# Equivalently:
pnorm(-1.19, lower.tail = FALSE)
```
:::
---

```{r e6-2-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(-4, 4, length.out = 300)
df <- data.frame(x = x, y = dnorm(x))
ggplot(df, aes(x, y)) +
  geom_line(color = "#2c4a6e", linewidth = 1) +
  geom_area(data = filter(df, x >= -1.19),
            aes(x, y), fill = "#2c4a6e", alpha = 0.4) +
  annotate("text", x = 1, y = 0.15, label = "0.8830", size = 5) +
  geom_vline(xintercept = -1.19, linetype = "dashed", color = "red") +
  labs(title = "P(Z > -1.19) = 0.8830", x = "z", y = "Density") +
  theme_minimal()
```

## Example 6-3 {#example-6-3}

:::{.callout-warning}
## Area Between Two z Values

Find the area between z = −1.37 and z = +1.68.
:::

## Example 6-3

:::{.callout-tip}
## Solution

**R implementation**

**Mathematical solution**

- Area left of z = 1.68: 0.9535
- Area left of z = −1.37: 0.0853
- Area between = 0.9535 − 0.0853 = **0.8682**

```{r e6-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# Area BETWEEN z = -1.37 and z = 1.68
pnorm(1.68) - pnorm(-1.37)
```
:::
---

```{r e6-3-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(-4, 4, length.out = 300)
df <- data.frame(x = x, y = dnorm(x))
ggplot(df, aes(x, y)) +
  geom_line(color = "#2c4a6e", linewidth = 1) +
  geom_area(data = filter(df, x >= -1.37 & x <= 1.68),
            aes(x, y), fill = "#2c4a6e", alpha = 0.4) +
  annotate("text", x = 0, y = 0.15, label = "0.8682", size = 5) +
  geom_vline(xintercept = c(-1.37, 1.68), linetype = "dashed", color = "red") +
  labs(title = "P(-1.37 < Z < 1.68) = 0.8682", x = "z", y = "Density") +
  theme_minimal()
```

## Example 6-4 {#example-6-4}

:::{.callout-warning}
## Multiple Probability Calculations

Find the following probabilities for the standard normal distribution:

**(a)** P(0 < Z < 2.32)   **(b)** P(Z < 1.65)   **(c)** P(Z > 1.91)
:::

## Example 6-4

:::{.callout-tip}
## Solution

**Solve**

Find the z value such that the area to the left is 0.7123.

:::

:::{.callout-tip}
## Solution

**Mathematical solution**

- Search Table E for the area closest to 0.7123
- The corresponding z value is **z = 0.56**
:::

```{r e6-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# Find z given left-tail area = 0.7123
qnorm(0.7123)
```

:::{.callout-important}
## Key Rule
`qnorm(p)` is the **inverse** of `pnorm(z)`. Use it whenever you need the z value for a given probability.
:::

# Section 6-2: Applications of the Normal Distribution {background-color="#2c4a6e"}

## Applying the Normal Distribution

To find probabilities for a **non-standard** normal variable X (with mean μ and standard deviation σ):

:::{.callout-note}
## Formula
$$z = \frac{X - \mu}{\sigma}$$

Convert X to a z score, then find the area using the standard normal.
:::

In R, use `pnorm(x, mean = μ, sd = σ)` directly — no conversion needed.

:::{.callout-important}
## Key Steps
1. Identify μ and σ
2. Draw a diagram and shade the desired region
3. Convert X to z (or use R directly)
4. Find the probability
:::

## Example 6-6 {#example-6-6}
:::{.callout-warning}
The average amount of money that Americans spend on summer travel is normally distributed with μ = $146.21 and σ = $29.44. Find the probability that a randomly selected American spends less than $160 on summer travel.
:::

:::{.callout-tip}
## Solution

$$z = \frac{160 - 146.21}{29.44} = \frac{13.79}{29.44} \approx 0.47$$

Area to the left of z = 0.47 → **0.6808** (68.08%)
:::

```{r e6-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mu <- 146.21; sigma <- 29.44
pnorm(160, mean = mu, sd = sigma)
```

---

```{r e6-6-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(mu - 4*sigma, mu + 4*sigma, length.out = 300)
df <- data.frame(x = x, y = dnorm(x, mu, sigma))
ggplot(df, aes(x, y)) +
  geom_line(color = "#2c4a6e", linewidth = 1) +
  geom_area(data = filter(df, x <= 160), aes(x, y),
            fill = "#2c4a6e", alpha = 0.4) +
  geom_vline(xintercept = 160, linetype = "dashed", color = "red") +
  annotate("text", x = 140, y = 0.008, label = "68.08%", size = 5) +
  labs(title = "P(X < 160): Summer Spending",
       subtitle = paste0("μ = ", mu, ", σ = ", sigma),
       x = "Amount ($)", y = "Density") +
  theme_minimal()
```

## Example 6-7 {#example-6-7}

:::{.callout-warning}
## Newspaper Recycling

Each month, an American household generates an average of 28 pounds of newspaper for recycling, with σ = 2 pounds. Assume a normal distribution.

**(a)** P(27 < X < 31)   **(b)** P(X > 30.2)
:::

## Example 6-7

:::{.callout-tip}
## Solution

**R implementation**

```{r e6-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mu <- 28; sigma <- 2

# (a) P(27 < X < 31)
pnorm(31, mu, sigma) - pnorm(27, mu, sigma)

# (b) P(X > 30.2)
pnorm(30.2, mu, sigma, lower.tail = FALSE)
```

**Answers:** (a) $\approx$ **62.47%**; (b) $\approx$ **13.57%**
:::
---

```{r e6-7-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(mu - 4*sigma, mu + 4*sigma, length.out = 300)
df <- data.frame(x = x, y = dnorm(x, mu, sigma))
p1 <- ggplot(df, aes(x, y)) +
  geom_line(color = "#2c4a6e", linewidth = 1) +
  geom_area(data = filter(df, x >= 27 & x <= 31),
            aes(x, y), fill = "#2c4a6e", alpha = 0.4) +
  labs(title = "(a) P(27 < X < 31)", x = "Pounds", y = "") +
  theme_minimal()
p2 <- ggplot(df, aes(x, y)) +
  geom_line(color = "#2c4a6e", linewidth = 1) +
  geom_area(data = filter(df, x >= 30.2),
            aes(x, y), fill = "#c0392b", alpha = 0.4) +
  labs(title = "(b) P(X > 30.2)", x = "Pounds", y = "") +
  theme_minimal()
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Example 6-8 {#example-6-8}

:::{.callout-warning}
## Coffee Consumption

The per-capita consumption of coffee is normally distributed with μ = 1.64 gallons per year and σ = 0.24 gallons. In a sample of 500 people, how many consume **less than 1 gallon** per year?
:::

## Example 6-8

:::{.callout-tip}
## Solution

**Solve**

To qualify for a police academy, applicants are tested on their physical fitness. The scores are normally distributed with μ = 200 and σ = 20. The top **10%** of applicants will be accepted. Find the minimum score needed.

:::

:::{.callout-tip}
## Solution

**Mathematical solution**

- Need the 90th percentile → find z such that P(Z < z) = 0.90
- $z \approx 1.28$
- X = μ + z·σ = 200 + 1.28(20) = **225.6 $\approx 226$**
:::

```{r e6-8-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mu <- 200; sigma <- 20
# Find the value X such that P(X > x) = 0.10
qnorm(0.90, mean = mu, sd = sigma)
```

**Answer:** An applicant needs a score of at least **226** to be in the top 10%.

## Example 6-10 {#example-6-10}
:::{.callout-warning}
For a certain group, systolic blood pressure is normally distributed with μ = 120 and σ = 8. Find the range of pressures that includes the **middle 60%** of readings.
:::

:::{.callout-tip}
## Solution

- Middle 60% → 20% in each tail → find z for P(Z < z) = 0.20 and P(Z < z) = 0.80
- $z_1 \approx -0.84$, $z_2 \approx 0.84$
- X₁ = 120 + (−0.84)(8) = **113.28**
- X₂ = 120 + (0.84)(8) = **126.72**
:::

```{r e6-10-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mu <- 120; sigma <- 8
qnorm(0.20, mean = mu, sd = sigma)
qnorm(0.80, mean = mu, sd = sigma)
```

**Answer:** The middle 60% of readings fall between **113.28** and **126.72** mmHg.

## Example 6-11 {#example-6-11}
:::{.callout-warning}
A researcher claims that the distribution of technology inventories at warehouses is approximately normal with μ = 110,000 and σ = 15,000 units. A warehouse is selected at random. Find the probability that it has **more than 132,500 units**.
:::

```{r e6-11-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mu <- 110000; sigma <- 15000

# P(X > 132500)
pnorm(132500, mean = mu, sd = sigma, lower.tail = FALSE)

# Equivalent z-score
z <- (132500 - mu) / sigma
z
pnorm(z, lower.tail = FALSE)
```

**Answer:** z = 1.50, P(X > 132,500) $\approx$ **0.0668** (about 6.68%).

## Example 6-12 {#example-6-12}
:::{.callout-warning}
The distribution of the duration of professional baseball games is **negatively skewed** with μ = 166 minutes and σ = 26 minutes. If a game is selected at random, find the probability that it lasts **less than 145 minutes**.
:::

:::{.callout-warning}
## Caution
This distribution is **not** normal — it is negatively skewed. Applying normal distribution methods is only appropriate when the distribution is approximately normal. Here the answer is an approximation.
:::

```{r e6-12-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mu <- 166; sigma <- 26
# Approximate probability (treating as normal)
z <- (145 - mu) / sigma
z
pnorm(145, mean = mu, sd = sigma)
```

**Note:** $z \approx -0.81$; P(X < 145) $\approx$ **0.2090** (approximately 20.9%). Since the data is skewed, this is only a rough estimate.

# Section 6-3: The Central Limit Theorem {background-color="#2c4a6e"}

## The Central Limit Theorem

:::{.callout-note}
## The Central Limit Theorem (CLT)
As sample size **n** increases, the sampling distribution of the sample mean $\bar{X}$ approaches a **normal distribution**, regardless of the shape of the population distribution, provided n is sufficiently large (typically **$n \geq 30$**).

$$\mu_{\bar{X}} = \mu \qquad \sigma_{\bar{X}} = \frac{\sigma}{\sqrt{n}}$$
:::

The **standard error** $\sigma_{\bar{X}} = \sigma/\sqrt{n}$ measures how much sample means vary from sample to sample.

## Applying the CLT

:::{.callout-important}
## z Score for Sample Means
$$z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}}$$
:::

**In R:** Use `pnorm(xbar, mean = μ, sd = σ/sqrt(n))`

The CLT explains why the normal distribution is so central to statistics:

- Survey results, poll averages, sample statistics all tend toward normality
- Enables inference even when the population shape is unknown

---

```{r e6-12-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# CLT demonstration: sampling from a skewed distribution
set.seed(42)
pop <- rexp(10000, rate = 0.5)  # skewed population
sample_means <- replicate(1000, mean(sample(pop, 30)))
hist(sample_means, breaks = 30, col = "#2c4a6e",
     main = "Sampling Distribution of X-bar (n=30)",
     xlab = "Sample Mean", freq = FALSE)
curve(dnorm(x, mean(pop), sd(pop)/sqrt(30)), add = TRUE,
      col = "red", lwd = 2)
```

## Example 6-13 {#example-6-13}

:::{.callout-warning}
## Children's TV Viewing

Children in the United States watch an average of μ = 25 hours of TV per week, with σ = 3 hours. A random sample of n = 20 children is selected. Find the probability that the sample mean exceeds 26.3 hours.
:::

## Example 6-13

:::{.callout-tip}
## Solution

**Solve**

The average age of vehicles registered in the United States is μ = 96 months, with σ = 16 months. Assume the distribution is approximately normal. A random sample of n = 36 vehicles is taken. Find P(90 < $\bar{X}$ < 100).

:::

```{r e6-13-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mu <- 96; sigma <- 16; n <- 36
se <- sigma / sqrt(n)
se

# P(90 < X-bar < 100)
pnorm(100, mean = mu, sd = se) - pnorm(90, mean = mu, sd = se)
```

**Answer:** P(90 < $\bar{X}$ < 100) $\approx$ **0.9211** (92.11%)

The sample mean is very likely to fall in this range because the standard error is only `r round(16/sqrt(36), 2)` months.

## Example 6-15 {#example-6-15}
:::{.callout-warning}
The average annual meat consumption per person is μ = 218.4 pounds, σ = 25 pounds.

**(a)** Find P(X < 224) for a **single individual**.
**(b)** Find P($\bar{X}$ < 224) for a **sample of n = 40** individuals.
:::

```{r e6-15-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mu <- 218.4; sigma <- 25; n <- 40

# (a) Individual: P(X < 224)
pnorm(224, mean = mu, sd = sigma)

# (b) Sample mean: P(X-bar < 224), n = 40
se <- sigma / sqrt(n)
pnorm(224, mean = mu, sd = se)
```

**Answers:** (a) $\approx$ **58.71%**; (b) $\approx$ **92.22%**

:::{.callout-important}
## Key Insight
The probability is much higher for the sample mean because the standard error (σ/√n) is smaller than σ — sample means are less variable than individual values.
:::

# Section 6-4: The Normal Approximation to the Binomial Distribution {background-color="#2c4a6e"}

## Why Use the Normal Approximation?

The binomial distribution is exact but computationally intensive for large n. The normal distribution provides a convenient approximation.

:::{.callout-note}
## When to Use the Normal Approximation
Use the normal approximation to the binomial when:
$$np \geq 5 \quad \text{and} \quad nq \geq 5$$
where q = 1 − p.

The approximating normal distribution has:
$$\mu = np \qquad \sigma = \sqrt{npq}$$
:::

## Correction for Continuity

:::{.callout-important}
## Correction for Continuity
Since we are approximating a **discrete** distribution with a **continuous** one, add or subtract **0.5** from X:

| Binomial Expression | Normal Approximation |
|---------------------|----------------------|
| P(X = a) | P(a − 0.5 < X < a + 0.5) |
| $P(X \geq a)$ | P(X > a − 0.5) |
| P(X > a) | P(X > a + 0.5) |
| $P(X \leq a)$ | P(X < a + 0.5) |
| P(X < a) | P(X < a − 0.5) |
:::

## Example 6-16 {#example-6-16}
:::{.callout-warning}
A survey found that 6% of American drivers read while driving. If 300 drivers are selected at random, find the approximate probability that **exactly 25** read while driving.
:::
---

**Check:** n = 300, p = 0.06, q = 0.94

- np = 300(0.06) = 18 $\geq 5$ ✓

- nq = 300(0.94) = 282 $\geq 5$ ✓

**Parameters:** μ = np = 18, σ = $\sqrt{npq}$ = $\sqrt{300 \times 0.06 \times 0.94}$

```{r e6-16-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 300; p <- 0.06; q <- 1 - p
mu <- n * p; sigma <- sqrt(n * p * q)
cat("mu =", mu, "  sigma =", round(sigma, 4), "\n")

# Normal approximation: P(X = 25) ≈ P(24.5 < X < 25.5)
pnorm(25.5, mean = mu, sd = sigma) - pnorm(24.5, mean = mu, sd = sigma)

# Exact binomial
dbinom(25, n, p)
```

**Answer (normal approx.):** $\approx$ **0.0227** (2.27%); Exact binomial: $\approx 0.0220.$

## Example 6-17 {#example-6-17}

:::{.callout-warning}
## Widowed Bowlers

Researchers found that 10% of people who bowl are widowed. If 200 bowlers are selected, find the probability that **10 or more** are widowed.
:::

## Example 6-17

:::{.callout-tip}
## Solution

**Solve**

In baseball, a .320 batting average means a player gets a hit 32% of the time. In 100 at-bats, find the probability of getting **26 or fewer** hits.

:::

**Check:** np = 32 $\geq 5$, nq = 68 $\geq 5$ ✓

**Parameters:** μ = 32, σ = $\sqrt{100 \times 0.32 \times 0.68}$

```{r e6-17-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 100; p <- 0.32; q <- 0.68
mu <- n * p; sigma <- sqrt(n * p * q)
cat("mu =", mu, "  sigma =", round(sigma, 4), "\n")

# P(X <= 26) with continuity correction: P(X < 26.5)
pnorm(26.5, mean = mu, sd = sigma)

# Exact binomial
pbinom(26, n, p)
```

**Answer (normal approx.):** $\approx$ **0.119** (11.9%); Exact: $\approx 0.116.$

## Example 6-19 {#example-6-19}
:::{.callout-warning}
A coin is flipped 10 times. Find the probability of getting **exactly 6 heads** using:

**(a)** The binomial distribution

**(b)** The normal approximation
:::

:::{.callout-warning}
## Caution
Here np = 5, nq = 5 — these are exactly at the minimum threshold. The normal approximation may be less accurate for small n.
:::
---

```{r e6-19-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 10; p <- 0.5; q <- 0.5
mu <- n * p; sigma <- sqrt(n * p * q)
cat("mu =", mu, "  sigma =", sigma, "\n")

# (a) Exact binomial
dbinom(6, n, p)

# (b) Normal approximation with continuity correction
pnorm(6.5, mean = mu, sd = sigma) - pnorm(5.5, mean = mu, sd = sigma)
```

**Comparison:** Exact = **0.2051**; Normal approx. = **0.2034**

The approximation is quite close even for this borderline case.

## Chapter 6 Summary {.smaller}

| Topic | Formula | R Function |
|-------|---------|------------|
| Standard normal area (left) | P(Z < z) | `pnorm(z)` |
| Standard normal area (right) | 1 − P(Z < z) | `pnorm(z, lower.tail=FALSE)` |
| Find z from area | — | `qnorm(p)` |
| Non-standard normal | z = (X − μ)/σ | `pnorm(x, mean, sd)` |
| Sample mean probability | z = (X̄ − μ)/(σ/√n) | `pnorm(xbar, mean, sd/sqrt(n))` |
| Normal approx. to binomial | μ=np, σ=$\sqrt{npq}$ | ±0.5 correction |

## Key Takeaways

1. The normal distribution is fully described by μ and σ
2. Convert to z scores to use the standard normal table
3. The CLT guarantees that sample means are approximately normal for $n \geq 30$
4. Use normal approximation to binomial when np $\geq 5$ **and** nq $\geq 5$
5. Always apply the **correction for continuity** in the normal approximation

## Acknowledgement
-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2011). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
