---
title: "Statistics"
subtitle: "Chapter 6: The Normal Distribution"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
language:
  callout-warning-title: "Problem"
  callout-important-title: "Key point"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch6-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_minimal())
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview
Chapter 6 introduces the **normal distribution** and its role in statistical inference.

| Section | Topics |
|---------|--------|
| 6-1 | Normal and standard normal distributions; areas under the curve |
| 6-2 | Applications; data values from given probabilities; normality |
| 6-3 | Central limit theorem; sample means; finite population correction |
| 6-4 | Normal approximation to the binomial distribution |

## Chapter Objectives

:::{.callout-important}
## After completing this chapter, you should be able to

1. Identify the properties of a normal distribution.
2. Identify distributions as symmetric or skewed.
3. Find the area under the standard normal distribution, given various $z$ values.
4. Find probabilities for a normally distributed variable by transforming it into a standard normal variable.
5. Find specific data values for given percentages, using the standard normal distribution.
6. Use the central limit theorem to solve problems involving sample means for large samples.
7. Use the normal approximation to compute probabilities for a binomial variable.
:::

# Section 6-1: Normal Distributions {background-color="#2c4a6e"}

## Introduction

Random variables can be **discrete** or **continuous**. Chapter 5 dealt with discrete variables; a **continuous** variable can assume all values between any two given values — heights of adult men, body temperatures of rats, cholesterol levels of adults.

Many continuous variables have bell-shaped distributions and are called **approximately normally distributed** variables. If a researcher measures the heights of 100 adult women and draws a histogram, then keeps increasing the sample size while decreasing the class width, the histogram approaches a smooth curve called a **normal distribution curve** (Figure 6-1). It is also called a **bell curve** or a **Gaussian distribution curve**, after Carl Friedrich Gauss (1777–1855).

No variable fits a normal distribution perfectly, because a normal distribution is a **theoretical** distribution. It is nevertheless useful because the deviations from it are very small.

---

## Introduction

**Figure 6-1** Histograms and normal model for the distribution of heights of adult women

```{r s6-1-1a, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
set.seed(6)
heights <- rnorm(20000, mean = 163, sd = 6.5)

ggplot(data.frame(height = heights[1:100]), aes(height)) +
  geom_histogram(binwidth = 5, color = "white") +
  labs(title = "(a) n = 100", x = "Height (cm)", y = "Frequency")
```

---

## Introduction

```{r s6-1-1b, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
ggplot(data.frame(height = heights[1:1000]), aes(height)) +
  geom_histogram(binwidth = 3, color = "white") +
  labs(title = "(b) larger n", x = "Height (cm)", y = "Frequency")
```

---

## Introduction

```{r s6-1-1c, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
ggplot(data.frame(height = heights), aes(height)) +
  geom_histogram(binwidth = 1, color = "white") +
  labs(title = "(c) larger n still", x = "Height (cm)", y = "Frequency")
```

---

## Introduction

```{r s6-1-1d, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
x <- seq(140, 186, by = 0.1)
curve <- data.frame(x, density = dnorm(x, mean = 163, sd = 6.5))

ggplot(curve, aes(x, density)) +
  geom_line() +
  labs(title = "(d) Normal distribution", x = "Height (cm)", y = "Density")
```

## Normal Distributions

:::{.callout-note}
## Normal Distribution

If a random variable has a probability distribution whose graph is **continuous**, **bell-shaped**, and **symmetric**, it is called a **normal distribution**. The graph is called a **normal distribution curve**.
:::

The mathematical equation for a normal distribution is

$$y = \frac{e^{-(X-\mu)^2/(2\sigma^2)}}{\sigma\sqrt{2\pi}}$$

where $e \approx 2.718$, $\pi \approx 3.14$, $\mu$ is the population mean and $\sigma$ is the population standard deviation. In applied statistics, tables or technology are used instead of the equation.

---

## Shapes of Normal Distributions

The shape and position of a normal distribution curve depend on two parameters: the **mean** and the **standard deviation** (Figure 6-3).

```{r s6-1-3, echo=T, warning=FALSE, out.width="70%", fig.align='center', fig.height=2.4}
x <- seq(-8, 10, by = 0.01)
curves <- data.frame(x, mu0_sd1 = dnorm(x, 0, 1), mu0_sd2 = dnorm(x, 0, 2),
                     mu2_sd2 = dnorm(x, 2, 2))

ggplot(curves, aes(x)) +
  geom_line(aes(y = mu0_sd1)) +
  geom_line(aes(y = mu0_sd2), linetype = "dashed") +
  geom_line(aes(y = mu2_sd2), linetype = "dotted") +
  labs(title = "Figure 6-3: shapes of normal distributions", x = "x", y = "Density")
```

The solid line has $\mu = 0$, $\sigma = 1$, the dashed line $\mu = 0$, $\sigma = 2$, and the dotted line $\mu = 2$, $\sigma = 2$. Solid and dashed have the same mean; dashed and dotted have the same standard deviation.

## Properties of a Normal Distribution

:::{.callout-important}
## Summary of the Properties of the Theoretical Normal Distribution

1. A normal distribution curve is **bell-shaped**.
2. The **mean, median, and mode are equal** and are located at the center of the distribution.
3. A normal distribution curve is **unimodal** (it has only one mode).
4. The curve is **symmetric about the mean**; its shape is the same on both sides of a vertical line through the center.
5. The curve is **continuous**: there are no gaps or holes. For each value of $X$ there is a corresponding value of $Y$.
6. The curve **never touches the $x$ axis**, no matter how far it extends in either direction — but it gets increasingly close.
7. The **total area** under a normal distribution curve is equal to $1.00$, or $100\%$.
8. The area within $1$ standard deviation of the mean is about $0.68$, within $2$ standard deviations about $0.95$, and within $3$ standard deviations about $0.997$.
:::

---

## Areas Under a Normal Distribution Curve

**Figure 6-4** — the areas in item 8 follow the **empirical rule** of Section 3-2.

```{r s6-1-4, echo=T, warning=FALSE, out.width="80%", fig.align='center', fig.height=3}
x <- seq(-4, 4, by = 0.01)
curve <- data.frame(x, density = dnorm(x))

ggplot(curve, aes(x, density)) +
  geom_line() +
  geom_vline(xintercept = -3:3) +
  labs(title = "About 68%, 95% and 99.7% within 1, 2 and 3 sigma",
       x = "z", y = "Density")
```

## Symmetric and Skewed Distributions

:::{.callout-note}
## Symmetric and Skewed Distributions

- When the data values are evenly distributed about the mean, the distribution is a **symmetric distribution**. A normal distribution is symmetric.
- When the majority of the data values fall to the right of the mean, the distribution is **negatively or left-skewed**: the mean is to the left of the median, and both are to the left of the mode.
- When the majority of the data values fall to the left of the mean, the distribution is **positively or right-skewed**: the mean falls to the right of the median, and both fall to the right of the mode.
:::

The "tail" of the curve indicates the direction of the skewness (right is positive, left is negative).

---

## Symmetric and Skewed Distributions

**Figure 6-5** Normal and skewed distributions

```{r s6-1-5, echo=T, warning=FALSE, out.width="90%", fig.align='center', fig.height=3}
x <- seq(0, 10, by = 0.1)
shapes <- data.frame(
  shape   = rep(c("(a) Normal", "(b) Negatively skewed", "(c) Positively skewed"),
                each = length(x)),
  x       = c(x, x, x),
  density = c(dnorm(x, 5, 1.5), dgamma(10 - x, shape = 3), dgamma(x, shape = 3))
)

ggplot(shapes, aes(x, density)) +
  geom_line() +
  facet_wrap(~ shape) +
  labs(y = "Density")
```

## The Standard Normal Distribution

Each normally distributed variable has its own mean and standard deviation, so a separate table of areas would be needed for every variable. To simplify this, statisticians use the **standard normal distribution**.

:::{.callout-note}
## Standard Normal Distribution

The **standard normal distribution** is a normal distribution with a **mean of 0** and a **standard deviation of 1**.
:::

The equation for the standard normal distribution is

$$y = \frac{e^{-z^2/2}}{\sqrt{2\pi}}$$

---

## The Standard Score

:::{.callout-important}
## Formula for the $z$ Score (Standard Score)

$$z = \frac{\text{value} - \text{mean}}{\text{standard deviation}} \qquad\text{or}\qquad z = \frac{X - \mu}{\sigma}$$
:::

- This is the same formula used in Section 3-3.
- Once the $X$ values are transformed, they are called **$z$ values**. The **$z$ value** (or **$z$ score**) is the number of standard deviations a particular $X$ value is away from the mean.
- **Table A-4** in Appendix A gives the area (to four decimal places) under the standard normal curve for any $z$ value from $-3.49$ to $3.49$.

---

## The Standard Normal Distribution

**Figure 6-6** Standard normal distribution

```{r s6-1-6, echo=T, warning=FALSE, out.width="75%", fig.align='center'}
ggplot(curve, aes(x, density)) +
  geom_line() +
  labs(title = "Standard normal distribution (mean 0, standard deviation 1)",
       x = "z", y = "Density")
```

## Finding Areas Under the Standard Normal Distribution Curve

A **two-step process** is recommended:

**Step 1** Draw the normal distribution curve and shade the area.

**Step 2** Find the appropriate figure in the Procedure Table and follow the directions given.

:::{.callout-important}
## Procedure Table — Finding the Area Under the Standard Normal Distribution Curve

1. **To the left of any $z$ value:** look up the $z$ value in the table and use the area given.
2. **To the right of any $z$ value:** look up the $z$ value and subtract the area from $1$.
3. **Between any two $z$ values:** look up both $z$ values and subtract the corresponding areas.
:::

There are three basic types of problems, and all three are summarized in the Procedure Table.

## Standard Normal Areas in R

This course uses R instead of Table A-4. The three cases of the Procedure Table become three R expressions.

:::{.callout-important}
## Key R Functions

- `pnorm(z)` — area to the **left** of $z$
- `1 - pnorm(z)` — area to the **right** of $z$
- `pnorm(z2) - pnorm(z1)` — area **between** $z_1$ and $z_2$
- `qnorm(p)` — the $z$ value whose left-tail area is $p$ (the inverse of `pnorm`)
:::

```{r s6-1-7, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
pnorm(1.58)   # the Table A-4 entry for z = 1.58
```

## Example 6-1 {#example-6-1}

:::{.callout-warning}
## Area to the Left of a $z$ Value

Find the area under the standard normal distribution curve to the left of $z = 1.72$.
:::

---

## Example 6-1

:::{.callout-tip}
## Solution

**Step 1** Draw the figure and shade the desired area (Figure 6-8).

**Step 2** This is the first case in the Procedure Table, so use the area directly. The area to the left of $z = 1.72$ is $0.9573$. Hence, $95.73\%$ of the area is to the left of $z = 1.72$.

```{r e6-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
pnorm(1.72)
```
:::

---

## Example 6-1

:::{.callout-tip}
## Solution

**Draw figure**

```{r e6-1-2, echo=T, warning=FALSE, out.width="55%", fig.align='center'}
ggplot(curve, aes(x, density)) +
  geom_line() +
  geom_area(data = filter(curve, x <= 1.72)) +
  labs(title = "Area to the left of z = 1.72", x = "z", y = "Density")
```
:::

## Example 6-2 {#example-6-2}

:::{.callout-warning}
## Area to the Right of a $z$ Value

Find the area under the standard normal distribution curve to the right of $z = -0.83$.
:::

---

## Example 6-2

:::{.callout-tip}
## Solution

**Step 1** Draw the figure and shade the desired area (Figure 6-9).

**Step 2** This is the second case. The area for $z = -0.83$ is $0.2033$. Subtract it from $1.0000$: $1.0000 - 0.2033 = 0.7967$. Hence, $79.67\%$ of the area lies to the right of $z = -0.83$.

```{r e6-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
pnorm(-0.83)
1 - pnorm(-0.83)
```
:::

---

## Example 6-2

:::{.callout-tip}
## Solution

**Draw figure**

```{r e6-2-2, echo=T, warning=FALSE, out.width="55%", fig.align='center'}
ggplot(curve, aes(x, density)) +
  geom_line() +
  geom_area(data = filter(curve, x >= -0.83)) +
  labs(title = "Area to the right of z = -0.83", x = "z", y = "Density")
```
:::

## Example 6-3 {#example-6-3}

:::{.callout-warning}
## Area Between Two $z$ Values

Find the area under the standard normal distribution curve between $z = 1.21$ and $z = -1.54$.
:::

---

## Example 6-3

:::{.callout-tip}
## Solution

**Step 1** Draw the figure (Figure 6-10). The desired area lies between the two $z$ values.

**Step 2** This is the third case: look up both areas and subtract the smaller from the larger. *(Do not subtract the $z$ values.)* The area for $z = 1.21$ is $0.8869$ and the area for $z = -1.54$ is $0.0618$, so the area between them is $0.8869 - 0.0618 = 0.8251$, or $82.51\%$.

```{r e6-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
pnorm(1.21) - pnorm(-1.54)
```
:::

---

## Example 6-3

:::{.callout-tip}
## Solution

**Draw figure**

```{r e6-3-2, echo=T, warning=FALSE, out.width="55%", fig.align='center'}
ggplot(curve, aes(x, density)) +
  geom_line() +
  geom_area(data = filter(curve, x >= -1.54, x <= 1.21)) +
  labs(title = "Area between z = -1.54 and z = 1.21", x = "z", y = "Density")
```
:::

## A Normal Distribution Curve as a Probability Distribution Curve

The area under the standard normal distribution curve can also be thought of as a **probability**, or as the **proportion of the population** with a given characteristic. If it were possible to select a $z$ value at random, the probability of choosing one between $0$ and $2.00$ would be the same as the area under the curve between $0$ and $2.00$, namely $0.4772$.

For probabilities, a special notation is used: the probability of a $z$ value between $0$ and $2.32$ is written $P(0 < z < 2.32)$.

:::{.callout-warning}
## Caution

In a **continuous** distribution the probability of any exact $z$ value is $0$, since the area would be a vertical line, which in theory has no area. Hence $P(a \leq z \leq b) = P(a < z < b)$.
:::

## Example 6-4 {#example-6-4}

:::{.callout-warning}
## Probabilities for the Standard Normal Distribution

Find the probability for each. (Assume this is a standard normal distribution.)

**(a)** $P(0 < z < 2.14)$ **(b)** $P(z < 1.35)$ **(c)** $P(z > 2.27)$
:::

---

## Example 6-4

:::{.callout-tip}
## Solution

**(a)** Find the area between $z = 0$ and $z = 2.14$. The area for $z = 2.14$ is $0.9838$ and the area for $z = 0$ is $0.5000$: $0.9838 - 0.5000 = 0.4838$, or $48.38\%$.

**(b)** Find the area to the left of $z = 1.35$. It is $0.9115$, so the probability is $0.9115$, or $91.15\%$.

**(c)** Find the area to the right of $z = 2.27$. The area for $z = 2.27$ is $0.9884$, so $1.0000 - 0.9884 = 0.0116$, or $1.16\%$.

```{r e6-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
pnorm(2.14) - pnorm(0)
pnorm(1.35)
1 - pnorm(2.27)
```
:::

## Example 6-5 {#example-6-5}

:::{.callout-warning}
## Finding a $z$ Value for a Given Area

Find the $z$ value such that the area under the standard normal distribution curve between $0$ and the $z$ value is $0.3078$.
:::

---

## Example 6-5

:::{.callout-tip}
## Solution

Draw the figure (Figure 6-14). Since Table A-4 is cumulative, add $0.5000$ to the given area of $0.3078$ to get the cumulative area $0.8078$. Looking up $0.8078$ in Table A-4 gives a row value of $0.8$ and a column value of $0.07$, so $z = 0.87$.

```{r e6-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
qnorm(0.5000 + 0.3078)
```

If the exact area cannot be found in the table, use the closest value: for a cumulative area of $0.8080$ the closest table area is $0.8078$, which again gives $z = 0.87$.
:::

# Section 6-2: Applications of the Normal Distribution {background-color="#2c4a6e"}

## Applications of the Normal Distribution

The standard normal distribution curve can be used to solve a wide variety of practical problems. The **only requirement** is that the variable be **normally or approximately normally distributed**. To solve such problems, transform the original variable to a standard normal variable:

$$z = \frac{\text{value} - \text{mean}}{\text{standard deviation}} \qquad\text{or}\qquad z = \frac{X - \mu}{\sigma}$$

:::{.callout-important}
## Procedure Table — Finding the Area Under Any Normal Curve

**Step 1** Draw a normal curve and shade the desired area.

**Step 2** Convert the values of $X$ to $z$ values, using the formula $z = \dfrac{X - \mu}{\sigma}$.

**Step 3** Find the corresponding area, using a table, calculator, or software.
:::

In R the conversion is optional: `pnorm(x, mean = mu, sd = sigma)` works directly.

## Example 6-6 {#example-6-6}

:::{.callout-warning}
## Food Delivery Times

On a food-delivery platform operating in Taipei, the time from placing an order to delivery at the door averages $27$ minutes. Assume the variable is normally distributed and has a standard deviation of $4$ minutes. Find the percentage of deliveries that arrive in less than $32$ minutes. *(hypothetical data)*
:::

---

## Example 6-6

:::{.callout-tip}
## Solution

**Step 1** Draw a normal curve and shade the desired area (Figure 6-18).

**Step 2** $z = \dfrac{X - \mu}{\sigma} = \dfrac{32 - 27}{4} = \dfrac{5}{4} = 1.25$

**Step 3** The area to the left of $z = 1.25$ is $0.8944$. Therefore $89.44\%$ of deliveries arrive in less than $32$ minutes.

```{r e6-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
pnorm(32, mean = 27, sd = 4)
```
:::

---

## Example 6-6

:::{.callout-tip}
## Solution

**Draw figure**

```{r e6-6-2, echo=T, warning=FALSE, out.width="65%", fig.align='center', fig.height=4}
x <- seq(11, 43, by = 0.1)
curve <- data.frame(x, density = dnorm(x, mean = 27, sd = 4))

ggplot(curve, aes(x, density)) +
  geom_line() +
  geom_area(data = filter(curve, x <= 32)) +
  labs(title = "Delivery time: P(X < 32)", x = "Minutes", y = "Density")
```
:::

## Example 6-7 {#example-6-7}

:::{.callout-warning}
## Order Values on an E-Commerce Platform

Completed orders on a Taiwanese e-commerce platform have an average value of $1{,}200$ NT dollars. Assume the variable is approximately normally distributed and the standard deviation is $150$ NT dollars. If one completed order is selected at random, find the probability that its value is

**(a)** between $1{,}050$ and $1{,}425$ NT dollars;  **(b)** more than $1{,}380$ NT dollars. *(hypothetical data)*
:::

---

## Example 6-7

:::{.callout-tip}
## Solution (a)

**Step 1** Draw a normal curve and shade the desired area (Figure 6-20).

**Step 2** $z_1 = \dfrac{1050 - 1200}{150} = -1.0$ and $z_2 = \dfrac{1425 - 1200}{150} = 1.5$

**Step 3** The area to the left of $z_2$ is $0.9332$ and the area to the left of $z_1$ is $0.1587$, so the area between them is $0.9332 - 0.1587 = 0.7745$. The probability is $77.45\%$.

```{r e6-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
pnorm(1425, mean = 1200, sd = 150) - pnorm(1050, mean = 1200, sd = 150)
```
:::

---

## Example 6-7

:::{.callout-tip}
## Solution (b)

**Step 1** Draw a normal curve and shade the desired area (Figure 6-22).

**Step 2** $z = \dfrac{1380 - 1200}{150} = \dfrac{180}{150} = 1.2$

**Step 3** The area to the left of $z = 1.2$ is $0.8849$, so the area to the right is $1.0000 - 0.8849 = 0.1151$. The probability that an order is worth more than $1{,}380$ NT dollars is $0.1151$, or $11.51\%$.

```{r e6-7-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
1 - pnorm(1380, mean = 1200, sd = 150)
```
:::

---

## Example 6-7

:::{.callout-tip}
## Solution

**Draw figure (a)**

```{r e6-7-3, echo=T, warning=FALSE, out.width="65%", fig.align='center', fig.height=4}
x <- seq(600, 1800, by = 1)
curve <- data.frame(x, density = dnorm(x, mean = 1200, sd = 150))

ggplot(curve, aes(x, density)) +
  geom_line() +
  geom_area(data = filter(curve, x >= 1050, x <= 1425)) +
  labs(title = "(a) P(1050 < X < 1425) = 0.7745", x = "NT dollars", y = "Density")
```
:::

---

## Example 6-7

:::{.callout-tip}
## Solution

**Draw figure (b)**

```{r e6-7-4, echo=T, warning=FALSE, out.width="65%", fig.align='center', fig.height=4}
ggplot(curve, aes(x, density)) +
  geom_line() +
  geom_area(data = filter(curve, x >= 1380)) +
  labs(title = "(b) P(X > 1380) = 0.1151", x = "NT dollars", y = "Density")
```
:::

## Example 6-8 {#example-6-8}

:::{.callout-warning}
## Order Picking Time

At a logistics centre in Taoyuan, the time needed to pick and pack one order averages $6.4$ minutes. Assume the variable is approximately normally distributed and the population standard deviation is $0.8$ minute. If $60$ orders are randomly selected, approximately how many will be picked and packed in less than $5.8$ minutes? *(hypothetical data)*
:::

---

## Example 6-8

:::{.callout-tip}
## Solution

**Step 1** Draw a normal curve and shade the desired area (Figure 6-24).

**Step 2** $z = \dfrac{5.8 - 6.4}{0.8} = -0.75$

**Step 3** The area corresponding to $z = -0.75$ is $0.2266$.

To find the number of orders, multiply $60 \times 0.2266 = 13.6$ and round to $14$. Approximately **14** of the $60$ randomly selected orders are picked and packed in less than $5.8$ minutes.

```{r e6-8-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
area <- pnorm(5.8, mean = 6.4, sd = 0.8)

area
60 * area
```

*Note:* for problems using percentages, write the percentage as a decimal before multiplying, and round the answer to the nearest whole number.
:::

## Finding Data Values Given Specific Probabilities

A normal distribution can also be used to find specific **data values** for given percentages. Starting from $z = \dfrac{X-\mu}{\sigma}$ and solving for $X$ gives:

:::{.callout-important}
## Formula for Finding the Value of a Normal Variable $X$

$$X = z \cdot \sigma + \mu$$
:::

:::{.callout-important}
## Procedure Table — Finding Data Values for Specific Probabilities

**Step 1** Draw a normal curve and shade the desired area that represents the probability, proportion, or percentile.

**Step 2** Find the $z$ value from the table that corresponds to the desired area.

**Step 3** Calculate the $X$ value by using the formula $X = z\sigma + \mu$.
:::

## Example 6-9 {#example-6-9}

:::{.callout-warning}
## Qualifying for a Rider Bonus

A delivery platform pays a quarterly bonus to the riders in the **top $10\%$** by number of orders completed per month. Assume the number of orders completed is normally distributed with a mean of $480$ and a standard deviation of $60$. Find the lowest number of completed orders needed to qualify. *(hypothetical data)*
:::

---

## Example 6-9

:::{.callout-tip}
## Solution

**Step 1** Work backward. Subtract $0.1000$ from $1.0000$ to get the area to the left of $X$: $1.0000 - 0.1000 = 0.9000$ (Figure 6-26).

**Step 2** Look up $0.9000$ in the area portion of Table A-4. The table rounds the answer to $z = 1.28$; computed exactly, $z = 1.2816$.

**Step 3** $X = z \cdot \sigma + \mu = 1.2816(60) + 480 = 556.89$.

Because a rider cannot complete a fraction of an order, **557** completed orders should be used as the cutoff: anybody completing $557$ or more qualifies for the bonus.

```{r e6-9-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
z <- qnorm(0.90)

z
z * 60 + 480
```
:::

## Example 6-10 {#example-6-10}

:::{.callout-warning}
## Daily Passengers at an MRT Station

A transport planner wishes to study the middle $60\%$ of weekdays at a Taipei MRT station. Assume the number of passengers entering the station on a weekday is normally distributed with a mean of $48$ thousand and a standard deviation of $5$ thousand.

**(a)** What is the passenger count for the lowest $20\%$ of weekdays?

**(b)** Find the lower and upper passenger counts for the middle $60\%$ of weekdays. *(hypothetical data)*
:::

---

## Example 6-10

:::{.callout-tip}
## Solution (a)

**Step 1** Draw the curve and shade the lower $20\%$ (Figure 6-28).

**Step 2** The $z$ value with area $0.2000$ to its left is $z = -0.8416$ (Table A-4 rounds it to $-0.84$).

**Step 3** $X = z\sigma + \mu = (-0.8416)(5) + 48 = 43.79$

The quietest $20\%$ of weekdays have fewer than $43.79$ thousand passengers.

```{r e6-10-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
z <- qnorm(0.20)

z
z * 5 + 48
```
:::

---

## Example 6-10

:::{.callout-tip}
## Solution (b)

**Step 1** Two values are needed, one above and one below the mean (Figure 6-29).

**Step 2** To get the area to the left of the positive $z$ value, add $0.5000 + 0.3000 = 0.8000$. The $z$ value for a left-tail area of $0.8000$ is $0.8416$, and the $z$ value for $0.2000$ is $-0.8416$.

**Step 3** $X_1 = (-0.8416)(5) + 48 = 43.79$ and $X_2 = (0.8416)(5) + 48 = 52.21$.

The middle $60\%$ of weekdays have between $43.79$ and $52.21$ thousand passengers.

```{r e6-10-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
qnorm(0.20) * 5 + 48
qnorm(0.80) * 5 + 48
```
:::

## Determining Normality

A bell-shaped distribution is only one of many shapes a distribution can assume, but it is very important since many statistical methods require the distribution of values to be normally or approximately normally shaped. Statisticians check normality in several ways.

:::{.callout-important}
## Checking a Data Set for Normality

1. **Histogram.** Draw a histogram and check its shape. If it is not approximately bell-shaped, the data are not normally distributed.
2. **Pearson coefficient (PC) of skewness**, also called **Pearson's index of skewness**:
$$PC = \frac{3(\bar{X} - \text{median})}{s}$$
If the index is **greater than or equal to $+1$** or **less than or equal to $-1$**, the data are significantly skewed.
3. **Outliers.** Check for outliers with the method of Chapter 3: a value more than $1.5(IQR)$ below $Q_1$ or above $Q_3$. Even one or two outliers can have a big effect on normality.
:::

## Example 6-11 {#example-6-11}

:::{.callout-warning}
## Supplier Lead Times

A trading company recorded the number of days of finished-goods inventory held by each of $18$ supplier factories. Determine if the data are approximately normally distributed. *(hypothetical data)*

```
  12    21    26    33    35    38    42    45    47
  49    52    56    59    63    68    74    81    93
```
:::

---

## Example 6-11

:::{.callout-tip}
## Solution

**Step 1** Construct a frequency distribution and draw a histogram (Figure 6-30).

```{r e6-11-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
lead_days <- c(12, 21, 26, 33, 35, 38, 42, 45, 47,
               49, 52, 56, 59, 63, 68, 74, 81, 93)
lead_breaks <- seq(9.5, 99.5, by = 15)

table(cut(lead_days, lead_breaks))
```
:::

---

## Example 6-11

:::{.callout-tip}
## Solution

**Draw figure**

```{r e6-11-2, echo=T, warning=FALSE, out.width="55%", fig.align='center'}
ggplot(data.frame(lead_days), aes(lead_days)) +
  geom_histogram(breaks = lead_breaks, color = "white") +
  labs(title = "Histogram for Example 6-11", x = "Days", y = "Frequency")
```

The histogram is approximately bell-shaped, so the distribution appears approximately normal.
:::

---

## Example 6-11

:::{.callout-tip}
## Solution

**Step 2** Check for skewness: $\bar{X} = 49.67$, median $= 48$, $s = 21.17$, so
$$PC = \frac{3(49.67 - 48)}{21.17} = 0.236$$

```{r e6-11-3, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mean(lead_days)
median(lead_days)
sd(lead_days)
3 * (mean(lead_days) - median(lead_days)) / sd(lead_days)
```

$PC$ is not greater than $+1$ nor less than $-1$, so the distribution is **not significantly skewed**.
:::

---

## Example 6-11

:::{.callout-tip}
## Solution

**Step 3** Check for outliers: $Q_1 = 35.75$ and $Q_3 = 62$, so $IQR = 26.25$. An outlier would be below $35.75 - 1.5(26.25) = -3.62$ or above $62 + 1.5(26.25) = 101.38$. There are **no outliers**.

```{r e6-11-4, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
quantile(lead_days, c(0.25, 0.75))
IQR(lead_days)
35.75 - 1.5 * 26.25
62 + 1.5 * 26.25
```

Bell-shaped, not significantly skewed, no outliers: the distribution is **approximately normally distributed**.
:::

## Example 6-12 {#example-6-12}

:::{.callout-warning}
## Customers Served at a Night-Market Stall

The data shown are the numbers of customers served by a night-market stall on $17$ randomly selected evenings. Determine if the data are approximately normally distributed. *(hypothetical data)*

```
  142    95   131   152   108   145
  149    25   139   118   151   135
  125    60   147    78    42
```
:::

---

## Example 6-12

:::{.callout-tip}
## Solution

**Step 1** Construct a frequency distribution and draw a histogram (Figure 6-31).

```{r e6-12-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
customers <- c(142,  95, 131, 152, 108, 145,
               149,  25, 139, 118, 151, 135,
               125,  60, 147,  78,  42)
customer_breaks <- seq(24.5, 174.5, by = 25)

# dig.lab = 4 prints boundaries such as 124.5 in full
table(cut(customers, customer_breaks, dig.lab = 4))
```
:::

---

## Example 6-12

:::{.callout-tip}
## Solution

**Draw figure**

```{r e6-12-2, echo=T, warning=FALSE, out.width="55%", fig.align='center'}
ggplot(data.frame(customers), aes(customers)) +
  geom_histogram(breaks = customer_breaks, color = "white") +
  labs(title = "Histogram for Example 6-12", x = "Customers", y = "Frequency")
```

The histogram shows that the frequency distribution is **negatively skewed**.
:::

---

## Example 6-12

:::{.callout-tip}
## Solution

**Step 2** Check for skewness: $\bar{X} = 114.24$, median $= 131$, $s = 40.37$, so
$$PC = \frac{3(114.24 - 131)}{40.37} = -1.25$$

```{r e6-12-3, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mean(customers)
median(customers)
sd(customers)
3 * (mean(customers) - median(customers)) / sd(customers)
```

Since $PC$ is less than $-1$, the distribution is **significantly skewed to the left**.
:::

---

## Example 6-12

:::{.callout-tip}
## Solution

**Step 3** Check for outliers: $Q_1 = 95$, $Q_3 = 145$, $IQR = 50$. Any value below $95 - 1.5(50) = 20$ or above $145 + 1.5(50) = 220$ is an outlier. There are **no outliers**.

```{r e6-12-4, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
quantile(customers, c(0.25, 0.75))
IQR(customers)
95 - 1.5 * 50
145 + 1.5 * 50
```

In summary, the distribution is **significantly negatively skewed**.
:::

## Normal Quantile Plots in R

Another method used to check normality is to draw a **normal quantile plot**. **Quantiles** (sometimes called **fractiles**) are values that separate the data set into approximately equal groups. A normal quantile plot graphs the data values against the $z$ values of the corresponding quantiles. **If the points do not lie in an approximately straight line, normality can be rejected.**

```{r s6-2-1, echo=T, warning=FALSE, out.width="60%", fig.align='center', fig.height=3.5}
ggplot(data.frame(lead_days), aes(sample = lead_days)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Example 6-11: supplier lead times",
       x = "Theoretical quantiles (z)", y = "Sample quantiles")
```

---

## Normal Quantile Plots in R

```{r s6-2-2, echo=T, warning=FALSE, out.width="60%", fig.align='center', fig.height=3.5}
ggplot(data.frame(customers), aes(sample = customers)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Example 6-12: customers served",
       x = "Theoretical quantiles (z)", y = "Sample quantiles")
```

Other methods include normal probability graph paper, the chi-square goodness-of-fit test (Chapter 11), the Kolmogorov-Smirnov test and the Lilliefors test.

# Section 6-3: The Central Limit Theorem {background-color="#2c4a6e"}

## Distribution of Sample Means

In addition to knowing how individual data values vary about the mean for a population, statisticians want to know how the **means of samples** of the same size taken from the same population vary about the population mean.

Suppose an analyst selects a sample of $30$ online orders and finds the mean order value to be $1{,}180$ NT dollars, then selects a second sample with mean $1{,}215$ NT dollars, and continues for $100$ samples. The mean becomes a random variable, and the sample means $1{,}180, 1{,}215, 1{,}164, \ldots, 1{,}232$ constitute a **sampling distribution of sample means**.

:::{.callout-note}
## Sampling Distribution of Sample Means

A **sampling distribution of sample means** is a distribution using the means computed from **all possible random samples** of a specific size taken from a population.
:::

---

## Sampling Error

If the samples are randomly selected **with replacement**, the sample means will for the most part be somewhat different from the population mean $\mu$. These differences are caused by **sampling error**.

:::{.callout-note}
## Sampling Error

**Sampling error** is the difference between the sample measure and the corresponding population measure, due to the fact that the sample is not a perfect representation of the population.
:::

:::{.callout-important}
## Properties of the Distribution of Sample Means

1. The **mean of the sample means** will be the same as the population mean.
2. The **standard deviation of the sample means** will be smaller than the standard deviation of the population, and it will be equal to the population standard deviation divided by the square root of the sample size.
:::

## A 10-Point Quiz Illustration

A lecturer gave a $10$-point quiz to a small class of four students; the results were $3$, $5$, $7$ and $9$. Treat the four students as the population.

```{r s6-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
quiz_scores <- c(3, 5, 7, 9)

mean(quiz_scores)
sigma <- sqrt(sum((quiz_scores - 6)^2) / 4)
sigma
```

The graph of the original distribution is a **uniform distribution** (Figure 6-32): each score occurs once.

---

## A 10-Point Quiz Illustration

All samples of size $2$ taken **with replacement**, and the mean of each sample:

```{r s6-3-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
samples <- expand.grid(first = quiz_scores, second = quiz_scores)
sample_means <- (samples$first + samples$second) / 2

table(sample_means)
```

---

## A 10-Point Quiz Illustration

```{r s6-3-3, echo=T, warning=FALSE, out.width="60%", fig.align='center', fig.height=2.5}
ggplot(data.frame(sample_means), aes(sample_means)) +
  geom_bar() +
  labs(title = "Figure 6-33: distribution of sample means",
       x = "Sample mean", y = "Frequency")
```

```{r s6-3-4, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mean(sample_means)
sqrt(sum((sample_means - 6)^2) / 16)
sigma / sqrt(2)
```

The histogram of the sample means appears approximately normal, the mean of the sample means equals the population mean, and the standard deviation of the sample means equals $\sigma/\sqrt{n}$.

## The Standard Error of the Mean

:::{.callout-note}
## Standard Error of the Mean

If all possible samples of size $n$ are taken with replacement from the same population, the mean of the sample means, denoted $\mu_{\bar{X}}$, equals the population mean $\mu$; and the standard deviation of the sample means, denoted $\sigma_{\bar{X}}$, equals $\sigma/\sqrt{n}$. The standard deviation of the sample means is called the **standard error of the mean**.
:::

:::{.callout-important}
## Mean and Standard Error of the Sample Means

$$\mu_{\bar{X}} = \mu \qquad\qquad \sigma_{\bar{X}} = \frac{\sigma}{\sqrt{n}}$$
:::

## The Central Limit Theorem

A third property of the sampling distribution of sample means concerns the **shape** of the distribution, and is explained by the central limit theorem.

:::{.callout-note}
## The Central Limit Theorem

As the sample size $n$ **increases without limit**, the shape of the distribution of the sample means taken with replacement from a population with mean $\mu$ and standard deviation $\sigma$ will approach a **normal distribution**. This distribution will have a mean $\mu$ and a standard deviation $\sigma/\sqrt{n}$.
:::

:::{.callout-important}
## $z$ Value for the Central Limit Theorem

$$z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}}$$

The denominator must be adjusted because means are used instead of individual data values; it is the standard deviation of the sample means.
:::

---

## Two Things to Remember

:::{.callout-important}
## When the Central Limit Theorem Applies

1. When the **original variable is normally distributed**, the distribution of the sample means will be normally distributed **for any sample size $n$**.
2. When the distribution of the original variable is **not normal**, a sample size of **$30$ or more** is needed to use a normal distribution to approximate the distribution of the sample means. The larger the sample, the better the approximation.
:::

**In R:** use `pnorm(xbar, mean = mu, sd = sigma / sqrt(n))`.

---

## The Central Limit Theorem in R

```{r s6-3-5, echo=T, warning=FALSE, out.width="65%", fig.align='center', fig.height=3}
set.seed(42)
# rate = 0.5 gives a skewed population
simulation <- data.frame(
  size        = rep(c("(a) n = 2", "(b) n = 10", "(c) n = 30"), each = 2000),
  sample_mean = c(replicate(2000, mean(rexp(2, rate = 0.5))),
                  replicate(2000, mean(rexp(10, rate = 0.5))),
                  replicate(2000, mean(rexp(30, rate = 0.5))))
)

ggplot(simulation, aes(sample_mean)) +
  geom_histogram(bins = 40, color = "white") +
  facet_wrap(~ size, scales = "free") +
  labs(title = "Sample means approach normality as n increases",
       x = "Sample mean", y = "Frequency")
```

## Example 6-13 {#example-6-13}

:::{.callout-warning}
## Length of Customer-Service Calls

Calls handled by the customer-service centre of an online retailer last an average of $4.5$ minutes. Assume the variable is normally distributed and the standard deviation is $1.25$ minutes. If $25$ calls are randomly selected, find the probability that their mean length will be greater than $5.0$ minutes. *(hypothetical data)*
:::

---

## Example 6-13

:::{.callout-tip}
## Solution

Since the variable is approximately normally distributed, the distribution of sample means is approximately normal with a mean of $4.5$ and standard error

$$\sigma_{\bar{X}} = \frac{\sigma}{\sqrt{n}} = \frac{1.25}{\sqrt{25}} = 0.25$$

**Step 1** Draw a normal curve and shade the desired area (Figure 6-35).

**Step 2** $z = \dfrac{\bar{X} - \mu}{\sigma/\sqrt{n}} = \dfrac{5.0 - 4.5}{0.25} = 2.00$

**Step 3** The area to the right of $2.00$ is $1.0000 - 0.9772 = 0.0228$, or $2.28\%$.

```{r e6-13-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
se <- 1.25 / sqrt(25)

se
1 - pnorm(5.0, mean = 4.5, sd = se)
```

The probability that the $25$ calls last more than $5.0$ minutes on average is **2.28%**.
:::

## Example 6-14 {#example-6-14}

:::{.callout-warning}
## Parcels Delivered per Van

A courier company's vans deliver an average of $120$ parcels per day. Assume the standard deviation is $18$ parcels. If a random sample of $36$ van-days is selected, find the probability that the sample mean is between $114$ and $123$ parcels. *(hypothetical data)*
:::

---

## Example 6-14

:::{.callout-tip}
## Solution

**Step 1** Draw a normal curve and shade the desired area (Figure 6-36). Since the sample is $30$ or more, the normality assumption is not necessary.

**Step 2** The standard error is $18/\sqrt{36} = 3$, so $z_1 = \dfrac{114 - 120}{3} = -2$ and $z_2 = \dfrac{123 - 120}{3} = 1$.

**Step 3** The area for $z = -2$ is $0.0228$ and the area for $z = 1$ is $0.8413$, so the area between them is $0.8413 - 0.0228 = 0.8186$, or $81.86\%$.

```{r e6-14-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
se <- 18 / sqrt(36)

se
pnorm(123, mean = 120, sd = se) - pnorm(114, mean = 120, sd = se)
```
:::

## Example 6-15 {#example-6-15}

:::{.callout-warning}
## Customs Clearance Time

Imported shipments arriving at a Taiwanese port clear customs in an average of $3.2$ days. Assume the distribution is approximately normal and has a standard deviation of $0.5$ day.

**(a)** Find the probability that one randomly selected shipment clears customs in fewer than $3.3$ days.

**(b)** If a sample of $36$ shipments is randomly selected, find the probability that the mean clearance time of the sample is less than $3.3$ days. *(hypothetical data)*
:::

---

## Example 6-15

:::{.callout-tip}
## Solution (a)

Since the question concerns an **individual** shipment, use $z = \dfrac{X - \mu}{\sigma}$ (Figure 6-37).

**Step 2** $z = \dfrac{3.3 - 3.2}{0.5} = 0.20$

**Step 3** The area to the left of $z = 0.20$ is $0.5793$.

The probability of selecting a shipment that clears customs in fewer than $3.3$ days is $0.5793$, or $57.93\%$.

```{r e6-15-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
pnorm(3.3, mean = 3.2, sd = 0.5)
```
:::

---

## Example 6-15

:::{.callout-tip}
## Solution (b)

Since the question concerns the **mean of a sample** of size $36$, use the central limit theorem formula $z = \dfrac{\bar{X} - \mu}{\sigma/\sqrt{n}}$ (Figure 6-38).

**Step 2** $z = \dfrac{3.3 - 3.2}{0.5/\sqrt{36}} = 1.20$

**Step 3** The area corresponding to $z = 1.20$ is $0.8849$, or $88.49\%$.

```{r e6-15-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
se <- 0.5 / sqrt(36)

se
pnorm(3.3, mean = 3.2, sd = se)
```

The gap of about $30.6$ percentage points between (a) and (b) arises because the distribution of sample means is **much less variable** than the distribution of individual data values: as the sample size increases, the standard deviation of the means decreases.
:::

## Table 6-1

:::{.callout-important}
## Table 6-1 Summary of Formulas and Their Uses

| Formula | Use |
|---|---|
| $z = \dfrac{X - \mu}{\sigma}$ | Used to gain information about an **individual data value** when the variable is normally distributed |
| $z = \dfrac{\bar{X} - \mu}{\sigma/\sqrt{n}}$ | Used to gain information, when applying the central limit theorem, about a **sample mean** when the variable is normally distributed or when the sample size is $30$ or more |
:::

Notice that the first formula contains $X$, an individual data value, while the second contains $\bar{X}$, the sample mean.

## Finite Population Correction Factor (Optional)

The standard error $\sigma/\sqrt{n}$ is exact for sampling **with replacement**, or from a very large (infinite) population. Sampling **without replacement from a finite population** needs a correction factor.

:::{.callout-important}
## Finite Population Correction Factor

$$\sqrt{\frac{N-n}{N-1}}$$

where $N$ = population size and $n$ = sample size. The standard error and $z$ become

$$\sigma_{\bar{X}} = \frac{\sigma}{\sqrt{n}} \cdot \sqrt{\frac{N-n}{N-1}} \qquad\qquad z = \frac{\bar{X} - \mu}{\dfrac{\sigma}{\sqrt{n}} \cdot \sqrt{\dfrac{N-n}{N-1}}}$$
:::

---

## Finite Population Correction Factor (Optional)

:::{.callout-warning}
## Caution

It is needed when the sample is relatively large — usually **more than $5\%$ of the population**. When the population is large and the sample small, the factor is very close to $1.00$ and is generally omitted.
:::

# Section 6-4: The Normal Approximation to the Binomial Distribution {background-color="#2c4a6e"}

## Why Use the Normal Approximation?

When $n$ is large (say, $100$), binomial calculations are too difficult by hand, so a normal distribution is used instead. Recall from Chapter 5 that a **binomial distribution** has these characteristics:

1. There must be a **fixed number of trials**.
2. The outcome of each trial must be **independent**.
3. Each trial can have only **two outcomes**, or outcomes reducible to two.
4. The **probability of a success must remain the same** for each trial.

As $n$ increases with $p$ near $0.5$, the binomial shape becomes similar to a normal distribution. **But when $p$ is close to $0$ or $1$ and $n$ is relatively small, a normal approximation is inaccurate.**

---

## The Conditions $np \geq 5$ and $nq \geq 5$

:::{.callout-important}
## Rule of Thumb

A normal approximation should be used only when $n \cdot p$ and $n \cdot q$ are **both greater than or equal to 5**, where $q = 1 - p$.

Also, $\mu = n \cdot p$ and $\sigma = \sqrt{n \cdot p \cdot q}$.
:::

For example, if $p = 0.3$ and $n = 10$, then $np = 3$, so a normal distribution should **not** be used; if $p = 0.5$ and $n = 10$, then $np = nq = 5$ and it **can** be used.

---

## The Conditions $np \geq 5$ and $nq \geq 5$ (cont.)

```{r s6-4-1, echo=T, warning=FALSE, out.width="65%", fig.align='center', fig.height=3}
binomial_shapes <- data.frame(
  label       = rep(c("n = 10, p = 0.3 (np = 3)", "n = 10, p = 0.5 (np = 5, nq = 5)"),
                    each = 11),
  x           = c(0:10, 0:10),
  probability = c(dbinom(0:10, 10, 0.3), dbinom(0:10, 10, 0.5))
)

ggplot(binomial_shapes, aes(x, probability)) +
  geom_col() +
  facet_wrap(~ label) +
  labs(title = "Figure 6-39: binomial distribution and a normal distribution",
       x = "X", y = "P(X)")
```

## Correction for Continuity

:::{.callout-note}
## Correction for Continuity

A **correction for continuity** is a correction employed when a **continuous** distribution is used to approximate a **discrete** distribution.
:::

The continuity correction means that for any specific value of $X$, say $8$, the **boundaries** of $X$ in the binomial distribution (in this case $7.5$ to $8.5$) must be used. For $P(X = 8)$ the correction is $P(7.5 < X < 8.5)$; for $P(X \leq 7)$ it is $P(X < 7.5)$; for $P(X \geq 3)$ it is $P(X > 2.5)$.

---

## Table 6-2

:::{.callout-important}
## Table 6-2 Summary of the Normal Approximation to the Binomial Distribution

| When finding (binomial) | Use (normal) |
|---|---|
| 1. $P(X = a)$ | $P(a - 0.5 < X < a + 0.5)$ |
| 2. $P(X \geq a)$ | $P(X > a - 0.5)$ |
| 3. $P(X > a)$ | $P(X > a + 0.5)$ |
| 4. $P(X \leq a)$ | $P(X < a + 0.5)$ |
| 5. $P(X < a)$ | $P(X < a - 0.5)$ |

For all cases, $\mu = n \cdot p$, $\sigma = \sqrt{n \cdot p \cdot q}$, $n \cdot p \geq 5$ and $n \cdot q \geq 5$.
:::

---

## Procedure Table

:::{.callout-important}
## Procedure for the Normal Approximation to the Binomial Distribution

**Step 1** Check to see whether the normal approximation can be used.

**Step 2** Find the mean $\mu$ and the standard deviation $\sigma$.

**Step 3** Write the problem in probability notation, using $X$.

**Step 4** Rewrite the problem by using the continuity correction factor, and show the corresponding area under the normal distribution.

**Step 5** Find the corresponding $z$ values.

**Step 6** Find the solution.
:::

## Example 6-16 {#example-6-16}

:::{.callout-warning}
## Convenience-Store Pickup

On a cross-border e-commerce platform, $30\%$ of buyers choose convenience-store pickup instead of home delivery. If $150$ buyers are selected at random, find the probability that exactly $52$ of them choose convenience-store pickup. *(hypothetical data)*
:::

---

## Example 6-16

:::{.callout-tip}
## Solution

Here $p = 0.30$, $q = 0.70$ and $n = 150$.

**Step 1** $np = (150)(0.30) = 45$ and $nq = (150)(0.70) = 105$. Since $np \geq 5$ and $nq \geq 5$, the normal distribution can be used.

**Step 2** $\mu = np = 45$ and $\sigma = \sqrt{npq} = \sqrt{(150)(0.30)(0.70)} = \sqrt{31.5} \approx 5.61$

**Step 3** $P(X = 52)$

**Step 4** Using approximation 1 in Table 6-2: $P(51.5 < X < 52.5)$ (Figure 6-40).

**Step 5** $z_1 = \dfrac{51.5 - 45}{5.61} \approx 1.16$ and $z_2 = \dfrac{52.5 - 45}{5.61} \approx 1.34$

**Step 6** The area between $z_1$ and $z_2$ is $0.0327$, or about $3.27\%$.
:::

---

## Example 6-16

:::{.callout-tip}
## Solution

**Compute in R**

```{r e6-16-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
150 * 0.30
150 * 0.70
sigma <- sqrt(150 * 0.30 * 0.70)
sigma
pnorm(52.5, mean = 45, sd = sigma) - pnorm(51.5, mean = 45, sd = sigma)
dbinom(52, 150, 0.30)
```

The normal approximation gives **0.0327**, and the exact binomial probability is $0.032$.
:::

## Example 6-17 {#example-6-17}

:::{.callout-warning}
## Returned Orders

Fifteen percent of the orders placed on an online marketplace are returned. If a random sample of $240$ orders is selected, find the probability that $24$ or more will be returned. *(hypothetical data)*
:::

---

## Example 6-17

:::{.callout-tip}
## Solution

**Step 1** Here $p = 0.15$, $q = 0.85$ and $n = 240$. Since $np = (240)(0.15) = 36$ and $nq = (240)(0.85) = 204$, the normal approximation can be used.

**Step 2** $\mu = np = 36$ and $\sigma = \sqrt{npq} = \sqrt{(240)(0.15)(0.85)} = \sqrt{30.6} \approx 5.53$

**Step 3** $P(X \geq 24)$

**Step 4** Using approximation 2 in Table 6-2: $P(X > 24 - 0.5) = P(X > 23.5)$ (Figure 6-41).

**Step 5** $z = \dfrac{23.5 - 36}{5.53} \approx -2.26$

**Step 6** The area to the right of that $z$ value is $0.9881$, or $98.81\%$.
:::

---

## Example 6-17

:::{.callout-tip}
## Solution

**Compute in R**

```{r e6-17-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
240 * 0.15
240 * 0.85
sigma <- sqrt(240 * 0.15 * 0.85)
sigma
1 - pnorm(23.5, mean = 36, sd = sigma)
1 - pbinom(23, 240, 0.15)
```

In a random sample of $240$ orders the probability of $24$ or more being returned is **98.81%**; the exact binomial probability is $0.991$.
:::

## Example 6-18 {#example-6-18}

:::{.callout-warning}
## Mobile Wallet Payments

A recent in-store survey found that $45\%$ of shoppers at a department store pay with a mobile wallet. Find the probability that in a random sample of $120$ shoppers, **at most $48$** will pay with a mobile wallet. *(hypothetical data)*
:::

---

## Example 6-18

:::{.callout-tip}
## Solution

**Step 1** Here $p = 0.45$, $q = 0.55$ and $n = 120$, so $np = 54$ and $nq = 66$. Since both values are greater than or equal to $5$, the normal approximation can be used.

**Step 2** $\mu = np = 54$ and $\sigma = \sqrt{npq} = \sqrt{(120)(0.45)(0.55)} \approx 5.45$

**Step 3** $P(X \leq 48)$, which with the correction factor becomes $P(X < 48 + 0.5) = P(X < 48.5)$ (Figure 6-42).

**Step 4** $z = \dfrac{48.5 - 54}{5.45} \approx -1.01$

**Step 5** The area to the left of that $z$ value is $0.1564$, or $15.64\%$.

The probability that at most $48$ of the $120$ shoppers pay with a mobile wallet is **15.64%**.
:::

---

## Example 6-18

:::{.callout-tip}
## Solution

**Compute in R**

```{r e6-18-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
120 * 0.45
120 * 0.55
sigma <- sqrt(120 * 0.45 * 0.55)
sigma
pnorm(48.5, mean = 54, sd = sigma)
pbinom(48, 120, 0.45)
```
:::

## Example 6-19 {#example-6-19}

:::{.callout-warning}
## Binomial versus Normal Approximation

When $n = 12$ and $p = 0.5$, use the binomial distribution table (Table A-2 in Appendix A) to find the probability that $X = 8$. Then use the normal approximation to find the probability that $X = 8$.
:::

---

## Example 6-19

:::{.callout-tip}
## Solution

From Table A-2, for $n = 12$, $p = 0.5$ and $X = 8$, the probability is $0.121$.

For a normal approximation, $\mu = np = (12)(0.5) = 6$ and $\sigma = \sqrt{npq} = \sqrt{(12)(0.5)(0.5)} = \sqrt{3} \approx 1.73$.

Now $X = 8$ is represented by the boundaries $7.5$ and $8.5$, so

$$z_1 = \frac{7.5 - 6}{1.73} \approx 0.87 \qquad z_2 = \frac{8.5 - 6}{1.73} \approx 1.44$$

The area between these boundaries is $0.1188$ — very close to the exact binomial probability $0.1208$ (Figure 6-43).
:::

---

## Example 6-19

:::{.callout-tip}
## Solution

**Compute in R**

```{r e6-19-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
12 * 0.5
12 * 0.5
sigma <- sqrt(12 * 0.5 * 0.5)
sigma
dbinom(8, 12, 0.5)
pnorm(8.5, mean = 6, sd = sigma) - pnorm(7.5, mean = 6, sd = sigma)
```
:::

---

## Example 6-19

:::{.callout-warning}
## Caution

Here $np = 6$ and $nq = 6$ — only just above the minimum threshold of $5$. The normal approximation may be less accurate for small $n$, yet $0.1188$ is still close to the exact value $0.1208$.

A normal distribution can also be used to approximate other distributions, such as the **Poisson** distribution (Table A-3 in Appendix A).
:::

## Important Terms

:::{.callout-important}
## Chapter 6 Vocabulary

central limit theorem · correction for continuity · negatively or left-skewed distribution · normal distribution · positively or right-skewed distribution · sampling distribution of sample means · sampling error · standard error of the mean · standard normal distribution · symmetric distribution · $z$ value ($z$ score)
:::

## Key Formulas

:::{.callout-important}
## The Normal Distribution

Equation of a normal distribution and of the standard normal distribution:
$$y = \frac{e^{-(X-\mu)^2/(2\sigma^2)}}{\sigma\sqrt{2\pi}} \qquad\qquad y = \frac{e^{-z^2/2}}{\sqrt{2\pi}}$$

Formula for the $z$ score (or standard score), and for finding a specific data value:
$$z = \frac{X - \mu}{\sigma} \qquad\qquad X = z \cdot \sigma + \mu$$
:::

---

## Key Formulas

:::{.callout-important}
## Sample Means and the Binomial Approximation

Mean of the sample means and standard error of the mean:
$$\mu_{\bar{X}} = \mu \qquad\qquad \sigma_{\bar{X}} = \frac{\sigma}{\sqrt{n}}$$

$z$ value for the central limit theorem, and the finite population correction factor:
$$z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}} \qquad\qquad \sqrt{\frac{N-n}{N-1}}$$

Mean and standard deviation for the binomial distribution:
$$\mu = n \cdot p \qquad\qquad \sigma = \sqrt{n \cdot p \cdot q}$$

Pearson coefficient of skewness: $PC = \dfrac{3(\bar{X} - \text{median})}{s}$
:::

## Key Takeaways

:::{.callout-important}
- A **normal distribution** is continuous, bell-shaped, unimodal and symmetric; mean, median and mode are equal, the total area is $1$, and the curve never touches the $x$ axis
- About $68\%$, $95\%$ and $99.7\%$ of the area lies within $1$, $2$ and $3$ standard deviations of the mean
- The **standard normal distribution** has $\mu = 0$, $\sigma = 1$; convert with $z = \dfrac{X-\mu}{\sigma}$ and recover a data value with $X = z\sigma + \mu$
- Areas under the standard normal curve are **probabilities**; the distribution is continuous, so $P(a \leq z \leq b) = P(a < z < b)$
- **Normality** is checked with a histogram, Pearson's index $PC = \dfrac{3(\bar{X}-\text{median})}{s}$ (significant when $PC \geq +1$ or $PC \leq -1$), outliers, and a normal quantile plot
:::

## Key Takeaways

:::{.callout-important}
- A **sampling distribution of sample means** uses the means of all samples of a given size; $\mu_{\bar{X}} = \mu$ and the **standard error** is $\sigma_{\bar{X}} = \sigma/\sqrt{n}$
- The **central limit theorem**: as $n$ increases without limit, the distribution of sample means approaches normal — for any $n$ if the population is normal, otherwise for $n \geq 30$
- The **finite population correction factor** $\sqrt{\dfrac{N-n}{N-1}}$ is needed for a sample larger than $5\%$ of the population drawn without replacement
- The normal approximation to the binomial applies when $np \geq 5$ **and** $nq \geq 5$, with $\mu = np$ and $\sigma = \sqrt{npq}$
- Always apply the **correction for continuity** ($\pm 0.5$) when a continuous distribution approximates a discrete one
:::

## Acknowledgement

-   **Copyright notice.** These teaching materials follow the organization and terminology of **Bluman, A. G. (2023). Elementary statistics: A step by step approach (11th ed.). McGraw Hill**. All rights in the original work are reserved by its authors and publishers.

-   **Original examples.** Every worked example, data set, and R script in these slides was written for this course. The data are hypothetical unless stated otherwise.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
