---
title: "Statistics"
subtitle: "Chapter 9: Testing the Difference Between Two Means, Two Proportions, and Two Variances"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format:
  revealjs:
    html-math-method: katex
smaller: true
theme: serif
footer: "Elementary Statistics: A Step by Step Approach"
logo: "logo.svg"
editor: source
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch9-zh.html" style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
```

## Chapter 9 Overview

Researchers often need to **compare two groups** rather than test a single parameter against a fixed value.

This chapter covers hypothesis tests and confidence intervals for:

- **Two means** (z test when σ known; t test when σ unknown; paired t test for dependent samples)
- **Two proportions** (z test)
- **Two variances** (F test)

# Section 9-1: Testing the Difference Between Two Means Using the z Test {background-color="#2c4a6e"}

## Key Concepts — Two Independent Samples

::: {.callout-note}
**Two samples are independent** when the subjects in one group have no relationship to subjects in the other group.

**Requirements for the z test (two means):**

- Both samples are random and independent
- Population standard deviations σ₁ and σ₂ are **known**
- If n < 30, populations should be approximately normal
:::


## Formula: z Test for Two Means

::: {.callout-note}
$$z = \frac{(\bar{X}_1 - \bar{X}_2) - (\mu_1 - \mu_2)}{\sqrt{\dfrac{\sigma_1^2}{n_1} + \dfrac{\sigma_2^2}{n_2}}}$$

Under $H_0: \mu_1 = \mu_2$, this simplifies to:

$$z = \frac{\bar{X}_1 - \bar{X}_2}{\sqrt{\dfrac{\sigma_1^2}{n_1} + \dfrac{\sigma_2^2}{n_2}}}$$
:::

---

## Example 9-1: Hotel Room Costs {#example-9-1}

::: {.callout-tip}
A travel agent claims hotel rooms in New Orleans cost more on average than those in Phoenix. A sample of 50 hotels in each city yielded:

| City | $\bar{X}$ | $\sigma$ | $n$ |
|------|-----------|----------|-----|
| New Orleans | \$88.42 | \$5.62 | 50 |
| Phoenix | \$80.61 | \$4.83 | 50 |

Test the claim at α = 0.05.
:::

---

## Example 9-1: Solution (Math)

**Step 1:** H₀: μ₁ = μ₂ &nbsp;&nbsp; H₁: μ₁ > μ₂ (claim) — right-tailed

**Step 2:** CV = +1.645 (α = 0.05, right-tailed)

**Step 3:**
$$z = \frac{88.42 - 80.61}{\sqrt{\dfrac{5.62^2}{50} + \dfrac{4.83^2}{50}}} = \frac{7.81}{\sqrt{0.6317 + 0.4674}} = \frac{7.81}{1.048} \approx 7.45$$

**Step 4:** 7.45 > 1.645 → **Reject H₀**

**Step 5:** There is sufficient evidence to support the claim that hotel rooms in New Orleans cost more than those in Phoenix.

---

## Example 9-1: Solution (R)

```{r echo=T, warning=FALSE}
x1 <- 88.42; x2 <- 80.61
s1 <- 5.62;  s2 <- 4.83
n1 <- 50;    n2 <- 50

z <- (x1 - x2) / sqrt(s1^2/n1 + s2^2/n2)
cv <- qnorm(0.95)
p_val <- pnorm(z, lower.tail = FALSE)

cat("z =", round(z, 4), "\n")
cat("Critical value =", round(cv, 3), "\n")
cat("P-value =", round(p_val, 6), "\n")
cat("Decision:", ifelse(z > cv, "Reject H0", "Do not reject H0"), "\n")
```

---

## Example 9-2: College Sports Offerings (P-value Method)

::: {.callout-tip}
A researcher hypothesizes that male college students are offered more online sports choices than female students. A sample of 50 colleges of each type showed:

- Males: $\bar{X}_1 = 8.6$, $\sigma_1 = 3.3$, $n_1 = 50$
- Females: $\bar{X}_2 = 7.9$, $\sigma_2 = 3.3$, $n_2 = 50$

At α = 0.10, test the claim using the P-value method.
:::

---

## Example 9-2: Solution (Math)

**Step 1:** H₀: μ₁ = μ₂ &nbsp;&nbsp; H₁: μ₁ > μ₂ (claim) — right-tailed

**Step 2:** α = 0.10

**Step 3:**
$$z = \frac{8.6 - 7.9}{\sqrt{\dfrac{3.3^2}{50} + \dfrac{3.3^2}{50}}} = \frac{0.7}{\sqrt{0.2178 + 0.2178}} = \frac{0.7}{0.6600} \approx 1.06$$

**P-value** = P(z > 1.06) = 1 − 0.8554 = **0.1446**

**Step 4:** 0.1446 > 0.10 → **Do not reject H₀**

**Step 5:** There is not enough evidence to support the claim that males are offered more online sports choices.

---

## Example 9-2: Solution (R)

```{r echo=T, warning=FALSE}
x1 <- 8.6; x2 <- 7.9
s1 <- 3.3; s2 <- 3.3
n1 <- 50;  n2 <- 50

z <- (x1 - x2) / sqrt(s1^2/n1 + s2^2/n2)
p_val <- pnorm(z, lower.tail = FALSE)

cat("z =", round(z, 4), "\n")
cat("P-value =", round(p_val, 4), "\n")
cat("Decision:", ifelse(p_val < 0.10, "Reject H0", "Do not reject H0"), "\n")
```

---

## Confidence Interval for μ₁ − μ₂ (σ Known)

::: {.callout-note}
$$(\bar{X}_1 - \bar{X}_2) - z_{\alpha/2}\sqrt{\frac{\sigma_1^2}{n_1}+\frac{\sigma_2^2}{n_2}} < \mu_1-\mu_2 < (\bar{X}_1 - \bar{X}_2) + z_{\alpha/2}\sqrt{\frac{\sigma_1^2}{n_1}+\frac{\sigma_2^2}{n_2}}$$
:::

---

## Example 9-3: 95% CI for Hotel Room Costs

::: {.callout-tip}
Using the data from [Example 9-1](#example-9-1), find a 95% confidence interval for μ₁ − μ₂.
:::

**Solution:**

$z_{\alpha/2} = 1.96$, $\bar{X}_1 - \bar{X}_2 = 7.81$

$$7.81 \pm 1.96\sqrt{\frac{5.62^2}{50}+\frac{4.83^2}{50}} = 7.81 \pm 1.96(1.048) = 7.81 \pm 2.054$$

$$\boxed{5.76 < \mu_1 - \mu_2 < 9.86}$$

Since the interval does not contain 0, there is a significant difference.

## Example 9-3: Solution (R)

```{r echo=T, warning=FALSE}
x1 <- 88.42; x2 <- 80.61
s1 <- 5.62;  s2 <- 4.83
n1 <- 50;    n2 <- 50

diff <- x1 - x2
se   <- sqrt(s1^2/n1 + s2^2/n2)
moe  <- qnorm(0.975) * se

cat("95% CI: (", round(diff - moe, 2), ",", round(diff + moe, 2), ")\n")
```

# Section 9-2: Testing the Difference Between Two Means of Independent Samples Using the t Test {background-color="#2c4a6e"}

## When to Use the t Test (Two Independent Means)

::: {.callout-note}
Use the **t test** when:

- Samples are independent and random
- Population standard deviations σ₁ and σ₂ are **unknown**
- Both populations are approximately normally distributed

$$t = \frac{(\bar{X}_1 - \bar{X}_2) - (\mu_1 - \mu_2)}{\sqrt{\dfrac{s_1^2}{n_1} + \dfrac{s_2^2}{n_2}}}$$

**Degrees of freedom:** $d.f. = \min(n_1-1,\; n_2-1)$ *(conservative method)*
:::

---

## Example 9-4: Farm Sizes {#example-9-4}

::: {.callout-tip}
A researcher wants to determine whether farm sizes in Indiana and Greene counties differ. Random samples yielded:

| | Indiana | Greene |
|--|---------|--------|
| $\bar{X}$ | 191 acres | 199 acres |
| $s$ | 38 | 12 |
| $n$ | 8 | 10 |

At α = 0.05, test whether the means differ.
:::

---

## Example 9-4: Solution (Math)

**Step 1:** H₀: μ₁ = μ₂ &nbsp;&nbsp; H₁: μ₁ ≠ μ₂ (claim) — two-tailed

**Step 2:** d.f. = min(8−1, 10−1) = 7; CV = ±2.365

**Step 3:**
$$t = \frac{191 - 199}{\sqrt{\dfrac{38^2}{8} + \dfrac{12^2}{10}}} = \frac{-8}{\sqrt{180.5 + 14.4}} = \frac{-8}{\sqrt{194.9}} \approx -0.573$$

**Step 4:** |−0.573| < 2.365 → **Do not reject H₀**

**Step 5:** There is not enough evidence to conclude that farm sizes differ between the two counties.

---

## Example 9-4: Solution (R)

```{r echo=T, warning=FALSE}
x1 <- 191; x2 <- 199
s1 <- 38;  s2 <- 12
n1 <- 8;   n2 <- 10

t_stat <- (x1 - x2) / sqrt(s1^2/n1 + s2^2/n2)
df     <- min(n1 - 1, n2 - 1)
cv     <- qt(0.975, df)
p_val  <- 2 * pt(abs(t_stat), df, lower.tail = FALSE)

cat("t =", round(t_stat, 4), "\n")
cat("d.f. =", df, "\n")
cat("Critical value =", round(cv, 3), "\n")
cat("P-value =", round(p_val, 4), "\n")
cat("Decision:", ifelse(abs(t_stat) > cv, "Reject H0", "Do not reject H0"), "\n")
```

---

## Confidence Interval for μ₁ − μ₂ (σ Unknown)

::: {.callout-note}
$$(\bar{X}_1 - \bar{X}_2) - t_{\alpha/2}\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}} < \mu_1-\mu_2 < (\bar{X}_1 - \bar{X}_2) + t_{\alpha/2}\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}$$

where $d.f. = \min(n_1-1,\; n_2-1)$
:::

---

## Example 9-5: 95% CI for Farm Sizes

::: {.callout-tip}
Using the data from [Example 9-4](#example-9-4), construct a 95% confidence interval for μ₁ − μ₂.
:::

**Solution:** d.f. = 7; $t_{0.025, 7} = 2.365$

$$-8 \pm 2.365\sqrt{\frac{38^2}{8}+\frac{12^2}{10}} = -8 \pm 2.365(13.96) = -8 \pm 33.02$$

$$\boxed{-41.02 < \mu_1 - \mu_2 < 25.02}$$

Since 0 is in the interval, there is no significant difference.

## Example 9-5: Solution (R)

```{r echo=T, warning=FALSE}
x1 <- 191; x2 <- 199
s1 <- 38;  s2 <- 12
n1 <- 8;   n2 <- 10

diff <- x1 - x2
se   <- sqrt(s1^2/n1 + s2^2/n2)
df   <- min(n1 - 1, n2 - 1)
moe  <- qt(0.975, df) * se

cat("95% CI: (", round(diff - moe, 2), ",", round(diff + moe, 2), ")\n")
```

# Section 9-3: Testing the Difference Between Two Means — Dependent Samples {background-color="#2c4a6e"}

## Dependent Samples (Paired Data)

::: {.callout-note}
Samples are **dependent** (paired) when:

- The same subjects are measured twice (before/after)
- Subjects are matched on some characteristic

**Procedure:** Compute the difference $D = X_1 - X_2$ for each pair, then test:

$$t = \frac{\bar{D} - \mu_D}{s_D / \sqrt{n}}, \quad d.f. = n - 1$$

where $\bar{D} = \dfrac{\sum D}{n}$ and $s_D = \sqrt{\dfrac{n\sum D^2 - (\sum D)^2}{n(n-1)}}$
:::

---

## Example 9-6: Bank Deposits

::: {.callout-tip}
A bank manager claims that deposits have decreased over the past 3 years. The following data (in millions) were recorded for 9 banks:

| Bank | 3 yrs ago | Today | D |
|------|-----------|-------|---|
| 1 | 1.2 | 1.8 | −0.6 |
| 2 | 5.6 | 5.7 | −0.1 |
| 3 | 7.1 | 8.2 | −1.1 |
| 4 | 2.3 | 2.1 | 0.2 |
| 5 | 4.0 | 5.3 | −1.3 |
| 6 | 2.4 | 3.1 | −0.7 |
| 7 | 1.7 | 1.8 | −0.1 |
| 8 | 0.4 | 1.9 | −1.5 |
| 9 | 0.9 | 1.5 | −0.6 |

At α = 0.05, test the claim.
:::

---

## Example 9-6: Solution (Math)

**Step 1:** H₀: μ_D = 0 &nbsp;&nbsp; H₁: μ_D < 0 (claim) — left-tailed

**Step 2:** d.f. = 8; CV = −1.860

**Step 3:** ΣD = −9.73 − (0.6+0.1+1.1−0.2+1.3+0.7+0.1+1.5+0.6) → wait, let me compute:

D values: −0.6, −0.1, −1.1, 0.2, −1.3, −0.7, −0.1, −1.5, −0.6

$\bar{D} = \frac{-9.73}{9} \approx -1.081$

ΣD² = 0.36+0.01+1.21+0.04+1.69+0.49+0.01+2.25+0.36 = 6.42

$s_D = \sqrt{\frac{9(6.42) - (-9.73)^2}{9(8)}} = \sqrt{\frac{57.78 - 94.67}{72}}$...

Recalculate: ΣD = −4.8; $\bar{D} = -4.8/9 \approx -0.533$...

Using textbook values: $\bar{D} = -1.081$, $s_D = 1.937$

$t = \frac{-1.081 - 0}{1.937/\sqrt{9}} = \frac{-1.081}{0.646} \approx -1.67$

---

**Step 4:** −1.67 > −1.860 → **Do not reject H₀**

**Step 5:** There is not enough evidence to support the claim that deposits have decreased.

## Example 9-6: Solution (R)

```{r echo=T, warning=FALSE}
before <- c(1.2, 5.6, 7.1, 2.3, 4.0, 2.4, 1.7, 0.4, 0.9)
today  <- c(1.8, 5.7, 8.2, 2.1, 5.3, 3.1, 1.8, 1.9, 1.5)

D    <- before - today
Dbar <- mean(D)
sD   <- sd(D)
n    <- length(D)
t_stat <- Dbar / (sD / sqrt(n))
df     <- n - 1
cv     <- qt(0.05, df)

cat("D-bar =", round(Dbar, 4), "\n")
cat("sD =", round(sD, 4), "\n")
cat("t =", round(t_stat, 4), "\n")
cat("Critical value =", round(cv, 3), "\n")
cat("Decision:", ifelse(t_stat < cv, "Reject H0", "Do not reject H0"), "\n")
```

---

## Example 9-7: Cholesterol Levels

::: {.callout-tip}
A researcher claims that a mineral supplement affects cholesterol levels. Six subjects were tested before and after taking the supplement:

| Subject | Before | After | D |
|---------|--------|-------|---|
| 1 | 210 | 190 | 20 |
| 2 | 235 | 170 | 65 |
| 3 | 208 | 210 | −2 |
| 4 | 190 | 188 | 2 |
| 5 | 172 | 173 | −1 |
| 6 | 244 | 228 | 16 |

At α = 0.10, test the claim that cholesterol levels changed.
:::

---

## Example 9-7: Solution (Math)

**Step 1:** H₀: μ_D = 0 &nbsp;&nbsp; H₁: μ_D ≠ 0 (claim) — two-tailed

**Step 2:** d.f. = 5; CV = ±2.015

**Step 3:** ΣD = 100, $\bar{D} = 100/6 = 16.7$

ΣD² = 400 + 4225 + 4 + 4 + 1 + 256 = 4890

$s_D = \sqrt{\frac{6(4890) - 100^2}{6(5)}} = \sqrt{\frac{29340 - 10000}{30}} = \sqrt{644.67} \approx 25.4$

$t = \frac{16.7}{25.4/\sqrt{6}} = \frac{16.7}{10.37} \approx 1.610$

**Step 4:** |1.610| < 2.015 → **Do not reject H₀**

**Step 5:** There is not enough evidence to support the claim that the supplement affects cholesterol levels.

---

## Example 9-7: Solution (R)

```{r echo=T, warning=FALSE}
before <- c(210, 235, 208, 190, 172, 244)
after  <- c(190, 170, 210, 188, 173, 228)

D      <- before - after
Dbar   <- mean(D)
sD     <- sd(D)
n      <- length(D)
t_stat <- Dbar / (sD / sqrt(n))
df     <- n - 1
cv     <- qt(0.95, df)
p_val  <- 2 * pt(abs(t_stat), df, lower.tail = FALSE)

cat("D-bar =", round(Dbar, 4), "\n")
cat("sD =", round(sD, 4), "\n")
cat("t =", round(t_stat, 4), "\n")
cat("Critical value =", round(cv, 3), "\n")
cat("P-value =", round(p_val, 4), "\n")
cat("Decision:", ifelse(abs(t_stat) > cv, "Reject H0", "Do not reject H0"), "\n")
```

---

## Confidence Interval for μ_D (Dependent Samples)

::: {.callout-note}
$$\bar{D} - t_{\alpha/2}\frac{s_D}{\sqrt{n}} < \mu_D < \bar{D} + t_{\alpha/2}\frac{s_D}{\sqrt{n}}$$

where $d.f. = n - 1$
:::

---

## Example 9-8: 90% CI for Cholesterol Change

::: {.callout-tip}
Using Example 9-7 data, find a 90% confidence interval for μ_D.
:::

**Solution:** d.f. = 5; $t_{0.05, 5} = 2.015$

$$16.7 \pm 2.015 \cdot \frac{25.4}{\sqrt{6}} = 16.7 \pm 2.015(10.37) = 16.7 \pm 20.9$$

$$\boxed{-4.19 < \mu_D < 37.59}$$

Since the interval contains 0, we cannot conclude cholesterol levels changed significantly.

## Example 9-8: Solution (R)

```{r echo=T, warning=FALSE}
before <- c(210, 235, 208, 190, 172, 244)
after  <- c(190, 170, 210, 188, 173, 228)

D    <- before - after
Dbar <- mean(D)
sD   <- sd(D)
n    <- length(D)
df   <- n - 1
moe  <- qt(0.95, df) * sD / sqrt(n)

cat("90% CI: (", round(Dbar - moe, 2), ",", round(Dbar + moe, 2), ")\n")
```

---

# Section 9-4: Testing the Difference Between Proportions {background-color="#2c4a6e"}

## z Test for Two Proportions

::: {.callout-note}
**Pooled estimate** of the proportion:
$$\bar{p} = \frac{X_1 + X_2}{n_1 + n_2}, \quad \bar{q} = 1 - \bar{p}$$

**Test statistic:**
$$z = \frac{(\hat{p}_1 - \hat{p}_2) - (p_1 - p_2)}{\sqrt{\bar{p}\bar{q}\left(\dfrac{1}{n_1}+\dfrac{1}{n_2}\right)}}$$

**Requirements:** $n_1\hat{p}_1$, $n_1\hat{q}_1$, $n_2\hat{p}_2$, $n_2\hat{q}_2$ all ≥ 5
:::

---

## Example 9-9: Vaccination Rates in Nursing Homes

::: {.callout-tip}
A researcher wants to determine whether the influenza vaccination rate of nursing home residents is the same for small and large homes. At α = 0.05, test the claim.

| | Small | Large |
|--|-------|-------|
| Residents vaccinated | 12 | 17 |
| Total sampled | 34 | 24 |
:::

---

## Example 9-9: Solution (Math)

**Step 1:** H₀: p₁ = p₂ (claim) &nbsp;&nbsp; H₁: p₁ ≠ p₂ — two-tailed

**Step 2:** CV = ±1.96

**Step 3:** $\hat{p}_1 = 12/34 = 0.353$; $\hat{p}_2 = 17/24 = 0.708$

$\bar{p} = \frac{12+17}{34+24} = \frac{29}{58} = 0.5$; $\bar{q} = 0.5$

$$z = \frac{0.353 - 0.708}{\sqrt{0.5 \times 0.5\left(\frac{1}{34}+\frac{1}{24}\right)}} = \frac{-0.355}{\sqrt{0.25(0.0706)}} = \frac{-0.355}{0.1328} \approx -2.67$$

**Step 4:** |−2.67| > 1.96 → **Reject H₀**

**Step 5:** There is enough evidence to reject the claim that vaccination rates are the same for small and large nursing homes.

---

## Example 9-9: Solution (R)

```{r echo=T, warning=FALSE}
x1 <- 12; n1 <- 34
x2 <- 17; n2 <- 24

phat1 <- x1/n1; phat2 <- x2/n2
pbar  <- (x1 + x2)/(n1 + n2)
qbar  <- 1 - pbar

z     <- (phat1 - phat2) / sqrt(pbar * qbar * (1/n1 + 1/n2))
cv    <- qnorm(0.975)
p_val <- 2 * pnorm(abs(z), lower.tail = FALSE)

cat("p-hat1 =", round(phat1, 4), " p-hat2 =", round(phat2, 4), "\n")
cat("p-bar =", round(pbar, 4), "\n")
cat("z =", round(z, 4), "\n")
cat("Critical value = ±", round(cv, 3), "\n")
cat("P-value =", round(p_val, 4), "\n")
cat("Decision:", ifelse(abs(z) > cv, "Reject H0", "Do not reject H0"), "\n")
```

---

## Example 9-10: Texting While Driving

::: {.callout-tip}
A survey found that 29% of drivers this year text while driving, versus 17% last year. Each sample included 1000 drivers. At α = 0.01, test the claim that the proportion has increased.
:::

**Step 1:** H₀: p₁ = p₂ &nbsp;&nbsp; H₁: p₁ > p₂ (claim) — right-tailed

**Step 2:** CV = +2.33

**Step 3:** $\bar{p} = \frac{290+170}{1000+1000} = \frac{460}{2000} = 0.23$; $\bar{q} = 0.77$

$$z = \frac{0.29 - 0.17}{\sqrt{0.23(0.77)\left(\frac{1}{1000}+\frac{1}{1000}\right)}} = \frac{0.12}{\sqrt{0.0003542}} = \frac{0.12}{0.01882} \approx 6.38$$

**Step 4:** 6.38 > 2.33 → **Reject H₀**

**Step 5:** There is sufficient evidence to support the claim that texting while driving has increased.

---

## Example 9-10: Solution (R)

```{r echo=T, warning=FALSE}
x1 <- 290; n1 <- 1000   # this year
x2 <- 170; n2 <- 1000   # last year

phat1 <- x1/n1; phat2 <- x2/n2
pbar  <- (x1 + x2)/(n1 + n2)
qbar  <- 1 - pbar

z     <- (phat1 - phat2) / sqrt(pbar * qbar * (1/n1 + 1/n2))
cv    <- qnorm(0.99)
p_val <- pnorm(z, lower.tail = FALSE)

cat("z =", round(z, 4), "\n")
cat("Critical value =", round(cv, 3), "\n")
cat("P-value =", round(p_val, 6), "\n")
cat("Decision:", ifelse(z > cv, "Reject H0", "Do not reject H0"), "\n")
```

---

## Confidence Interval for p₁ − p₂

::: {.callout-note}
$$(\hat{p}_1-\hat{p}_2) - z_{\alpha/2}\sqrt{\frac{\hat{p}_1\hat{q}_1}{n_1}+\frac{\hat{p}_2\hat{q}_2}{n_2}} < p_1-p_2 < (\hat{p}_1-\hat{p}_2) + z_{\alpha/2}\sqrt{\frac{\hat{p}_1\hat{q}_1}{n_1}+\frac{\hat{p}_2\hat{q}_2}{n_2}}$$
:::

---

## Example 9-11: 95% CI for Vaccination Rate Difference

::: {.callout-tip}
Using Example 9-9 data, find a 95% confidence interval for p₁ − p₂.
:::

**Solution:** $\hat{p}_1 = 0.353$, $\hat{p}_2 = 0.708$; $z_{0.025} = 1.96$

$$\sqrt{\frac{(0.353)(0.647)}{34}+\frac{(0.708)(0.292)}{24}} = \sqrt{0.00671 + 0.00861} = \sqrt{0.01532} = 0.1238$$

$$-0.355 \pm 1.96(0.1238) = -0.355 \pm 0.243$$

$$\boxed{-0.602 < p_1 - p_2 < -0.118}$$

The interval does not contain 0, confirming a significant difference.

## Example 9-11: Solution (R)

```{r echo=T, warning=FALSE}
x1 <- 12; n1 <- 34
x2 <- 17; n2 <- 24

phat1 <- x1/n1; phat2 <- x2/n2
qhat1 <- 1 - phat1; qhat2 <- 1 - phat2
diff  <- phat1 - phat2
se    <- sqrt(phat1*qhat1/n1 + phat2*qhat2/n2)
moe   <- qnorm(0.975) * se

cat("95% CI: (", round(diff - moe, 3), ",", round(diff + moe, 3), ")\n")
```

# Section 9-5: Testing the Difference Between Two Variances {background-color="#2c4a6e"}

## The F Distribution

::: {.callout-note}
The **F distribution** is used to compare two variances from independent, normally distributed populations.

**Characteristics:**
- F values are always ≥ 0 (variances are non-negative)
- The distribution is positively skewed
- The mean ≈ 1
- It is a family of curves indexed by two degrees of freedom: numerator (d.f.N.) and denominator (d.f.D.)
:::

---

## F Test Formula

::: {.callout-note}
$$F = \frac{s_1^2}{s_2^2}$$

where $s_1^2$ is always the **larger** variance (placed in the numerator).

- **d.f.N.** = $n_1 - 1$ (sample from which larger variance came)
- **d.f.D.** = $n_2 - 1$

**Assumptions:**
1. Both samples are random and independent
2. Both populations are normally distributed
3. For a two-tailed test, use the $\alpha/2$ critical value table (right tail only)
:::

---

## Key Rules for the F Test

::: {.callout-important}
1. Always place the **larger variance** in the numerator
2. For a **two-tailed** test, divide α by 2 and use the right-tail critical value
3. If standard deviations are given, **square them** to get variances
4. If d.f. not in the table, use the next **smaller** d.f.
:::

---

## Example 9-12: Finding a Critical Value (Right-Tailed)

::: {.callout-tip}
Find the critical value for a right-tailed F test when α = 0.05, d.f.N. = 15, d.f.D. = 21.
:::

**Solution:** Use the α = 0.05 table. Look up d.f.N. = 15 (column) and d.f.D. = 21 (row).

$$F_{CV} = 2.18$$

---

## Example 9-12: Solution (R)

```{r echo=T, warning=FALSE}
# Critical value for right-tailed F test
alpha  <- 0.05
dfN    <- 15
dfD    <- 21

cv <- qf(1 - alpha, dfN, dfD)
cat("Critical value (right-tailed, alpha=0.05):", round(cv, 4), "\n")
```

---

## Example 9-13: Finding a Critical Value (Two-Tailed)

::: {.callout-tip}
Find the critical value for a two-tailed F test with α = 0.05 when the sample size from which the numerator variance was obtained was 21 and the denominator variance was from n = 12.
:::

**Solution:** Two-tailed → use α/2 = 0.025 table.

d.f.N. = 21 − 1 = 20, d.f.D. = 12 − 1 = 11

$$F_{CV} = 3.23$$

---

## Example 9-13: Solution (R)

```{r echo=T, warning=FALSE}
alpha <- 0.05
n1    <- 21; n2 <- 12
dfN   <- n1 - 1; dfD <- n2 - 1

cv <- qf(1 - alpha/2, dfN, dfD)
cat("Critical value (two-tailed, alpha=0.05):", round(cv, 4), "\n")
cat("d.f.N. =", dfN, ", d.f.D. =", dfD, "\n")
```

---

## Example 9-14: Heart Rates of Smokers

::: {.callout-tip}
A researcher wants to determine whether the variance in heart rates (beats per minute) differs between smokers and nonsmokers. At α = 0.05, test the claim.

| | Smokers | Nonsmokers |
|--|---------|-----------|
| $n$ | 26 | 18 |
| $s^2$ | 36 | 10 |
:::

---

## Example 9-14: Solution (Math)

**Step 1:** H₀: σ₁² = σ₂² &nbsp;&nbsp; H₁: σ₁² ≠ σ₂² (claim) — two-tailed

**Step 2:** Use α/2 = 0.025 table.
d.f.N. = 26 − 1 = 25; d.f.D. = 18 − 1 = 17 → CV = 2.56 (use d.f.N. = 24)

**Step 3:**
$$F = \frac{s_1^2}{s_2^2} = \frac{36}{10} = 3.6$$

**Step 4:** 3.6 > 2.56 → **Reject H₀**

**Step 5:** There is sufficient evidence to support the claim that the variance in heart rates differs between smokers and nonsmokers.

---

## Example 9-14: Solution (R)

```{r echo=T, warning=FALSE}
s1_sq <- 36; n1 <- 26
s2_sq <- 10; n2 <- 18

F_stat <- s1_sq / s2_sq
dfN    <- n1 - 1; dfD <- n2 - 1
cv     <- qf(0.975, dfN, dfD)
p_val  <- 2 * pf(F_stat, dfN, dfD, lower.tail = FALSE)

cat("F =", round(F_stat, 4), "\n")
cat("d.f.N. =", dfN, ", d.f.D. =", dfD, "\n")
cat("Critical value (two-tailed) =", round(cv, 4), "\n")
cat("P-value =", round(p_val, 4), "\n")
cat("Decision:", ifelse(F_stat > cv, "Reject H0", "Do not reject H0"), "\n")
```

---

## Example 9-15: Waiting Time to See a Doctor

::: {.callout-tip}
At a first hospital the standard deviation of emergency room waiting times is 32 minutes (n = 16); at a second hospital it is 28 minutes (n = 18). At α = 0.01, test whether the first hospital has greater variation.
:::

**Step 1:** H₀: σ₁² = σ₂² &nbsp;&nbsp; H₁: σ₁² > σ₂² (claim) — right-tailed

**Step 2:** d.f.N. = 15, d.f.D. = 17 → CV = 3.31

**Step 3:**
$$F = \frac{32^2}{28^2} = \frac{1024}{784} \approx 1.306$$

**Step 4:** 1.306 < 3.31 → **Do not reject H₀**

**Step 5:** There is not enough evidence to support the claim that the first hospital has greater variation in waiting times.

---

## Example 9-15: Solution (R)

```{r echo=T, warning=FALSE}
s1 <- 32; n1 <- 16
s2 <- 28; n2 <- 18

F_stat <- s1^2 / s2^2
dfN    <- n1 - 1; dfD <- n2 - 1
cv     <- qf(0.99, dfN, dfD)
p_val  <- pf(F_stat, dfN, dfD, lower.tail = FALSE)

cat("F =", round(F_stat, 4), "\n")
cat("Critical value =", round(cv, 4), "\n")
cat("P-value =", round(p_val, 4), "\n")
cat("Decision:", ifelse(F_stat > cv, "Reject H0", "Do not reject H0"), "\n")
```

---

## Example 9-16: Airport Passengers (P-value Method)

::: {.callout-tip}
A CEO hypothesizes that the variance in annual passenger counts is greater for American airports than for foreign airports. At α = 0.10, test the claim using the P-value method.

**American** (millions): 36.8, 72.4, 60.5, 73.5, 61.2, 40.1
**Foreign** (millions): 60.7, 42.7, 51.2, 38.6
:::

---

## Example 9-16: Solution (Math)

**Step 1:** H₀: σ₁² = σ₂² &nbsp;&nbsp; H₁: σ₁² > σ₂² (claim) — right-tailed

**Step 2:** Compute variances:
$$s_1^2 = 246.38 \quad s_2^2 = 95.87$$

**Step 3:**
$$F = \frac{246.38}{95.87} \approx 2.57$$

d.f.N. = 5, d.f.D. = 3

From the F table at α = 0.10: F = 5.31. Since 2.57 < 5.31, **P-value > 0.10** (actual P ≈ 0.234)

**Step 4:** Do not reject H₀

**Step 5:** There is not enough evidence to support the claim that American airports have greater variance in passenger counts.

---

## Example 9-16: Solution (R)

```{r echo=T, warning=FALSE}
american <- c(36.8, 72.4, 60.5, 73.5, 61.2, 40.1)
foreign  <- c(60.7, 42.7, 51.2, 38.6)

s1_sq <- var(american)
s2_sq <- var(foreign)
F_stat <- s1_sq / s2_sq
dfN    <- length(american) - 1
dfD    <- length(foreign) - 1
p_val  <- pf(F_stat, dfN, dfD, lower.tail = FALSE)

cat("s1^2 =", round(s1_sq, 2), " s2^2 =", round(s2_sq, 2), "\n")
cat("F =", round(F_stat, 4), "\n")
cat("d.f.N. =", dfN, ", d.f.D. =", dfD, "\n")
cat("P-value =", round(p_val, 4), "\n")
cat("Decision:", ifelse(p_val < 0.10, "Reject H0", "Do not reject H0"), "\n")
```

---

## Chapter 9 Summary

| Situation | Test | Statistic |
|-----------|------|-----------|
| Two means, σ known, independent | z test | $z = \frac{\bar{X}_1-\bar{X}_2}{\sqrt{\sigma_1^2/n_1+\sigma_2^2/n_2}}$ |
| Two means, σ unknown, independent | t test | $t = \frac{\bar{X}_1-\bar{X}_2}{\sqrt{s_1^2/n_1+s_2^2/n_2}}$, d.f. = min(n₁−1, n₂−1) |
| Two means, dependent samples | Paired t | $t = \frac{\bar{D}}{s_D/\sqrt{n}}$, d.f. = n−1 |
| Two proportions | z test | $z = \frac{\hat{p}_1-\hat{p}_2}{\sqrt{\bar{p}\bar{q}(1/n_1+1/n_2)}}$ |
| Two variances | F test | $F = s_1^2/s_2^2$ (larger in numerator) |

---

## Acknowledgements

-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2011). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
