---
title: "Statistics"
subtitle: "Chapter 8: Hypothesis Testing"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
language:
  callout-warning-title: "Problem"
  callout-important-title: "Key point"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch8-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
```

## Overview

Statistical **hypothesis testing** is a decision-making procedure used to evaluate a claim about a population parameter.

:::{.callout-note}
## Key Vocabulary

- **Statistical hypothesis**: A conjecture about a population parameter
- **Null hypothesis ($H_0$)**: States no difference or no effect — the status quo
- **Alternative hypothesis ($H_1$)**: The research claim; what we try to support
- **Test value**: A computed statistic used to decide between $H_0$ and $H_1$
:::

## Chapter Sections

- **8-1** Steps in Hypothesis Testing — Traditional Method
- **8-2** z Test for a Mean
- **8-3** t Test for a Mean
- **8-4** z Test for a Proportion
- **8-5** $\chi^2$ Test for a Variance or Standard Deviation
- **8-6** Additional Topics Regarding Hypothesis Testing

# Section 8-1: Steps in Hypothesis Testing {background-color="#2c4a6e"}

## Null and Alternative Hypotheses

:::{.callout-note}
**Hypothesis Structure**

- **$H_0$** (null hypothesis): Contains equality ($=$, $\leq$, $\geq$)
- **$H_1$** (alternative/research hypothesis): Contains inequality ($\neq$, $<$, $>$)
- The **claim** may be in either $H_0$ or $H_1$
:::

The direction of the alternative hypothesis determines the **type of test**:

| $H_1$ contains | Test type |
|---|---|
| > | Right-tailed |
| < | Left-tailed |
| $\neq$ | Two-tailed |

## Types of Errors

:::{.callout-important}
**Type I and Type II Errors**

| Decision | $H_0$ True | $H_0$ False |
|---|---|---|
| Reject $H_0$ | **Type I error** ($\alpha$) | Correct decision (1 − $\beta$) |
| Do not reject $H_0$ | Correct decision (1 − $\alpha$) | **Type II error** ($\beta$) |

- **$\alpha$** = P(Type I error) = level of significance
- **$\beta$** = P(Type II error)
- Common values of $\alpha$: 0.10, 0.05, 0.01
:::

## Critical Region and Critical Value

:::{.callout-note}
**Key Terms**

- **Critical (rejection) region**: Range of values that lead to rejecting $H_0$
- **Noncritical region**: Range of values that lead to not rejecting $H_0$
- **Critical value**: Boundary separating the two regions
:::

## Example 8-1 {#example-8-1}

:::{.callout-warning}
## Stating Hypotheses

State $H_0$ and $H_1$ for each claim:

**(a)** A researcher claims newborns weigh more than 8.6 lb on average.

**(b)** A manufacturer claims defective disks average fewer than 18 per 1000.

**(c)** A researcher claims background soft music changes exam scores from the mean of 73.
:::

## Example 8-1

:::{.callout-tip}
## Solution

**(a)** $H_0$: $\mu = 8.6$ and $H_1$: $\mu > 8.6$ *(claim; right-tailed)*

**(b)** $H_0$: $\mu = 18$ and $H_1$: $\mu < 18$ *(claim; left-tailed)*

**(c)** $H_0$: $\mu = 73$ and $H_1$: $\mu \neq 73$ *(claim; two-tailed)*
:::

## Example 8-2 {#example-8-2}

:::{.callout-warning}
## Finding Critical Values

Find the critical value(s) for each situation:

**(a)** Left-tailed, $\alpha$ = 0.10
**(b)** Two-tailed, $\alpha$ = 0.02
**(c)** Right-tailed, $\alpha$ = 0.005
:::

## Example 8-2

:::{.callout-tip}
## Solution

```{r e8-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# (a) Left-tailed, alpha = 0.10
qnorm(0.10)

# (b) Two-tailed, alpha = 0.02 -> alpha/2 = 0.01 on each side
qnorm(0.01)  # left; right is +2.326

# (c) Right-tailed, alpha = 0.005
qnorm(1 - 0.005)
```

- (a) CV = **−1.28** (left-tailed)
- (b) CV = **$\pm$2.33** (two-tailed)
- (c) CV = **+2.58** (right-tailed)
:::

## Five-Step Hypothesis Testing Procedure

:::{.callout-important}
**Traditional Method**

1. **State** the hypotheses and identify the claim
2. **Find** the critical value(s) from the appropriate table
3. **Compute** the test value
4. **Make** the decision (reject or do not reject $H_0$)
5. **Summarize** the results in plain language
:::

# Section 8-2: z Test for a Mean {background-color="#2c4a6e"}

## The z Test for a Mean

:::{.callout-note}
**Formula**

$$z = \frac{\bar{X} - \mu}{\sigma / \sqrt{n}}$$

**Requirements:**
- Population standard deviation $\sigma$ is **known**
- If n < 30, the population must be approximately normally distributed
:::

Use the standard normal (z) distribution to find critical values and P-values.

## Example 8-3 {#example-8-3}

:::{.callout-warning}
## Days on Dealers' Lots (Traditional)

A researcher claims that cars sit on dealers' lots more than 29 days on average. A random sample of 30 cars gives X̄ = 30.1 days; $\sigma$ = 3.8. Use $\alpha$ = 0.05.
:::

## Example 8-3

:::{.callout-tip}
## Solution

**Step 1:** $H_0$: $\mu$ = 29 and $H_1$: $\mu$ > 29 *(claim; right-tailed)*

**Step 2:** CV = +1.65

**Step 3:**
$$z = \frac{\bar{X} - \mu}{\sigma/\sqrt{n}} = \frac{30.1 - 29}{3.8/\sqrt{30}} \approx 1.59$$

```{r e8-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 30.1; mu <- 29; sigma <- 3.8; n <- 30
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)
```

**Step 4:** Since 1.59 < 1.65 (does not fall in critical region), **do not reject $H_0$**.

**Step 5:** There is not enough evidence to support the claim that the mean time is greater than 29 days.
:::

## Example 8-4 {#example-8-4}

:::{.callout-warning}
## Costs of Men's Athletic Shoes

A researcher claims the average price of men's athletic shoes is less than \$80. A sample of 36 shoes gives $\bar{X}$ = \$75, $\sigma$ = \$19.2. Use $\alpha$ = 0.10.
:::

## Example 8-4

:::{.callout-tip}
## Solution

**Step 1:** $H_0$: $\mu$ = 80 and $H_1$: $\mu$ < 80 *(claim; left-tailed)*

**Step 2:** CV = −1.28

```{r e8-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 75; mu <- 80; sigma <- 19.2; n <- 36
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)
```

**Step 4:** Since −1.56 < −1.28 (falls in critical region), **reject $H_0$**.

**Step 5:** There is enough evidence to support the claim that the average price is less than \$80.
:::

## Example 8-5 {#example-8-5}
:::{.callout-warning}
A researcher claims the average rehabilitation cost for stroke patients differs from $24,672. A sample of 35 patients gives X̄ = $26,343, $\sigma$ = $3251. Use \alpha  = 0.01.$

**Step 1:** $H_0$: $\mu$ = 24672 and $H_1$: $\mu \neq 24672$ *(claim; two-tailed)*

**Step 2:** CV = $\pm$2.58

```{r e8-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 26343; mu <- 24672; sigma <- 3251; n <- 35
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)
```

**Step 4:** Since 3.04 > 2.58 (falls in critical region), **reject $H_0$**.

**Step 5:** There is enough evidence to conclude that the rehabilitation cost differs from $24,672.$

:::

## The P-Value Method

:::{.callout-note}
**P-Value Decision Rule**

- Compute the **P-value**: the probability of obtaining a test statistic at least as extreme as the one computed, assuming $H_0$ is true
- **If P-value $\leq \alpha$ → reject $H_0$**
- **If P-value > $\alpha$ → do not reject $H_0$**

Steps:
1. State hypotheses and identify claim
2. Compute the test value
3. Find the P-value
4. Make the decision
5. Summarize the results
:::

## Example 8-6 {#example-8-6}
:::{.callout-warning}
A researcher claims average tuition exceeds $5700. A sample of 36 schools gives X̄ = $5950, $\sigma$ = $659. Use \alpha  = 0.05.$

**Step 1:** $H_0$: $\mu$ = 5700 and $H_1$: $\mu$ > 5700 *(claim; right-tailed)*

```{r e8-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 5950; mu <- 5700; sigma <- 659; n <- 36
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)

# P-value for right-tailed test
p_val <- 1 - pnorm(z)
round(p_val, 4)
```

**Step 4:** Since P = 0.0114 < $\alpha$ = 0.05, **reject $H_0$**.

**Step 5:** There is enough evidence to support the claim that average tuition exceeds $5700.$

:::

## Example 8-7 {#example-8-7}
:::{.callout-warning}
A researcher claims the average wind speed in a city is 8 mph. A sample of 32 days gives X̄ = 8.2 mph, $\sigma$ = 0.6. Use $\alpha$ = 0.05.

**Step 1:** $H_0$: $\mu$ = 8 *(claim)* and $H_1$: $\mu \neq 8$ *(two-tailed)*

```{r e8-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 8.2; mu <- 8; sigma <- 0.6; n <- 32
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)

# Two-tailed P-value
p_val <- 2 * (1 - pnorm(abs(z)))
round(p_val, 4)
```

**Step 4:** Since P = 0.0593 > $\alpha$ = 0.05, **do not reject $H_0$**.

**Step 5:** There is not enough evidence to reject the claim that the average wind speed is 8 mph.

:::

# Section 8-3: t Test for a Mean {background-color="#2c4a6e"}

## The t Test for a Mean

:::{.callout-note}
**Formula**

$$t = \frac{\bar{X} - \mu}{s / \sqrt{n}}, \quad \text{d.f.} = n - 1$$

**Use when:**
- $\sigma$ is **unknown** (use sample s instead)
- If n < 30, the population must be approximately normally distributed
:::

Critical values come from the **t distribution** with n − 1 degrees of freedom.

## Examples 8-8 through 8-11: Finding t Critical Values {#example-8-8 .smaller}

:::{.callout-warning}
**Problem**
Find the t critical value for each:

- **8-8**: $\alpha$ = 0.05, d.f. = 16, right-tailed
- **8-9**: $\alpha$ = 0.01, d.f. = 22, left-tailed
- **8-10**: $\alpha$ = 0.10, d.f. = 18, two-tailed
- **8-11**: $\alpha$ = 0.05, d.f. = 28, right-tailed
:::
---

```{r e8-8-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
qt(1 - 0.05, df = 16)   # 8-8: right-tailed
qt(0.01, df = 22)        # 8-9: left-tailed
qt(1 - 0.10/2, df = 18) # 8-10: two-tailed (±)
qt(1 - 0.05, df = 28)   # 8-11: right-tailed
```

**Answers:** +1.746 | −2.508 | $\pm$1.734 | +1.701

## Example 8-12 {#example-8-12}

:::{.callout-warning}
## Hospital Infections (Traditional)

A hospital administrator claims the average number of infections per week is 16.3. A sample of 10 weeks gives X̄ = 17.7, s = 1.8. At $\alpha$ = 0.05, is there enough evidence to reject the claim?
:::

## Example 8-12

:::{.callout-tip}
## Solution

**Step 1:** $H_0$: $\mu$ = 16.3 *(claim)* and $H_1$: $\mu \neq 16.3$ *(two-tailed)*

**Step 2:** d.f. = 9; CV = $\pm$2.262

**Step 3:**
$$t = \frac{\bar{X} - \mu}{s/\sqrt{n}} = \frac{17.7 - 16.3}{1.8/\sqrt{10}} \approx 2.46$$

```{r e8-12-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 17.7; mu <- 16.3; s <- 1.8; n <- 10
t <- (xbar - mu) / (s / sqrt(n))
round(t, 2)
qt(1 - 0.05/2, df = n - 1)  # critical value
```

**Step 4:** Since 2.46 > 2.262 (falls in critical region), **reject $H_0$**.

**Step 5:** There is enough evidence to reject the claim that the average number of infections per week is 16.3.
:::

## Example 8-13 {#example-8-13}

:::{.callout-warning}
## Substitute Teachers' Salaries

A researcher claims the average salary for substitute teachers in a region is less than \$60/day. Data: 60, 56, 60, 55, 70, 55, 60, 55. Use $\alpha$ = 0.10.
:::

## Example 8-13

:::{.callout-tip}
## Solution

**Step 1:** $H_0$: $\mu$ = 60 and $H_1$: $\mu$ < 60 *(claim; left-tailed)*

**Step 2:** d.f. = 7; CV = −1.415

```{r e8-13-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
data <- c(60, 56, 60, 55, 70, 55, 60, 55)
n <- length(data)
xbar <- mean(data); s <- sd(data)
t <- (xbar - 60) / (s / sqrt(n))
cat("X-bar:", round(xbar, 2), "  s:", round(s, 2), "  t:", round(t, 3))
qt(0.10, df = 7)  # critical value
```

**Step 4:** Since −0.626 > −1.415 (does not fall in critical region), **do not reject $H_0$**.

**Step 5:** There is not enough evidence to support the claim that the average salary is less than \$60.
:::

## Examples 8-14 & 8-15: P-Value Intervals for t {#example-8-14 .smaller}

:::{.callout-warning}
**Problem**
Determine the P-value interval using Table F (t table):

- **8-14**: t = 2.056, n = 11, right-tailed
- **8-15**: t = 2.983, n = 6, two-tailed
:::

---

```{r e8-14-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# 8-14: right-tailed, df = 10
p14 <- pt(2.056, df = 10, lower.tail = FALSE)
cat("8-14 P-value:", round(p14, 4), " → Interval: 0.025 < P < 0.05\n")

# 8-15: two-tailed, df = 5
p15 <- 2 * pt(2.983, df = 5, lower.tail = FALSE)
cat("8-15 P-value:", round(p15, 4), " → Interval: 0.02 < P < 0.05")
```

:::{.callout-note}
From the t table, locate d.f. row, find where test statistic falls, and read off the corresponding $\alpha$ values as bounds for the P-value.
:::

## Example 8-16 {#example-8-16}

:::{.callout-warning}
## Jogger's Oxygen Uptake (P-Value)

A researcher claims that joggers' maximum oxygen uptake exceeds 36.7 mL/kg/min. A sample of 15 joggers gives X̄ = 40.6, s = 6. Use $\alpha$ = 0.05.
:::

## Example 8-16

:::{.callout-tip}
## Solution

**Step 1:** $H_0$: $\mu$ = 36.7 and $H_1$: $\mu$ > 36.7 *(claim; right-tailed)*

**Step 2:**
$$t = \frac{\bar{X} - \mu}{s/\sqrt{n}} = \frac{40.6 - 36.7}{6/\sqrt{15}} \approx 2.517$$

**Step 3:** d.f. = 14; from Table F, 2.517 falls between 2.145 and 2.624, so **0.01 < P < 0.025**.

```{r e8-16-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 40.6; mu <- 36.7; s <- 6; n <- 15
t <- (xbar - mu) / (s / sqrt(n))
round(t, 3)
p_val <- pt(t, df = n - 1, lower.tail = FALSE)
round(p_val, 4)
```

**Step 4:** Since P $\approx$ 0.0123 < $\alpha$ = 0.05, **reject $H_0$**.

**Step 5:** There is enough evidence to support the claim that joggers' maximum oxygen uptake exceeds 36.7 mL/kg/min.
:::

# Section 8-4: z Test for a Proportion {background-color="#2c4a6e"}

## The z Test for a Proportion

:::{.callout-note}
**Formula**

$$z = \frac{\hat{p} - p}{\sqrt{pq/n}}$$

where $\hat{p} = X/n$ (sample proportion), $p$ = hypothesized population proportion, $q = 1 - p$

**Requirements:**
- Random sample
- Binomial conditions satisfied
- np $\geq 5$ and nq $\geq 5$
:::

## Example 8-17 {#example-8-17}
:::{.callout-warning}
A dietitian claims 60% of people try to avoid trans fats. A survey of 200 people finds 128 trying to avoid them. At $\alpha$ = 0.05, is there enough evidence to reject the claim?

**Step 1:** $H_0$: p = 0.60 *(claim)* and $H_1$: p $\neq 0.60$ *(two-tailed)*

**Step 2:** CV = $\pm$1.96

```{r e8-17-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 128; n <- 200; p <- 0.60; q <- 1 - p
p_hat <- X / n
z <- (p_hat - p) / sqrt(p * q / n)
cat("p-hat:", p_hat, "  z:", round(z, 2))
```

**Step 4:** Since 1.15 is between −1.96 and +1.96, **do not reject $H_0$**.

**Step 5:** There is not enough evidence to reject the claim that 60% of people try to avoid trans fats.

:::

## Example 8-18 {#example-8-18}
:::{.callout-warning}
In 2000, 60% said the Family and Medical Leave Act was very easy to use. A researcher thinks this percentage has changed. A sample of 100 people finds p̂ = 53% found it easy. At $\alpha$ = 0.01, has the percentage changed?

**Step 1:** $H_0$: p = 0.60 and $H_1$: p $\neq 0.60$ *(claim; two-tailed)*

**Step 2:** CV = $\pm$2.58

```{r e8-18-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
p_hat <- 0.53; p <- 0.60; q <- 0.40; n <- 100
z <- (p_hat - p) / sqrt(p * q / n)
round(z, 2)
```

**Step 4:** Since −1.43 is in the noncritical region (between $\pm$2.58), **do not reject $H_0$**.

**Step 5:** There is not enough evidence to support the claim that the percentage has changed.

:::

## Example 8-19 {#example-8-19}
:::{.callout-warning}
A claim states at least 77% of the population opposes replacing $1 bills with $1 coins. A sample of 80 people finds 55 opposed. At $\alpha$ = 0.01, test the claim.

**Step 1:** $H_0$: p = 0.77 *(claim)* and $H_1$: p < 0.77 *(left-tailed)*

**Step 2:** CV = −2.33

```{r e8-19-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 55; n <- 80; p <- 0.77; q <- 1 - p
p_hat <- X / n
z <- (p_hat - p) / sqrt(p * q / n)
cat("p-hat:", round(p_hat, 4), "  z:", round(z, 2))
```

**Step 4:** Since −1.75 > −2.33 (does not fall in critical region), **do not reject $H_0$**.

**Step 5:** There is not enough evidence to reject the claim that at least 77% of the population opposes replacing $1 bills with $1 coins.

:::

## Example 8-20 {#example-8-20}
:::{.callout-warning}
An attorney claims more than 25% of all lawyers advertise. A sample of 200 lawyers finds 63 who advertised. At $\alpha$ = 0.05, is there enough evidence to support the claim?

**Step 1:** $H_0$: p = 0.25 and $H_1$: p > 0.25 *(claim; right-tailed)*

```{r e8-20-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 63; n <- 200; p <- 0.25; q <- 0.75
p_hat <- X / n
z <- (p_hat - p) / sqrt(p * q / n)
cat("p-hat:", round(p_hat, 4), "  z:", round(z, 2))

p_val <- 1 - pnorm(z)
cat("  P-value:", round(p_val, 4))
```

**Step 4:** Since P $\approx 0.017$ < $\alpha$ = 0.05, **reject $H_0$**.

**Step 5:** There is enough evidence to support the claim that more than 25% of lawyers advertise.

:::

# Section 8-5: $\chi^2$ Test for a Variance or Standard Deviation {background-color="#2c4a6e"}

## The Chi-Square Test for Variance

:::{.callout-note}
**Formula**

$$\chi^2 = \frac{(n-1)s^2}{\sigma^2}, \quad \text{d.f.} = n - 1$$

**Assumptions:**
- Random sample
- Population is normally distributed
- Observations are independent

**Three test types:** right-tailed ($\sigma^2$ > value), left-tailed ($\sigma^2$ < value), two-tailed ($\sigma^2 \neq$ value)
:::

## Examples 8-21 through 8-23: Finding $\chi^2$ Critical Values {#example-8-21 .smaller}

:::{.callout-warning}
**Problem**
Find the critical chi-square value(s):

- **8-21**: d.f. = 15, $\alpha$ = 0.05, right-tailed
- **8-22**: d.f. = 10, $\alpha$ = 0.05, left-tailed
- **8-23**: d.f. = 22, $\alpha$ = 0.05, two-tailed
:::
---

```{r e8-21-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# 8-21: right-tailed
qchisq(1 - 0.05, df = 15)

# 8-22: left-tailed (use 1 - α for left tail)
qchisq(0.05, df = 10)

# 8-23: two-tailed (α/2 on each side)
qchisq(1 - 0.025, df = 22)  # right CV
qchisq(0.025, df = 22)       # left CV
```

**Answers:**

- 8-21: **24.996**
- 8-22: **3.940**
- 8-23: **36.781** and **10.982**

## Example 8-24 {#example-8-24}

:::{.callout-warning}
## Variation of Test Scores (Traditional)

An instructor claims the variance in her class of 23 students ($s^2$ = 198) is less than the population variance ($\sigma^2$ = 225). At $\alpha$ = 0.05, is there enough evidence to support the claim?
:::

## Example 8-24

:::{.callout-tip}
## Solution

**Step 1:** $H_0$: $\sigma^2$ = 225 and $H_1$: $\sigma^2$ < 225 *(claim; left-tailed)*

**Step 2:** d.f. = 22; CV = 12.338

**Step 3:**
$$\chi^2 = \frac{(n-1)s^2}{\sigma^2} = \frac{(23-1)(198)}{225} = 19.36$$

```{r e8-24-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 23; s2 <- 198; sigma2 <- 225
chi2 <- (n - 1) * s2 / sigma2
round(chi2, 2)
qchisq(0.05, df = n - 1)  # CV (left-tailed)
```

**Step 4:** Since 19.36 > 12.338 (does not fall in critical region), **do not reject $H_0$**.

**Step 5:** There is not enough evidence to support the claim that the variation in test scores is less than the population variance.
:::

## Example 8-25 {#example-8-25}

:::{.callout-warning}
## Outpatient Surgery

A hospital administrator claims the standard deviation of outpatient surgery patients per day is greater than 8. A random sample of 15 days gives data below. At $\alpha$ = 0.10, test the claim.

Data: 25, 30, 5, 15, 18, 42, 16, 9, 10, 12, 12, 38, 8, 14, 27
:::

## Example 8-25

:::{.callout-tip}
## Solution

**Step 1:** $H_0$: $\sigma$ = 8 and $H_1$: $\sigma$ > 8 *(claim; right-tailed)*

```{r e8-25-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
data <- c(25, 30, 5, 15, 18, 42, 16, 9, 10, 12, 12, 38, 8, 14, 27)
n <- length(data); s <- sd(data)
cat("s =", round(s, 1))

chi2 <- (n - 1) * s^2 / 8^2
cat("  chi² =", round(chi2, 2))

qchisq(1 - 0.10, df = n - 1)  # CV
```

**Step 4:** Since 27.45 > 21.064, **reject $H_0$**.

**Step 5:** There is enough evidence to support the claim that the standard deviation is greater than 8.
:::

## Example 8-26 {#example-8-26}

:::{.callout-warning}
## Nicotine Content of Cigarettes (Traditional)

A manufacturer claims the variance in nicotine content is 0.644. A sample of 20 cigarettes has s = 1.00. At $\alpha$ = 0.05, is there enough evidence to reject the claim?
:::

## Example 8-26

:::{.callout-tip}
## Solution

**Step 1:** $H_0$: $\sigma^2$ = 0.644 *(claim)* and $H_1$: $\sigma^2 \neq 0.644$ *(two-tailed)*

**Step 2:** d.f. = 19; CVs = 8.907 and 32.852

**Step 3:**
$$\chi^2 = \frac{(n-1)s^2}{\sigma^2} = \frac{(20-1)(1.00)^2}{0.644} \approx 29.50$$

```{r e8-26-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 20; s <- 1.00; sigma2 <- 0.644
chi2 <- (n - 1) * s^2 / sigma2
round(chi2, 2)
qchisq(0.025, df = n - 1)      # left CV
qchisq(1 - 0.025, df = n - 1)  # right CV
```

**Step 4:** Since 8.907 < 29.50 < 32.852 (falls in noncritical region), **do not reject $H_0$**.

**Step 5:** There is not enough evidence to reject the claim that the variance of the nicotine content is 0.644.
:::

## Examples 8-27 & 8-28: P-Value Intervals for $\chi^2$ {#example-8-27 .smaller}

:::{.callout-warning}
**Problem**
Find the P-value interval:

- **8-27**: $\chi^2$ = 19.274, n = 8, right-tailed
- **8-28**: $\chi^2$ = 3.823, n = 13, left-tailed
:::
---

```{r e8-27-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# 8-27: right-tailed, df = 7
p27 <- pchisq(19.274, df = 7, lower.tail = FALSE)
cat("8-27 P-value:", round(p27, 4), " → Interval: 0.005 < P < 0.01\n")

# 8-28: left-tailed, df = 12
p28 <- pchisq(3.823, df = 12, lower.tail = TRUE)
cat("8-28 P-value:", round(p28, 4), " → Interval: 0.01 < P < 0.025")
```

## Example 8-29 {#example-8-29}
:::{.callout-warning}
The standard deviation for car inspection time is known to be 16.8 min. A sample of 24 cars gives s = 12.5 min. At $\alpha$ = 0.05, has the standard deviation changed?
:::
---

**Step 1:** $H_0$: $\sigma$ = 16.8 and $H_1$: $\sigma$ $\neq 16.8$ *(claim; two-tailed)*

```{r e8-29-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 24; s <- 12.5; sigma <- 16.8
chi2 <- (n - 1) * s^2 / sigma^2
round(chi2, 3)

# Two-tailed P-value
p_left <- pchisq(chi2, df = n - 1, lower.tail = TRUE)
p_two <- 2 * p_left  # test statistic on left side
cat("P-value ≈", round(p_two, 4), " → Interval: 0.05 < P < 0.10")
```

**Step 4:** Since P > $\alpha$ = 0.05, **do not reject $H_0$**.

**Step 5:** There is not enough evidence to conclude that the standard deviation of inspection time has changed.

# Section 8-6: Additional Topics Regarding Hypothesis Testing {background-color="#2c4a6e"}

## Confidence Intervals and Hypothesis Testing

:::{.callout-note}
**Key Relationship**

For two-tailed hypothesis tests:

- If $H_0$ is **rejected** at $\alpha$ → the (1 − $\alpha$)·100% confidence interval will **NOT** contain the hypothesized value
- If $H_0$ is **not rejected** at $\alpha$ → the (1 − $\alpha$)·100% confidence interval **WILL** contain the hypothesized value

This means a hypothesis test and confidence interval at the same level lead to the **same conclusion**.
:::

## Example 8-30 {#example-8-30}

:::{.callout-warning}
## Sugar Production

Sugar bags are supposed to contain 5 lb. A sample of 50 bags gives X̄ = 4.6 lb, s = 0.7. At $\alpha$ = 0.05, is there evidence the bags differ from 5 lb? Also find the 95% CI.
:::

## Example 8-30

:::{.callout-tip}
## Solution

**Step 1:** $H_0$: $\mu$ = 5 and $H_1$: $\mu \neq 5$ *(claim; two-tailed)*

**Step 2:** d.f. = 49; CV = $\pm$2.010

**Step 3:**
$$t = \frac{\bar{X} - \mu}{s/\sqrt{n}} = \frac{4.6 - 5}{0.7/\sqrt{50}} \approx -4.04$$

```{r e8-30-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 4.6; mu <- 5; s <- 0.7; n <- 50; alpha <- 0.05
t_val <- (xbar - mu) / (s / sqrt(n))
cat("t =", round(t_val, 2))

t_crit <- qt(1 - alpha/2, df = n - 1)
cat("  CV = ±", round(t_crit, 3))

# 95% CI
ME <- t_crit * s / sqrt(n)
cat("  95% CI: (", round(xbar - ME, 1), ",", round(xbar + ME, 1), ")")
```

**Decision:** |−4.04| > 2.010 → **reject $H_0$**. The 95% CI (4.4, 4.8) does **not** contain $\mu$ = 5, confirming the rejection: there is enough evidence that the bags differ from 5 lb.
:::

## Example 8-31 {#example-8-31}

:::{.callout-warning}
## Hog Weights

A researcher claims hogs fed a special diet will average 200 lb. A sample of 10 hogs gives $\bar{X}$ = 198.2, s = 3.3. At $\alpha$ = 0.05, can the claim be rejected? Find the 95% CI.
:::

## Example 8-31

:::{.callout-tip}
## Solution

**Step 1:** $H_0$: $\mu$ = 200 *(claim)* and $H_1$: $\mu \neq 200$ *(two-tailed)*

```{r e8-31-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
xbar <- 198.2; mu <- 200; s <- 3.3; n <- 10; alpha <- 0.05
t_val <- (xbar - mu) / (s / sqrt(n))
cat("t =", round(t_val, 2))

t_crit <- qt(1 - alpha/2, df = n - 1)
cat("  CV = ±", round(t_crit, 3))

# 95% CI
ME <- t_crit * s / sqrt(n)
cat("  95% CI: (", round(xbar - ME, 1), ",", round(xbar + ME, 1), ")")
```

**Decision:** |−1.72| < 2.262 → **do not reject $H_0$**.

The 95% CI (195.8, 200.6) **contains** $\mu$ = 200, confirming we cannot reject the claim.
:::

## Type II Error and the Power of a Test

:::{.callout-note}
**Power of a Test**

$$\text{Power} = 1 - \beta$$

where $\beta$ = P(Type II error) = probability of failing to reject a false $H_0$

**Increasing power:**
- Increase $\alpha$ (raises sensitivity but increases Type I error risk)
- Increase sample size n (reduces standard error, most reliable method)

**Practical note:** Failing to reject $H_0$ does not prove it true — the test may simply lack sufficient power to detect a real difference.
:::

## Summary of Decision Rules

:::{.callout-note}
**Traditional vs. P-Value Method**

**Traditional:** Compare test value to critical value from table

**P-value:** Compare computed P-value to $\alpha$

- **Reject $H_0$** if test value in critical region *or* P-value $\leq \alpha$
- **Do not reject $H_0$** otherwise

**Confidence Interval Method (two-tailed only):** If CI contains the hypothesized value → do not reject $H_0$; if it does not → reject $H_0$
:::

## Key Takeaways

:::{.callout-important}
**Key Tests**

| Situation | Test | Formula |
|---|---|---|
| $\mu$, $\sigma$ known | z test | $z = \frac{\bar{X}-\mu}{\sigma/\sqrt{n}}$ |
| $\mu$, $\sigma$ unknown | t test | $t = \frac{\bar{X}-\mu}{s/\sqrt{n}}$ |
| Proportion p | z test | $z = \frac{\hat{p}-p}{\sqrt{pq/n}}$ |
| Variance $\sigma^2$ | $\chi^2$ test | $\chi^2 = \frac{(n-1)s^2}{\sigma^2}$ |
:::

## Acknowledgement
-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2012). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.