---
title: "Statistics"
subtitle: "Chapter 8: Hypothesis Testing"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "Elementary Statistics: A Step by Step Approach"
logo: "logo.svg"
editor: source
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
```

## Overview

Statistical **hypothesis testing** is a decision-making procedure used to evaluate a claim about a population parameter.

::: {.callout-note}
### Key Vocabulary

- **Statistical hypothesis**: A conjecture about a population parameter
- **Null hypothesis (H₀)**: States no difference or no effect — the status quo
- **Alternative hypothesis (H₁)**: The research claim; what we try to support
- **Test value**: A computed statistic used to decide between H₀ and H₁
:::

## Chapter Sections

- **8-1** Steps in Hypothesis Testing — Traditional Method
- **8-2** z Test for a Mean
- **8-3** t Test for a Mean
- **8-4** z Test for a Proportion
- **8-5** χ² Test for a Variance or Standard Deviation
- **8-6** Additional Topics Regarding Hypothesis Testing

# Section 8-1: Steps in Hypothesis Testing {background-color="#2c4a6e"}

## Null and Alternative Hypotheses

::: {.callout-note}
### Hypothesis Structure

- **H₀** (null hypothesis): Contains equality (=, ≤, ≥)
- **H₁** (alternative/research hypothesis): Contains inequality (≠, <, >)
- The **claim** may be in either H₀ or H₁
:::

The direction of the alternative hypothesis determines the **type of test**:

| H₁ contains | Test type |
|---|---|
| > | Right-tailed |
| < | Left-tailed |
| ≠ | Two-tailed |

## Types of Errors

::: {.callout-important}
### Type I and Type II Errors

| Decision | H₀ True | H₀ False |
|---|---|---|
| Reject H₀ | **Type I error** (α) | Correct decision (1 − β) |
| Do not reject H₀ | Correct decision (1 − α) | **Type II error** (β) |

- **α** = P(Type I error) = level of significance
- **β** = P(Type II error)
- Common values of α: 0.10, 0.05, 0.01
:::

## Critical Region and Critical Value

::: {.callout-note}
### Key Terms

- **Critical (rejection) region**: Range of values that lead to rejecting H₀
- **Noncritical region**: Range of values that lead to not rejecting H₀
- **Critical value**: Boundary separating the two regions
:::

## Example 8-1: Stating Hypotheses {.smaller}

::: {.callout-tip}
### Problem
State H₀ and H₁ for each claim:

**(a)** A researcher claims newborns weigh more than 8.6 lb on average.

**(b)** A manufacturer claims defective disks average fewer than 18 per 1000.

**(c)** A researcher claims background soft music changes exam scores from the mean of 73.
:::

**Solutions:**

**(a)** Claim: μ > 8.6 → H₀: μ = 8.6 and H₁: μ > 8.6 *(right-tailed)*

**(b)** Claim: μ < 18 → H₀: μ = 18 and H₁: μ < 18 *(left-tailed)*

**(c)** Claim: μ ≠ 73 → H₀: μ = 73 and H₁: μ ≠ 73 *(two-tailed)*

## Example 8-2: Finding Critical Values {.smaller}

::: {.callout-tip}
### Problem
Find the critical value(s) for each situation:

**(a)** Left-tailed, α = 0.10
**(b)** Two-tailed, α = 0.02
**(c)** Right-tailed, α = 0.005
:::

```{r echo=T, warning=FALSE}
# (a) Left-tailed, α = 0.10
qnorm(0.10)

# (b) Two-tailed, α = 0.02 → α/2 = 0.01 on each side
qnorm(0.01)  # left; right is +2.326

# (c) Right-tailed, α = 0.005
qnorm(1 - 0.005)
```

---

**Answers:**

- (a) CV = **−1.28** (left-tailed)
- (b) CV = **±2.33** (two-tailed)
- (c) CV = **+2.58** (right-tailed)

## Five-Step Hypothesis Testing Procedure

::: {.callout-important}
### Traditional Method

1. **State** the hypotheses and identify the claim
2. **Find** the critical value(s) from the appropriate table
3. **Compute** the test value
4. **Make** the decision (reject or do not reject H₀)
5. **Summarize** the results in plain language
:::

# Section 8-2: z Test for a Mean {background-color="#2c4a6e"}

## The z Test for a Mean

::: {.callout-note}
### Formula

$$z = \frac{\bar{X} - \mu}{\sigma / \sqrt{n}}$$

**Requirements:**
- Population standard deviation σ is **known**
- If n < 30, the population must be approximately normally distributed
:::

Use the standard normal (z) distribution to find critical values and P-values.

## Example 8-3: Days on Dealers' Lots (Traditional) {.smaller}

::: {.callout-tip}
### Problem
A researcher claims that cars sit on dealers' lots more than 29 days on average. A random sample of 30 cars gives X̄ = 30.1 days; σ = 3.8. Use α = 0.05.
:::

**Step 1:** H₀: μ = 29 and H₁: μ > 29 *(claim; right-tailed)*

**Step 2:** CV = +1.65

**Step 3:** Compute the test value.

```{r echo=T, warning=FALSE}
xbar <- 30.1; mu <- 29; sigma <- 3.8; n <- 30
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)
```

**Step 4:** Since 1.59 < 1.65, **do not reject H₀**.

**Step 5:** There is not enough evidence to support the claim that cars sit more than 29 days.

## Example 8-4: Cost of Men's Athletic Shoes (Traditional) {.smaller}

::: {.callout-tip}
### Problem
A researcher claims the average price of men's athletic shoes is less than $80. A sample of 36 shoes gives X̄ = $75, σ = $19.2. Use α = 0.10.
:::

**Step 1:** H₀: μ = 80 and H₁: μ < 80 *(claim; left-tailed)*

**Step 2:** CV = −1.28

```{r echo=T, warning=FALSE}
xbar <- 75; mu <- 80; sigma <- 19.2; n <- 36
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)
```

**Step 4:** Since −1.56 < −1.28 (falls in critical region), **reject H₀**.

**Step 5:** There is enough evidence to support the claim that the average price is less than $80.

## Example 8-5: Rehabilitation Cost (Traditional) {.smaller}

::: {.callout-tip}
### Problem
A researcher claims the average rehabilitation cost for stroke patients differs from $24,672. A sample of 35 patients gives X̄ = $26,343, σ = $3251. Use α = 0.01.
:::

**Step 1:** H₀: μ = 24672 and H₁: μ ≠ 24672 *(claim; two-tailed)*

**Step 2:** CV = ±2.58

```{r echo=T, warning=FALSE}
xbar <- 26343; mu <- 24672; sigma <- 3251; n <- 35
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)
```

**Step 4:** Since 3.04 > 2.58 (falls in critical region), **reject H₀**.

**Step 5:** There is enough evidence to conclude that the rehabilitation cost differs from $24,672.

## The P-Value Method

::: {.callout-note}
### P-Value Decision Rule

- Compute the **P-value**: the probability of obtaining a test statistic at least as extreme as the one computed, assuming H₀ is true
- **If P-value ≤ α → reject H₀**
- **If P-value > α → do not reject H₀**

Steps:
1. State hypotheses and identify claim
2. Compute the test value
3. Find the P-value
4. Make the decision
5. Summarize the results
:::

## Example 8-6: College Tuition (P-Value) {.smaller}

::: {.callout-tip}
### Problem
A researcher claims average tuition exceeds $5700. A sample of 36 schools gives X̄ = $5950, σ = $659. Use α = 0.05.
:::

**Step 1:** H₀: μ = 5700 and H₁: μ > 5700 *(claim; right-tailed)*

```{r echo=T, warning=FALSE}
xbar <- 5950; mu <- 5700; sigma <- 659; n <- 36
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)

# P-value for right-tailed test
p_val <- 1 - pnorm(z)
round(p_val, 4)
```

**Step 4:** Since P = 0.0113 < α = 0.05, **reject H₀**.

**Step 5:** There is enough evidence to support the claim that average tuition exceeds $5700.

## Example 8-7: Wind Speed (P-Value) {.smaller}

::: {.callout-tip}
### Problem
A researcher claims the average wind speed in a city is 8 mph. A sample of 32 days gives X̄ = 8.2 mph, σ = 0.6. Use α = 0.05.
:::

**Step 1:** H₀: μ = 8 *(claim)* and H₁: μ ≠ 8 *(two-tailed)*

```{r echo=T, warning=FALSE}
xbar <- 8.2; mu <- 8; sigma <- 0.6; n <- 32
z <- (xbar - mu) / (sigma / sqrt(n))
round(z, 2)

# Two-tailed P-value
p_val <- 2 * (1 - pnorm(abs(z)))
round(p_val, 4)
```

**Step 4:** Since P = 0.0588 > α = 0.05, **do not reject H₀**.

**Step 5:** There is not enough evidence to reject the claim that the average wind speed is 8 mph.

# Section 8-3: t Test for a Mean {background-color="#2c4a6e"}

## The t Test for a Mean

::: {.callout-note}
### Formula

$$t = \frac{\bar{X} - \mu}{s / \sqrt{n}}, \quad \text{d.f.} = n - 1$$

**Use when:**
- σ is **unknown** (use sample s instead)
- If n < 30, the population must be approximately normally distributed
:::

Critical values come from the **t distribution** with n − 1 degrees of freedom.

## Examples 8-8 through 8-11: Finding t Critical Values {.smaller}

::: {.callout-tip}
### Problem
Find the t critical value for each:

- **8-8**: α = 0.05, d.f. = 16, right-tailed
- **8-9**: α = 0.01, d.f. = 22, left-tailed
- **8-10**: α = 0.10, d.f. = 18, two-tailed
- **8-11**: α = 0.05, d.f. = 28, right-tailed
:::

---

```{r echo=T, warning=FALSE}
qt(1 - 0.05, df = 16)   # 8-8: right-tailed
qt(0.01, df = 22)        # 8-9: left-tailed
qt(1 - 0.10/2, df = 18) # 8-10: two-tailed (±)
qt(1 - 0.05, df = 28)   # 8-11: right-tailed
```

**Answers:** +1.746 | −2.508 | ±1.734 | +1.701

## Example 8-12: Hospital Infections (Traditional) {.smaller}

::: {.callout-tip}
### Problem
A hospital administrator claims the average number of infections per week is 16.3. A sample of 10 weeks gives X̄ = 17.7, s = 1.8. At α = 0.05, is there enough evidence to reject the claim?
:::

**Step 1:** H₀: μ = 16.3 *(claim)* and H₁: μ ≠ 16.3 *(two-tailed)*

**Step 2:** d.f. = 9; CV = ±2.262

```{r echo=T, warning=FALSE}
xbar <- 17.7; mu <- 16.3; s <- 1.8; n <- 10
t <- (xbar - mu) / (s / sqrt(n))
round(t, 2)
qt(1 - 0.05/2, df = 9)  # critical value
```

**Step 4:** Since 2.46 > 2.262, **reject H₀**.

**Step 5:** There is enough evidence to reject the claim that the average number of infections is 16.3 per week.

## Example 8-13: Substitute Teachers' Salaries (Traditional) {.smaller}

::: {.callout-tip}
### Problem
A researcher claims the average salary for substitute teachers in a region is less than $60/day. Data: 60, 56, 60, 55, 70, 55, 60, 55. Use α = 0.10.
:::

---

**Step 1:** H₀: μ = 60 and H₁: μ < 60 *(claim; left-tailed)*

**Step 2:** d.f. = 7; CV = −1.415

```{r echo=T, warning=FALSE}
data <- c(60, 56, 60, 55, 70, 55, 60, 55)
n <- length(data)
xbar <- mean(data); s <- sd(data)
t <- (xbar - 60) / (s / sqrt(n))
cat("X-bar:", round(xbar, 2), "  s:", round(s, 2), "  t:", round(t, 3))
qt(0.10, df = 7)  # critical value
```

**Step 4:** Since −0.624 > −1.415 (does not fall in critical region), **do not reject H₀**.

**Step 5:** There is not enough evidence to support the claim that the average salary is less than $60.

## Examples 8-14 & 8-15: P-Value Intervals for t {.smaller}

::: {.callout-tip}
### Problem
Determine the P-value interval using Table F (t table):

- **8-14**: t = 2.056, n = 11, right-tailed
- **8-15**: t = 2.983, n = 6, two-tailed
:::

---

```{r echo=T, warning=FALSE}
# 8-14: right-tailed, df = 10
p14 <- pt(2.056, df = 10, lower.tail = FALSE)
cat("8-14 P-value:", round(p14, 4), " → Interval: 0.025 < P < 0.05\n")

# 8-15: two-tailed, df = 5
p15 <- 2 * pt(2.983, df = 5, lower.tail = FALSE)
cat("8-15 P-value:", round(p15, 4), " → Interval: 0.02 < P < 0.05")
```

::: {.callout-note}
From the t table, locate d.f. row, find where test statistic falls, and read off the corresponding α values as bounds for the P-value.
:::

## Example 8-16: Jogger's Oxygen Uptake (P-Value) {.smaller}

::: {.callout-tip}
### Problem
A researcher claims that joggers' maximum oxygen uptake exceeds 36.7 mL/kg/min. A sample of 15 joggers gives X̄ = 40.6, s = 6. Use α = 0.05.
:::

**Step 1:** H₀: μ = 36.7 and H₁: μ > 36.7 *(claim; right-tailed)*

```{r echo=T, warning=FALSE}
xbar <- 40.6; mu <- 36.7; s <- 6; n <- 15
t <- (xbar - mu) / (s / sqrt(n))
round(t, 3)

# P-value
p_val <- pt(t, df = n - 1, lower.tail = FALSE)
round(p_val, 4)
```

**Step 4:** Since P ≈ 0.012 < α = 0.05, **reject H₀**.

**Step 5:** There is enough evidence to support the claim that joggers' maximum oxygen uptake exceeds 36.7 mL/kg/min.

# Section 8-4: z Test for a Proportion {background-color="#2c4a6e"}

## The z Test for a Proportion

::: {.callout-note}
### Formula

$$z = \frac{\hat{p} - p}{\sqrt{pq/n}}$$

where $\hat{p} = X/n$ (sample proportion), $p$ = hypothesized population proportion, $q = 1 - p$

**Requirements:**
- Random sample
- Binomial conditions satisfied
- np ≥ 5 and nq ≥ 5
:::

## Example 8-17: Trans Fats (Traditional, Two-Tailed) {.smaller}

::: {.callout-tip}
### Problem
A dietitian claims 60% of people try to avoid trans fats. A survey of 200 people finds 128 trying to avoid them. At α = 0.05, is there enough evidence to reject the claim?
:::

**Step 1:** H₀: p = 0.60 *(claim)* and H₁: p ≠ 0.60 *(two-tailed)*

**Step 2:** CV = ±1.96

```{r echo=T, warning=FALSE}
X <- 128; n <- 200; p <- 0.60; q <- 1 - p
p_hat <- X / n
z <- (p_hat - p) / sqrt(p * q / n)
cat("p-hat:", p_hat, "  z:", round(z, 2))
```

**Step 4:** Since 1.15 is between −1.96 and +1.96, **do not reject H₀**.

**Step 5:** There is not enough evidence to reject the claim that 60% of people try to avoid trans fats.

## Example 8-18: Family and Medical Leave Act (Traditional) {.smaller}

::: {.callout-tip}
### Problem
In 2000, 60% said the Family and Medical Leave Act was very easy to use. A researcher thinks this percentage has changed. A sample of 100 people finds p̂ = 53% found it easy. At α = 0.01, has the percentage changed?
:::

**Step 1:** H₀: p = 0.60 and H₁: p ≠ 0.60 *(claim; two-tailed)*

**Step 2:** CV = ±2.58

```{r echo=T, warning=FALSE}
p_hat <- 0.53; p <- 0.60; q <- 0.40; n <- 100
z <- (p_hat - p) / sqrt(p * q / n)
round(z, 2)
```

**Step 4:** Since −1.43 is in the noncritical region (between ±2.58), **do not reject H₀**.

**Step 5:** There is not enough evidence to support the claim that the percentage has changed.

## Example 8-19: Replacing $1 Bills with $1 Coins (Traditional) {.smaller}

::: {.callout-tip}
### Problem
A claim states at least 77% of the population opposes replacing $1 bills with $1 coins. A sample of 80 people finds 55 opposed. At α = 0.01, test the claim.
:::

**Step 1:** H₀: p = 0.77 *(claim)* and H₁: p < 0.77 *(left-tailed)*

**Step 2:** CV = −2.33

```{r echo=T, warning=FALSE}
X <- 55; n <- 80; p <- 0.77; q <- 1 - p
p_hat <- X / n
z <- (p_hat - p) / sqrt(p * q / n)
cat("p-hat:", round(p_hat, 4), "  z:", round(z, 2))
```

**Step 4:** Since −1.75 > −2.33 (does not fall in critical region), **do not reject H₀**.

**Step 5:** There is not enough evidence to reject the claim that at least 77% of the population opposes replacing $1 bills with $1 coins.

## Example 8-20: Attorney Advertisements (P-Value) {.smaller}

::: {.callout-tip}
### Problem
An attorney claims more than 25% of all lawyers advertise. A sample of 200 lawyers finds 63 who advertised. At α = 0.05, is there enough evidence to support the claim?
:::

**Step 1:** H₀: p = 0.25 and H₁: p > 0.25 *(claim; right-tailed)*

```{r echo=T, warning=FALSE}
X <- 63; n <- 200; p <- 0.25; q <- 0.75
p_hat <- X / n
z <- (p_hat - p) / sqrt(p * q / n)
cat("p-hat:", round(p_hat, 4), "  z:", round(z, 2))

p_val <- 1 - pnorm(z)
cat("  P-value:", round(p_val, 4))
```

**Step 4:** Since P ≈ 0.017 < α = 0.05, **reject H₀**.

**Step 5:** There is enough evidence to support the claim that more than 25% of lawyers advertise.

# Section 8-5: χ² Test for a Variance or Standard Deviation {background-color="#2c4a6e"}

## The Chi-Square Test for Variance

::: {.callout-note}
### Formula

$$\chi^2 = \frac{(n-1)s^2}{\sigma^2}, \quad \text{d.f.} = n - 1$$

**Assumptions:**
- Random sample
- Population is normally distributed
- Observations are independent

**Three test types:** right-tailed (σ² > value), left-tailed (σ² < value), two-tailed (σ² ≠ value)
:::

## Examples 8-21 through 8-23: Finding χ² Critical Values {.smaller}

::: {.callout-tip}
### Problem
Find the critical chi-square value(s):

- **8-21**: d.f. = 15, α = 0.05, right-tailed
- **8-22**: d.f. = 10, α = 0.05, left-tailed
- **8-23**: d.f. = 22, α = 0.05, two-tailed
:::

---

```{r echo=T, warning=FALSE}
# 8-21: right-tailed
qchisq(1 - 0.05, df = 15)

# 8-22: left-tailed (use 1 - α for left tail)
qchisq(0.05, df = 10)

# 8-23: two-tailed (α/2 on each side)
qchisq(1 - 0.025, df = 22)  # right CV
qchisq(0.025, df = 22)       # left CV
```

**Answers:**

- 8-21: **24.996**
- 8-22: **3.940**
- 8-23: **36.781** and **10.982**

## Example 8-24: Variation of Test Scores (Traditional) {.smaller}

::: {.callout-tip}
### Problem
An instructor claims the variance in her class of 23 students (s² = 198) is less than the population variance (σ² = 225). At α = 0.05, is there enough evidence to support the claim?
:::

**Step 1:** H₀: σ² = 225 and H₁: σ² < 225 *(claim; left-tailed)*

**Step 2:** d.f. = 22; CV = 12.338

```{r echo=T, warning=FALSE}
n <- 23; s2 <- 198; sigma2 <- 225
chi2 <- (n - 1) * s2 / sigma2
round(chi2, 2)

qchisq(0.05, df = n - 1)  # left-tailed critical value
```

**Step 4:** Since 19.36 > 12.338 (falls in noncritical region), **do not reject H₀**.

**Step 5:** There is not enough evidence to support the claim that the class variance is less than 225.

## Example 8-25: Outpatient Surgery (Traditional) {.smaller}

::: {.callout-tip}
### Problem
A hospital administrator claims the standard deviation of outpatient surgery patients per day is greater than 8. A random sample of 15 days gives data below. At α = 0.10, test the claim.

Data: 25, 30, 5, 15, 18, 42, 16, 9, 10, 12, 12, 38, 8, 14, 27
:::

---

**Step 1:** H₀: σ = 8 and H₁: σ > 8 *(claim; right-tailed)*

```{r echo=T, warning=FALSE}
data <- c(25, 30, 5, 15, 18, 42, 16, 9, 10, 12, 12, 38, 8, 14, 27)
n <- length(data); s <- sd(data)
cat("s =", round(s, 1))

chi2 <- (n - 1) * s^2 / 8^2
cat("  chi² =", round(chi2, 2))

qchisq(1 - 0.10, df = n - 1)  # CV
```

**Step 4:** Since 27.44 > 21.064, **reject H₀**.

**Step 5:** There is enough evidence to support the claim that the standard deviation is greater than 8.

## Example 8-26: Nicotine Content of Cigarettes (Traditional) {.smaller}

::: {.callout-tip}
### Problem
A manufacturer claims the variance in nicotine content is 0.644. A sample of 20 cigarettes has s = 1.00. At α = 0.05, is there enough evidence to reject the claim?
:::

**Step 1:** H₀: σ² = 0.644 *(claim)* and H₁: σ² ≠ 0.644 *(two-tailed)*

**Step 2:** d.f. = 19; CV = 32.852 and 8.907

```{r echo=T, warning=FALSE}
n <- 20; s <- 1.00; sigma2 <- 0.644
chi2 <- (n - 1) * s^2 / sigma2
round(chi2, 1)

c(qchisq(0.975, df = 19), qchisq(0.025, df = 19))  # right, left CVs
```

**Step 4:** Since 8.907 < 29.5 < 32.852, **do not reject H₀**.

**Step 5:** There is not enough evidence to reject the manufacturer's claim that the variance is 0.644.

## Examples 8-27 & 8-28: P-Value Intervals for χ² {.smaller}

::: {.callout-tip}
### Problem
Find the P-value interval:

- **8-27**: χ² = 19.274, n = 8, right-tailed
- **8-28**: χ² = 3.823, n = 13, left-tailed
:::

```{r echo=T, warning=FALSE}
# 8-27: right-tailed, df = 7
p27 <- pchisq(19.274, df = 7, lower.tail = FALSE)
cat("8-27 P-value:", round(p27, 4), " → Interval: 0.005 < P < 0.01\n")

# 8-28: left-tailed, df = 12
p28 <- pchisq(3.823, df = 12, lower.tail = TRUE)
cat("8-28 P-value:", round(p28, 4), " → Interval: 0.01 < P < 0.025")
```

## Example 8-29: Car Inspection Times (P-Value) {.smaller}

::: {.callout-tip}
### Problem
The standard deviation for car inspection time is known to be 16.8 min. A sample of 24 cars gives s = 12.5 min. At α = 0.05, has the standard deviation changed?
:::

---

**Step 1:** H₀: σ = 16.8 and H₁: σ ≠ 16.8 *(claim; two-tailed)*

```{r echo=T, warning=FALSE}
n <- 24; s <- 12.5; sigma <- 16.8
chi2 <- (n - 1) * s^2 / sigma^2
round(chi2, 3)

# Two-tailed P-value
p_left <- pchisq(chi2, df = n - 1, lower.tail = TRUE)
p_two <- 2 * p_left  # test statistic on left side
cat("P-value ≈", round(p_two, 4), " → Interval: 0.05 < P < 0.10")
```

**Step 4:** Since P > α = 0.05, **do not reject H₀**.

**Step 5:** There is not enough evidence to conclude that the standard deviation of inspection time has changed.

# Section 8-6: Additional Topics Regarding Hypothesis Testing {background-color="#2c4a6e"}

## Confidence Intervals and Hypothesis Testing

::: {.callout-note}
### Key Relationship

For two-tailed hypothesis tests:

- If H₀ is **rejected** at α → the (1 − α)·100% confidence interval will **NOT** contain the hypothesized value
- If H₀ is **not rejected** at α → the (1 − α)·100% confidence interval **WILL** contain the hypothesized value

This means a hypothesis test and confidence interval at the same level lead to the **same conclusion**.
:::

## Example 8-30: Sugar Production {.smaller}

::: {.callout-tip}
### Problem
Sugar bags are supposed to contain 5 lb. A sample of 50 bags gives X̄ = 4.6 lb, s = 0.7. At α = 0.05, is there evidence the bags differ from 5 lb? Also find the 95% CI.
:::

**Step 1:** H₀: μ = 5 and H₁: μ ≠ 5 *(claim; two-tailed)*

```{r echo=T, warning=FALSE}
xbar <- 4.6; mu <- 5; s <- 0.7; n <- 50; alpha <- 0.05
# t-test (σ unknown)
t_val <- (xbar - mu) / (s / sqrt(n))
cat("t =", round(t_val, 2))

t_crit <- qt(1 - alpha/2, df = n - 1)
cat("  CV =", round(t_crit, 3))

# 95% CI
ME <- t_crit * s / sqrt(n)
cat("  95% CI: (", round(xbar - ME, 2), ",", round(xbar + ME, 2), ")")
```

**Decision:** |−4.04| > 2.010 → **reject H₀**.

The 95% CI is (4.4, 4.8) — does **not** contain μ = 5, confirming the test result.

## Example 8-31: Hog Weights {.smaller}

::: {.callout-tip}
### Problem
A researcher claims hogs fed a special diet will average 200 lb. A sample of 10 hogs gives X̄ = 198.2, s = 3.3. At α = 0.05, can the claim be rejected? Find the 95% CI.
:::

**Step 1:** H₀: μ = 200 *(claim)* and H₁: μ ≠ 200 *(two-tailed)*

```{r echo=T, warning=FALSE}
xbar <- 198.2; mu <- 200; s <- 3.3; n <- 10; alpha <- 0.05
t_val <- (xbar - mu) / (s / sqrt(n))
cat("t =", round(t_val, 2))

t_crit <- qt(1 - alpha/2, df = n - 1)
cat("  CV = ±", round(t_crit, 3))

# 95% CI
ME <- t_crit * s / sqrt(n)
cat("  95% CI: (", round(xbar - ME, 1), ",", round(xbar + ME, 1), ")")
```

**Decision:** |−1.72| < 2.262 → **do not reject H₀**.

The 95% CI (195.8, 200.6) **contains** μ = 200, confirming we cannot reject the claim.

## Type II Error and the Power of a Test

::: {.callout-note}
### Power of a Test

$$\text{Power} = 1 - \beta$$

where β = P(Type II error) = probability of failing to reject a false H₀

**Increasing power:**
- Increase α (raises sensitivity but increases Type I error risk)
- Increase sample size n (reduces standard error, most reliable method)

**Practical note:** Failing to reject H₀ does not prove it true — the test may simply lack sufficient power to detect a real difference.
:::

## Summary of Chapter 8

::: {.callout-important}
### Key Tests

| Situation | Test | Formula |
|---|---|---|
| μ, σ known | z test | $z = \frac{\bar{X}-\mu}{\sigma/\sqrt{n}}$ |
| μ, σ unknown | t test | $t = \frac{\bar{X}-\mu}{s/\sqrt{n}}$ |
| Proportion p | z test | $z = \frac{\hat{p}-p}{\sqrt{pq/n}}$ |
| Variance σ² | χ² test | $\chi^2 = \frac{(n-1)s^2}{\sigma^2}$ |
:::

## Summary of Decision Rules

::: {.callout-note}
### Traditional vs. P-Value Method

**Traditional:** Compare test value to critical value from table

**P-value:** Compare computed P-value to α

- **Reject H₀** if test value in critical region *or* P-value ≤ α
- **Do not reject H₀** otherwise

**Confidence Interval Method (two-tailed only):** If CI contains the hypothesized value → do not reject H₀; if it does not → reject H₀
:::

## Acknowledgements

-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2011). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
