---
title: "Statistics"
subtitle: "Chapter 8: Hypothesis Testing"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
language:
  callout-warning-title: "Problem"
  callout-important-title: "Key point"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch8-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Chapter 8 is about **hypothesis testing** — using sample data to evaluate a claim about a population parameter.

| Section | Topics |
|---------|--------|
| 8-1 | Null and alternative hypotheses; test statistic; type I and II errors; $\alpha$; critical value and region; one- and two-tailed tests; traditional and P-value methods |
| 8-2 | $z$ test for a mean, $\sigma$ known; assumptions; traditional and P-value methods |
| 8-3 | $t$ test for a mean, $\sigma$ unknown; critical $t$ values; P-value intervals; $z$ versus $t$ |

---

## Overview (continued)

| Section | Topics |
|---------|--------|
| 8-4 | $z$ test for a proportion; assumptions $np \geq 5$, $nq \geq 5$ |
| 8-5 | $\chi^2$ test for a variance or standard deviation; critical $\chi^2$ values |
| 8-6 | Confidence intervals and hypothesis testing; type II error, $\beta$, power |

## Chapter Objectives

:::{.callout-important}
## After completing this chapter, you should be able to

1. Understand the definitions used in hypothesis testing.
2. State the null and alternative hypotheses.
3. Find critical values for the $z$ test.
4. State the five steps used in hypothesis testing.
5. Test means when $\sigma$ is known, using the $z$ test.
6. Test means when $\sigma$ is unknown, using the $t$ test.
7. Test proportions, using the $z$ test.
8. Test variances or standard deviations, using the chi-square test.
9. Test hypotheses, using confidence intervals.
10. Explain the relationship between type I and type II errors and the power of a test.
:::

# Section 8-1: Steps in Hypothesis Testing—Traditional Method {background-color="#2c4a6e"}

## Hypothesis Testing

**Statistical hypothesis testing** is a decision-making process for evaluating claims about a population. The researcher defines the population, states the hypotheses, gives the significance level, selects a sample, collects the data, performs the calculations required for the statistical test, and reaches a conclusion.

:::{.callout-important}
## Three Methods Used to Test Hypotheses

1. The **traditional method** — compare the test statistic with a critical value.
2. The **P-value method** — compare the P-value with $\alpha$.
3. The **confidence interval method** — see whether the interval contains the hypothesized value (Section 8-6).

The traditional method has been used since hypothesis testing was formulated; the P-value method became popular with modern computers and statistical calculators.
:::

---

## Statistical Hypotheses

:::{.callout-note}
## Statistical Hypothesis

A **statistical hypothesis** is a conjecture about a population parameter. This conjecture may or may not be true.
:::

:::{.callout-note}
## Null and Alternative Hypotheses

The **null hypothesis**, symbolized by $H_0$, is a statistical hypothesis that states that there is no difference between a parameter and a specific value, or that there is no difference between two parameters.

The **alternative hypothesis**, symbolized by $H_1$, is a statistical hypothesis that states the existence of a difference between a parameter and a specific value, or states that there is a difference between two parameters.
:::

---

## Three Illustrative Situations

**Situation A.** A medical researcher asks whether a new medication changes the pulse rate; the population mean is 82 beats per minute. $H_0: \mu = 82$ and $H_1: \mu \neq 82$ — a **two-tailed** test.

**Situation B.** A chemist invents an additive to increase battery life; the mean lifetime without it is 36 months. $H_0: \mu = 36$ and $H_1: \mu > 36$ — a **right-tailed** test.

**Situation C.** A contractor wishes to lower the average monthly heating bill of \$78 with special insulation. $H_0: \mu = 78$ and $H_1: \mu < 78$ — a **left-tailed** test.

:::{.callout-important}
## Stating the Hypotheses Together

| | Two-tailed test | Right-tailed test | Left-tailed test |
|---|---|---|---|
| Mean | $H_0: \mu = k$, $H_1: \mu \neq k$ | $H_0: \mu = k$, $H_1: \mu > k$ | $H_0: \mu = k$, $H_1: \mu < k$ |
| Proportion | $H_0: p = k$, $H_1: p \neq k$ | $H_0: p = k$, $H_1: p > k$ | $H_0: p = k$, $H_1: p < k$ |

In this book the null hypothesis is **always** stated with an equals sign. Because a researcher generally looks for evidence to *support* a claim, the claim is usually stated as the alternative hypothesis — which is why $H_1$ is sometimes called the **research hypothesis**.
:::

---

## Table 8-1: Hypothesis-Testing Common Phrases

:::{.callout-important}
## Translating Words into Symbols

| $>$ | $<$ |
|---|---|
| Is greater than | Is less than |
| Is above | Is below |
| Is higher than | Is lower than |
| Is longer than | Is shorter than |
| Is bigger than | Is smaller than |
| Is increased | Is decreased or reduced from |

| $=$ | $\neq$ |
|---|---|
| Is equal to | Is not equal to |
| Is no different from | Is different from |
| Has not changed from | Has changed from |
| Is the same as | Is not the same as |
:::

---

## "At Least" and "At Most"

:::{.callout-warning}
## Caution

A claim stated with the words **"at least"** or **"at most"** is a *null* hypothesis.

- "You must be at least 21 years of age" means $\mu \geq 21$, written as $\mu = 21$ for $H_0$; then $H_1: \mu < 21$.
- "A theater can seat at most 115 persons" means $\mu \leq 115$, written as $\mu = 115$ for $H_0$; then $H_1: \mu > 115$.
:::

A claim can be stated as either the null or the alternative hypothesis, but statistical evidence can only **support** the claim when it is the alternative hypothesis, and can only **reject** the claim when it is the null hypothesis.

---

## Example 8-1 {#example-8-1}

:::{.callout-warning}
## Stating the Null and Alternative Hypotheses

State the null and alternative hypotheses for each conjecture.

**(a)** A logistics manager thinks that the average picking time per order at a distribution centre is less than 42 seconds.

**(b)** A hotel group believes that the average length of stay of a foreign guest in Taipei is greater than 2.8 nights.

**(c)** An analyst thinks that the average weekday number of passengers entering Taipei Main Station has changed from 320 thousand.

**(d)** Forty-six percent of online shoppers in Taiwan pay by cash on delivery at a convenience store.

**(e)** More than 35% of the export containers leaving Kaohsiung are bound for Southeast Asia.

**(f)** Less than 58% of food-delivery orders are placed on a weekday.
:::

---

## Example 8-1

:::{.callout-tip}
## Solution

**(a)** $H_0: \mu = 42$ and $H_1: \mu < 42$

**(b)** $H_0: \mu = 2.8$ and $H_1: \mu > 2.8$

**(c)** $H_0: \mu = 320$ and $H_1: \mu \neq 320$

**(d)** $H_0: p = 0.46$ and $H_1: p \neq 0.46$

**(e)** $H_0: p = 0.35$ and $H_1: p > 0.35$

**(f)** $H_0: p = 0.58$ and $H_1: p < 0.58$
:::

---

## Statistical Test and Test Statistic

:::{.callout-note}
## Statistical Test

A **statistical test** uses the data obtained from a sample to make a decision about whether the null hypothesis should be rejected.

The numerical value obtained from a statistical test is called the **test statistic**.
:::

Even when $H_0$ is true, the sample mean will rarely equal the population mean exactly. Either the difference is due to chance, or the sample came from a population whose mean is some other, unknown value. The farther the sample mean is from the hypothesized mean, the more evidence there is for rejecting $H_0$.

---

## Four Possible Outcomes

:::{.callout-important}
## Figure 8-2: Possible Outcomes of a Hypothesis Test

| Decision | $H_0$ true | $H_0$ false |
|---|---|---|
| Reject $H_0$ | **Type I error** | Correct decision |
| Do not reject $H_0$ | Correct decision | **Type II error** |
:::

:::{.callout-note}
## Type I and Type II Errors

A **type I error** occurs if you reject the null hypothesis when it is true.

A **type II error** occurs if you do not reject the null hypothesis when it is false.
:::

---

## Hypothesis Testing and a Jury Trial

The hypotheses are $H_0$: the defendant is innocent; $H_1$: the defendant is not innocent.

:::{.callout-important}
## Figure 8-3: The Results of a Trial

| Decision | $H_0$ true (innocent) | $H_0$ false (not innocent) |
|---|---|---|
| Reject $H_0$ (convict) | Type I error | Correct decision |
| Do not reject $H_0$ (acquit) | Correct decision | Type II error |

A type I error puts an innocent person in jail; a type II error lets a guilty person go free.
:::

The verdict does not *prove* that the defendant did or did not commit the crime — it is based on the evidence presented. Likewise, rejecting or not rejecting $H_0$ proves nothing; the decision is made on the basis of probabilities.

---

## Level of Significance

:::{.callout-note}
## Level of Significance

The **level of significance** is the maximum probability of committing a type I error. This probability is symbolized by $\alpha$. That is, $P(\text{type I error}) = \alpha$.

The probability of a type II error is symbolized by $\beta$: $P(\text{type II error}) = \beta$.
:::

:::{.callout-important}
## Three Conventional Levels

Statisticians generally agree on the **0.10, 0.05 and 0.01** levels. When $\alpha = 0.10$ there is a 10% chance of rejecting a true null hypothesis; when $\alpha = 0.05$, a 5% chance; when $\alpha = 0.01$, a 1% chance.

In most hypothesis-testing situations $\beta$ cannot be easily computed; $\alpha$ and $\beta$ are related in that decreasing one increases the other.
:::

---

## Critical Region, Noncritical Region and Critical Value

:::{.callout-note}
## Three Definitions

The **critical or rejection region** is the range of test statistics that indicates that there is a significant difference and that the null hypothesis should be rejected.

The **noncritical or nonrejection region** is the range of test statistics that indicates that the difference was probably due to chance and that the null hypothesis should not be rejected.

The **critical value** separates the critical region from the noncritical region. The symbol for critical value is C.V.
:::

---

## One-Tailed and Two-Tailed Tests

:::{.callout-note}
## One-Tailed Test

A **one-tailed test** indicates that the null hypothesis should be rejected when the test statistic is in the critical region on one side of the mean. A one-tailed test is either a right-tailed test or a left-tailed test, depending on the direction of the inequality of the alternative hypothesis.
:::

:::{.callout-note}
## Two-Tailed Test

In a **two-tailed test**, the null hypothesis should be rejected when the test statistic is in either of the two critical regions. The critical region must be split into two equal parts, each with area $\alpha/2$.
:::

---

## Figure 8-9: Summary of Hypothesis Testing and Critical Values

:::{.callout-important}
## Critical Values for the z Test

| Test | $\alpha = 0.10$ | $\alpha = 0.05$ | $\alpha = 0.01$ |
|---|---|---|---|
| Left-tailed: $H_0: \mu = k$, $H_1: \mu < k$ | C.V. $= -1.282$ | C.V. $= -1.645$ | C.V. $= -2.326$ |
| Right-tailed: $H_0: \mu = k$, $H_1: \mu > k$ | C.V. $= +1.282$ | C.V. $= +1.645$ | C.V. $= +2.326$ |
| Two-tailed: $H_0: \mu = k$, $H_1: \mu \neq k$ | C.V. $= \pm 1.645$ | C.V. $= \pm 1.960$ | C.V. $= \pm 2.576$ |

Table A-4 in Appendix A is a convenient way to look these values up by hand. Every critical value quoted in this chapter is the value `qnorm()` returns.
:::

---

## Procedure Table: Finding Critical Values for Specific $\alpha$ Values

:::{.callout-important}
## Using Table A-4

**Step 1.** Draw the figure and indicate the appropriate area.

  a. If the test is **left-tailed**, the critical region, with area $\alpha$, is on the left side of the mean.
  b. If the test is **right-tailed**, the critical region, with area $\alpha$, is on the right side of the mean.
  c. If the test is **two-tailed**, $\alpha$ must be divided by 2; half the area is on each side.

**Step 2.**

  a. For a left-tailed test, use the $z$ score that corresponds to the area equivalent to $\alpha$.
  b. For a right-tailed test, use the $z$ score that corresponds to the area equivalent to $1 - \alpha$.
  c. For a two-tailed test, use the $z$ score corresponding to $\alpha/2$ for the (negative) left value, and to $1 - \alpha/2$ for the (positive) right value.
:::

---

## Example 8-2 {#example-8-2}

:::{.callout-warning}
## Finding the Critical Value(s)

Find the critical value(s) for each situation and draw the appropriate figure, showing the critical region. Table A-4 in Appendix A gives the same values by hand.

**(a)** A right-tailed test with $\alpha = 0.02$.

**(b)** A two-tailed test with $\alpha = 0.05$.

**(c)** A left-tailed test with $\alpha = 0.10$.
:::

---

## Example 8-2

:::{.callout-tip}
## Solution

**Step 1 — draw the figure and locate the area.** In (a) the area 0.02 sits in the right tail; in (b) two areas of $\alpha/2 = 0.025$ sit in the two tails; in (c) the area 0.10 sits in the left tail.

**Step 2 — read the $z$ score.** In R, `qnorm()` plays the role of Table A-4.

```{r e8-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
alpha_right <- 0.02
alpha_two   <- 0.05
alpha_left  <- 0.10

round(qnorm(1 - alpha_right), 2)
round(qnorm(c(alpha_two / 2, 1 - alpha_two / 2)), 2)
round(qnorm(alpha_left), 2)
```

**Answers.** (a) $+2.05$; (b) $-1.96$ and $+1.96$; (c) $-1.28$.
:::

---

## Critical and Noncritical Regions in R

**Draw figure** — the right-tailed case of Figure 8-5, with $\alpha = 0.01$.

```{r s8-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
critical_value <- qnorm(0.99)

normal_curve <- tibble(z = seq(-4, 4, length.out = 400)) |>
  mutate(density = dnorm(z))

region_plot <- ggplot(normal_curve, aes(z, density)) +
  geom_area(data = filter(normal_curve, z >= critical_value),
            fill = "firebrick", alpha = 0.6) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = critical_value, linetype = "dashed") +
  annotate("text", x =  3.0, y = 0.03, label = "critical region 0.01") +
  annotate("text", x = -0.4, y = 0.15, label = "noncritical region 0.99") +
  labs(title = "Critical and Noncritical Regions, Right-Tailed Test",
       x = "z", y = "Density") +
  theme_minimal()
```

---

## Critical and Noncritical Regions in R

**Output figure**

```{r s8-1-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
region_plot
```

The shaded area to the right of the critical value $z = 2.326$ is the **critical (rejection) region**; everything to the left is the **noncritical (nonrejection) region**.

---

## The Five Steps of Hypothesis Testing

In general a study proceeds by (a) stating the hypotheses, (b) designing the study, (c) conducting the study and collecting the data, (d) evaluating the data, and (e) summarizing the results. Because designing the study and collecting the data are omitted here, a simplified version is used.

:::{.callout-important}
## Procedure Table: Solving Hypothesis-Testing Problems (Traditional Method)

**Step 1.** State the hypotheses and identify the claim.

**Step 2.** Find the critical value(s) from the appropriate table in Appendix A.

**Step 3.** Compute the test statistic.

**Step 4.** Make the decision to reject or not reject the null hypothesis.

**Step 5.** Summarize the results.
:::

---

## Summarizing the Results

:::{.callout-important}
## Wording the Conclusion

| Decision | Claim is $H_0$ | Claim is $H_1$ |
|---|---|---|
| Reject $H_0$ | There is enough evidence to **reject** the claim. | There is enough evidence to **support** the claim. |
| Do not reject $H_0$ | There is not enough evidence to **reject** the claim. | There is not enough evidence to **support** the claim. |

Nothing is proved true or false. The statistician only states that there is, or is not, enough evidence to say that a claim is probably true or false.
:::

---

## The P-Value Method for Hypothesis Testing

:::{.callout-note}
## P-Value

The **P-value** (or probability value) is the probability of getting a sample statistic (such as the sample mean) or a more extreme sample statistic in the direction of the alternative hypothesis when the null hypothesis is true.
:::

The P-value is the actual area under the curve of the appropriate distribution representing the probability of the observed sample statistic, or a more extreme one, occurring if $H_0$ is true.

---

## Comparing $\alpha$ Values and P-Values

Suppose $H_1: \mu > 50$ and a sample gives $\bar{X} = 52$. If software reports a P-value of 0.0356, then the probability of getting a sample mean of 52 or greater is 0.0356 when the true population mean is 50. For $P = 0.0356$, $H_0$ would be rejected at $\alpha = 0.05$ but **not** at $\alpha = 0.01$.

:::{.callout-warning}
## Caution

When the test is two-tailed, the area in one tail must be **doubled**. If the area in one tail is 0.0356, the P-value is $2(0.0356) = 0.0712$, so $H_0$ would not be rejected at $\alpha = 0.05$.

Most calculators and computers report the exact two-tailed P-value already, so it should **not** be doubled again.
:::

---

## Decision Rule When Using a P-Value

:::{.callout-important}
## The Rule

If P-value $\leq \alpha$, **reject** the null hypothesis.

If P-value $> \alpha$, **do not reject** the null hypothesis.
:::

The $\alpha$ value is chosen by the researcher **before** the statistical test is conducted; the P-value is computed **after** the sample statistic has been found.

# Section 8-2: z Test for a Mean {background-color="#2c4a6e"}

## The General Form of a Test Statistic

:::{.callout-important}
## General Formula

$$\text{Test statistic} = \frac{(\text{observed value}) - (\text{expected value})}{\text{standard error}}$$

The **observed value** is the statistic computed from the sample data; the **expected value** is the parameter you would expect if $H_0$ were true; the denominator is the standard error of the statistic being tested.
:::

---

## The z Test

:::{.callout-note}
## z Test

The **$z$ test** is a statistical test for the mean of a population. It can be used either when $n \geq 30$ or when the population is normally distributed and $\sigma$ is known. The formula for the $z$ test statistic is

$$z = \frac{\bar{X} - \mu}{\sigma / \sqrt{n}}$$

where $\bar{X}$ is the sample mean, $\mu$ the hypothesized population mean, $\sigma$ the population standard deviation and $n$ the sample size.
:::

---

## Assumptions and Rounding

:::{.callout-important}
## Assumptions for the z Test for a Mean When $\sigma$ Is Known

1. The sample is a **random sample**.
2. Either $n \geq 30$, or the population is normally distributed when $n < 30$.
:::

:::{.callout-warning}
## Caution

Test statistics are **reported** to two or three decimal places so that they are easy to read, but every decision in this chapter is made with the full-precision value R computes. Never round an intermediate result and then keep calculating with the rounded number.
:::

---

## Example 8-3 {#example-8-3}

:::{.callout-warning}
## Basket Size at a Convenience-Store Chain

A category manager believes that the average basket size at her convenience-store chain is now more than the NT\$118 reported for last year. Assume the population standard deviation is NT\$12. A random sample of 35 transactions has a mean of NT\$122.5 *(hypothetical data)*. Test the claim at $\alpha = 0.05$.
:::

---

## Example 8-3

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \mu = 118$ and $H_1: \mu > 118$ *(claim; right-tailed)*.

**Step 2.** Since $\alpha = 0.05$ and the test is right-tailed, the critical value is $z = +1.645$.

**Step 3 — compute the test statistic.**

```{r e8-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_mean <- 122.5
hypothesized_mean <- 118
population_sd <- 12
sample_size <- 35

test_statistic <- (sample_mean - hypothesized_mean) /
  (population_sd / sqrt(sample_size))

round(c(z = test_statistic, critical_value = qnorm(0.95)), 3)
```

**Step 4.** Since $2.219$ is greater than the critical value $1.645$, the decision is to **reject $H_0$**.

**Step 5.** There is enough evidence to support the claim that the average basket size is greater than NT\$118.
:::

---

## Example 8-4 {#example-8-4}

:::{.callout-warning}
## Island-Wide Parcel Delivery Time

An e-commerce platform states that island-wide parcels arrive on average 48 hours after dispatch. The variable is normally distributed and the population standard deviation is 9 hours. A category manager wishes to see whether parcels in her category arrive significantly faster. She randomly selected 36 parcels and recorded the delivery time in hours, as shown *(hypothetical data)*. At $\alpha = 0.10$, is there enough evidence to support the claim?

```
46   52   44   49   41   55
50   43   48   47   39   53
45   51   42   56   38   49
72   44   47   50   46   43
26   48   52   45   49   47
80   41   44   50   30   48
```
:::

---

## Example 8-4

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \mu = 48$ and $H_1: \mu < 48$ *(claim; left-tailed)*.

**Step 2.** Since $\alpha = 0.10$ and the test is left-tailed, the critical value is $-1.282$.

**Step 3 — prepare data and compute the test statistic.**

```{r e8-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
delivery_hours <- c(46, 52, 44, 49, 41, 55,
                    50, 43, 48, 47, 39, 53,
                    45, 51, 42, 56, 38, 49,
                    72, 44, 47, 50, 46, 43,
                    26, 48, 52, 45, 49, 47,
                    80, 41, 44, 50, 30, 48)

hypothesized_mean <- 48
population_sd <- 9
sample_mean <- mean(delivery_hours)
test_statistic <- (sample_mean - hypothesized_mean) /
  (population_sd / sqrt(length(delivery_hours)))

round(c(mean = sample_mean, z = test_statistic,
        critical_value = qnorm(0.10)), 3)
```
:::

---

## Example 8-4

:::{.callout-tip}
## Solution

**Step 4.** The test statistic $-0.333$ falls in the noncritical region, so the decision is to **not reject $H_0$**.

**Step 5.** There is not enough evidence to support the claim that parcels in this category arrive faster than the platform average.

**Comment.** Even though the sample mean 47.5 hours is lower than the hypothesized mean 48 hours, it is not *significantly* lower; the difference may be due to chance. When $H_0$ is not rejected, a type II error is still possible, and $H_0$ cannot be accepted as true — there is merely not enough evidence to say that it is false.
:::

---

## Example 8-5 {#example-8-5}

:::{.callout-warning}
## Checked Baggage on a Regional Route

An airline's operations report states that checked bags on its Taipei-Tokyo route weigh 18.0 kilograms on average. A load planner selects a random sample of 40 checked bags and finds a mean weight of 19.1 kilograms *(hypothetical data)*. Assume the population standard deviation is 3.0 kilograms. At $\alpha = 0.01$, can it be concluded that the mean weight is not equal to 18.0 kilograms?
:::

---

## Example 8-5

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \mu = 18$ and $H_1: \mu \neq 18$ *(claim; two-tailed)*.

**Step 2.** Since $\alpha = 0.01$ and the test is two-tailed, the critical values are $+2.576$ and $-2.576$.

**Step 3 — compute the test statistic.**

```{r e8-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_mean <- 19.1
hypothesized_mean <- 18
population_sd <- 3.0
sample_size <- 40

test_statistic <- (sample_mean - hypothesized_mean) /
  (population_sd / sqrt(sample_size))

round(c(z = test_statistic, lower_cv = qnorm(0.005),
        upper_cv = qnorm(0.995)), 3)
```

**Step 4.** The test statistic $2.319$ falls in the noncritical region, so **do not reject $H_0$**.

**Step 5.** There is not enough evidence to support the claim that the average weight of a checked bag is not equal to 18.0 kilograms.
:::

---

## Procedure Table: The P-Value Method

:::{.callout-important}
## Solving Hypothesis-Testing Problems (P-Value Method)

**Step 1.** State the hypotheses and identify the claim.

**Step 2.** Compute the test statistic.

**Step 3.** Find the P-value.

**Step 4.** Make the decision.

**Step 5.** Summarize the results.

Recall the decision rule: if P-value $\leq \alpha$, reject $H_0$; if P-value $> \alpha$, do not reject $H_0$.
:::

---

## Example 8-6 {#example-8-6}

:::{.callout-warning}
## Monthly Revenue of a Coffee-Chain Store

A franchise consultant wishes to test the claim that the average monthly revenue of a single coffee-chain store is greater than NT\$860 thousand. She selects a random sample of 36 stores and finds the mean to be NT\$902 thousand *(hypothetical data)*. The population standard deviation is NT\$118 thousand. Is there evidence to support the claim at $\alpha = 0.05$? Use the P-value method.
:::

---

## Example 8-6

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \mu = 860$ and $H_1: \mu > 860$ *(claim; right-tailed)*.

**Step 2 and Step 3 — compute the test statistic and the P-value.**

```{r e8-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_mean <- 902
hypothesized_mean <- 860
population_sd <- 118
sample_size <- 36

test_statistic <- (sample_mean - hypothesized_mean) /
  (population_sd / sqrt(sample_size))
p_value <- pnorm(test_statistic, lower.tail = FALSE)

round(c(z = test_statistic, p_value = p_value), 4)
```

**Step 4.** Since the P-value $0.0164$ is less than 0.05, **reject $H_0$**.

**Step 5.** There is enough evidence to support the claim that the average monthly revenue of a store is greater than NT\$860 thousand. Had $\alpha = 0.01$ been chosen, $H_0$ would not have been rejected.
:::

---

## Example 8-7 {#example-8-7}

:::{.callout-warning}
## Courier Pick-Up Time

A food-delivery platform claims that the average time a courier needs to pick up an order from a restaurant is 12 minutes. A sample of 32 pick-ups has an average time of 12.6 minutes *(hypothetical data)*. The standard deviation of the population is 1.9 minutes. At $\alpha = 0.05$, is there enough evidence to reject the claim? Use the P-value method.
:::

---

## Example 8-7

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \mu = 12$ *(claim)* and $H_1: \mu \neq 12$ *(two-tailed)*.

**Step 2 and Step 3 — compute the test statistic and the P-value.**

```{r e8-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_mean <- 12.6
hypothesized_mean <- 12
population_sd <- 1.9
sample_size <- 32

test_statistic <- (sample_mean - hypothesized_mean) /
  (population_sd / sqrt(sample_size))
p_value <- 2 * pnorm(test_statistic, lower.tail = FALSE)

round(c(z = test_statistic, p_value = p_value), 4)
```

**Step 4.** The decision is to **not reject $H_0$**, since the P-value $0.0740$ is greater than 0.05.

**Step 5.** There is not enough evidence to reject the claim that the average pick-up time is 12 minutes.
:::

---

## Guidelines for P-Values

:::{.callout-important}
## When No $\alpha$ Level Is Chosen

Some researchers report the P-value and let the reader decide. These guidelines can then be used.

- If P-value $\leq 0.01$, reject the null hypothesis. The difference is **highly significant**.
- If P-value $> 0.01$ but P-value $\leq 0.05$, reject the null hypothesis. The difference is **significant**.
- If P-value $> 0.05$ but P-value $\leq 0.10$, consider the consequences of a type I error before rejecting the null hypothesis.
:::

# Section 8-3: t Test for a Mean {background-color="#2c4a6e"}

## The $t$ Distribution

When the population standard deviation is unknown, the $z$ test is not typically used for testing hypotheses involving means; the **$t$ test** is used instead, and the distribution of the variable should be approximately normal.

:::{.callout-note}
## How the $t$ Distribution Compares with the Standard Normal

**Similarities.** It is bell-shaped; it is symmetric about the mean; the mean, median and mode are equal to 0 and are located at the centre of the distribution; the curve approaches but never touches the $x$ axis.

**Differences.** The variance is greater than 1; it is a family of curves based on the degrees of freedom; as the sample size increases, the $t$ distribution approaches the normal distribution.
:::

---

## The t Test

:::{.callout-note}
## t Test

The **$t$ test** is a statistical test for the mean of a population and is used when the population is normally or approximately normally distributed and $\sigma$ is unknown. The formula for the $t$ test statistic is

$$t = \frac{\bar{X} - \mu}{s / \sqrt{n}}$$

The degrees of freedom are $\text{d.f.} = n - 1$.
:::

The formula is the same as for the $z$ test, except that the sample standard deviation $s$ replaces the unknown $\sigma$.

---

## Reading Table A-5

:::{.callout-warning}
## Caution

Table A-5 gives degrees of freedom for values from 1 through 30, then at intervals above 30. When a critical value is looked up by hand and the degrees of freedom are above 30, **round down** to the nearest table value — for example, if $\text{d.f.} = 59$, read $\text{d.f.} = 55$. This is a conservative approach.

As the degrees of freedom get larger, the critical values approach the $z$ scores. In R, `qt()` returns the critical value for the **exact** degrees of freedom, and that is the value quoted on every slide in this chapter; the round-down rule is needed only when the table is read by hand.
:::

---

## Example 8-8 {#example-8-8}

:::{.callout-warning}
## Critical t Value, Right-Tailed

Find the critical $t$ value for $\alpha = 0.05$ with $\text{d.f.} = 19$ for a right-tailed $t$ test.
:::

---

## Example 8-8

:::{.callout-tip}
## Solution

Find the 0.05 column in the top row labelled *One tail* and 19 in the left-hand column. The critical value is $+1.729$.

```{r e8-8-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(qt(1 - 0.05, df = 19), 3)
```
:::

---

## Example 8-9 {#example-8-9}

:::{.callout-warning}
## Critical t Value, Left-Tailed

Find the critical $t$ value for $\alpha = 0.01$ with $\text{d.f.} = 17$ for a left-tailed $t$ test.
:::

---

## Example 8-9

:::{.callout-tip}
## Solution

Find the critical value in the 0.01 column in the row labelled *One tail*, and find 17 in the left column. The critical value is $-2.567$.

```{r e8-9-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(qt(0.01, df = 17), 3)
```
:::

---

## Example 8-10 {#example-8-10}

:::{.callout-warning}
## Critical t Values, Two-Tailed

Find the critical values for $\alpha = 0.05$ with $\text{d.f.} = 23$ for a two-tailed test.
:::

---

## Example 8-10

:::{.callout-tip}
## Solution

Find the 0.05 column in the row labelled *Two tails* and 23 in the column labelled d.f. The critical values are $-2.069$ and $+2.069$.

```{r e8-10-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(qt(c(0.05 / 2, 1 - 0.05 / 2), df = 23), 3)
```
:::

---

## Example 8-11 {#example-8-11}

:::{.callout-warning}
## Critical t Value, Right-Tailed

Find the critical value for $\alpha = 0.10$ with $\text{d.f.} = 12$ for a right-tailed test.
:::

---

## Example 8-11

:::{.callout-tip}
## Solution

Find the critical value in the 0.10 column in the *One tail* row and 12 in the d.f. column. The critical value is $+1.356$.

```{r e8-11-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(qt(1 - 0.10, df = 12), 3)
```
:::

---

## Assumptions for the t Test

:::{.callout-important}
## Assumptions for the t Test for a Mean When $\sigma$ Is Unknown

1. The sample is a **random sample**.
2. Either $n \geq 30$, or the population is normally distributed when $n < 30$.
:::

:::{.callout-important}
## Traditional Method with the t Test

The same five steps are used as for the $z$ test, except that critical values come from **Table A-5**: state the hypotheses and identify the claim; find the critical value(s); compute the test statistic; make the decision; summarize the results.
:::

---

## Example 8-12 {#example-8-12}

:::{.callout-warning}
## Call Handling Time

A customer-service centre claims that the average time an agent needs to handle one call is 4.5 minutes. A random sample of 12 calls had a mean handling time of 5.3 minutes with a standard deviation of 1.1 minutes *(hypothetical data)*. Is there enough evidence to reject the claim at $\alpha = 0.05$? Assume the variable is normally distributed.
:::

---

## Example 8-12

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \mu = 4.5$ *(claim)* and $H_1: \mu \neq 4.5$ *(two-tailed)*.

**Step 2.** The critical values are $+2.201$ and $-2.201$ for $\alpha = 0.05$ and $\text{d.f.} = 11$.

**Step 3 — compute the test statistic.**

```{r e8-12-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_mean <- 5.3
hypothesized_mean <- 4.5
sample_sd <- 1.1
sample_size <- 12

test_statistic <- (sample_mean - hypothesized_mean) /
  (sample_sd / sqrt(sample_size))

round(c(t = test_statistic,
        critical_value = qt(1 - 0.05 / 2, df = sample_size - 1)), 3)
```

**Step 4.** **Reject $H_0$** since $2.519 > 2.201$.

**Step 5.** There is enough evidence to reject the claim that the average call handling time is 4.5 minutes.
:::

---

## Example 8-13 {#example-8-13}

:::{.callout-warning}
## Days to Payment on Export Invoices

An industry report states that small exporters wait on average 27 days to be paid. A trade-finance analyst believes that the exporters using her platform are paid in less time. A random sample of 10 invoices was selected and the number of days to payment is shown *(hypothetical data)*. Test the claim at $\alpha = 0.10$. Assume the variable is normally distributed.

```
24   31   19   27   22   30   18   25   21   26
```
:::

---

## Example 8-13

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \mu = 27$ and $H_1: \mu < 27$ *(claim; left-tailed)*.

**Step 2.** At $\alpha = 0.10$ and $\text{d.f.} = 9$, the critical value is $-1.383$.

**Step 3 — prepare data and compute the test statistic.** With the raw data in hand, `t.test()` does the arithmetic.

```{r e8-13-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
days_to_payment <- c(24, 31, 19, 27, 22, 30, 18, 25, 21, 26)

payment_test <- t.test(days_to_payment, mu = 27, alternative = "less")

round(c(mean = mean(days_to_payment), sd = sd(days_to_payment)), 3)
round(c(t = unname(payment_test$statistic),
        critical_value = qt(0.10, df = length(days_to_payment) - 1)), 3)
```

**Step 4.** **Reject $H_0$** since $-1.953$ falls in the critical region.

**Step 5.** There is enough evidence to support the claim that exporters on the platform are paid in fewer than 27 days.
:::

---

## P-Value Intervals for the t Test

P-values for $t$ tests cannot be read directly from Table A-5, since only selected values of $\alpha$ are given. Only an **interval** for the P-value can be found from the table: locate the row for the degrees of freedom, find the two table values the test statistic falls between, and read the corresponding $\alpha$ values from the *One tail* row (one-tailed test) or the *Two tails* row (two-tailed test).

:::{.callout-warning}
## Caution

R's `pt()` gives the exact P-value. Table A-5 gives only an interval, and the exact value always falls inside that interval. Every decision on these slides is made with the exact value.
:::

---

## Example 8-14 {#example-8-14}

:::{.callout-warning}
## P-Value Interval, Right-Tailed

Find the P-value when the $t$ test statistic is 2.462, the sample size is 14, and the test is right-tailed.
:::

---

## Example 8-14

:::{.callout-tip}
## Solution

Across the row with $\text{d.f.} = 13$, the value 2.462 falls between 2.160 and 2.650, which correspond to $\alpha = 0.025$ and $\alpha = 0.01$ in the *One tail* row. Hence

$$0.01 < P\text{-value} < 0.025$$

```{r e8-14-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(pt(2.462, df = 13, lower.tail = FALSE), 4)
```

The exact P-value is $0.0143$, which lies inside the interval read from the table. If $\alpha$ were 0.05, $H_0$ would be rejected; if $\alpha$ were 0.01, it would not.
:::

---

## Example 8-15 {#example-8-15}

:::{.callout-warning}
## P-Value Interval, Two-Tailed

Find the P-value when the $t$ test statistic is 2.575, the sample size is 9, and the test is two-tailed.
:::

---

## Example 8-15

:::{.callout-tip}
## Solution

Across the row with $\text{d.f.} = 8$, the value 2.575 falls between 2.306 and 2.896, which correspond to $\alpha = 0.05$ and $\alpha = 0.02$ in the *Two tails* row. Hence

$$0.02 < P\text{-value} < 0.05$$

```{r e8-15-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(2 * pt(2.575, df = 8, lower.tail = FALSE), 4)
```

The exact P-value is $0.0329$. At $\alpha = 0.05$ the null hypothesis can be rejected; at $\alpha = 0.01$ it cannot.
:::

---

## Example 8-16 {#example-8-16}

:::{.callout-warning}
## Containers Shipped by Programme Members

A trade association claims that firms which joined its export-promotion programme ship more containers per month than the industry average of 24. A random sample of 15 member firms has a mean of 27.8 containers per month and a standard deviation of 6.2 containers *(hypothetical data)*. Is there enough evidence to support the claim at $\alpha = 0.05$? Assume the variable is normally distributed. Use the P-value method.
:::

---

## Example 8-16

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \mu = 24$ and $H_1: \mu > 24$ *(claim; right-tailed)*.

**Step 2 and Step 3 — compute the test statistic and the P-value.**

```{r e8-16-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_mean <- 27.8
hypothesized_mean <- 24
sample_sd <- 6.2
sample_size <- 15

test_statistic <- (sample_mean - hypothesized_mean) /
  (sample_sd / sqrt(sample_size))
p_value <- pt(test_statistic, df = sample_size - 1, lower.tail = FALSE)

round(c(t = test_statistic, p_value = p_value), 4)
```

From Table A-5 with $\text{d.f.} = 14$, the value 2.3738 falls between 2.145 and 2.624, so $0.01 < P\text{-value} < 0.025$; the exact P-value is $0.0162$.

**Step 4.** **Reject $H_0$** since P-value $< \alpha$.

**Step 5.** There is enough evidence to support the claim that member firms ship more than 24 containers per month.
:::

---

## Choosing Between the z Test and the t Test

:::{.callout-important}
## Figure 8-23: The Rules

1. If $\sigma$ is **known**, use the $z$ test. The variable must be normally distributed if $n < 30$.
2. If $\sigma$ is **unknown** but $n \geq 30$, use the $t$ test.
3. If $\sigma$ is **unknown** and $n < 30$, use the $t$ test; the population must be approximately normally distributed.

These are the same rules that were used for confidence intervals in Chapter 7.
:::

# Section 8-4: z Test for a Proportion {background-color="#2c4a6e"}

## Testing a Proportion

Many hypothesis-testing situations involve **proportions**; a proportion is the same as a percentage of the population. A hypothesis test involving a population proportion can be considered a binomial experiment when there are only two outcomes and the probability of a success does not change from trial to trial.

Since a normal distribution can be used to approximate the binomial distribution when $np \geq 5$ and $nq \geq 5$, the standard normal distribution can be used to test hypotheses for proportions.

---

## Formula for the z Test for Proportions

:::{.callout-important}
## The Test Statistic

$$z = \frac{\hat{p} - p}{\sqrt{pq/n}}$$

where $\hat{p} = X/n$ is the sample proportion, $p$ is the hypothesized population proportion, $q = 1 - p$ and $n$ is the sample size.

The formula follows the general form: $\hat{p}$ is the observed value, $p$ the expected value, and $\sqrt{pq/n}$ the standard error. It is derived from the normal approximation to the binomial by substituting $\mu = np$ and $\sigma = \sqrt{npq}$ and dividing numerator and denominator by $n$.
:::

---

## Assumptions for Testing a Proportion

:::{.callout-important}
## Three Assumptions

1. The sample is a **random sample**.
2. The conditions for a **binomial experiment** are satisfied.
3. $np \geq 5$ **and** $nq \geq 5$.
:::

Table A-4 is used to find both critical values and P-values. Examples 8-17 to 8-19 use the traditional method; Example 8-20 uses the P-value method.

---

## Example 8-17 {#example-8-17}

:::{.callout-warning}
## Mobile-Wallet Payments

A market-research note claims that 30% of online shoppers in Taiwan pay with a mobile wallet. To test the claim, a student randomly selected 120 online shoppers and found that 42 of them paid with a mobile wallet *(hypothetical data)*. At $\alpha = 0.05$, is there enough evidence to reject the claim?
:::

---

## Example 8-17

:::{.callout-tip}
## Solution

**Step 1.** $H_0: p = 0.30$ *(claim)* and $H_1: p \neq 0.30$ *(two-tailed)*.

**Step 2.** Since $\alpha = 0.05$ and the test is two-tailed, the critical values are $\pm 1.960$.

**Step 3 — compute the test statistic.**

```{r e8-17-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
successes <- 42
sample_size <- 120
hypothesized_p <- 0.30

sample_proportion <- successes / sample_size
standard_error <- sqrt(hypothesized_p * (1 - hypothesized_p) / sample_size)
test_statistic <- (sample_proportion - hypothesized_p) / standard_error

round(c(p_hat = sample_proportion, z = test_statistic,
        critical_value = qnorm(0.975)), 3)
```

**Step 4.** **Do not reject $H_0$**, since the test statistic $1.195$ falls in the noncritical region.

**Step 5.** There is not enough evidence to reject the claim that 30% of online shoppers pay with a mobile wallet.
:::

---

## Example 8-18 {#example-8-18}

:::{.callout-warning}
## Night-Market Stalls Accepting Mobile Payment

An industry survey reports that 40% of night-market stalls accept mobile payment. A researcher believes that the percentage is lower at one particular night market. She randomly selected 60 stalls there and found that 20 accept mobile payment *(hypothetical data)*. At $\alpha = 0.01$, is there enough evidence to support the claim?
:::

---

## Example 8-18

:::{.callout-tip}
## Solution

**Step 1.** $H_0: p = 0.40$ and $H_1: p < 0.40$ *(claim; left-tailed)*.

**Step 2.** Since $\alpha = 0.01$ and the test is one-tailed, the critical value is $-2.326$.

**Step 3 — compute the test statistic.**

```{r e8-18-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
successes <- 20
sample_size <- 60
hypothesized_p <- 0.40

sample_proportion <- successes / sample_size
standard_error <- sqrt(hypothesized_p * (1 - hypothesized_p) / sample_size)
test_statistic <- (sample_proportion - hypothesized_p) / standard_error

round(c(p_hat = sample_proportion, z = test_statistic,
        critical_value = qnorm(0.01)), 3)
```

**Step 4.** **Do not reject $H_0$**, since the test statistic $-1.054$ does not fall in the critical region.

**Step 5.** There is not enough evidence to support the claim that fewer than 40% of the stalls at this night market accept mobile payment.
:::

---

## Example 8-19 {#example-8-19}

:::{.callout-warning}
## Booking Accommodation Online

A tourism report states that at least 72% of inbound visitors to Taiwan book their accommodation online. To see whether this claim is valid, a researcher selected a random sample of 90 inbound visitors and found that 60 of them had booked online *(hypothetical data)*. At $\alpha = 0.01$, test the claim that at least 72% of inbound visitors book their accommodation online.
:::

---

## Example 8-19

:::{.callout-tip}
## Solution

**Step 1.** $H_0: p = 0.72$ *(claim)* and $H_1: p < 0.72$ *(left-tailed)*.

**Step 2.** Since $\alpha = 0.01$ and the test is left-tailed, the critical value is $-2.326$.

**Step 3 — compute the test statistic.**

```{r e8-19-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
successes <- 60
sample_size <- 90
hypothesized_p <- 0.72

sample_proportion <- successes / sample_size
standard_error <- sqrt(hypothesized_p * (1 - hypothesized_p) / sample_size)
test_statistic <- (sample_proportion - hypothesized_p) / standard_error

round(c(p_hat = sample_proportion, z = test_statistic,
        critical_value = qnorm(0.01)), 3)
```

**Step 4.** **Do not reject $H_0$**, since the test statistic $-1.127$ does not fall in the critical region.

**Step 5.** There is not enough evidence to reject the claim that at least 72% of inbound visitors book their accommodation online.
:::

---

## Example 8-20 {#example-8-20}

:::{.callout-warning}
## Cross-Border Orders Shipped to Japan

A payments company claims that more than 20% of the cross-border e-commerce orders placed from Taiwan are shipped to Japan. A random sample of 250 cross-border orders showed that 63 were shipped to Japan *(hypothetical data)*. At $\alpha = 0.05$, is there enough evidence to support the claim? Use the P-value method.
:::

---

## Example 8-20

:::{.callout-tip}
## Solution

**Step 1.** $H_0: p = 0.20$ and $H_1: p > 0.20$ *(claim; right-tailed)*.

**Step 2 and Step 3 — compute the test statistic and the P-value.**

```{r e8-20-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
successes <- 63
sample_size <- 250
hypothesized_p <- 0.20

sample_proportion <- successes / sample_size
standard_error <- sqrt(hypothesized_p * (1 - hypothesized_p) / sample_size)
test_statistic <- (sample_proportion - hypothesized_p) / standard_error
p_value <- pnorm(test_statistic, lower.tail = FALSE)

round(c(p_hat = sample_proportion, z = test_statistic, p_value = p_value), 4)
```

**Step 4.** **Reject $H_0$**, since the P-value $0.0199$ is less than 0.05.

**Step 5.** There is enough evidence to support the claim that more than 20% of the cross-border orders placed from Taiwan are shipped to Japan.
:::

---

## Finding Chi-Square Critical Values

:::{.callout-important}
## Three Cases Using Table A-6

Table A-6 gives the areas **to the right** of the critical value.

1. **Right-tailed test** — use the area under the $\alpha$ value for the specific degrees of freedom.
2. **Left-tailed test** — subtract $\alpha$ from 1, then use the area for $1 - \alpha$.
3. **Two-tailed test** — divide $\alpha$ by 2; use the area for $\alpha/2$ for the right critical value and the area for $1 - \alpha/2$ for the left critical value.

In R, `qchisq(area, df)` returns the value with `area` to its **left**, which is why the arguments below are $1 - \alpha$, $\alpha$, and so on.
:::

---

## Example 8-21 {#example-8-21}

:::{.callout-warning}
## Critical Chi-Square Value, Right-Tailed

Find the critical chi-square value for 18 degrees of freedom when $\alpha = 0.05$ and the test is right-tailed.
:::

---

## Example 8-21

:::{.callout-tip}
## Solution

Find the $\alpha$ value at the top of Table A-6 and the degrees of freedom in the left column. The critical value is located where the two columns meet: **28.869**.

```{r e8-21-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(qchisq(1 - 0.05, df = 18), 3)
```
:::

---

## Example 8-22 {#example-8-22}

:::{.callout-warning}
## Critical Chi-Square Value, Left-Tailed

Find the critical chi-square value for 14 degrees of freedom when $\alpha = 0.05$ and the test is left-tailed.
:::

---

## Example 8-22

:::{.callout-tip}
## Solution

When the test is left-tailed, the $\alpha$ value must be subtracted from 1: $1 - 0.05 = 0.95$. The chi-square table gives the area to the right of the critical value, and chi-square values cannot be negative, so 95% of the area will be to the right of the value. For 0.95 and 14 degrees of freedom the critical value is **6.571**.

```{r e8-22-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(qchisq(0.05, df = 14), 3)
```
:::

---

## Example 8-23 {#example-8-23}

:::{.callout-warning}
## Critical Chi-Square Values, Two-Tailed

Find the critical chi-square values for 19 degrees of freedom when $\alpha = 0.10$ and a two-tailed test is conducted.
:::

---

## Example 8-23

:::{.callout-tip}
## Solution

The area must be split: the area to the right of the larger value is $\alpha/2 = 0.05$, and the area to the right of the smaller value is $1 - \alpha/2 = 0.95$. With 19 degrees of freedom the critical values are **10.117** and **30.144**.

```{r e8-23-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(qchisq(c(0.10 / 2, 1 - 0.10 / 2), df = 19), 3)
```
:::

---

## Degrees of Freedom Beyond the Table

:::{.callout-warning}
## Caution

Table A-6 gives values only up to 30 degrees of freedom. When the degrees of freedom exceed those specified in the table, **use the table value for 30 degrees of freedom**. This guideline keeps the type I error equal to or below the $\alpha$ value.
:::

:::{.callout-important}
## Three Test Situations for a Variance

If a researcher believes the variance of a population is greater than 225: $H_0: \sigma^2 = 225$ and $H_1: \sigma^2 > 225$ — right-tailed.

If less than 225: $H_0: \sigma^2 = 225$ and $H_1: \sigma^2 < 225$ — left-tailed.

If no direction is specified: $H_0: \sigma^2 = 225$ and $H_1: \sigma^2 \neq 225$ — two-tailed.
:::

---

## The Chi-Square Test for a Single Variance

:::{.callout-important}
## Formula

$$\chi^2 = \frac{(n-1)s^2}{\sigma^2}$$

with degrees of freedom equal to $n - 1$, where $n$ is the sample size, $s^2$ the sample variance and $\sigma^2$ the population variance.
:::

:::{.callout-important}
## Assumptions for the Chi-Square Test for a Single Variance

1. The sample must be **randomly selected** from the population.
2. The population must be **normally distributed** for the variable under study.
3. The observations must be **independent** of one another.
:::

---

## Why Test a Variance?

Wherever consistency is required, the smallest possible variation is wanted. When bolts are manufactured, the variation in diameters must be kept to a minimum, or the nuts will not fit properly. In education, a test used repeatedly on the same students should produce approximately the same grades, so the variance of each student's grades should be small — but a test used to judge learning should have a large overall standard deviation, so that those who have learned can be distinguished from those who have not.

:::{.callout-warning}
## Caution

The methods used to test a hypothesis about a population variance or standard deviation are **not robust**. If the data come from a population that is not normally distributed, these methods cannot be used.
:::

---

## Example 8-24 {#example-8-24}

:::{.callout-warning}
## Fill Weight of Snack Packs

A quality engineer wishes to see whether the variance of the fill weight of a snack pack on one production line is less than the process standard of $\sigma^2 = 36$ (grams squared). A random sample of 12 packs from that line had a variance of $s^2 = 27$ *(hypothetical data)*. Test her claim at $\alpha = 0.05$.
:::

---

## Example 8-24

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \sigma^2 = 36$ and $H_1: \sigma^2 < 36$ *(claim; left-tailed)*.

**Step 2.** Since this is a left-tailed test and $\alpha = 0.05$, the value $1 - 0.05 = 0.95$ is used with $\text{d.f.} = 11$. The critical value is $4.575$.

**Step 3 — compute the test statistic.**

```{r e8-24-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_variance <- 27
hypothesized_variance <- 36
sample_size <- 12
degrees_freedom <- sample_size - 1

test_statistic <- degrees_freedom * sample_variance / hypothesized_variance

round(c(chi_square = test_statistic,
        critical_value = qchisq(0.05, df = degrees_freedom)), 3)
```

**Step 4.** Since $8.250$ falls in the noncritical region, **do not reject $H_0$**.

**Step 5.** There is not enough evidence to support the claim that the variance of the fill weight on this line is less than 36.
:::

---

## Example 8-25 {#example-8-25}

:::{.callout-warning}
## Orders Picked per Hour

A distribution-centre manager believes that the standard deviation of the number of orders picked per hour is greater than 6. A random sample of 15 hours is selected and the number of orders picked in each hour is shown *(hypothetical data)*. At $\alpha = 0.10$, is there enough evidence to support the manager's claim? Assume the variable is normally distributed.

```
38   52   45   61   40
55   33   48   57   42
36   50   65   44   39
```
:::

---

## Example 8-25

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \sigma = 6$ and $H_1: \sigma > 6$ *(claim; right-tailed)*. Since the standard deviation is given, it is squared to get the variance.

**Step 2.** Right-tailed with $\text{d.f.} = 14$ and $\alpha = 0.10$: the critical value is $21.064$.

**Step 3 — prepare data and compute the test statistic.**

```{r e8-25-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
orders_per_hour <- c(38, 52, 45, 61, 40,
                     55, 33, 48, 57, 42,
                     36, 50, 65, 44, 39)

hypothesized_sd <- 6
degrees_freedom <- length(orders_per_hour) - 1
test_statistic <- degrees_freedom * var(orders_per_hour) / hypothesized_sd^2

round(c(sd = sd(orders_per_hour), chi_square = test_statistic,
        critical_value = qchisq(1 - 0.10, df = degrees_freedom)), 3)
```

**Step 4.** **Reject $H_0$**, since $35.222 > 21.064$.

**Step 5.** There is enough evidence to support the claim that the standard deviation of the number of orders picked per hour is greater than 6.
:::

---

## Example 8-26 {#example-8-26}

:::{.callout-warning}
## Net Weight of Instant-Noodle Packs

A contract manufacturer wishes to test the claim that the variance of the net weight of its instant-noodle packs is 2.25. Net weight is measured in grams, and assume that it is normally distributed. A random sample of 20 packs has a standard deviation of 1.8 grams *(hypothetical data)*. At $\alpha = 0.05$, is there enough evidence to reject the manufacturer's claim?
:::

---

## Example 8-26

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \sigma^2 = 2.25$ *(claim)* and $H_1: \sigma^2 \neq 2.25$ *(two-tailed)*.

**Step 2.** Two-tailed at $\alpha = 0.05$, so the critical values for 0.025 and 0.975 are needed. With $\text{d.f.} = 19$ they are $8.907$ and $32.852$.

**Step 3 — compute the test statistic.** Since the sample standard deviation is given, it must be squared for the formula.

```{r e8-26-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_sd <- 1.8
hypothesized_variance <- 2.25
sample_size <- 20
degrees_freedom <- sample_size - 1

test_statistic <- degrees_freedom * sample_sd^2 / hypothesized_variance

round(c(chi_square = test_statistic,
        lower_cv = qchisq(0.025, df = degrees_freedom),
        upper_cv = qchisq(0.975, df = degrees_freedom)), 3)
```

**Step 4.** **Do not reject $H_0$**, since $8.907 < 27.360 < 32.852$.

**Step 5.** There is not enough evidence to reject the manufacturer's claim that the variance of the net weight is equal to 2.25.
:::

---

## P-Value Intervals for the Chi-Square Test

Approximate P-values for the chi-square test can be found from Table A-6, but the procedure is more complicated than for the $z$ and $t$ tests, since the chi-square distribution is not symmetric and $\chi^2$ values cannot be negative.

:::{.callout-important}
## Reading the Interval

- **Right-tailed.** Find the two table values the test statistic falls between and read the corresponding $\alpha$ values directly.
- **Left-tailed.** Read the corresponding table values, then **subtract each from 1**.
- **Two-tailed.** Find the interval as above, then **double both endpoints**.
:::

---

## Example 8-27 {#example-8-27}

:::{.callout-warning}
## P-Value Interval, Right-Tailed

Find the P-value when $\chi^2 = 21.5$, $n = 11$, and the test is right-tailed.
:::

---

## Example 8-27

:::{.callout-tip}
## Solution

Across the row with $\text{d.f.} = 10$, the value 21.5 falls between 20.483 and 23.209, corresponding to 0.025 and 0.01 respectively. Hence

$$0.01 < P\text{-value} < 0.025$$

```{r e8-27-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(pchisq(21.5, df = 10, lower.tail = FALSE), 4)
```

The exact P-value is $0.0179$, which lies inside that interval.
:::

---

## Example 8-28 {#example-8-28}

:::{.callout-warning}
## P-Value Interval, Left-Tailed

Find the P-value when $\chi^2 = 5.892$, $n = 15$, and the test is left-tailed.
:::

---

## Example 8-28

:::{.callout-tip}
## Solution

Across the row with $\text{d.f.} = 14$, the value 5.892 falls between 5.629 and 6.571, corresponding to 0.975 and 0.95. Because the test statistic falls on the left side, each value must be subtracted from 1: $1 - 0.975 = 0.025$ and $1 - 0.95 = 0.05$. Hence

$$0.025 < P\text{-value} < 0.05$$

```{r e8-28-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
round(pchisq(5.892, df = 14), 4)
```

The exact P-value is $0.0309$. Had the test been two-tailed, the interval would be doubled: $0.05 < P\text{-value} < 0.10$.
:::

---

## Example 8-29 {#example-8-29}

:::{.callout-warning}
## Customs Clearance Times

A customs broker knows from past records that the standard deviation of the time needed to clear an import container is 9.5 hours. A random sample of 25 containers is selected and timed; the standard deviation is 7.0 hours *(hypothetical data)*. At $\alpha = 0.05$, can it be concluded that the standard deviation has changed? Use the P-value method. Assume the variable is normally distributed.
:::

---

## Example 8-29

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \sigma = 9.5$ and $H_1: \sigma \neq 9.5$ *(claim; two-tailed)*.

**Step 2 and Step 3 — compute the test statistic and the P-value.**

```{r e8-29-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_sd <- 7.0
hypothesized_sd <- 9.5
sample_size <- 25
degrees_freedom <- sample_size - 1

test_statistic <- degrees_freedom * sample_sd^2 / hypothesized_sd^2
p_value <- 2 * pchisq(test_statistic, df = degrees_freedom)

round(c(chi_square = test_statistic, p_value = p_value), 4)
```

With $\text{d.f.} = 24$, the value 13.0305 falls between 12.401 and 13.848, corresponding to 0.975 and 0.95. Subtracting from 1 gives 0.025 and 0.05, and doubling for the two tails gives $0.05 < P\text{-value} < 0.10$; the exact P-value is $0.0688$.

**Step 4.** **Do not reject $H_0$**, since P-value $> \alpha$.

**Step 5.** There is not enough evidence to support the claim that the standard deviation of the customs clearance time has changed.
:::

---

## Example 8-30 {#example-8-30}

:::{.callout-warning}
## Value of Air-Freight Shipments

A trade magazine reports that the average declared value of an air-freight shipment leaving Taoyuan is NT\$68 thousand. A freight forwarder randomly selected 50 recent shipments and found that the average declared value was NT\$63.8 thousand and the standard deviation was NT\$9.6 thousand *(hypothetical data)*. Is there enough evidence at $\alpha = 0.05$ to conclude that the average value is not NT\$68 thousand? Assume the variable is normally distributed. Also find the 95% confidence interval of the true mean.
:::

---

## Example 8-30

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \mu = 68$ and $H_1: \mu \neq 68$ *(claim; two-tailed)*.

**Step 2.** At $\alpha = 0.05$ with $\text{d.f.} = 49$, `qt()` gives the critical values $+2.010$ and $-2.010$.

**Step 3 — compute the test statistic.**

```{r e8-30-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_mean <- 63.8
hypothesized_mean <- 68
sample_sd <- 9.6
sample_size <- 50

test_statistic <- (sample_mean - hypothesized_mean) /
  (sample_sd / sqrt(sample_size))
critical_value <- qt(1 - 0.05 / 2, df = sample_size - 1)

round(c(t = test_statistic, critical_value = critical_value), 3)
```

**Step 4.** Since $-3.094 < -2.010$, the decision is to **reject $H_0$**.

**Step 5.** There is enough evidence to support the claim that the average declared value is not NT\$68 thousand.
:::

---

## Example 8-30

:::{.callout-tip}
## Solution

**The 95% confidence interval.**

$$\bar{X} - t_{\alpha/2}\frac{s}{\sqrt{n}} < \mu < \bar{X} + t_{\alpha/2}\frac{s}{\sqrt{n}}$$

```{r e8-30-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
margin_of_error <- critical_value * sample_sd / sqrt(sample_size)

round(c(lower = sample_mean - margin_of_error,
        upper = sample_mean + margin_of_error), 2)
```

The interval $61.07 < \mu < 66.53$ does **not** contain the hypothesized value $\mu = 68$. There is agreement between the hypothesis test and the confidence interval.
:::

---

## Example 8-31 {#example-8-31}

:::{.callout-warning}
## Net Weight of Tea-Bag Cartons

A supplier claims that its cartons of tea bags have an average net weight of 120 grams. A random sample of 10 cartons has an average net weight of 118.6 grams and a standard deviation of 2.8 grams *(hypothetical data)*. At $\alpha = 0.05$, can the claim be rejected? Also, find the 95% confidence interval of the true mean. Assume the variable is normally distributed.
:::

---

## Example 8-31

:::{.callout-tip}
## Solution

**Step 1.** $H_0: \mu = 120$ *(claim)* and $H_1: \mu \neq 120$ *(two-tailed)*.

**Step 2.** At $\alpha = 0.05$ and $\text{d.f.} = 9$, the critical values are $+2.262$ and $-2.262$.

**Step 3 — compute the test statistic and the interval.**

```{r e8-31-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
sample_mean <- 118.6
hypothesized_mean <- 120
sample_sd <- 2.8
sample_size <- 10

test_statistic <- (sample_mean - hypothesized_mean) /
  (sample_sd / sqrt(sample_size))
critical_value <- qt(1 - 0.05 / 2, df = sample_size - 1)
margin_of_error <- critical_value * sample_sd / sqrt(sample_size)

round(c(t = test_statistic, critical_value = critical_value), 3)
round(c(lower = sample_mean - margin_of_error,
        upper = sample_mean + margin_of_error), 2)
```

**Step 4 and Step 5.** **Do not reject $H_0$**. There is not enough evidence to reject the claim that the mean net weight is 120 grams. The 95% interval $116.6 < \mu < 120.6$ **does** contain $\mu = 120$; again there is agreement.
:::

---

## Type II Error and the Power of a Test

:::{.callout-important}
## Figure 8-40: Possibilities in Hypothesis Testing

| Decision | $H_0$ true | $H_0$ false |
|---|---|---|
| Reject $H_0$ | Type I error, $\alpha$ | Correct decision, $1 - \beta$ |
| Do not reject $H_0$ | Correct decision, $1 - \alpha$ | Type II error, $\beta$ |
:::

The value of $\beta$ depends on $\alpha$, the sample size, the population standard deviation, and the actual difference between the hypothesized parameter and the true parameter. Since that last difference is unknown, $\beta$ cannot be computed.

---

## The Power of a Test

:::{.callout-note}
## Power of a Test

The **power of a statistical test** measures the sensitivity of the test to detect a real difference in parameters if one actually exists. The power of a test is equal to

$$\text{Power} = 1 - \beta$$

Power is a probability and can take values from 0 to 1. The closer the power is to 1, the better the test is at rejecting the null hypothesis when the null hypothesis is in fact false. If $\beta = 0.04$, the power is $1 - 0.04 = 0.96$, or 96%.
:::

---

## Increasing the Power of a Test

:::{.callout-important}
## Two Practical Methods

1. **Increase $\alpha$** — for example use $\alpha = 0.05$ instead of $\alpha = 0.01$. As $\alpha$ increases, $\beta$ decreases, so $1 - \beta$ increases.
2. **Select a larger sample size** — a larger $n$ makes the standard error of the mean smaller and consequently reduces $\beta$.

Neither should be used at the researcher's whim. Before increasing $\alpha$, consider the consequences of a type I error; increasing the sample size costs money and time.

When a choice of statistical tests exists, use the test with the highest power for the data — provided its assumptions can be met. If they cannot, a test with lower power must be used.
:::

---

## Power in R

**Draw figure** — the distribution of $\bar{X}$ under $H_0: \mu = 50$ and under the true value $\mu = 54$, with $\sigma = 10$, $n = 25$ and a right-tailed test at $\alpha = 0.05$.

```{r s8-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mu_null <- 50
mu_alternative <- 54
standard_error <- 10 / sqrt(25)
critical_mean <- mu_null + qnorm(1 - 0.05) * standard_error

sampling_curves <- tibble(x = seq(40, 68, length.out = 400)) |>
  mutate(under_h0 = dnorm(x, mu_null, standard_error),
         under_h1 = dnorm(x, mu_alternative, standard_error))

power_plot <- ggplot(sampling_curves, aes(x)) +
  geom_area(data = filter(sampling_curves, x >= critical_mean),
            aes(y = under_h1), fill = "steelblue", alpha = 0.5) +
  geom_line(aes(y = under_h0)) +
  geom_line(aes(y = under_h1), linetype = "dashed") +
  geom_vline(xintercept = critical_mean, linetype = "dotted") +
  labs(title = "Power of a Right-Tailed Test", x = "Sample mean", y = "Density") +
  theme_minimal()
```

---

## Power in R

**Output figure**

```{r s8-6-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
power_plot

power <- pnorm(critical_mean, mu_alternative, standard_error,
               lower.tail = FALSE)
round(c(critical_mean = critical_mean, beta = 1 - power, power = power), 3)
```

The shaded area is the power $1 - \beta = 0.639$; the unshaded part of the dashed curve is $\beta = 0.361$. Increasing $n$ narrows both curves and increases the power.

---

## A Final Comment

:::{.callout-warning}
## Caution

When the researcher fails to reject the null hypothesis, this does **not** mean that there is not enough evidence to support alternative hypotheses. It may be that the null hypothesis is false but the statistical test has too low a power to detect the real difference. One can conclude only that in this study there is not enough evidence to reject the null hypothesis.

Whether the decision is to reject $H_0$ or not to reject $H_0$, there is in either case a chance of being wrong. The goal is to keep the probabilities of type I and type II errors as small as possible.
:::

## Important Terms

:::{.callout-important}
## Chapter 8 Vocabulary

$\alpha$ (alpha) · alternative hypothesis · $\beta$ (beta) · chi-square test · critical or rejection region · critical value · hypothesis testing · left-tailed test · level of significance · noncritical or nonrejection region · null hypothesis · one-tailed test · power of a test · P-value · research hypothesis · right-tailed test · statistical hypothesis · statistical test · test statistic · $t$ test · two-tailed test · type I error · type II error · $z$ test
:::

## Key Formulas

:::{.callout-important}
## Tests for Means

The general form of a test statistic:

$$\text{Test statistic} = \frac{(\text{observed value}) - (\text{expected value})}{\text{standard error}}$$

The $z$ test for means — if $n < 30$, the variable must be normally distributed:

$$z = \frac{\bar{X} - \mu}{\sigma / \sqrt{n}}$$

The $t$ test for means, $\text{d.f.} = n - 1$ — if $n < 30$, the variable must be normally distributed:

$$t = \frac{\bar{X} - \mu}{s / \sqrt{n}}$$
:::

---

## Key Formulas

:::{.callout-important}
## Tests for Proportions, Variances and Standard Deviations

The $z$ test for proportions, valid if $np \geq 5$ and $nq \geq 5$, where $\hat{p} = X/n$ and $q = 1 - p$:

$$z = \frac{\hat{p} - p}{\sqrt{pq/n}}$$

The chi-square test for a variance or standard deviation, $\text{d.f.} = n - 1$ — the variable must be normally distributed:

$$\chi^2 = \frac{(n-1)s^2}{\sigma^2}$$

The power of a test:

$$\text{Power} = 1 - \beta$$
:::

## Key Takeaways

:::{.callout-important}
- A **statistical hypothesis** is a conjecture about a population parameter; $H_0$ states no difference and always carries the equals sign, while $H_1$ states a difference and is also called the **research hypothesis**
- Evidence can **support** a claim only when the claim is $H_1$, and can **reject** a claim only when the claim is $H_0$; nothing is ever proved
- A **type I error** rejects a true $H_0$ and has probability $\alpha$, the **level of significance**; a **type II error** fails to reject a false $H_0$ and has probability $\beta$ — decreasing one increases the other
- The **critical value** separates the **critical (rejection) region** from the **noncritical (nonrejection) region**; the direction of $H_1$ decides whether the test is left-tailed, right-tailed or two-tailed
- **Traditional method:** state the hypotheses and identify the claim, find the critical value(s), compute the test statistic, make the decision, summarize the results
:::

## Key Takeaways

:::{.callout-important}
- **P-value method:** state the hypotheses and identify the claim, compute the test statistic, find the P-value, make the decision, summarize the results — reject $H_0$ when P-value $\leq \alpha$
- Use the $z$ test when $\sigma$ is known, the $t$ test with $\text{d.f.} = n - 1$ when $\sigma$ is unknown; test a proportion with $z$ when $np \geq 5$ and $nq \geq 5$, and a variance with $\chi^2$ when the population is normal
- For a two-tailed test, the $(1 - \alpha)$ confidence interval contains the hypothesized value exactly when $H_0$ is not rejected
- The **power** of a test is $1 - \beta$; it is raised by increasing $\alpha$ or, better, by increasing the sample size
:::

## Acknowledgement

-   **Copyright notice.** These teaching materials follow the organization and terminology of **Bluman, A. G. (2023). Elementary statistics: A step by step approach (11th ed.). McGraw Hill**. All rights in the original work are reserved by its authors and publishers.

-   **Original examples.** Every worked example, data set, and R script in these slides was written for this course. The data are hypothetical unless stated otherwise.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
