---
title: "Statistics"
subtitle: "Chapter 1: The Nature of Probability and Statistics"
author: "Yu-You Liou (NTU)"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
format: revealjs
smaller: true
theme: serif
footer: "Elementary Statistics: A Step-by-Step Approach"
logo: "logo.svg"
editor: source
bibliography: bib.bib
slide-number: true
scrollable: false
transition: "none"
menu:
  useTextContentForMissingTitles: false
  hideMissingTitles: true
  side: 'left'
  titleSelector: 'h1'
---

# Introduction

## Why Study Statistics?

Statistics is woven into nearly every domain of modern life — sports analytics, public health surveillance, economic forecasting, clinical drug trials, and even the algorithms behind social media feeds. Whenever you read that "30 minutes of exercise three times a week raises HDL cholesterol by 10–15%" or that "the median credit-card debt for college students is $3,173," you are encountering a statistical claim.

::: {.callout-note icon=false}
## Definition: Statistics
**Statistics** is the science of conducting studies to collect, organise, summarise, analyse, and draw conclusions from data.
:::

------------------------------------------------------------------------

**Three reasons every student should learn statistics**

1. **Reading research critically** — Every profession produces statistical reports. Understanding the vocabulary and logic of statistical studies allows you to evaluate whether a claim is well-supported or misleading.

2. **Conducting your own research** — Designing experiments, gathering data, analysing results, and communicating findings honestly are skills rooted in statistical thinking.

3. **Being an informed citizen and consumer** — From interpreting government budgets to evaluating product safety claims, statistical reasoning empowers better everyday decisions.

------------------------------------------------------------------------

## Core Vocabulary

Before exploring the two branches of statistics, three foundational terms must be defined.

::: {.callout-note icon=false}
## Definition: Variable
A **variable** is any characteristic or attribute that can assume different values across individuals, objects, or time points.
:::

::: {.callout-note icon=false}
## Definition: Data
**Data** are the specific values — measurements or observations — that a variable takes on. A collection of such values is called a **data set**; each individual entry is a **datum**. When the value a variable takes is governed by chance, the variable is called a **random variable**.
:::

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE}
# Illustrating variables and data types in R

# A variable: student height (continuous quantitative)
heights_in <- c(65, 70, 68, 72, 60, 67, 71, 63, 69, 74)

# A variable: blood type (qualitative)
blood_type <- c("A", "O", "B", "AB", "O", "A", "B", "O", "A", "B")

cat("Variable: student height (inches)\n")
cat("Data set:", heights_in, "\n")
cat("n =", length(heights_in), "observations\n\n")

cat("Variable: blood type (qualitative)\n")
print(table(blood_type))
```

# Section 1–1

## Descriptive and Inferential Statistics

All statistical work falls into one of two broad branches, distinguished by what the analyst is trying to accomplish with the data.

::: {.callout-note icon=false}
## Definition: Descriptive Statistics
**Descriptive statistics** encompasses the methods used to collect, organise, summarise, and present data in a clear and informative form — through tables, charts, averages, and other numerical measures. The goal is to describe *what the data show*, without extending conclusions beyond the data set itself.
:::

::: {.callout-note icon=false}
## Definition: Inferential Statistics
**Inferential statistics** uses information gathered from a *sample* to draw conclusions (inferences) about a larger *population*. It relies on probability theory and includes estimation, hypothesis testing, identifying relationships among variables, and making predictions.
:::

------------------------------------------------------------------------

**The two branches in practice**

| Branch | Goal | Typical output |
|---|---|---|
| Descriptive | Summarise the data at hand | Mean, median, table, chart |
| Inferential | Generalise to a population | Confidence interval, p-value |

*Example — Descriptive:* The U.S. Census Bureau surveys all residents every 10 years and reports the average household income, age distribution, and educational attainment. These summaries describe the population directly from a complete count.

*Example — Inferential:* A pharmaceutical company tests a new pain reliever on 200 volunteers and concludes that it is effective for all adults — a generalisation beyond the trial participants.

------------------------------------------------------------------------

## Population and Sample

::: {.callout-note icon=false}
## Definition: Population
A **population** consists of *all* subjects — people, objects, or events — that belong to the group under investigation. Populations can be finite (all students enrolled at a university) or theoretically infinite (all possible coin flips).
:::

::: {.callout-note icon=false}
## Definition: Sample
A **sample** is a subset of subjects selected from the population for the purpose of study. A properly chosen sample should possess characteristics similar to those of the full population, allowing valid inferences to be drawn.
:::

------------------------------------------------------------------------

**Why use a sample instead of a census?**

Studying an entire population is often:

- **Too expensive** — surveying every household nationwide
- **Too time-consuming** — testing every manufactured item for defects
- **Physically impossible** — measuring every star in a galaxy
- **Destructive** — testing every firework to see if it ignites

A carefully designed sample provides nearly as much information as a full census at a fraction of the cost.

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE}
# Descriptive statistics: summarising a sample
set.seed(42)
exam_scores <- c(78, 85, 90, 72, 88, 95, 60, 77, 83, 91,
                 68, 74, 89, 93, 80, 76, 87, 70, 82, 96)

cat("=== Descriptive Statistics (n =", length(exam_scores), ") ===\n")
cat("Mean   :", round(mean(exam_scores), 2), "\n")
cat("Median :", median(exam_scores), "\n")
cat("Std Dev:", round(sd(exam_scores), 2), "\n")
cat("Min    :", min(exam_scores), "\n")
cat("Max    :", max(exam_scores), "\n")
```

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE}
# Inferential statistics: testing a claim about the population mean
# Claim: the population mean exam score exceeds 75
result <- t.test(exam_scores, mu = 75, alternative = "greater")

cat("=== One-Sample t-Test (H0: μ = 75  vs  Ha: μ > 75) ===\n\n")
cat("t-statistic :", round(result$statistic, 3), "\n")
cat("Degrees of freedom:", result$parameter, "\n")
cat("p-value     :", round(result$p.value, 4), "\n")
cat("95% lower bound:", round(result$conf.int[1], 2), "\n\n")

if (result$p.value < 0.05) {
  cat("Decision: Reject H0 — evidence supports μ > 75.\n")
} else {
  cat("Decision: Fail to reject H0.\n")
}
```

The sample result (descriptive) is used to make an inference about all students who might take this exam (inferential).

------------------------------------------------------------------------

## Applying the Concepts 1–1

::: {.callout-tip icon=false}
## Attendance and Grades

A study at Manatee Community College found a clear pattern linking attendance to final grades:

- **95–100% attendance** → typically earned an **A**
- **80–90% attendance** → typically earned a **B or C**
- **Below 80% attendance** → typically earned a **D, F**, or withdrew

Answer the following questions based on this information.

1. What are the **variables** under study?
2. What are the **data** in this study?
3. Are **descriptive, inferential, or both** types of statistics used?
4. What is the **population** under study?
5. Was a **sample** collected? If so, from where?
6. Comment on the **relationship** between the two variables.
:::

------------------------------------------------------------------------

**Solution — Applying the Concepts 1–1**

```{r,echo=T,warning=FALSE,message=FALSE,out.width="72%",fig.align='center'}
library(ggplot2)

attendance_cat <- factor(c("95–100%","80–90%","Below 80%"),
                          levels = c("95–100%","80–90%","Below 80%"))
grade_cat      <- c("A", "B or C", "D / F / Withdrew")
pct            <- c(85, 70, 40)   # approximate grade-A rate within each group

df <- data.frame(Attendance = attendance_cat,
                 Approx_A_Rate = pct)

ggplot(df, aes(Attendance, Approx_A_Rate, fill = Attendance)) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Attendance Category vs Approximate A-Rate",
       y = "% of Students Earning A", x = "Attendance Range") +
  theme_minimal(base_size = 13)
```

------------------------------------------------------------------------

**Answers — Applying the Concepts 1–1**

| Question | Answer |
|---|---|
| Variables | Attendance percentage; course grade |
| Data | Specific attendance records and letter grades |
| Type of statistics | Primarily **descriptive**; generalising to all colleges would be inferential |
| Population | All students at Manatee Community College |
| Sample | The students included in the study (a subset) |
| Relationship | Positive — higher attendance is associated with higher grades |

# Section 1–2

## Variables and Types of Data

Variables are classified along two dimensions: their **nature** (qualitative vs. quantitative) and, for quantitative variables, their **continuity** (discrete vs. continuous).

::: {.callout-note icon=false}
## Definition: Qualitative Variable
A **qualitative** (or categorical) variable places each observation into a non-numerical category defined by some attribute or characteristic. Examples: gender, blood type, political affiliation, eye colour.
:::

::: {.callout-note icon=false}
## Definition: Quantitative Variable
A **quantitative** variable is inherently numerical and allows meaningful arithmetic operations. Observations can be ordered and compared by magnitude. Examples: age, height, income, temperature.
:::

------------------------------------------------------------------------

## Discrete and Continuous Variables

::: {.callout-note icon=false}
## Definition: Discrete Variable
A **discrete** variable can assume only isolated, countable values — typically non-negative integers. There are no possible values between consecutive counts. Examples: number of children in a family, number of defective items in a shipment, number of emergency calls per hour.
:::

::: {.callout-note icon=false}
## Definition: Continuous Variable
A **continuous** variable can assume any value within an interval — including fractions and decimals — because it is obtained by *measurement* rather than counting. In principle, infinitely many values exist between any two observations. Examples: height, weight, temperature, elapsed time.
:::

------------------------------------------------------------------------

**Classification tree for data**

```
Data
├── Qualitative  →  categories, labels, names
└── Quantitative
    ├── Discrete   →  counted values (0, 1, 2, …)
    └── Continuous →  measured values (any real number in a range)
```

| Variable | Type | Sub-type |
|---|---|---|
| Blood type (A, B, O, AB) | Qualitative | — |
| Number of siblings | Quantitative | Discrete |
| Body temperature (°F) | Quantitative | Continuous |
| Zip code | Qualitative | — |
| Household income ($) | Quantitative | Continuous |

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE}
# Classifying variables in R

# Qualitative — stored as a factor
blood <- factor(c("A","O","B","AB","O","A","B","O","AB","A"))
cat("Qualitative variable (blood type):\n")
print(table(blood))

# Discrete — whole-number counts
siblings <- c(0, 2, 1, 3, 0, 4, 1, 2, 0, 3)
cat("\nDiscrete variable (siblings):\n")
cat("  Values:", siblings, "\n")
cat("  Mean:", mean(siblings), " | Can be non-integer (average), but\n")
cat("  individual values must be whole numbers.\n")
```

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE,out.width="72%",fig.align='center'}
# Continuous — measured on a scale
body_temp <- c(98.2, 99.1, 97.8, 98.6, 100.2, 98.4, 97.5, 99.0, 98.9, 98.1)

cat("Continuous variable (body temperature, °F):\n")
cat("  Values:", body_temp, "\n")
cat("  Mean:", round(mean(body_temp), 2), "\n")
cat("  Any value between 97 and 101 is theoretically possible.\n\n")

hist(body_temp,
     main = "Distribution of Body Temperatures",
     xlab = "Temperature (°F)", col = "steelblue",
     border = "white", breaks = 6)
```

------------------------------------------------------------------------

## Boundaries of Continuous Data

Because continuous measurements are always rounded to the nearest unit of precision, every recorded value represents an *interval* of possible true values. These intervals are called **class boundaries**.

::: {.callout-note icon=false}
## Rule: Computing Boundaries
Boundaries are found by subtracting and adding **half of one unit of precision** from the recorded value.
If data are recorded to whole numbers → boundaries extend by **±0.5**
If data are recorded to tenths → boundaries extend by **±0.05**
:::

------------------------------------------------------------------------

**Boundary examples**

| Variable | Recorded Value | Boundaries |
|---|---|---|
| Length | 15 cm | 14.5 – 15.5 cm |
| Temperature | 86 °F | 85.5 – 86.5 °F |
| Time | 0.43 sec | 0.425 – 0.435 sec |
| Mass | 1.6 g | 1.55 – 1.65 g |

```{r,echo=T,warning=FALSE}
# Computing boundaries automatically
recorded <- c(15, 86, 0.43, 1.6)
precision <- c(1, 1, 0.01, 0.1)

boundaries <- data.frame(
  Recorded  = recorded,
  Precision = precision,
  Lower     = recorded - precision / 2,
  Upper     = recorded + precision / 2
)
print(boundaries)
```

------------------------------------------------------------------------

## Levels of Measurement

Beyond qualitative/quantitative, variables are characterised by four **measurement scales** that determine which mathematical operations and statistical procedures are appropriate.

| Level | Properties | Example |
|---|---|---|
| **Nominal** | Categories only; no order | Eye colour, zip code |
| **Ordinal** | Ordered categories; gaps unequal | Letter grade (A > B > C …) |
| **Interval** | Equal spacing; no true zero | Temperature (°F), IQ |
| **Ratio** | Equal spacing + true zero | Height, weight, income |

Each higher level inherits all properties of the levels below it.

------------------------------------------------------------------------

::: {.callout-note icon=false}
## Definition: Nominal Level
The **nominal level** classifies data into mutually exclusive, exhaustive categories in which **no ordering or ranking** can be meaningfully imposed. Numbers may be used as labels (e.g., jersey numbers, zip codes), but arithmetic operations on them are meaningless.
:::

::: {.callout-note icon=false}
## Definition: Ordinal Level
The **ordinal level** classifies data into categories that **can be ranked**, but the *differences between ranks are not guaranteed to be equal*. Knowing that "Excellent" > "Good" > "Poor" does not reveal how much better one rating is than another.
:::

------------------------------------------------------------------------

::: {.callout-note icon=false}
## Definition: Interval Level
The **interval level** has all ordinal properties **plus equal, meaningful spacing between units**. However, there is **no true zero point** — a score of 0 does not mean complete absence of the attribute. Therefore, ratios are not interpretable (80°F is not "twice as warm" as 40°F).
:::

::: {.callout-note icon=false}
## Definition: Ratio Level
The **ratio level** possesses all interval properties **plus a genuine zero point**, making ratios meaningful and interpretable. A person earning $80,000 earns twice as much as one earning $40,000. A weight of 0 kg means no weight at all.
:::

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE}
# Demonstrating measurement levels in R

# --- Nominal: no meaningful order ---
nationality <- factor(c("Taiwanese","Japanese","American","Korean","Thai"))
cat("Nominal (nationality) — levels have no inherent order:\n")
print(levels(nationality))

# --- Ordinal: ranked categories ---
service_rating <- factor(c("Poor","Excellent","Good","Good","Poor","Excellent"),
                         levels = c("Poor","Good","Excellent"), ordered = TRUE)
cat("\nOrdinal (service rating) — order is meaningful:\n")
print(service_rating)
cat("Is 'Excellent' > 'Good'?", service_rating[2] > service_rating[3], "\n")
```

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE}
# --- Interval: equal spacing, no true zero ---
temp_F <- c(32, 50, 68, 86, 104)
cat("Interval (temperature °F) — equal spacing of", diff(temp_F)[1], "°F\n")
cat("0°F does NOT mean 'no heat'; ratios are not meaningful.\n")
cat("32°F is NOT 'twice as hot' as 16°F.\n\n")

# --- Ratio: true zero, ratios valid ---
salaries <- c(30000, 45000, 60000, 90000)
cat("Ratio (annual salary $) — true zero exists.\n")
cat("Salaries:", salaries, "\n")
cat("$60,000 is", 60000/30000, "times $30,000 — ratio is meaningful.\n")
```

------------------------------------------------------------------------

## Applying the Concepts 1–2

::: {.callout-tip icon=false}
## Safe Travel — Transportation Injury Data

The table below shows the number of job-related injuries for five U.S. transportation industries in 1998.

| Industry | Injuries |
|---|---|
| Railroad | 4,520 |
| Intercity bus | 5,100 |
| Subway | 6,850 |
| Trucking | 7,144 |
| Airline | 9,950 |

1. What are the **variables** under study?
2. Classify each variable as **qualitative or quantitative**.
3. Classify each quantitative variable as **discrete or continuous**.
4. Identify the **level of measurement** for each variable.
5. Does fewer injuries necessarily mean railroads are *safer*? Explain.
6. What factors beyond safety might influence a traveller's choice?
:::

------------------------------------------------------------------------

**Solution — Applying the Concepts 1–2**

```{r,echo=T,warning=FALSE,out.width="72%",fig.align='center'}
library(ggplot2)

industry  <- c("Railroad","Intercity Bus","Subway","Trucking","Airline")
injuries  <- c(4520, 5100, 6850, 7144, 9950)
df <- data.frame(Industry  = factor(industry, levels = industry),
                 Injuries  = injuries)

ggplot(df, aes(x = reorder(Industry, Injuries), y = Injuries, fill = Industry)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_brewer(palette = "Oranges") +
  labs(title = "Job-Related Injuries by Transportation Industry (1998)",
       x = NULL, y = "Number of Injuries") +
  theme_minimal(base_size = 13)
```

------------------------------------------------------------------------

**Answers — Applying the Concepts 1–2**

| Question | Answer |
|---|---|
| Variables | Industry type; number of injuries |
| Classification | Industry = qualitative; injuries = quantitative |
| Discrete / continuous | Number of injuries = **discrete** (counted) |
| Measurement level | Industry = **nominal**; injuries = **ratio** |
| Caution on "safest" | Railroads employ far fewer workers → raw injury counts are not comparable rates |
| Other factors | Cost, convenience, schedule, comfort, geographic availability |

# Section 1–3

## Data Collection and Sampling Techniques

Researchers gather data in several ways: telephone surveys, mailed questionnaires, personal interviews, direct observation, and examination of existing records. Each approach involves a trade-off of cost, accuracy, coverage, and potential for bias.

**Common survey methods**

| Method | Advantage | Disadvantage |
|---|---|---|
| Telephone survey | Fast; inexpensive | Misses unlisted numbers; interviewer tone can bias |
| Mailed questionnaire | Wide geographic reach; anonymous | Low response rate; literacy barriers |
| Personal interview | In-depth responses | Costly; interviewer selection bias |

------------------------------------------------------------------------

## Why Sampling Matters

Gathering data from every member of a population (a **census**) is rarely feasible. Instead, researchers rely on **samples** — and the method used to draw those samples determines whether the results can be trusted.

A sample drawn **haphazardly** introduces **bias**: systematically favoring certain kinds of subjects while excluding others. Surveying only mall shoppers on a Tuesday morning excludes working adults, shift workers, home-makers, and many others.

The solution is to use a **probability-based sampling method** that gives every member of the population a calculable, non-zero chance of being selected.

------------------------------------------------------------------------

::: {.callout-note icon=false}
## Definition: Random Sampling
In **simple random sampling**, every possible sample of the desired size has an *equal probability* of being selected. Subjects are numbered, and a computer or random-number table is used to choose which numbers are included.
:::

::: {.callout-note icon=false}
## Definition: Systematic Sampling
In **systematic sampling**, subjects are numbered from 1 to *N*. A starting point is chosen at random between 1 and *k* (where *k = N ÷ n*, the desired sample size), and then every *k*th subject is selected thereafter.
:::

------------------------------------------------------------------------

::: {.callout-note icon=false}
## Definition: Stratified Sampling
In **stratified sampling**, the population is first divided into non-overlapping, exhaustive subgroups called **strata** (based on a characteristic relevant to the study). An independent random sample is then drawn from each stratum, ensuring all important subgroups are represented.
:::

::: {.callout-note icon=false}
## Definition: Cluster Sampling
In **cluster sampling**, the population is divided into naturally occurring **clusters** (e.g., city blocks, hospital wards, school classes). A random selection of clusters is made, and *all members* of the chosen clusters form the sample. Most useful when the population is geographically spread out.
:::

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE}
set.seed(101)
N <- 100           # population of 100 subjects

# 1. Simple Random Sample  (n = 10)
srs <- sort(sample(1:N, size = 10, replace = FALSE))
cat("Simple Random Sample (n=10):\n ", srs, "\n\n")

# 2. Systematic Sample  (k = 10, n = 10)
k     <- 10
start <- sample(1:k, 1)               # random start in [1, k]
sys   <- seq(start, N, by = k)
cat("Systematic Sample (k=10, start=", start, "):\n ", sys, "\n")
```

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE}
set.seed(101)

# 3. Stratified Sample — 5 per gender stratum
pop <- data.frame(id     = 1:100,
                  gender = rep(c("Female","Male"), each = 50))

strat <- do.call(rbind,
           lapply(split(pop, pop$gender),
                  function(g) g[sample(nrow(g), 5), ]))
strat <- strat[order(strat$id), ]
cat("Stratified Sample (5 per gender):\n")
print(strat, row.names = FALSE)
```

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE}
set.seed(101)

# 4. Cluster Sample — randomly select 2 of 10 classrooms
classrooms <- split(1:100, rep(1:10, each = 10))
chosen     <- sample(1:10, size = 2)
cluster_members <- sort(unlist(classrooms[chosen]))

cat("Cluster Sample:\n")
cat("  Classrooms selected:", chosen, "\n")
cat("  Members:", cluster_members, "\n")
cat("  Total subjects:", length(cluster_members), "\n")
```

------------------------------------------------------------------------

**Sampling methods at a glance**

| Method | Selection rule | Strength |
|---|---|---|
| Random | Every subject equally likely | Unbiased; theoretically ideal |
| Systematic | Every *k*th after random start | Simple to implement |
| Stratified | Random within predefined groups | Guarantees group representation |
| Cluster | All members of randomly chosen clusters | Practical for spread-out populations |
| Convenience | Whoever is available | Quick but prone to bias — not recommended for inference |

------------------------------------------------------------------------

## Applying the Concepts 1–3

::: {.callout-tip icon=false}
## American Culture and Drug Abuse

Imagine you are a member of the Family Research Council studying how American media (television, film, music, magazines) may influence illegal drug use among the general public. You design a survey of 2,250 adults and adolescents selected from across the country.

A consumer advocacy group asks you the following questions about your design:

1. What **type of survey** did you use, and why?
2. What are the **advantages and disadvantages** of the methods you did not use?
3. What **type of scale** (nominal, ordinal, interval, ratio) did you use? Why?
4. Did you use a **random** selection method? Which one?
5. Why was that method more appropriate than alternatives?
6. How would results be affected if only **adolescents** were surveyed?
:::

------------------------------------------------------------------------

**Solution — Applying the Concepts 1–3**

```{r,echo=T,warning=FALSE,out.width="72%",fig.align='center'}
library(ggplot2)

# Illustrate a 5-point Likert scale used in such a survey
responses  <- c("Strongly\nAgree","Agree","Neutral","Disagree","Strongly\nDisagree")
counts_all <- c(510, 720, 480, 360, 180)   # hypothetical all-ages data
counts_ado <- c(680, 600, 300, 170, 500)   # hypothetical adolescent-only data

df <- data.frame(
  Response = factor(rep(responses, 2), levels = responses),
  Count    = c(counts_all, counts_ado),
  Group    = rep(c("All ages (stratified)","Adolescents only"), each = 5)
)

ggplot(df, aes(Response, Count, fill = Group)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("steelblue","tomato")) +
  labs(title = "Survey Responses: Full Sample vs Adolescent-Only",
       x = NULL, y = "Count", fill = NULL) +
  theme_minimal(base_size = 12)
```

------------------------------------------------------------------------

**Answers — Applying the Concepts 1–3**

- **Survey type**: Stratified telephone survey — cost-effective, national reach, and ensures age/region representation
- **Scale**: Ordinal (Likert 1–5: *strongly disagree* to *strongly agree*) — captures ranked attitudes without claiming equal spacing
- **Sampling method**: Stratified by region and age group ensures that both adults and adolescents from all parts of the country appear in the sample
- **Adolescent-only bias**: Adolescents have different media exposure and social pressures than adults; restricting the sample would produce results that cannot be generalised to the broader population

# Section 1–4

## Observational and Experimental Studies

All statistical studies can be categorised by how the researcher relates to the data-generating process.

::: {.callout-note icon=false}
## Definition: Observational Study
In an **observational study**, the researcher watches, records, and analyses events *without intervening or manipulating any variable*. Because no variable is controlled, cause-and-effect relationships cannot be established — only associations can be noted.
:::

::: {.callout-note icon=false}
## Definition: Experimental Study
In an **experimental study**, the researcher *actively manipulates* at least one variable and measures the effect on another. Properly designed experiments — with random assignment and control groups — permit causal conclusions.
:::

------------------------------------------------------------------------

## Key Components of an Experiment

::: {.callout-note icon=false}
## Definition: Independent Variable (Explanatory Variable)
The **independent variable** is the factor the researcher deliberately changes or controls. It is the presumed *cause* in a cause-and-effect relationship.
:::

::: {.callout-note icon=false}
## Definition: Dependent Variable (Outcome Variable)
The **dependent variable** is the response that is measured to determine whether the manipulation had an effect. It is the presumed *effect*.
:::

::: {.callout-note icon=false}
## Definition: Confounding Variable
A **confounding variable** is one that influences the dependent variable but was not separated from the independent variable — making it impossible to determine which factor drove the observed change.
:::

------------------------------------------------------------------------

## Example — The Sit-Up Study

::: {.callout-tip icon=false}
## The Sit-Up Study (Virginia Polytechnic Institute)

Female undergraduates were randomly divided into two groups and performed as many sit-ups as possible in 90 seconds, once per day for four days.

- **Control group**: told only to *"do your best"*
- **Treatment group**: told to *increase their daily count by 10%*

**Independent variable**: type of instruction (vague vs. specific goal)
**Dependent variable**: number of sit-ups completed per session

After four days, the control group averaged **43 sit-ups**; the treatment group averaged **56 sit-ups**.

*Does setting specific, measurable goals improve athletic performance?*
:::

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE,message=FALSE,out.width="78%",fig.align='center'}
library(ggplot2)

day   <- rep(1:4, 2)
group <- rep(c("Control (vague goal)","Treatment (10% goal)"), each = 4)
avg   <- c(38, 40, 41, 43,   # control trajectory
           44, 48, 52, 56)   # treatment trajectory

df <- data.frame(Day = day, Group = group, Avg_Situps = avg)

ggplot(df, aes(Day, Avg_Situps, color = Group, group = Group)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 4) +
  scale_color_manual(values = c("steelblue","tomato")) +
  labs(title = "Average Sit-ups Over Four Days",
       subtitle = "Specific goals produced a 30% performance advantage by Day 4",
       x = "Day", y = "Average Sit-ups Completed", color = NULL) +
  theme_minimal(base_size = 13)
```

------------------------------------------------------------------------

**Conclusion of the sit-up study**

```{r,echo=T,warning=FALSE}
# Simulating and testing the Day-4 group difference
set.seed(7)
control   <- round(rnorm(20, mean = 43, sd = 6))
treatment <- round(rnorm(20, mean = 56, sd = 6))

cat("Control group   — Day 4  | Mean:", round(mean(control), 1),
    "| SD:", round(sd(control), 1), "\n")
cat("Treatment group — Day 4  | Mean:", round(mean(treatment), 1),
    "| SD:", round(sd(treatment), 1), "\n\n")

result <- t.test(treatment, control, alternative = "greater")
cat("Two-sample t-test (treatment > control):\n")
cat("  t =", round(result$statistic, 3),
    "| p-value =", round(result$p.value, 5), "\n")
cat("  Conclusion:", ifelse(result$p.value < 0.05,
    "Significant — specific goals improve performance.",
    "Not significant at α = 0.05."), "\n")
```

------------------------------------------------------------------------

## Hazards in Experimental Design

**The Hawthorne Effect** — Discovered at the Western Electric plant in 1924: participants who *know* they are being studied may change their behaviour regardless of the treatment. This makes it difficult to isolate the true effect of the independent variable.

**Confounding** — Example: subjects placed on a new exercise regimen may simultaneously improve their diet. If health improves, is it exercise, diet, or both? The two factors are *confounded*.

**Unnatural setting** — Lab results may not transfer to real-world behaviour ("This mouthwash kills 10,000 germs in a test tube — but how many in my mouth?").

------------------------------------------------------------------------

## Applying the Concepts 1–4

::: {.callout-tip icon=false}
## Just a Pinch Between Your Cheek and Gum — Smokeless Tobacco

As cigarette warnings grew, some people turned to *smokeless tobacco* (chewing tobacco placed between cheek and gum) as an alternative. Researchers studied its cardiovascular effects using **40 university students** randomly assigned to two groups:

- **Treatment group (n = 20)**: chewed real smokeless tobacco
- **Control group (n = 20)**: chewed a look-alike substance with no active ingredients

Blood pressure and heart rate were recorded *before* and *20 minutes after* chewing. A significant increase in heart rate was observed only in the treatment group.

1. What **type of study** is this?
2. What are the **independent and dependent variables**?
3. Which group is the **treatment group**?
4. Could knowledge of being in the study affect blood pressure?
5. List possible **confounding variables**.
6. Is this a sound study design?
:::

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE,out.width="76%",fig.align='center'}
library(ggplot2)

set.seed(42)
group     <- rep(c("Control","Treatment"), each = 20)
hr_before <- c(round(rnorm(20, 70, 5)), round(rnorm(20, 71, 5)))
hr_after  <- c(round(rnorm(20, 71, 5)), round(rnorm(20, 80, 6)))

df <- data.frame(Group = group, Before = hr_before, After = hr_after)
df$Change <- df$After - df$Before

ggplot(df, aes(Group, Change, fill = Group)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  scale_fill_manual(values = c("steelblue","tomato")) +
  labs(title = "Change in Heart Rate After Chewing (Simulated)",
       y = "Heart Rate Change (bpm)", x = NULL) +
  theme_minimal(base_size = 13)
```

------------------------------------------------------------------------

**Answers — Applying the Concepts 1–4**

```{r,echo=T,warning=FALSE}
# Test whether the heart-rate change differs between groups
set.seed(42)
ctrl_change  <- rnorm(20, mean = 1, sd = 4)
treat_change <- rnorm(20, mean = 9, sd = 5)

res <- t.test(treat_change, ctrl_change, alternative = "greater")
cat("Treatment mean change:", round(mean(treat_change), 1), "bpm\n")
cat("Control mean change  :", round(mean(ctrl_change),  1), "bpm\n")
cat("p-value:", round(res$p.value, 5),
    "— Smokeless tobacco significantly raises heart rate.\n")
```

| Question | Answer |
|---|---|
| Study type | **Experimental** — researcher assigned groups and controlled the treatment |
| Independent variable | Type of substance chewed (tobacco vs. placebo) |
| Dependent variables | Heart rate and blood pressure |
| Treatment group | The 20 students given real smokeless tobacco |
| Hawthorne effect | Possible — awareness of being monitored can itself alter physiology |
| Confounding variables | Student anxiety, baseline fitness, caffeinated beverages consumed |

# Section 1–5

## Uses and Misuses of Statistics

Statistical methods, applied honestly and rigorously, are among the most powerful tools available to researchers, policy-makers, and citizens. But applied carelessly — or dishonestly — they can mislead just as powerfully.

> *"There are three kinds of lies: lies, damned lies, and statistics."*
> — popularised by Mark Twain

> *"Figures don't lie, but liars figure."*

The goal of this section is not to make you cynical about statistics, but to sharpen your critical reading of any claim that invokes numerical evidence.

------------------------------------------------------------------------

## Suspect Samples

::: {.callout-note icon=false}
## Warning: Small or Biased Samples
"Three out of four doctors recommend Brand X." — If only *four* doctors were surveyed, this result could easily arise by chance. The key questions to always ask: **How large was the sample?** and **How were subjects selected?**
:::

Common sources of sample bias:

- **Volunteer samples** — people who self-select into a study tend to have stronger opinions or unusual characteristics not representative of the general population
- **Convenience samples** — students in a single classroom, shoppers at one mall, callers to a radio station
- **Online polls** — only those who happen to visit a website and feel strongly enough to respond

------------------------------------------------------------------------

## Ambiguous Averages

There are **four** distinct measures commonly called an "average": **mean, median, mode**, and **midrange**. For the same data set, these can differ dramatically. A dishonest communicator can choose whichever measure best supports a predetermined conclusion — without technically lying.

```{r,echo=T,warning=FALSE}
# A skewed salary distribution (in $1,000s)
salaries <- c(28, 32, 35, 38, 40, 42, 45, 50, 55, 380)

cat("Salary data ($thousands):", salaries, "\n\n")
cat("Mean   :", mean(salaries),   "  ← pulled up by the outlier\n")
cat("Median :", median(salaries), "  ← resistant to the outlier\n")
cat("Mode   : none (all values unique)\n")
cat("Midrange:", (min(salaries)+max(salaries))/2, "← highly sensitive to extremes\n")
```

A company with one executive earning $380K and nine staff averaging ~$40K could truthfully advertise an "average salary of $74,500" — technically correct, deeply misleading.

------------------------------------------------------------------------

## Detached Statistics, Implied Connections, and Misleading Graphs

::: {.callout-note icon=false}
## Warning: Detached Statistics
A **detached statistic** makes a comparison without naming what it is being compared to. *"Our crackers contain one-third fewer calories."* — Fewer than what? Always ask: **Compared to what?**
:::

::: {.callout-note icon=false}
## Warning: Implied Connections
Hedging language signals that no causal relationship has been established: *"may reduce," "studies suggest," "in some people," "might help."* These phrases allow the speaker to retreat from any claim. Read them carefully.
:::

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE,message=FALSE,out.width="80%",fig.align='center'}
library(ggplot2); library(patchwork)

revenue <- data.frame(Year = 2020:2024,
                      Rev  = c(100, 101, 102, 103, 104))

p1 <- ggplot(revenue, aes(Year, Rev)) +
  geom_col(fill = "steelblue") +
  coord_cartesian(ylim = c(0, 120)) +
  labs(title = "Honest Y-axis (starts at 0)",
       y = "Revenue ($M)", x = NULL) +
  theme_minimal(base_size = 12)

p2 <- ggplot(revenue, aes(Year, Rev)) +
  geom_col(fill = "tomato") +
  coord_cartesian(ylim = c(99, 105)) +
  labs(title = "Truncated Y-axis (starts at 99)",
       y = "Revenue ($M)", x = NULL) +
  theme_minimal(base_size = 12)

p1 | p2
```

Both panels display identical data. The truncated axis makes a modest 4% growth appear as a quadrupling. Always inspect the **axis origin** before interpreting a bar chart.

------------------------------------------------------------------------

## Faulty Survey Questions

The phrasing of a survey question can dramatically shift responses, even when both versions ask about the same underlying topic.

::: {.callout-tip icon=false}
## Example — Question Wording Effect

**Version A**: *"Do you think the North Huntingdon School District should build a new football stadium?"*

**Version B**: *"Do you favour increasing school taxes so that the North Huntingdon School District can build a new football stadium?"*

Both questions ask about the *same* proposal. Version B makes the cost explicit, which typically depresses support significantly. The choice of wording can make the same policy appear popular or unpopular.
:::

```{r,echo=T,warning=FALSE}
# Simulating how wording affects support
set.seed(5)
support_A <- rbinom(200, 1, prob = 0.65)  # 65% support (no cost mention)
support_B <- rbinom(200, 1, prob = 0.38)  # 38% support (cost made explicit)

cat("Version A — support rate:", round(mean(support_A) * 100, 1), "%\n")
cat("Version B — support rate:", round(mean(support_B) * 100, 1), "%\n\n")
prop.test(c(sum(support_A), sum(support_B)),
          c(200, 200))
```

# Section 1–6

## Computers and Calculators in Statistics

Before the digital era, statisticians performed all computations by hand — an error-prone and time-consuming process. Modern software has eliminated this bottleneck. This course uses **R**, a free, open-source language that has become the global standard for statistical analysis and data science.

::: {.callout-note icon=false}
## Key Principle
**Software computes; you interpret.**
R produces numbers. Understanding *what those numbers mean*, *when to trust them*, and *how to communicate them honestly* is the statistician's job. Doing problems step-by-step by hand first — before turning to software — builds the intuition needed for correct interpretation.
:::

------------------------------------------------------------------------

## Core R Workflow

::: {.callout-tip icon=false}
## Example 1-R: Entering, Sorting, and Summarising Data

Enter the following weight measurements (in lbs) into R and compute descriptive summaries:

**Data**: 213, 208, 203, 215, 222
:::

```{r,echo=T,warning=FALSE}
# Step 1 — Enter the data
weights <- c(213, 208, 203, 215, 222)
cat("Data:", weights, "\n\n")

# Step 2 — Sort ascending and descending
cat("Sorted (ascending) :", sort(weights), "\n")
cat("Sorted (descending):", sort(weights, decreasing = TRUE), "\n\n")

# Step 3 — Compute summary statistics
cat("Mean   :", mean(weights), "\n")
cat("Median :", median(weights), "\n")
cat("Std Dev:", round(sd(weights), 3), "\n")
cat("Range  :", diff(range(weights)), "\n")
cat("Min    :", min(weights), "| Max:", max(weights), "\n")
```

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE,out.width="74%",fig.align='center'}
# Step 4 — Visualise the data
df_w <- data.frame(weight = weights)

ggplot(df_w, aes(x = "", y = weight)) +
  geom_boxplot(fill = "lightblue", color = "navy", width = 0.35) +
  geom_jitter(width = 0.05, color = "darkred", size = 4) +
  labs(title = "Distribution of Five Weight Measurements",
       x = NULL, y = "Weight (lbs)") +
  theme_minimal(base_size = 13)
```

The box shows the interquartile range (middle 50%), the horizontal line marks the median, and the red dots show individual data values.

------------------------------------------------------------------------

## Random Number Generation in R

::: {.callout-tip icon=false}
## Example 1-R2: Selecting a Random Sample

Use R to select a random sample of **15 subjects** from a population of **85 subjects**, mirroring the random-number-table approach described in the textbook.
:::

```{r,echo=T,warning=FALSE}
set.seed(2024)                         # ensures reproducibility

population <- 1:85                     # subjects numbered 01–85
sample_15  <- sort(sample(population,
                          size    = 15,
                          replace = FALSE))

cat("Random sample of 15 from population of 85:\n")
cat(sample_15, "\n\n")

cat("Compare with textbook's table method:\n")
cat("Textbook: 12 27 75 62 57 13 31 06 16 49 46 71 53 41 02\n")
cat("R result : both are equally valid random samples.\n")
```

------------------------------------------------------------------------

## Sampling Error and Repeated Sampling

::: {.callout-tip icon=false}
## Example 1-R3: Demonstrating Sampling Error

A population of 500 individuals has a true mean IQ of approximately 100. Draw repeated random samples of size 30 and observe how much sample means vary around the true population mean.
:::

```{r,echo=T,warning=FALSE}
set.seed(2024)
population_iq <- round(rnorm(500, mean = 100, sd = 15))
pop_mean      <- mean(population_iq)

# One sample
one_sample <- sample(population_iq, 30)
cat("Population mean (true) :", round(pop_mean, 2), "\n")
cat("One sample mean        :", round(mean(one_sample), 2), "\n")
cat("Sampling error         :",
    round(abs(mean(one_sample) - pop_mean), 2), "\n")
```

------------------------------------------------------------------------

```{r,echo=T,warning=FALSE,out.width="78%",fig.align='center'}
set.seed(2024)
pop <- round(rnorm(500, mean = 100, sd = 15))
pop_mean <- mean(pop)

# Repeat sampling 1,000 times
sample_means <- replicate(1000, mean(sample(pop, 30)))

hist(sample_means, breaks = 35, col = "steelblue", border = "white",
     main = "Distribution of 1,000 Sample Means (n = 30)",
     xlab = "Sample Mean", ylab = "Frequency")
abline(v = pop_mean, col = "red", lwd = 2.5, lty = 2)
legend("topright",
       legend = paste("True population mean =", round(pop_mean, 1)),
       col = "red", lty = 2, lwd = 2)
```

This bell-shaped sampling distribution is a preview of the **Central Limit Theorem** (Chapter 6): regardless of the population shape, sample means become approximately normally distributed as *n* grows.

# Summary

## Chapter 1 Summary

::: {.callout-note icon=false}
## Section 1–1 — Descriptive and Inferential Statistics
- **Descriptive** statistics: collect, organise, summarise, present data
- **Inferential** statistics: generalise from sample to population using probability
- Key terms: **variable**, **data**, **population**, **sample**, **random variable**
:::

::: {.callout-note icon=false}
## Section 1–2 — Variables and Types of Data
- Variables are **qualitative** or **quantitative** (discrete / continuous)
- Four measurement scales: **nominal → ordinal → interval → ratio**
- Higher levels inherit all properties of lower levels
:::

------------------------------------------------------------------------

::: {.callout-note icon=false}
## Section 1–3 — Data Collection and Sampling
- Main survey methods: telephone, mail, personal interview
- Four probability sampling strategies: **random, systematic, stratified, cluster**
- Convenience sampling is susceptible to bias and should not be used for inference
:::

::: {.callout-note icon=false}
## Section 1–4 — Observational and Experimental Studies
- **Observational**: no manipulation → can identify associations, not causes
- **Experimental**: researcher controls the independent variable → can support causal claims
- Watch for: confounding variables, Hawthorne effect, unnatural settings
:::

------------------------------------------------------------------------

::: {.callout-note icon=false}
## Section 1–5 — Uses and Misuses of Statistics
- Misuses include: suspect samples, ambiguous averages, detached statistics, implied connections, misleading graphs, faulty survey questions
- Critical reading of any statistical claim is essential
:::

::: {.callout-note icon=false}
## Section 1–6 — Computers and Calculators
- R handles computation; the analyst provides interpretation
- Core tasks: enter data → sort → summarise → visualise → sample
- Sampling error is inevitable; understanding its distribution is the foundation of inferential statistics
:::

------------------------------------------------------------------------

## Looking Ahead — Chapter 2

With the vocabulary and conceptual framework of Chapter 1 established, the course turns to **organising and displaying data** (Chapter 2):

- Constructing **frequency distributions** (categorical and grouped)
- Drawing **histograms**, **frequency polygons**, and **ogives**
- Building **bar charts**, **Pareto charts**, **time series graphs**, and **pie charts**
- Reading and constructing **stem-and-leaf plots**

```{r,echo=T,warning=FALSE,out.width="76%",fig.align='center'}
# Preview of Chapter 2: a simple frequency table and bar chart
categories <- c("Strongly\nAgree","Agree","Neutral","Disagree","Strongly\nDisagree")
freq       <- c(45, 30, 15, 7, 3)
df_prev <- data.frame(Response = factor(categories, levels = categories),
                      Frequency = freq)

ggplot(df_prev, aes(Response, Frequency, fill = Response)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = Frequency), vjust = -0.4, size = 4) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Survey Response Frequency Distribution (Ch. 2 Preview)",
       x = NULL, y = "Frequency") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(size = 10))
```

## Acknowledgement

-   **Textbook source.** All chapter content is derived from *Elementary Statistics: A Step-by-Step Approach* (8th ed.) by Allan G. Bluman. McGraw-Hill Education.

-   **Non-commercial use only.** These slides are prepared strictly for educational purposes and must not be used for commercial gain.

-   **Proper attribution.** Any reproduction or distribution of these materials must provide full attribution to the original textbook and author.

## References
