---
title: "Statistics"
subtitle: "Chapter 5: Discrete Probability Distributions"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
language:
  callout-warning-title: "Problem"
  callout-important-title: "Key point"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch5-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
theme_set(theme_minimal())
```

## Overview
Chapter 5 explains **probability distributions** for discrete random variables — how to build them, how to describe them, and the special distributions used most often.

| Section | Topics |
|------|--------------------------------------------|
| 5-1 | Random variables; discrete probability distributions; graphs; the two requirements |
| 5-2 | Mean, variance, standard deviation, expectation for a discrete random variable |
| 5-3 | Binomial experiments; binomial probability formula (factorial and combination forms); binomial mean, variance, standard deviation |
| 5-4 | Other distributions: multinomial, Poisson, hypergeometric, geometric |

## Chapter Objectives

:::{.callout-important}
## After completing this chapter, you should be able to

1. Construct a probability distribution for a random variable.
2. Find the mean, variance, standard deviation, and expected value for a discrete random variable.
3. Find the exact probability for $X$ successes in $n$ trials of a binomial experiment.
4. Find the mean, variance, and standard deviation for the variable of a binomial distribution.
5. Find probabilities for outcomes of variables, using the Poisson, hypergeometric, geometric, and multinomial distributions.
:::

# Section 5-1: Probability Distributions {background-color="#2c4a6e"}

## Random Variables

A variable was defined in Chapter 1 as a characteristic or attribute that can assume different values. Because the variables in this chapter are associated with probability, they are called **random variables**.

:::{.callout-note}
## Random Variable

A **random variable** is a variable whose values are determined by chance.
:::

:::{.callout-note}
## Discrete and Continuous Random Variables

- **Discrete variables** have a finite number of possible values or an infinite number of values that can be *counted* — enumerated using the numbers $1, 2, 3, \ldots$
- **Continuous variables** can assume all values in the interval between any two given values; they are obtained from data that are *measured* rather than counted.
:::

Only discrete random variables are used in this chapter; Chapter 6 explains continuous random variables.

## Constructing a Probability Distribution

Toss three coins. The sample space is TTT, TTH, THT, HTT, HHT, HTH, THH, HHH, and if $X$ is the number of heads, then $X$ assumes the value $0$, $1$, $2$ or $3$.

| Number of heads $X$ | 0 | 1 | 2 | 3 |
|---|---|---|---|---|
| Probability $P(X)$ | $\frac{1}{8}$ | $\frac{3}{8}$ | $\frac{3}{8}$ | $\frac{1}{8}$ |

:::{.callout-note}
## Discrete Probability Distribution

A **discrete probability distribution** consists of the values a random variable can assume and the corresponding probabilities of the values. The probabilities are determined theoretically or by observation.
:::

## Procedure Table

:::{.callout-important}
## Constructing a Probability Distribution

**Step 1** Make a frequency distribution for the outcomes of the variable.

**Step 2** Find the probability for each outcome by dividing the frequency of the outcome by the sum of the frequencies.

**Step 3** If a graph is required, place the outcomes on the $x$ axis and the probabilities on the $y$ axis, and draw vertical bars for each outcome and its corresponding probability.
:::

Discrete probability distributions can be shown by using a graph, a table, or a formula.

## Example 5-1 {#example-5-1}

:::{.callout-warning}
## Rolling an Eight-Sided Die

A board game uses an eight-sided die with the numbers 1 through 8 on its faces. Construct a probability distribution for the number that shows when the die is rolled once.
:::

---

## Example 5-1

:::{.callout-tip}
## Solution

**Prepare data**

The sample space is $1, 2, 3, 4, 5, 6, 7, 8$ and each outcome has probability $\frac{1}{8} = 0.125$, so the distribution is as shown.

```{r e5-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
data.frame(X = 1:8, P = rep(1/8, 8))
```
:::

## Example 5-2 {#example-5-2}

:::{.callout-warning}
## Tossing Four Coins

Represent graphically the probability distribution of the number of heads obtained when four coins are tossed.

```
Number of heads X      0      1      2      3      4
Probability  P(X)    1/16   4/16   6/16   4/16   1/16
```

Back to [Example 5-7](#example-5-7)
:::

---

## Example 5-2

:::{.callout-tip}
## Solution

**Prepare data**

The values that $X$ assumes are located on the $x$ axis, and the values for $P(X)$ are located on the $y$ axis. The 16 equally likely outcomes give $1, 4, 6, 4, 1$ ways of obtaining $0, 1, 2, 3, 4$ heads.

```{r e5-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
coin_dist <- data.frame(X = 0:4, P = c(1, 4, 6, 4, 1) / 16)
coin_dist
```
:::

---

## Example 5-2

:::{.callout-tip}
## Solution

**Draw figure** (Figure 5-1: Probability Distribution for Example 5-2)

```{r e5-2-2, echo=T, warning=FALSE, out.width="65%", fig.align='center', fig.height=3.5}
ggplot(coin_dist, aes(X, P)) +
  geom_col() +
  labs(x = "Number of heads", y = "P(X)")
```
:::

## Example 5-3 {#example-5-3}

:::{.callout-warning}
## Late Deliveries per Courier

A food-delivery platform in Kaohsiung tracked its 50 scooter couriers during one evening shift and recorded how many deliveries each courier arrived late for. Construct a probability distribution for the number of late deliveries per courier *(hypothetical data)*.

```
Late deliveries    1     2     3     4
Frequency         10    20    15     5
```

Back to [Example 5-8](#example-5-8)
:::

---

## Example 5-3

:::{.callout-tip}
## Solution

**Step 1 and Step 2** — make the frequency distribution and divide each frequency by 50.

```{r e5-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
courier_dist <- data.frame(X = 1:4, P = c(10, 20, 15, 5) / 50)
courier_dist
```
:::

---

## Example 5-3

:::{.callout-tip}
## Solution

**Step 3 — draw the graph** (Figure 5-2: Probability Distribution for Example 5-3)

```{r e5-3-2, echo=T, warning=FALSE, out.width="65%", fig.align='center', fig.height=3.5}
ggplot(courier_dist, aes(X, P)) +
  geom_col() +
  labs(x = "Late deliveries", y = "Probability")
```
:::

## Two Requirements for a Probability Distribution

:::{.callout-important}
## Two Requirements for a Probability Distribution

1. The sum of the probabilities of all the events in the sample space must equal 1; that is, $\Sigma P(X) = 1$.
2. The probability of each event in the sample space must be between or equal to 0 and 1. That is, $0 \leq P(X) \leq 1$.
:::

The sum cannot be less than 1 or greater than 1, since the sample space includes all possible outcomes of the probability experiment. A probability cannot be a negative number or greater than 1.

## Example 5-4 {#example-5-4}

:::{.callout-warning}
## Probability Distributions

Determine whether each distribution is a probability distribution.

```
a.  X      2     4     6     8
    P(X)  0.3   0.3   0.2   0.4

b.  X    -$10    $0   $10   $20
    P(X)  1/8   1/2   1/4   1/8

c.  X      0     1     2     3     4
    P(X)  0.2   0.2   0.2   0.2   0.2

d.  X     10    20    30    40
    P(X)  0.5   0.7  -0.3   0.1
```
:::

---

## Example 5-4

:::{.callout-tip}
## Solution

**Check both requirements for each distribution**

```{r e5-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
prob_a <- c(0.3, 0.3, 0.2, 0.4)
prob_b <- c(1/8, 1/2, 1/4, 1/8)

sum(prob_a)
all(prob_a >= 0 & prob_a <= 1)
sum(prob_b)
all(prob_b >= 0 & prob_b <= 1)
```
:::

---

## Example 5-4

:::{.callout-tip}
## Solution

**Check the remaining two distributions**

```{r e5-4-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
prob_c <- rep(0.2, 5)
prob_d <- c(0.5, 0.7, -0.3, 0.1)

sum(prob_c)
all(prob_c >= 0 & prob_c <= 1)
sum(prob_d)
all(prob_d >= 0 & prob_d <= 1)
```

- **a.** No. The sum of the probabilities is greater than 1.
- **b.** Yes. The sum equals 1 and every probability lies between 0 and 1.
- **c.** Yes. The sum equals 1 and every probability lies between 0 and 1.
- **d.** No. The sum equals 1, but one of the probabilities is less than 0.
:::

# Section 5-2: Mean, Variance, Standard Deviation, and Expectation {background-color="#2c4a6e"}

## The Mean of a Probability Distribution

The sample mean $\bar{X} = \frac{\Sigma X}{n}$ and the population mean $\mu = \frac{\Sigma X}{N}$ cannot be used for a random variable of a probability distribution, because the die would have to be rolled an infinite number of times. A new formula gives the **exact theoretical** value.

:::{.callout-important}
## Formula for the Mean of a Probability Distribution

The mean of a random variable with a discrete probability distribution is

$$\mu = X_1 \cdot P(X_1) + X_2 \cdot P(X_2) + \cdots + X_n \cdot P(X_n) = \Sigma X \cdot P(X)$$

where $X_1, X_2, \ldots, X_n$ are the outcomes and $P(X_1), P(X_2), \ldots, P(X_n)$ are the corresponding probabilities.
:::

## Rounding Rule

:::{.callout-important}
## Rounding Rule for the Mean, Variance, and Standard Deviation

The mean, variance, and standard deviation of a probability distribution should be rounded to **one more decimal place** than the outcome $X$. When fractions are used, they should be reduced to lowest terms.
:::

## Example 5-5 {#example-5-5}

:::{.callout-warning}
## Rolling a Four-Sided Die

Find the mean of the number that shows when a four-sided die, with the numbers 1 through 4 on its faces, is rolled.

Back to [Example 5-9](#example-5-9)
:::

---

## Example 5-5

:::{.callout-tip}
## Solution

**Solve**

$$\mu = \Sigma X \cdot P(X) = 1 \cdot \tfrac{1}{4} + 2 \cdot \tfrac{1}{4} + 3 \cdot \tfrac{1}{4} + 4 \cdot \tfrac{1}{4} = \frac{10}{4} = 2.5$$

```{r e5-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 1:4
P <- rep(1/4, 4)

X * P
sum(X * P)
```

When the die is rolled many times, the theoretical mean is 2.5, even though the die cannot show 2.5.
:::

## Example 5-6 {#example-5-6}

:::{.callout-warning}
## Barcode Scanners

An ageing sorting line in a distribution centre has 3 independent barcode scanners, each with a 50% chance of still being in working order (W) rather than broken (B) at the end of a year. Find the mean number of working scanners on the line.

```
BBB   WBB
BBW   WBW
BWB   WWB
BWW   WWW
```
:::

---

## Example 5-6

:::{.callout-tip}
## Solution

**Prepare data** — there are 8 equally likely outcomes, so the probability distribution of the number of working scanners $X$ is

```{r e5-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 0:3
P <- c(1, 3, 3, 1) / 8

P
X * P
sum(X * P)
```

Hence, the mean number of working scanners is 1.5.
:::

## Example 5-7 {#example-5-7}

:::{.callout-warning}
## Tossing Four Coins

If four coins are tossed, find the mean of the number of heads that occur. (See the distribution in [Example 5-2](#example-5-2).)
:::

---

## Example 5-7

:::{.callout-tip}
## Solution

**Solve**

$$\mu = \Sigma X \cdot P(X) = 0 \cdot \tfrac{1}{16} + 1 \cdot \tfrac{4}{16} + 2 \cdot \tfrac{6}{16} + 3 \cdot \tfrac{4}{16} + 4 \cdot \tfrac{1}{16} = \frac{32}{16} = 2$$

```{r e5-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 0:4
P <- c(1, 4, 6, 4, 1) / 16

X * P
sum(X * P)
```

The mean of 2 heads is the long-run or theoretical average of many sets of four tosses.
:::

## Example 5-8 {#example-5-8}

:::{.callout-warning}
## Late Deliveries per Courier

Find the mean number of late deliveries per courier for the evening shift, using the distribution built in [Example 5-3](#example-5-3).

```
Outcome X         1      2      3      4
Probability P(X) 0.20   0.40   0.30   0.10
```
:::

---

## Example 5-8

:::{.callout-tip}
## Solution

**Solve**

$$\mu = \Sigma X \cdot P(X) = 1(0.20) + 2(0.40) + 3(0.30) + 4(0.10) = 2.3$$

```{r e5-8-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 1:4
P <- c(0.20, 0.40, 0.30, 0.10)

X * P
sum(X * P)
```

Hence, the mean number of late deliveries per courier is 2.3.
:::

## Variance and Standard Deviation

The mean describes the long-run average, but it says nothing about the spread of the distribution. The definition formula is $\sigma^2 = \Sigma[(X - \mu)^2 \cdot P(X)]$; the algebraically equivalent **shortcut formula** below is used in the examples.

:::{.callout-important}
## Formula for the Variance of a Probability Distribution

$$\sigma^2 = \Sigma[X^2 \cdot P(X)] - \mu^2$$

The standard deviation of a probability distribution is

$$\sigma = \sqrt{\sigma^2} \qquad\text{or}\qquad \sigma = \sqrt{\Sigma[X^2 \cdot P(X)] - \mu^2}$$

Remember that the variance and standard deviation cannot be negative.
:::

## Example 5-9 {#example-5-9}

:::{.callout-warning}
## Rolling a Four-Sided Die

Compute the variance and standard deviation for the probability distribution in [Example 5-5](#example-5-5).
:::

---

## Example 5-9

:::{.callout-tip}
## Solution

**Solve**

$$\sigma^2 = \left(1^2 \cdot \tfrac{1}{4} + 2^2 \cdot \tfrac{1}{4} + 3^2 \cdot \tfrac{1}{4} + 4^2 \cdot \tfrac{1}{4}\right) - (2.5)^2 = 7.5 - 6.25 = 1.25$$

```{r e5-9-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 1:4
P <- rep(1/4, 4)
variance <- sum(X^2 * P) - 2.5^2

variance
sqrt(variance)
```

Hence, the standard deviation for rolling a four-sided die is about 1.118.
:::

## Example 5-10 {#example-5-10}

:::{.callout-warning}
## Selecting Numbered Tokens

A bag holds 8 tokens. Three are numbered 2, two are numbered 4, and three are numbered 6. The tokens are mixed and one is drawn at random. Its number is recorded and the token is replaced. If the experiment is repeated many times, find the variance and standard deviation of the numbers on the tokens.
:::

---

## Example 5-10

:::{.callout-tip}
## Solution

**Prepare data and solve**

$$\mu = 2 \cdot \tfrac{3}{8} + 4 \cdot \tfrac{2}{8} + 6 \cdot \tfrac{3}{8} = 4 \qquad
\sigma^2 = 19 - 16 = 3$$

```{r e5-10-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- c(2, 4, 6)
P <- c(3, 2, 3) / 8
mu       <- sum(X * P)
variance <- sum(X^2 * P) - mu^2

mu
variance
sqrt(variance)
```
:::

## Example 5-11 {#example-5-11}

:::{.callout-warning}
## Escalated Calls

On a randomly selected day, a customer-service centre in Taichung recorded how many calls each of its 40 agents had to escalate to a supervisor. Twelve agents escalated no calls, 14 escalated one call, 8 escalated two calls, 4 escalated three calls, and 2 escalated four calls. Find the variance and standard deviation of the number of escalated calls per agent *(hypothetical data)*.
:::

---

## Example 5-11

:::{.callout-tip}
## Solution

**Prepare data** — divide each frequency by 40 to get $P(X)$.

```{r e5-11-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 0:4
P <- c(12, 14, 8, 4, 2) / 40

P
```
:::

---

## Example 5-11

:::{.callout-tip}
## Solution

**Solve**

```{r e5-11-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
mu       <- sum(X * P)
variance <- sum(X^2 * P) - mu^2

mu
variance
sqrt(variance)
```

The mean is 1.25, the variance is 1.2875, and the standard deviation is about 1.135 escalated calls.
:::

## Expectation

:::{.callout-note}
## Expected Value

The **expected value** of a discrete random variable of a probability distribution is the theoretical average of the variable. The formula is

$$\mu = E(X) = \Sigma X \cdot P(X)$$

The symbol $E(X)$ is used for the expected value.
:::

The formula for the expected value is the same as the formula for the theoretical mean, so $E(X) = \mu$. When expected value problems involve money, round the answer to the nearest cent.

## Example 5-12 {#example-5-12}

:::{.callout-warning}
## Raffle Tickets

A student association sells 500 raffle tickets at NT\$100 each for a tablet valued at NT\$12,000. What is the expected value of the gain if you buy one ticket?

```
                    Win          Lose
Gain X           NT$11,900     -NT$100
Probability P(X)   1/500        499/500
```
:::

---

## Example 5-12

:::{.callout-tip}
## Solution

**Solve** — for a win the net gain is NT\$11,900, since the NT\$100 ticket price is not returned; for a loss the gain is $-$NT\$100.

$$E(X) = 11900 \cdot \tfrac{1}{500} + (-100) \cdot \tfrac{499}{500} = -76$$

```{r e5-12-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- c(11900, -100)
P <- c(1, 499) / 500

sum(X * P)
12000 * (1 / 500) - 100
```

A buyer loses, on average, NT\$76 on each ticket purchased.
:::

## Example 5-13 {#example-5-13}

:::{.callout-warning}
## Prize Coupons

A night-market game booth draws prize coupons from a drum holding 20 coupons: 10 are worth NT\$50, 5 are worth NT\$100, 3 are worth NT\$200, 1 is worth NT\$500, and 1 is worth NT\$1,000. If one coupon is drawn at random, find the expected value of the coupon.
:::

---

## Example 5-13

:::{.callout-tip}
## Solution

**Solve**

$$\mu = 50 \cdot \tfrac{10}{20} + 100 \cdot \tfrac{5}{20} + 200 \cdot \tfrac{3}{20} + 500 \cdot \tfrac{1}{20} + 1000 \cdot \tfrac{1}{20} = 155$$

```{r e5-13-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- c(50, 100, 200, 500, 1000)
P <- c(10, 5, 3, 1, 1) / 20

X * P
sum(X * P)
```

The expected value of a coupon is NT\$155.
:::

## Example 5-14 {#example-5-14}

:::{.callout-warning}
## Choosing a Bond

An investment adviser asks a client to put NT\$400,000 into one of two corporate bonds. Bond A pays a return of 5% and has a default rate of 1%. Bond B pays a return of 8% and has a default rate of 4%. Find the expected rate of return and decide which bond is the better investment. When a bond defaults, the investor loses the whole investment *(hypothetical data)*.
:::

---

## Example 5-14

:::{.callout-tip}
## Solution

**Solve**

$$E(A) = 20000(0.99) - 400000(0.01) = 15800 \qquad E(B) = 32000(0.96) - 400000(0.04) = 14720$$

```{r e5-14-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
20000 * 0.99 - 400000 * 0.01
32000 * 0.96 - 400000 * 0.04
```

Bond A is the better investment, since its expected gain of NT\$15,800 exceeds the NT\$14,720 expected from Bond B.
:::

## Expectation and Games of Chance

:::{.callout-warning}
## Caution

In gambling games, if the expected value of the game is **zero**, the game is said to be *fair*. If the expected value is **positive**, the game is in favor of the player; if it is **negative**, the game is in favor of the house — in the long run the players will lose money.
:::

# Section 5-3: The Binomial Distribution {background-color="#2c4a6e"}

## Binomial Experiments

Many probability problems have only two outcomes, or can be reduced to two outcomes: a coin lands heads or tails, a team wins or loses, a treatment is effective or ineffective, an answer is correct or incorrect. Each repetition of the experiment is called a **trial**.

:::{.callout-note}
## Binomial Experiment

A **binomial experiment** is a probability experiment that satisfies the following four requirements:

1. There must be a **fixed number of trials**.
2. Each trial can have only **two outcomes**, or outcomes that can be reduced to two outcomes. These outcomes can be considered as either *success* or *failure*.
3. The outcomes of each trial must be **independent** of one another.
4. The **probability of a success must remain the same** for each trial.
:::

The word *success* does not imply that something good or positive has occurred.

## Example 5-15 {#example-5-15}

:::{.callout-warning}
## Binomial Experiments

Decide whether each experiment is a binomial experiment. If not, state the reason why.

```
a. Selecting 30 ETP students and recording their year of study
b. Selecting 30 shoppers at a department store and recording whether each
   one paid with a mobile wallet
c. Drawing four cards from a deck without replacement and recording whether
   each card is a face card
d. Selecting eight employees from a large trading firm and asking each one
   whether they hold a TOEIC certificate
e. Recording the number of items in each of 60 randomly selected online orders
```
:::

---

## Example 5-15

:::{.callout-tip}
## Solution

**Check the four requirements for each experiment**

- **a.** No. There are four possible outcomes: first year, second year, third year, and fourth year.
- **b.** Yes. All four requirements are met.
- **c.** No. Since the cards are not replaced, the events are not independent.
- **d.** Yes. All four requirements are met.
- **e.** No. An order can contain any number of items, so there are more than two outcomes.
:::

## The Binomial Distribution

:::{.callout-note}
## Binomial Distribution

The outcomes of a binomial experiment and the corresponding probabilities of these outcomes are called a **binomial distribution**.
:::

:::{.callout-important}
## Notation for the Binomial Distribution

| Symbol | Meaning |
|---|---|
| $P(S)$ | The symbol for the probability of success |
| $P(F)$ | The symbol for the probability of failure |
| $p$ | The numerical probability of a success |
| $q$ | The numerical probability of a failure |
| $n$ | The number of trials |
| $X$ | The number of successes in $n$ trials |

$P(S) = p$ and $P(F) = 1 - p = q$. Note that $0 \leq X \leq n$ and $X = 0, 1, 2, 3, \ldots, n$.
:::

## Binomial Probability Formula

:::{.callout-important}
## Binomial Probability Formula

In a binomial experiment, the probability of exactly $X$ successes in $n$ trials is

$$P(X) = \frac{n!}{(n - X)!\,X!} \cdot p^X \cdot q^{n-X}
\qquad\text{or}\qquad
P(X) = {}_{n}C_{X} \cdot p^X \cdot q^{n-X}$$
:::

The two forms are the same formula: the number of ways to get $X$ successes from $n$ trials **without regard to order** is the combination

$$_{n}C_{X} = \frac{n!}{(n-X)!\,X!}$$

Each success has probability $p$ and can occur $X$ times; each failure has probability $q$ and can occur $n - X$ times.

## Example 5-16 {#example-5-16}

:::{.callout-warning}
## Tossing Four Coins

A coin is tossed 4 times. Find the probability of getting exactly three heads.

```
HHHH  HHHT  HHTH  HTHH  THHH  HHTT  HTHT  HTTH
THHT  THTH  TTHH  HTTT  THTT  TTHT  TTTH  TTTT
```

Back to [Example 5-19](#example-5-19)
:::

---

## Example 5-16

:::{.callout-tip}
## Solution

**Solve using the sample space**

There are four ways to get three heads out of sixteen equally likely outcomes, so the answer is $\frac{4}{16}$, or 0.25.

```{r e5-16-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
4 / 16
```
:::

---

## Example 5-16

:::{.callout-tip}
## Solution

**Solve using the binomial formula** — here $n = 4$, $X = 3$, $p = \frac{1}{2}$, $q = \frac{1}{2}$.

$$P(3\text{ heads}) = {}_{4}C_{3} \left(\tfrac{1}{2}\right)^3 \left(\tfrac{1}{2}\right)^1
= \frac{4!}{(4-3)!\,3!} \left(\tfrac{1}{2}\right)^3 \left(\tfrac{1}{2}\right)^1 = 0.25$$

```{r e5-16-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
factorial(4) / (factorial(4 - 3) * factorial(3)) * 0.5^3 * 0.5^1
choose(4, 3) * 0.5^3 * 0.5^1
```
:::

## Sampling Without Replacement

:::{.callout-warning}
## Caution

When sampling is done without replacement, such as in surveys, the events are **dependent**. However, the events can be considered independent if the size of the sample is no more than 5% of the size of the population, that is, $n \leq 0.05N$.
:::

## Example 5-17 {#example-5-17}

:::{.callout-warning}
## Mobile Wallet Payments

A payments study of a Taiwanese online marketplace found that 3 out of 10 orders are paid with a mobile wallet. If 12 orders are selected at random, find the probability that exactly three of them were paid with a mobile wallet *(hypothetical data)*.
:::

---

## Example 5-17

:::{.callout-tip}
## Solution

**Solve with the factorial form** — here $n = 12$, $X = 3$, $p = \frac{3}{10}$, $q = \frac{7}{10}$.

$$P(3) = \frac{12!}{(12-3)!\,3!}\left(\tfrac{3}{10}\right)^3\left(\tfrac{7}{10}\right)^9 = 220(0.027)(0.040353607) \approx 0.2397$$

```{r e5-17-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
factorial(12) / (factorial(12 - 3) * factorial(3))
220 * 0.3^3 * 0.7^9
```
:::

---

## Example 5-17

:::{.callout-tip}
## Solution

**Alternate solution — using the combination formula**

$$P(3) = {}_{12}C_{3}\left(\tfrac{3}{10}\right)^3\left(\tfrac{7}{10}\right)^9 \approx 0.2397$$

```{r e5-17-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
choose(12, 3)
choose(12, 3) * 0.3^3 * 0.7^9
dbinom(3, 12, 0.3)
```

There is about a 0.2397 probability that exactly 3 of 12 randomly selected orders were paid with a mobile wallet.
:::

## Example 5-18 {#example-5-18}

:::{.callout-warning}
## Use of Freight Forwarders

A trade study of Taiwanese small and medium-sized exporters found that 40% of them hand their shipments to a freight forwarder rather than booking space themselves. If 6 exporters are selected at random, find the probability that at least 4 of them use a freight forwarder *(hypothetical data)*.
:::

---

## Example 5-18

:::{.callout-tip}
## Solution

**Solve with the factorial form** — find $P(4)$, $P(5)$ and $P(6)$, then add.

$$P(4) = \frac{6!}{(6-4)!\,4!}(0.4)^4(0.6)^2 \approx 0.138 \qquad
P(5) = \frac{6!}{(6-5)!\,5!}(0.4)^5(0.6)^1 \approx 0.037$$

$$P(6) = \frac{6!}{(6-6)!\,6!}(0.4)^6(0.6)^0 \approx 0.004$$

```{r e5-18-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 4:6
P <- factorial(6) / (factorial(6 - X) * factorial(X)) * 0.4^X * 0.6^(6 - X)

P
sum(P)
```
:::

---

## Example 5-18

:::{.callout-tip}
## Solution

**Alternate solution — using the combination formula**

$$P(4) = {}_{6}C_{4}(0.4)^4(0.6)^2 \qquad P(5) = {}_{6}C_{5}(0.4)^5(0.6)^1 \qquad P(6) = {}_{6}C_{6}(0.4)^6(0.6)^0$$

```{r e5-18-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
choose(6, X) * 0.4^X * 0.6^(6 - X)
sum(dbinom(X, 6, 0.4))
```

$P(\text{at least four exporters use a freight forwarder}) = 0.138 + 0.037 + 0.004 = 0.179$
:::

## Using the Binomial Table

Computing probabilities with the binomial formula can be tedious, so **Table A-2** in Appendix A gives the probabilities for individual events for selected values of $n$ and $p$. In R the same values come from `dbinom(X, n, p)` and cumulative probabilities from `pbinom(X, n, p)`.

## Example 5-19 {#example-5-19}

:::{.callout-warning}
## Tossing Four Coins

Solve the problem in [Example 5-16](#example-5-16) by using Table A-2.
:::

---

## Example 5-19

:::{.callout-tip}
## Solution

**Look up the value** — since $n = 4$, $X = 3$, and $p = 0.5$, the value 0.25 is found in Table A-2 (Figure 5-3).

```{r e5-19-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
dbinom(3, 4, 0.5)
dbinom(0:4, 4, 0.5)
```
:::

## Example 5-20 {#example-5-20}

:::{.callout-warning}
## Repairs on Returned Handsets

A repair centre in Hsinchu finds that 10% of the handsets returned under warranty need a new mainboard. If a random sample of 15 returned handsets is selected, find these probabilities by using the binomial table in Appendix A *(hypothetical data)*.

```
a. Exactly 4 of the handsets in the sample need a new mainboard.
b. At most 3 of the handsets in the sample need a new mainboard.
c. At least 3 of the handsets in the sample need a new mainboard.
```
:::

---

## Example 5-20

:::{.callout-tip}
## Solution

**Solve** — $n = 15$, $p = 0.10$. "At most 3" means 0, 1, 2 or 3; "at least 3" is best found as $1 - P(X \leq 2)$.

```{r e5-20-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
dbinom(4, 15, 0.10)
pbinom(3, 15, 0.10)
1 - pbinom(2, 15, 0.10)
```

$P(X = 4) = 0.043$, $P(X \leq 3) = 0.944$, and $P(X \geq 3) = 0.184$.
:::

## Example 5-21 {#example-5-21}

:::{.callout-warning}
## Honoured Hotel Bookings

A boutique hotel in Taipei finds that 80% of its confirmed bookings are honoured; the rest are no-shows. If a random sample of 25 confirmed bookings is selected, find the probability that between 18 and 22, inclusive, of them are honoured *(hypothetical data)*.
:::

---

## Example 5-21

:::{.callout-tip}
## Solution

**Solve** — $n = 25$, $p = 0.80$, and $X = 18, 19, 20, 21, 22$.

```{r e5-21-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
dbinom(18:22, 25, 0.80)
sum(dbinom(18:22, 25, 0.80))
```

$P(18 \leq X \leq 22) = 0.111 + 0.163 + 0.196 + 0.187 + 0.136 = 0.793$
:::

## Binomial Distributions in R

Changing $n$ and $p$ changes the centre and the shape of a binomial distribution. **Prepare data**

```{r s5-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
binomials <- data.frame(
  X = rep(0:20, 3),
  P = c(dbinom(0:20, 10, 0.3), dbinom(0:20, 10, 0.5), dbinom(0:20, 20, 0.3)),
  setting = rep(c("n = 10, p = 0.3", "n = 10, p = 0.5", "n = 20, p = 0.3"), each = 21)
)
```

---

## Binomial Distributions in R

**Output figure**

```{r s5-3-2, echo=T, warning=FALSE, out.width="75%", fig.align='center', fig.height=3.5}
ggplot(binomials, aes(X, P)) +
  geom_col() +
  facet_wrap(~ setting) +
  labs(y = "P(X)")
```

## Mean, Variance, and Standard Deviation for the Binomial Distribution

:::{.callout-important}
## Binomial Mean, Variance, and Standard Deviation

$$\mu = n \cdot p \qquad \sigma^2 = n \cdot p \cdot q \qquad \sigma = \sqrt{n \cdot p \cdot q}$$
:::

These shorter formulas are mathematically equivalent to the general formulas $\mu = \Sigma X \cdot P(X)$ and $\sigma^2 = \Sigma[X^2 \cdot P(X)] - \mu^2$ of Section 5-2.

## Example 5-22 {#example-5-22}

:::{.callout-warning}
## Tossing a Coin

A coin is tossed 16 times. Find the mean, variance, and standard deviation of the number of heads that will be obtained.
:::

---

## Example 5-22

:::{.callout-tip}
## Solution

**Solve** — with $n = 16$, $p = \frac{1}{2}$, and $q = \frac{1}{2}$:

$$\mu = n \cdot p = 16 \cdot \tfrac{1}{2} = 8 \qquad \sigma^2 = n \cdot p \cdot q = 16 \cdot \tfrac{1}{2} \cdot \tfrac{1}{2} = 4 \qquad \sigma = \sqrt{4} = 2$$

```{r e5-22-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
16 * 0.5
16 * 0.5 * 0.5
sqrt(16 * 0.5 * 0.5)
```
:::

---

## Example 5-22

:::{.callout-tip}
## Solution

**Check with the Section 5-2 formulas** — apply $\mu = \Sigma X \cdot P(X)$ and $\sigma^2 = \Sigma[X^2 \cdot P(X)] - \mu^2$ to the full binomial distribution.

```{r e5-22-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X  <- 0:16
P  <- dbinom(X, 16, 0.5)
mu <- sum(X * P)

mu
sum(X^2 * P) - mu^2
```

The simplified binomial formulas give the same results.
:::

## Example 5-23 {#example-5-23}

:::{.callout-warning}
## Rolling a Die

A 10-sided die (with the numbers 1 through 10 on its faces) is rolled 500 times. Find the mean, variance, and standard deviation of the number of 7s that will be rolled.
:::

---

## Example 5-23

:::{.callout-tip}
## Solution

**Solve** — this is a binomial experiment with $n = 500$, $p = \frac{1}{10}$, and $q = \frac{9}{10}$.

$$\mu = 500 \cdot \tfrac{1}{10} = 50 \qquad \sigma^2 = 500 \cdot \tfrac{1}{10} \cdot \tfrac{9}{10} = 45 \qquad \sigma = \sqrt{45} \approx 6.708$$

```{r e5-23-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
500 * 1/10
500 * 1/10 * 9/10
sqrt(500 * 1/10 * 9/10)
```
:::

## Example 5-24 {#example-5-24}

:::{.callout-warning}
## Cross-Border Online Shopping

A consumer study reports that 75% of Taiwanese online shoppers have placed at least one order on a cross-border e-commerce site. If 1,200 shoppers are selected at random, find the mean, variance, and standard deviation of the number who have shopped across borders *(hypothetical data)*.
:::

---

## Example 5-24

:::{.callout-tip}
## Solution

**Solve** — this is a binomial situation, since a shopper either has or has not ordered from a cross-border site.

$$\mu = (1200)(0.75) = 900 \qquad \sigma^2 = (1200)(0.75)(0.25) = 225 \qquad \sigma = \sqrt{225} = 15$$

```{r e5-24-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
1200 * 0.75
1200 * 0.75 * 0.25
sqrt(1200 * 0.75 * 0.25)
```
:::

# Section 5-4: Other Types of Distributions {background-color="#2c4a6e"}

## Four More Discrete Distributions

In addition to the binomial distribution, four of the most commonly used distributions are the **multinomial**, the **Poisson**, the **hypergeometric** and the **geometric** distribution.

| Distribution | When to use it |
|---|---|
| Multinomial | More than two outcomes per trial |
| Poisson | $n$ large, $p$ small; occurrences over time, area or volume |
| Hypergeometric | Two outcomes, sampling **without** replacement |
| Geometric | Two outcomes, repeated **until** the first success |

## The Multinomial Distribution

If each trial in an experiment has more than two outcomes — "approve", "disapprove", "no opinion" — the binomial distribution cannot be used.

:::{.callout-note}
## Multinomial Experiment

A **multinomial experiment** is a probability experiment that satisfies the following four requirements:

1. There must be a fixed number of trials.
2. Each trial has a specific — but not necessarily the same — number of outcomes.
3. The trials are independent.
4. The probability of a particular outcome remains the same.
:::

## Formula for the Multinomial Distribution

:::{.callout-important}
## Formula for the Multinomial Distribution

If $X$ consists of events $E_1, E_2, \ldots, E_k$ with corresponding probabilities $p_1, p_2, \ldots, p_k$, and $X_i$ is the number of times $E_i$ occurs, then

$$P(X) = \frac{n!}{X_1! \cdot X_2! \cdot X_3! \cdots X_k!} \cdot p_1^{X_1} \cdot p_2^{X_2} \cdots p_k^{X_k}$$

where $X_1 + X_2 + \cdots + X_k = n$ and $p_1 + p_2 + \cdots + p_k = 1$.
:::

The multinomial distribution is a **general** distribution, and the binomial distribution is a special case of it with $k = 2$.

## Example 5-25 {#example-5-25}

:::{.callout-warning}
## Monthly Orders on an E-Commerce Platform

An e-commerce platform finds that the probabilities an active customer places 1, 2, 3, or 4 orders in a month are 0.45, 0.30, 0.20, and 0.05. If 10 active customers are selected, find the probability that 4 place one order, 3 place two orders, 2 place three orders, and 1 places four orders *(hypothetical data)*.
:::

---

## Example 5-25

:::{.callout-tip}
## Solution

**Solve** — $n = 10$; $X_1 = 4$, $X_2 = 3$, $X_3 = 2$, $X_4 = 1$; $p_1 = 0.45$, $p_2 = 0.30$, $p_3 = 0.20$, $p_4 = 0.05$.

$$P(X) = \frac{10!}{4!\,3!\,2!\,1!}(0.45)^4(0.30)^3(0.20)^2(0.05)^1 \approx 0.0279$$

```{r e5-25-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
factorial(10) / (factorial(4) * factorial(3) * factorial(2) * factorial(1))
12600 * 0.45^4 * 0.30^3 * 0.20^2 * 0.05^1
dmultinom(c(4, 3, 2, 1), prob = c(0.45, 0.30, 0.20, 0.05))
```

There is about a 0.0279 probability that the ten customers split in exactly this way.
:::

## Example 5-26 {#example-5-26}

:::{.callout-warning}
## Bubble-Tea Orders

A bubble-tea chain finds that the probabilities a customer entering a store buys 0, 1, 2, or 3 drinks are 0.20, 0.50, 0.20, and 0.10, respectively. If 9 customers enter the store, find the probability that 2 buy nothing, 4 buy 1 drink, 2 buy 2 drinks, and 1 buys 3 drinks *(hypothetical data)*.
:::

---

## Example 5-26

:::{.callout-tip}
## Solution

**Solve** — $n = 9$; $X_1 = 2$, $X_2 = 4$, $X_3 = 2$, $X_4 = 1$; $p_1 = 0.20$, $p_2 = 0.50$, $p_3 = 0.20$, $p_4 = 0.10$.

$$P(X) = \frac{9!}{2!\,4!\,2!\,1!}(0.20)^2(0.50)^4(0.20)^2(0.10)^1 = 0.0378$$

```{r e5-26-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
factorial(9) / (factorial(2) * factorial(4) * factorial(2) * factorial(1))
dmultinom(c(2, 4, 2, 1), prob = c(0.20, 0.50, 0.20, 0.10))
```

There is a 0.0378 probability that the results will occur as described.
:::

## Example 5-27 {#example-5-27}

:::{.callout-warning}
## Selecting Colored Marbles

A box contains 5 red marbles, 3 white marbles, and 2 black marbles. A marble is selected at random, its color is written down, and it is replaced each time. Find the probability that if 6 marbles are selected, 3 are red, 2 are white, and 1 is black.
:::

---

## Example 5-27

:::{.callout-tip}
## Solution

**Solve** — $n = 6$; $X_1 = 3$, $X_2 = 2$, $X_3 = 1$; $p_1 = \frac{5}{10}$, $p_2 = \frac{3}{10}$, $p_3 = \frac{2}{10}$.

$$P(X) = \frac{6!}{3!\,2!\,1!}\left(\tfrac{5}{10}\right)^3\left(\tfrac{3}{10}\right)^2\left(\tfrac{2}{10}\right)^1 = 60(0.00225) = 0.135$$

```{r e5-27-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
factorial(6) / (factorial(3) * factorial(2) * factorial(1))
dmultinom(c(3, 2, 1), prob = c(5, 3, 2) / 10)
```
:::

## The Poisson Distribution

:::{.callout-note}
## Poisson Experiment

A **Poisson experiment** is a probability experiment that satisfies the following requirements:

1. The random variable $X$ is the number of occurrences of an event over some interval (length, area, volume, period of time, etc.).
2. The occurrences occur randomly.
3. The occurrences are independent of one another.
4. The average number of occurrences over an interval is known.
:::

The **Poisson distribution** is useful when $n$ is large and $p$ is small and the independent variables occur over a period of time, or when a density of items is distributed over a given area or volume.

## Formula for the Poisson Distribution

:::{.callout-important}
## Formula for the Poisson Distribution

The probability of $X$ occurrences in an interval of time, volume, area, etc., for a variable where $\lambda$ is the mean number of occurrences per unit is

$$P(X;\, \lambda) = \frac{e^{-\lambda}\lambda^X}{X!} \qquad\text{where } X = 0, 1, 2, \ldots$$

The letter $e$ is a constant approximately equal to 2.7183. Round the answers to four decimal places.
:::

**Table A-3** in Appendix A gives $P$ for various values of $\lambda$ and $X$; in R the same values come from `dpois(X, lambda)`.

## Example 5-28 {#example-5-28}

:::{.callout-warning}
## Fabric Flaws

A textile mill in Changhua finds 120 flaws randomly distributed along a 400-metre roll of fabric. Find the probability that a given metre of the roll contains exactly 2 flaws *(hypothetical data)*.
:::

---

## Example 5-28

:::{.callout-tip}
## Solution

**Solve** — first find the mean number of flaws per metre, $\lambda = \frac{120}{400} = 0.3$; then substitute $X = 2$.

$$P(X;\, \lambda) = \frac{(2.7183)^{-0.3}(0.3)^2}{2!} \approx 0.0333$$

```{r e5-28-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
120 / 400
exp(-0.3) * 0.3^2 / factorial(2)
dpois(2, 0.3)
```

About 3.3% of the metres along the roll contain exactly 2 flaws.
:::

## Example 5-29 {#example-5-29}

:::{.callout-warning}
## Urgent Restock Requests

A logistics warehouse in Kaohsiung receives, on average, 2 urgent restock requests per hour. For any given hour, find the probability that it will receive the following *(hypothetical data)*.

```
a. At most 3 requests    b. At least 3 requests    c. 5 or more requests
```
:::

---

## Example 5-29

:::{.callout-tip}
## Solution

**Solve** — with $\lambda = 2$, "at most 3" means 0, 1, 2 or 3 requests; the other two parts are easier to obtain from the complement.

```{r e5-29-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
dpois(0:4, 2)
ppois(3, 2)
1 - ppois(1, 2)
1 - ppois(4, 2)
```

$P(X \leq 3) = 0.8571$, $P(X \geq 3) = 0.5940$ and $P(X \geq 5) = 0.0527$; the part **a** event is most likely to occur, and the part **c** event is least likely to occur.
:::

## Poisson Approximation to the Binomial

:::{.callout-important}
## When the Poisson Approximates the Binomial

The Poisson distribution can also be used to approximate the binomial distribution when the expected value $\lambda = n \cdot p$ is **less than 5**. (The same is true when $n \cdot q < 5$.)
:::

## Example 5-30 {#example-5-30}

:::{.callout-warning}
## Container Inspections

Customs selects approximately 1.5% of the containers arriving at Taichung port for a physical inspection. If a vessel discharges 300 containers, find the probability that exactly 6 of them are inspected *(hypothetical data)*.
:::

---

## Example 5-30

:::{.callout-tip}
## Solution

**Solve** — since $\lambda = n \cdot p = (300)(0.015) = 4.5$,

$$P(X;\, \lambda) = \frac{(2.7183)^{-4.5}(4.5)^6}{6!} \approx 0.1281$$

```{r e5-30-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
dpois(6, 300 * 0.015)
dbinom(6, 300, 0.015)
```

The exact binomial value ${}_{300}C_{6}(0.015)^6(0.985)^{294} \approx 0.1289$; the small difference reflects that the Poisson distribution is an approximation.
:::

## The Hypergeometric Distribution

When sampling is done **without replacement**, the binomial distribution does not give exact probabilities, since the trials are not independent. Suppose a committee of 4 people is to be selected from 7 women and 5 men: the number of ways to get 3 women and 1 man is ${}_{7}C_{3} \cdot {}_{5}C_{1} = 35 \cdot 5 = 175$, and ${}_{12}C_{4} = 495$, so $P(X) = \frac{175}{495} = \frac{35}{99}$.

:::{.callout-note}
## Hypergeometric Experiment

A **hypergeometric experiment** is a probability experiment that satisfies the following requirements:

1. There are a fixed number of trials.
2. There are two outcomes, and they can be classified as success or failure.
3. The sample is selected without replacement.
:::

## Formula for the Hypergeometric Distribution

:::{.callout-important}
## Formula for the Hypergeometric Distribution

Given a population with only two types of objects such that there are $a$ items of one kind and $b$ items of another kind and $a + b$ equals the total population, the probability $P(X)$ of selecting without replacement a sample of size $n$ with $X$ items of type $a$ and $n - X$ items of type $b$ is

$$P(X) = \frac{{}_{a}C_{X} \cdot {}_{b}C_{n-X}}{{}_{a+b}C_{n}}$$
:::

## Example 5-31 {#example-5-31}

:::{.callout-warning}
## Internship Applicants

Twelve students apply for a summer internship at an international trading company. Seven of them have studied abroad and five have not. If the manager selects 4 applicants at random, find the probability that all 4 have studied abroad.
:::

---

## Example 5-31

:::{.callout-tip}
## Solution

**Solve** — $a = 7$ applicants who studied abroad, $b = 5$ who did not, $n = 4$, $X = 4$, and $n - X = 0$.

$$P(X) = \frac{{}_{7}C_{4} \cdot {}_{5}C_{0}}{{}_{12}C_{4}} = \frac{35}{495} = \frac{7}{99} \approx 0.071$$

```{r e5-31-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
choose(7, 4) * choose(5, 0) / choose(12, 4)
# dhyper(X, a, b, n)
dhyper(4, 7, 5, 4)
```
:::

## Example 5-32 {#example-5-32}

:::{.callout-warning}
## Sensors Out of Calibration

A quality engineer knows that 4 of the 12 sensors in an incoming shipment are out of calibration. If 5 sensors are selected at random for testing, find the probability that exactly 2 of them are out of calibration.
:::

---

## Example 5-32

:::{.callout-tip}
## Solution

**Solve** — $a = 4$, $b = 8$, $n = 5$, and $X = 2$.

$$P(X) = \frac{{}_{4}C_{2} \cdot {}_{8}C_{3}}{{}_{12}C_{5}} = \frac{6 \cdot 56}{792} \approx 0.424$$

```{r e5-32-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
choose(4, 2) * choose(8, 3) / choose(12, 5)
dhyper(2, 4, 8, 5)
```
:::

## Example 5-33 {#example-5-33}

:::{.callout-warning}
## Defective Power Banks

A batch of 20 power banks is checked before shipment. Four of them are tested; if 1 or more of the 4 is defective, the whole batch is rejected. Find the probability that the batch will be rejected if there are actually 5 defective power banks in the batch.
:::

---

## Example 5-33

:::{.callout-tip}
## Solution

**Solve** — find the probability that none of the 4 tested is defective and subtract it from 1. Here $a = 5$, $b = 15$, $n = 4$, and $X = 0$.

$$P(0) = \frac{{}_{5}C_{0} \cdot {}_{15}C_{4}}{{}_{20}C_{4}} = \frac{1365}{4845} \approx 0.282$$

```{r e5-33-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
dhyper(0, 5, 15, 4)
1 - dhyper(0, 5, 15, 4)
```

There is a 0.718, or 71.8%, probability that the batch will be rejected when 5 of the 20 power banks are defective.
:::

## The Geometric Distribution

The **geometric distribution** is used when an experiment that has two outcomes is repeated **until a successful outcome is obtained** — flipping a coin until a head appears, or rolling a die until a 6 appears. The geometric probability distribution tells us when the success is likely to occur.

:::{.callout-note}
## Geometric Experiment

A **geometric experiment** is a probability experiment if it satisfies the following requirements:

1. Each trial has two outcomes that can be either success or failure.
2. The outcomes are independent of each other.
3. The probability of a success is the same for each trial.
4. The experiment continues until a success is obtained.
:::

## Formula for the Geometric Distribution

:::{.callout-important}
## Formula for the Geometric Distribution

If $p$ is the probability of a success on each trial and $n$ is the number of the trial at which the first success occurs, then the probability of getting the first success on the $n$th trial is

$$P(n) = p(1 - p)^{n-1} \qquad\text{where } n = 1, 2, 3, \ldots$$
:::

## Example 5-34 {#example-5-34}

:::{.callout-warning}
## Rolling a Die

A die is rolled repeatedly. Find the probability that the first 1 appears on the third roll.
:::

---

## Example 5-34

:::{.callout-tip}
## Solution

**Solve** — to get the first 1 on the third roll, the first two rolls must show something other than a 1:

$$P(\text{not }1,\ \text{not }1,\ 1) = \tfrac{5}{6} \cdot \tfrac{5}{6} \cdot \tfrac{1}{6} = \frac{25}{216} \approx 0.116$$

$$P(n) = p(1-p)^{n-1} = \tfrac{1}{6}\left(1 - \tfrac{1}{6}\right)^{3-1} = \tfrac{1}{6}\left(\tfrac{5}{6}\right)^2 = \frac{25}{216}$$

```{r e5-34-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
1/6 * (1 - 1/6)^(3 - 1)
# dgeom() uses the number of failures before the first success
dgeom(3 - 1, 1/6)
```
:::

## Example 5-35 {#example-5-35}

:::{.callout-warning}
## Booking Through an Online Travel Agency

About 35% of Taiwanese outbound travellers book their trip through an online travel agency. If travellers are contacted one at a time, find the probability that the fifth traveller contacted is the first one who booked through an online travel agency *(hypothetical data)*.
:::

---

## Example 5-35

:::{.callout-tip}
## Solution

**Solve** — let $p = 0.35$ and $n = 5$.

$$P(5) = (0.35)(1 - 0.35)^{5-1} = (0.35)(0.65)^4 \approx 0.0625$$

```{r e5-35-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
0.35 * (1 - 0.35)^(5 - 1)
dgeom(5 - 1, 0.35)
```

There is a 0.0625 probability that the fifth traveller contacted is the first one who booked through an online travel agency.
:::

## Table 5-1: Summary of Discrete Distributions

```{r s5-4-1, echo=F, warning=FALSE, out.width="80%", fig.align='center'}
summary_df <- data.frame(
  Distribution = c("Binomial", "Multinomial", "Poisson",
                   "Hypergeometric", "Geometric"),
  `Used when` = c(
    "Two outcomes, fixed number of independent trials, constant p",
    "More than two outcomes, constant probabilities, independent, fixed n",
    "n large and p small; occurrences over time, area or volume",
    "Two outcomes and sampling is done without replacement",
    "Two outcomes; the first success occurs on the nth trial"),
  check.names = FALSE)
knitr::kable(summary_df, align = "l")
```

## Table 5-1: Summary of Discrete Distributions (cont.)

Probability formulas — binomial and multinomial:

$$P(X) = {}_{n}C_{X}\,p^X q^{n-X} \qquad
P(X) = \frac{n!}{X_1!\cdots X_k!}p_1^{X_1}\cdots p_k^{X_k}$$

Poisson and hypergeometric:

$$P(X;\lambda) = \frac{e^{-\lambda}\lambda^X}{X!} \qquad
P(X) = \frac{{}_{a}C_{X} \cdot {}_{b}C_{n-X}}{{}_{a+b}C_{n}}$$

Geometric:

$$P(n) = p(1-p)^{n-1}$$

## Important Terms

:::{.callout-important}
## Chapter 5 Vocabulary

binomial distribution · binomial experiment · discrete probability distribution · expected value · geometric distribution · geometric experiment · hypergeometric distribution · hypergeometric experiment · multinomial distribution · multinomial experiment · Poisson distribution · Poisson experiment · random variable
:::

## Key Formulas

:::{.callout-important}
## Probability Distributions

Mean of a probability distribution: $\mu = \Sigma X \cdot P(X)$

Variance and standard deviation of a probability distribution:
$$\sigma^2 = \Sigma[X^2 \cdot P(X)] - \mu^2 \qquad \sigma = \sqrt{\Sigma[X^2 \cdot P(X)] - \mu^2}$$

Expected value: $E(X) = \Sigma X \cdot P(X)$
:::

---

## Key Formulas

:::{.callout-important}
## The Binomial Distribution

$$P(X) = \frac{n!}{(n-X)!\,X!} \cdot p^X \cdot q^{n-X} \qquad\text{or}\qquad P(X) = {}_{n}C_{X} \cdot p^X \cdot q^{n-X}$$

$$\mu = n \cdot p \qquad \sigma^2 = n \cdot p \cdot q \qquad \sigma = \sqrt{n \cdot p \cdot q}$$
:::

---

## Key Formulas

:::{.callout-important}
## Other Distributions

$$P(X) = \frac{n!}{X_1! \cdot X_2! \cdots X_k!} \cdot p_1^{X_1} \cdot p_2^{X_2} \cdots p_k^{X_k}
\qquad P(X;\, \lambda) = \frac{e^{-\lambda}\lambda^X}{X!}$$

$$P(X) = \frac{{}_{a}C_{X} \cdot {}_{b}C_{n-X}}{{}_{a+b}C_{n}} \qquad P(n) = p(1-p)^{n-1}$$
:::

## Key Takeaways

:::{.callout-important}
- A **discrete probability distribution** consists of the values a random variable can assume and the corresponding probabilities; it requires $\Sigma P(X) = 1$ and $0 \leq P(X) \leq 1$
- **Mean**, **variance** and **standard deviation** of a probability distribution: $\mu = \Sigma X \cdot P(X)$, $\sigma^2 = \Sigma[X^2 \cdot P(X)] - \mu^2$, $\sigma = \sqrt{\sigma^2}$
- The **expected value** $E(X) = \mu$ is the theoretical average; a game is *fair* when $E(X) = 0$
- A **binomial experiment** has a fixed number of trials, two outcomes, independent trials and a constant $p$
- The binomial probability can be computed either with the **factorial form** $\frac{n!}{(n-X)!X!}p^Xq^{n-X}$ or with the **combination formula** ${}_{n}C_{X}\,p^Xq^{n-X}$ — the two are identical
- For a binomial variable, $\mu = np$, $\sigma^2 = npq$ and $\sigma = \sqrt{npq}$
- The **multinomial** distribution generalises the binomial to more than two outcomes per trial
- The **Poisson** distribution counts occurrences over an interval and approximates the binomial when $\lambda = np < 5$
- The **hypergeometric** distribution applies when sampling is done without replacement
- The **geometric** distribution gives the probability that the first success occurs on the $n$th trial: $P(n) = p(1-p)^{n-1}$
:::

## Acknowledgement

-   **Copyright notice.** These teaching materials follow the organization and terminology of **Bluman, A. G. (2023). Elementary statistics: A step by step approach (11th ed.). McGraw Hill**. All rights in the original work are reserved by its authors and publishers.

-   **Original examples.** Every worked example, data set, and R script in these slides was written for this course. The data are hypothetical unless stated otherwise.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
