---
title: "Statistics"
subtitle: "Chapter 5: Discrete Probability Distributions"
author: "Yu-You Liou"
institute: "Shih Chien University"
date: "`r Sys.Date()`"
language:
  callout-warning-title: "Problem"
  callout-important-title: "Key point"
format:
  revealjs:
    html-math-method: katex
    smaller: true
    theme: serif
    footer: "Elementary Statistics: A Step by Step Approach"
    logo: "logo.svg"
    editor: source
    slide-number: true
    scrollable: false
    transition: "none"
    menu:
      useTextContentForMissingTitles: false
      hideMissingTitles: true
      side: 'left'
      titleSelector: 'h1'
title-slide-attributes:
  data-notes: ""
include-after-body:
  text: |
    <script>
    Reveal.on('ready', function() {
      var title = Reveal.getSlides()[0];
      if (title) {
        title.insertAdjacentHTML('beforeend',
          '<p style="font-size:0.5em; margin-top:1.5em;">' +
          '<a href="ch5-zh.html"  style="color:#555;">[中文版由此去]</a>' +
          '</p>');
      }
    });
    </script>
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
```

## Overview

**Sections covered:**

- **5-1** Probability Distributions
- **5-2** Mean, Variance, Standard Deviation, and Expectation
- **5-3** The Binomial Distribution
- **5-4** Other Types of Distributions *(Optional)*

**Learning objectives:**

- Construct and interpret discrete probability distributions
- Compute the mean, variance, and expected value of a discrete variable
- Apply the binomial probability formula and identify binomial experiments
- Use the multinomial, Poisson, and hypergeometric distributions

# Section 5-1: Probability Distributions {background-color="#2c4a6e"}

## Random Variables

:::{.callout-note}
## Key Definitions

A **random variable** is a variable whose value is determined by the outcome of a probability experiment.

A **discrete random variable** has a finite or countable number of possible values.

A **continuous random variable** can take on any value in an interval on the number line.
:::

**Examples of discrete random variables:**

- Number of heads in 3 coin tosses: 0, 1, 2, 3
- Number of children in a family: 0, 1, 2, 3, …
- Number of defects in a batch of products

**Examples of continuous random variables:**

- Height, weight, temperature, time

## Discrete Probability Distributions

:::{.callout-note}
## Definition

A **discrete probability distribution** consists of all possible values a discrete random variable can take on, along with their corresponding probabilities.
:::

**Two requirements for a valid probability distribution:**

1. The sum of all probabilities must equal 1: $\sum P(X) = 1$
2. Each probability must be between 0 and 1: $0 \leq P(X) \leq 1$

Probability distributions can be presented as tables, formulas, or graphs.

## Example 5-1 {#example-5-1}

:::{.callout-warning}
## Rolling a Die

Construct a probability distribution for the outcomes of rolling a single fair die.
:::

## Example 5-1

:::{.callout-tip}
## Solution

**R implementation**

```{r e5-1-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 1:6
PX <- rep(1/6, 6)
dist <- data.frame(X = X, `P(X)` = round(PX, 4), check.names = FALSE)
```
:::
---

```{r e5-1-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
knitr::kable(dist, caption = "Probability Distribution: Rolling a Die")
```

Each outcome has an equal probability of $\frac{1}{6} \approx 0.167$. Note that $\sum P(X) = 1$.

## Example 5-2 {#example-5-2}

:::{.callout-warning}
## Tossing Three Coins

Represent the probability distribution for the number of heads when 3 fair coins are tossed, and draw a graph.
:::

## Example 5-2

:::{.callout-tip}
## Solution

**R implementation**

```{r e5-2-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 0:3
PX <- c(1/8, 3/8, 3/8, 1/8)
dist <- data.frame(X = X, `P(X)` = PX, check.names = FALSE)
knitr::kable(dist, caption = "Three-Coin Toss: Number of Heads")
```
:::
---

```{r e5-2-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
ggplot(dist, aes(x = factor(X), y = `P(X)`)) +
  geom_col(fill = "steelblue", width = 0.5) +
  labs(title = "Probability Distribution: 3 Coin Tosses",
       x = "Number of Heads (X)", y = "P(X)") +
  theme_minimal()
```

## Example 5-3 {#example-5-3}

:::{.callout-warning}
## Baseball World Series

The following data show the number of games played in 40 World Series. Construct a probability distribution and draw a graph.

| Games played | 4 | 5 | 6 | 7 |
|---|---|---|---|---|
| Frequency | 8 | 7 | 9 | 16 |
:::
---

```{r e5-3-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 4:7
freq <- c(8, 7, 9, 16)
PX <- freq / 40
dist <- data.frame(X = X, Frequency = freq, `P(X)` = PX, check.names = FALSE)
knitr::kable(dist, caption = "World Series: Games Played in 40 Series")
```

---
  
```{r e5-3-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
ggplot(dist, aes(x = factor(X), y = `P(X)`)) +
  geom_col(fill = "darkorange", width = 0.5) +
  labs(title = "World Series Games Played",
       x = "Number of Games (X)", y = "P(X)") +
  theme_minimal()
```

## Example 5-4 {#example-5-4}

:::{.callout-warning}
## Checking Valid Distributions

Determine whether each of the following is a valid probability distribution.
:::

## Example 5-4

:::{.callout-tip}
## Solution

**R implementation**

```{r e5-4-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# (a) Valid
a_X <- 0:4; a_P <- c(0.5, 0.3, 0.15, 0.04, 0.01)
cat("(a) Sum =", sum(a_P), "| All in [0,1]:", all(a_P >= 0 & a_P <= 1),
    "→", ifelse(sum(a_P) == 1 & all(a_P >= 0 & a_P <= 1), "VALID", "INVALID"), "\n")

# (b) Not valid: sum ≠ 1
b_X <- 1:3; b_P <- c(1/3, 1/3, 1/4)
cat("(b) Sum =", round(sum(b_P), 4), "→", ifelse(abs(sum(b_P)-1) < 1e-9, "VALID", "INVALID"), "\n")
```
:::
---

```{r e5-4-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# (c) Valid
c_X <- 1:5; c_P <- c(0.1, 0.3, 0.4, 0.1, 0.1)
cat("(c) Sum =", sum(c_P), "→", ifelse(sum(c_P) == 1 & all(c_P >= 0 & c_P <= 1), "VALID", "INVALID"), "\n")

# (d) Not valid: negative probability
d_X <- c(2, 4, 6); d_P <- c(-0.1, 0.6, 0.5)
cat("(d) Has negative P:", any(d_P < 0), "→ INVALID\n")
```

:::{.callout-important}
A distribution is valid only when **both** conditions hold: $\sum P(X) = 1$ **and** $0 \leq P(X) \leq 1$ for every value.
:::

# Section 5-2: Mean, Variance, Standard Deviation, and Expectation {background-color="#2c4a6e"}

## Formulas for Mean, Variance, and Standard Deviation

:::{.callout-note}
## Formulas

**Mean (Expected Value):**
$$\mu = E(X) = \sum X \cdot P(X)$$

**Variance:**
$$\sigma^2 = \sum \left[X^2 \cdot P(X)\right] - \mu^2$$

**Standard Deviation:**
$$\sigma = \sqrt{\sigma^2}$$
:::

The mean describes the long-run average outcome of the experiment. The variance and standard deviation measure the spread of the distribution around the mean.

## Example 5-5 {#example-5-5}

:::{.callout-warning}
## Rolling a Die

Find the mean of the number of spots that appear when a fair die is rolled.
:::

## Example 5-5

:::{.callout-tip}
## Solution

Each outcome $1, 2, \ldots, 6$ has probability $\frac{1}{6}$, so

$$\mu = \sum X \cdot P(X) = 1 \cdot \frac{1}{6} + 2 \cdot \frac{1}{6} + \cdots + 6 \cdot \frac{1}{6} = \frac{21}{6} = 3.5$$

:::

```{r e5-5-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 1:6
PX <- rep(1/6, 6)
mu <- sum(X * PX)
df <- data.frame(X = X, `P(X)` = round(PX, 4), `X·P(X)` = round(X * PX, 4), check.names = FALSE)
knitr::kable(df, caption = "Rolling a Die: Number of Spots")
```

----

```{r e5-5-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
cat("Mean μ =", mu)
```

When a die is tossed many times, the theoretical mean is **3.5** — even though the die can never show 3.5.

## Example 5-6 {#example-5-6}

:::{.callout-warning}
## Children in a Family

In a family with two children, find the mean of the number of children who will be girls.
:::

## Example 5-6

:::{.callout-tip}
## Solution

The sample space is BB, BG, GB, GG, so $P(0) = \frac{1}{4}$, $P(1) = \frac{2}{4}$, $P(2) = \frac{1}{4}$.

$$\mu = \sum X \cdot P(X) = 0 \cdot \frac{1}{4} + 1 \cdot \frac{2}{4} + 2 \cdot \frac{1}{4} = 1$$

:::

```{r e5-6-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 0:2
PX <- c(1/4, 2/4, 1/4)
mu <- sum(X * PX)
df <- data.frame(X = X, `P(X)` = PX, `X·P(X)` = X * PX, check.names = FALSE)
knitr::kable(df, caption = "Number of Girls in Two-Child Family")
```

----

```{r e5-6-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
cat("Mean μ =", mu)
```

On average, a two-child family will have **1 girl**.

## Example 5-7 {#example-5-7}
:::{.callout-warning}
If 3 coins are tossed, find the mean of the number of heads.
:::

```{r e5-7-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 0:3
PX <- c(1/8, 3/8, 3/8, 1/8)
mu <- sum(X * PX)
df <- data.frame(X = X, `P(X)` = PX, `X·P(X)` = X * PX, check.names = FALSE)
knitr::kable(df, caption = "Three Coins: Number of Heads")
```

----

```{r e5-7-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
cat("Mean μ =", mu)
```

When 3 coins are tossed many times, the average number of heads is **1.5**.

## Example 5-8 {#example-5-8}
:::{.callout-warning}
The probability distribution below shows the number of trips of five or more nights that American adults take per year. Find the mean.

| X | 0 | 1 | 2 | 3 | 4 |
|---|---|---|---|---|---|
| P(X) | 0.08 | 0.60 | 0.25 | 0.04 | 0.03 |
:::

----

```{r e5-8-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 0:4
PX <- c(0.08, 0.60, 0.25, 0.04, 0.03)
mu <- sum(X * PX)
df <- data.frame(X = X, `P(X)` = PX, `X·P(X)` = X * PX, check.names = FALSE)
knitr::kable(df)
cat("Mean μ =", mu, "trips per year")
```

American adults take an average of approximately **1.34** overnight trips of 5+ nights per year.

## Example 5-9 {#example-5-9}
:::{.callout-warning}
Find the variance and standard deviation for the probability distribution of rolling a fair die ([Example 5-5](#example-5-5)).
:::

```{r e5-9-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 1:6
PX <- rep(1/6, 6)
mu <- sum(X * PX)
sigma2 <- sum(X^2 * PX) - mu^2
sigma <- sqrt(sigma2)
df <- data.frame(X = X, `P(X)` = PX, `X²·P(X)` = X^2 * PX, check.names = FALSE)
```

----

```{r e5-9-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
knitr::kable(df)
cat("μ =", mu, "\nσ² = Σ[X²·P(X)] - μ² =", round(sum(X^2*PX),4), "-", mu^2, "=", round(sigma2,4),
    "\nσ =", round(sigma, 2))
```

## Example 5-10 {#example-5-10}
:::{.callout-warning}
A bag contains 5 balls numbered 3, 3, 4, 5, and 5. A ball is selected at random. Find the mean, variance, and standard deviation.
:::

```{r e5-10-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- c(3, 4, 5)
PX <- c(2/5, 1/5, 2/5)
mu <- sum(X * PX)
sigma2 <- sum(X^2 * PX) - mu^2
sigma <- sqrt(sigma2)
df <- data.frame(X = X, `P(X)` = PX, `X·P(X)` = X*PX, `X²·P(X)` = X^2*PX, check.names = FALSE)
knitr::kable(df)
cat("μ =", mu, "| σ² =", round(sigma2, 4), "| σ =", round(sigma, 4))
```

## Example 5-11 {#example-5-11}
:::{.callout-warning}
The probability distribution below shows the number of callers on hold for a radio talk show. Find the mean, variance, and standard deviation.

| X | 0 | 1 | 2 | 3 | 4 |
|---|---|---|---|---|---|
| P(X) | 0.1 | 0.3 | 0.3 | 0.2 | 0.1 |
:::

---

```{r e5-11-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- 0:4
PX <- c(0.1, 0.3, 0.3, 0.2, 0.1)
mu <- sum(X * PX)
sigma2 <- sum(X^2 * PX) - mu^2
sigma <- sqrt(sigma2)
df <- data.frame(X = X, `P(X)` = PX, `X·P(X)` = X*PX, `X²·P(X)` = X^2*PX, check.names = FALSE)
knitr::kable(df)
cat("μ =", mu, "| σ² =", round(sigma2, 4), "| σ =", round(sigma, 4))
```

## Expected Value

:::{.callout-note}
## Expected Value

The **expected value** $E(X)$ of a discrete random variable equals its mean:
$$E(X) = \mu = \sum X \cdot P(X)$$

Expected value is used to evaluate the long-run average gain or loss in games of chance, investments, and insurance.
:::

:::{.callout-warning}
## Caution
If $E(X) < 0$, the player or investor expects a **net loss** in the long run.
:::

## Example 5-12 {#example-5-12}

:::{.callout-warning}
## Winning Lottery Tickets

A lottery offers one $1000 prize, two $500 prizes, and five $100 prizes. One thousand tickets are sold at $3 each. Find the expected value of a ticket.
:::
---

```{r e5-12-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- c(1000-3, 500-3, 100-3, -3)   # net gain
PX <- c(1/1000, 2/1000, 5/1000, 992/1000)
EX <- sum(X * PX)
df <- data.frame(`Net Gain X` = X, `P(X)` = PX, `X·P(X)` = X*PX, check.names = FALSE)
knitr::kable(df)
cat("E(X) = $", round(EX, 4))
```

The expected value is **negative**, meaning a player can expect to lose about $0.50 per ticket on average.

## Example 5-13 {#example-5-13}

:::{.callout-warning}
## Special Die

A special die has faces labeled 1, 4, 4, 6, 6, 6. Find the expected value when rolling it.
:::

## Example 5-13

:::{.callout-tip}
## Solution

The die shows 6 spots on 3 sides, 4 spots on 2 sides, and 1 spot on 1 side, so $P(6) = \frac{3}{6}$, $P(4) = \frac{2}{6}$, and $P(1) = \frac{1}{6}$.

$$E(X) = 1 \cdot \frac{1}{6} + 4 \cdot \frac{2}{6} + 6 \cdot \frac{3}{6} = \frac{27}{6} = 4.5$$

:::

```{r e5-13-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
X <- c(1, 4, 6)
PX <- c(1/6, 2/6, 3/6)
EX <- sum(X * PX)
cat("E(X) =", EX)
```

Only 1, 4, or 6 spots can occur on a single roll, but over many rolls the average is **4.5**.

## Example 5-14 {#example-5-14}

:::{.callout-warning}
## Bond Investment

An investor is considering two bonds. Bond X: 70% chance of earning $120, 30% chance of losing $40. Bond Y: 60% chance of earning $90, 40% chance of losing $30. Which bond has the better expected return?
:::

## Example 5-14

:::{.callout-tip}
## Solution

$$E(\text{Bond X}) = 120(0.70) + (-40)(0.30) = 84 - 12 = 72$$

$$E(\text{Bond Y}) = 90(0.60) + (-30)(0.40) = 54 - 12 = 42$$

:::

```{r e5-14-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# Bond X
X_x <- c(120, -40); P_x <- c(0.70, 0.30)
EX <- sum(X_x * P_x)

# Bond Y
X_y <- c(90, -30); P_y <- c(0.60, 0.40)
EY <- sum(X_y * P_y)

cat("E(Bond X) = $", EX, "\n")
cat("E(Bond Y) = $", EY, "\n")
cat("Better investment:", ifelse(EX > EY, "Bond X", "Bond Y"))
```

**Bond X** has the higher expected return ($72 vs $42).

# Section 5-3: The Binomial Distribution {background-color="#2c4a6e"}

## Binomial Experiments

:::{.callout-note}
## Definition

A **binomial experiment** satisfies all four of the following conditions:

1. There is a **fixed number** of trials $n$.
2. Each trial has only **two possible outcomes**: success or failure.
3. The outcomes of each trial are **independent** of one another.
4. The **probability of success** $p$ remains the same for every trial.
:::

**Examples:** Tossing a coin $n$ times; surveying $n$ people with a yes/no question; testing $n$ components with replacement.

## Notation for the Binomial Distribution

:::{.callout-note}
## Notation

| Symbol | Meaning |
|---|---|
| $P(S)$ | Probability of success |
| $P(F)$ | Probability of failure |
| $p$ | Numerical probability of success |
| $q = 1 - p$ | Numerical probability of failure |
| $n$ | Number of trials |
| $X$ | Number of successes in $n$ trials |

Note: $0 \leq X \leq n$ and $X = 0, 1, 2, \ldots, n$
:::

## Binomial Probability Formula

:::{.callout-note}
## Formula

In a binomial experiment, the probability of exactly $X$ successes in $n$ trials is:

$$P(X) = \frac{n!}{(n-X)!\,X!} \cdot p^X \cdot q^{n-X}$$

where $q = 1 - p$.
:::

- $\dfrac{n!}{(n-X)!\,X!} = {}_nC_X$ counts the number of ways to arrange $X$ successes among $n$ trials
- $p^X$ is the probability of $X$ successes
- $q^{n-X}$ is the probability of $n - X$ failures

## Example 5-15: Tossing Coins {#example-5-15}

:::{.callout-warning}
A coin is tossed 3 times. Find the probability of getting exactly 2 heads.
:::

```{r e5-15-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 3; X <- 2; p <- 0.5; q <- 1 - p
prob <- choose(n, X) * p^X * q^(n - X)
cat("n =", n, ", X =", X, ", p =", p, ", q =", q, "\n")
cat("P(2 heads) = C(3,2) × (0.5)² × (0.5)¹ =",
    choose(n,X), "×", p^X, "×", q^(n-X), "=", prob)

# Verification using dbinom
cat("\nVerification via dbinom:", dbinom(2, size = 3, prob = 0.5))
```

:::{.callout-important}
The coin-toss problem meets all 4 binomial conditions: fixed $n=3$ trials, two outcomes (H/T), independent tosses, constant $p=0.5$.
:::

## Example 5-16 {#example-5-16}
:::{.callout-warning}
A survey found that 1 out of 5 Americans has visited a doctor in any given month. If 10 people are selected at random, find the probability that exactly 3 visited a doctor last month.
:::

```{r e5-16-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 10; X <- 3; p <- 1/5; q <- 1 - p
prob <- choose(n, X) * p^X * q^(n - X)
cat("n =", n, ", X =", X, ", p =", round(p, 2), ", q =", round(q, 2), "\n")
cat("P(3) = C(10,3) × (1/5)³ × (4/5)⁷ =", round(prob, 3))
cat("\ndbinom check:", round(dbinom(X, n, p), 3))
```

## Example 5-17 {#example-5-17}
:::{.callout-warning}
A survey found that 30% of teenage consumers receive spending money from part-time jobs. If 5 teenagers are selected at random, find the probability that **at least 3** have part-time jobs.
:::

```{r e5-17-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 5; p <- 0.30; q <- 0.70
P3 <- choose(n,3) * p^3 * q^2
P4 <- choose(n,4) * p^4 * q^1
P5 <- choose(n,5) * p^5 * q^0
total <- P3 + P4 + P5
```

---

```{r e5-17-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
cat("P(3) =", round(P3, 3), "\n")
cat("P(4) =", round(P4, 3), "\n")
cat("P(5) =", round(P5, 3), "\n")
cat("P(at least 3) = P(3)+P(4)+P(5) =", round(total, 3))

# Using pbinom complement
cat("\nVerification: 1 - P(X ≤ 2) =", round(1 - pbinom(2, n, p), 3))
```

## Example 5-18 {#example-5-18}

:::{.callout-warning}
## Using the Binomial Table

Solve [Example 5-15](#example-5-15) (3 coins, P(exactly 2 heads)) using the binomial probability table (Table B) or R.
:::

## Example 5-18

:::{.callout-tip}
## Solution

**R implementation**

```{r e5-18-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# Table B: look up n=3, X=2, p=0.5
result <- dbinom(2, size = 3, prob = 0.5)
cat("From binomial table (or dbinom): P(X=2 | n=3, p=0.5) =", result)
```
:::
---

```{r e5-18-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# Full distribution for n=3, p=0.5
dist <- data.frame(
  X = 0:3,
  `P(X)` = dbinom(0:3, 3, 0.5),
  check.names = FALSE
)
knitr::kable(dist, digits = 4, caption = "Binomial Distribution: n=3, p=0.5")
```

## Example 5-19 {#example-5-19}

:::{.callout-warning}
## Fear of Being Home Alone at Night

A survey reported that 5% of Americans are afraid of being alone at night. If 20 Americans are randomly selected, find the probability that:

a. Exactly 5 are afraid

b. At most 3 are afraid

c. At least 3 are afraid
:::
---

```{r e5-19-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 20; p <- 0.05

# (a) P(X = 5)
a <- dbinom(5, n, p)
cat("(a) P(X = 5) =", round(a, 4), "\n")

# (b) P(X ≤ 3)
b <- sum(dbinom(0:3, n, p))
cat("(b) P(X ≤ 3) = P(0)+P(1)+P(2)+P(3) =", round(b, 4), "\n")

# (c) P(X ≥ 3) = 1 - P(X ≤ 2)
c <- 1 - sum(dbinom(0:2, n, p))
cat("(c) P(X ≥ 3) = 1 - P(X ≤ 2) =", round(c, 4))
```

## Example 5-20 {#example-5-20}

:::{.callout-warning}
## Driving While Intoxicated

Reports indicate that 70% of single-vehicle traffic fatalities at night on weekends involve an intoxicated driver. If 15 such fatalities are selected, find the probability that exactly 12 involve an intoxicated driver.
:::

## Example 5-20

:::{.callout-tip}
## Solution

Here $n = 15$, $p = 0.70$, and $X = 12$:

$$P(12) = \frac{15!}{(15-12)!\,12!}(0.70)^{12}(0.30)^{3} \approx 0.170$$

:::

```{r e5-20-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 15; X <- 12; p <- 0.70; q <- 1 - p
prob <- choose(n, X) * p^X * q^(n - X)
cat("P(X = 12) = C(15,12) × (0.7)^12 × (0.3)^3 =", round(prob, 4))

# Verification using dbinom
cat("\nVerification via dbinom:", round(dbinom(12, 15, 0.7), 4))
```

The probability that exactly 12 of the 15 fatalities involve an intoxicated driver is about **0.17**.

## Mean, Variance, and Standard Deviation for the Binomial Distribution

:::{.callout-note}
## Formulas

$$\mu = n \cdot p$$

$$\sigma^2 = n \cdot p \cdot q$$

$$\sigma = \sqrt{n \cdot p \cdot q}$$

These are algebraically equivalent to the general formulas $\mu = \sum X \cdot P(X)$ and $\sigma^2 = \sum[X^2 \cdot P(X)] - \mu^2$, but are much simpler to use for binomial variables.
:::

## Example 5-21 {#example-5-21}
:::{.callout-warning}
A coin is tossed 4 times. Find the mean, variance, and standard deviation of the number of heads.
:::

---

```{r e5-21-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 4; p <- 0.5; q <- 0.5
mu <- n * p
sigma2 <- n * p * q
sigma <- sqrt(sigma2)
cat("μ = n·p =", n, "×", p, "=", mu, "\n")
cat("σ² = n·p·q =", n, "×", p, "×", q, "=", sigma2, "\n")
cat("σ = √σ² = √", sigma2, "=", sigma, "\n")

# Verification using general formulas
X <- 0:4
PX <- dbinom(X, n, p)
mu_check <- sum(X * PX)
sigma2_check <- sum(X^2 * PX) - mu_check^2
cat("\nVerification: μ =", mu_check, "| σ² =", sigma2_check)
```

## Example 5-22 {#example-5-22}

:::{.callout-warning}
## Rolling a Die 480 Times

A die is rolled 480 times. Find the mean, variance, and standard deviation for the number of 3s rolled.
:::

## Example 5-22

:::{.callout-tip}
## Solution

Getting a 3 is a success, so $n = 480$, $p = \frac{1}{6}$, and $q = \frac{5}{6}$:

$$\mu = n \cdot p = 480 \cdot \frac{1}{6} = 80$$

$$\sigma^2 = n \cdot p \cdot q = 480 \cdot \frac{1}{6} \cdot \frac{5}{6} \approx 66.67$$

$$\sigma = \sqrt{66.67} \approx 8.16$$

:::

```{r e5-22-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 480; p <- 1/6; q <- 5/6
mu <- n * p
sigma2 <- n * p * q
sigma <- sqrt(sigma2)
cat("μ = n·p =", mu, "\n")
cat("σ² = n·p·q =", round(sigma2, 2), "\n")
cat("σ = √σ² =", round(sigma, 2))
```

## Example 5-23 {#example-5-23}

:::{.callout-warning}
## Likelihood of Twins

Approximately 2% of all American births result in twins. If a random sample of 8000 births is taken, find the mean, variance, and standard deviation of the number of twin births.
:::

## Example 5-23

:::{.callout-tip}
## Solution

This is a binomial situation with $n = 8000$, $p = 0.02$, and $q = 0.98$:

$$\mu = n \cdot p = 8000 \cdot 0.02 = 160$$

$$\sigma^2 = n \cdot p \cdot q = 8000 \cdot 0.02 \cdot 0.98 = 156.8$$

$$\sigma = \sqrt{156.8} \approx 12.5$$

:::

```{r e5-23-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 8000; p <- 0.02; q <- 0.98
mu <- n * p
sigma2 <- n * p * q
sigma <- sqrt(sigma2)
cat("μ = n·p =", mu, "births\n")
cat("σ² = n·p·q =", sigma2, "\n")
cat("σ = √σ² =", round(sigma, 1))
```

:::{.callout-important}
In a sample of 8000 births, we expect on average **160** twin births, with $\sigma \approx 12.5.$
:::

## Binomial Distribution: Visualization

```{r e5-22-2, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
# Plot binomial distributions for different p values
params <- list(
  list(n=10, p=0.3, label="n=10, p=0.3"),
  list(n=10, p=0.5, label="n=10, p=0.5"),
  list(n=20, p=0.3, label="n=20, p=0.3")
)

df_all <- do.call(rbind, lapply(params, function(par) {
  x <- 0:par$n
  data.frame(X = x, PX = dbinom(x, par$n, par$p), Label = par$label)
}))
```

---

```{r e5-22-3, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
ggplot(df_all, aes(x = X, y = PX, fill = Label)) +
  geom_col(position = "dodge", width = 0.7) +
  facet_wrap(~Label, scales = "free_x") +
  labs(title = "Binomial Distributions", x = "X", y = "P(X)") +
  theme_minimal() +
  theme(legend.position = "none")
```

# Section 5-4: Other Types of Distributions (Optional) {background-color="#2c4a6e"}

## Overview of Other Discrete Distributions

When the binomial model does not apply, other discrete distributions are available:

| Distribution | When to Use |
|---|---|
| **Multinomial** | More than two outcomes per trial |
| **Poisson** | Events over time/area/volume; large $n$, small $p$ |
| **Hypergeometric** | Sampling **without** replacement |

## The Multinomial Distribution

:::{.callout-note}
## Formula

If an experiment has $k$ possible outcomes $E_1, E_2, \ldots, E_k$ with probabilities $p_1, p_2, \ldots, p_k$, and $X_i$ is the number of times outcome $E_i$ occurs in $n$ independent trials, then:

$$P(X) = \frac{n!}{X_1!\, X_2!\, X_3!\, \cdots\, X_k!} \cdot p_1^{X_1} \cdot p_2^{X_2} \cdots p_k^{X_k}$$

where $X_1 + X_2 + \cdots + X_k = n$ and $p_1 + p_2 + \cdots + p_k = 1$.
:::

The multinomial distribution generalizes the binomial — the binomial is a special case with $k = 2$.

## Example 5-24 {#example-5-24}

:::{.callout-warning}
## Leisure Activities

In a large city, 50% of residents choose a movie, 30% choose dinner and a play, and 20% choose shopping as a leisure activity. If 5 people are randomly selected, find the probability that 3 plan to see a movie, 1 to a play, and 1 to a shopping mall.
:::

## Example 5-24

:::{.callout-tip}
## Solution

Here $n = 5$, $X_1 = 3$, $X_2 = 1$, $X_3 = 1$, $p_1 = 0.50$, $p_2 = 0.30$, and $p_3 = 0.20$:

$$P(X) = \frac{5!}{3!\,1!\,1!}(0.50)^3(0.30)^1(0.20)^1 = 0.15$$

:::

```{r e5-24-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 5
X1 <- 3; X2 <- 1; X3 <- 1
p1 <- 0.50; p2 <- 0.30; p3 <- 0.20

prob <- factorial(n) / (factorial(X1)*factorial(X2)*factorial(X3)) *
        p1^X1 * p2^X2 * p3^X3

cat("P(X) = 5!/(3!·1!·1!) × (0.5)³ × (0.3)¹ × (0.2)¹ =", round(prob, 4))
```

## Example 5-25 {#example-5-25}

:::{.callout-warning}
## Coffee Shop Customers

A coffee shop manager found that customers buy 0, 1, 2, or 3 cups of coffee with probabilities 0.3, 0.5, 0.15, and 0.05, respectively. If 8 customers enter, find the probability that 2 buy nothing, 4 buy 1 cup, 1 buys 2 cups, and 1 buys 3 cups.
:::

## Example 5-25

:::{.callout-tip}
## Solution

Here $n = 8$, $X_1 = 2$, $X_2 = 4$, $X_3 = 1$, $X_4 = 1$, with $p_1 = 0.3$, $p_2 = 0.5$, $p_3 = 0.15$, and $p_4 = 0.05$:

$$P(X) = \frac{8!}{2!\,4!\,1!\,1!}(0.3)^2(0.5)^4(0.15)^1(0.05)^1 \approx 0.0354$$

:::

```{r e5-25-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 8
X1 <- 2; X2 <- 4; X3 <- 1; X4 <- 1
p1 <- 0.30; p2 <- 0.50; p3 <- 0.15; p4 <- 0.05

prob <- factorial(n) / (factorial(X1)*factorial(X2)*factorial(X3)*factorial(X4)) *
        p1^X1 * p2^X2 * p3^X3 * p4^X4

cat("n =", n, "\n")
cat("X₁=2 (0 cups), X₂=4 (1 cup), X₃=1 (2 cups), X₄=1 (3 cups)\n")
cat("P(X) =", round(prob, 4))
```

## Example 5-26 {#example-5-26}
:::{.callout-warning}
A box contains 4 white, 3 red, and 3 blue balls. A ball is selected, its color recorded, and replaced. If 5 balls are selected, find the probability that 2 are white, 2 are red, and 1 is blue.
:::

```{r e5-26-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 5
X1 <- 2; X2 <- 2; X3 <- 1
p1 <- 4/10; p2 <- 3/10; p3 <- 3/10

prob <- factorial(n) / (factorial(X1)*factorial(X2)*factorial(X3)) *
        p1^X1 * p2^X2 * p3^X3

cat("p₁ = 4/10 (white), p₂ = 3/10 (red), p₃ = 3/10 (blue)\n")
cat("P(X) = 5!/(2!·2!·1!) × (4/10)² × (3/10)² × (3/10)¹\n")
cat("     =", round(prob, 4), "(= 81/625)")
```

## The Poisson Distribution

:::{.callout-note}
## Definition & Formula

The **Poisson distribution** applies when events occur randomly over a continuous interval of time, area, or volume — especially when $n$ is large and $p$ is small.

$$P(X;\, \lambda) = \frac{e^{-\lambda}\, \lambda^X}{X!}, \quad X = 0, 1, 2, \ldots$$

- $\lambda$ (lambda) = mean number of occurrences per unit interval
- $e \approx 2.71828$ (Euler's number)
- $X$ = number of occurrences in the interval

Also used to **approximate the binomial** when $\lambda = n \cdot p < 5$.
:::

## Example 5-27 {#example-5-27}
:::{.callout-warning}
A 500-page manuscript contains 200 randomly distributed typographical errors. Find the probability that a given page contains exactly 3 errors.
:::

```{r e5-27-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
lambda <- 200 / 500   # mean errors per page
X <- 3
prob <- (exp(-lambda) * lambda^X) / factorial(X)
cat("λ = 200/500 =", lambda, "errors per page\n")
cat("P(X=3; λ=0.4) = e^(-0.4) × (0.4)³ / 3!\n")
cat("              =", round(exp(-lambda), 6), "×", lambda^X, "/", factorial(X),
    "\n              =", round(prob, 4))

# Verification
cat("\nVerification via dpois:", round(dpois(X, lambda), 4))
```

## Example 5-28 {#example-5-28}
:::{.callout-warning}
A sales firm receives an average of 3 calls per hour on its toll-free number. For any given hour, find the probability of:

a. At most 3 calls

b. At least 3 calls

c. 5 or more calls
:::

---

```{r e5-28-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
lambda <- 3

# (a) P(X ≤ 3)
a <- ppois(3, lambda)
cat("(a) P(X ≤ 3) = P(0)+P(1)+P(2)+P(3) =", round(a, 4), "\n")

# (b) P(X ≥ 3) = 1 - P(X ≤ 2)
b <- 1 - ppois(2, lambda)
cat("(b) P(X ≥ 3) = 1 - P(X ≤ 2) =", round(b, 4), "\n")

# (c) P(X ≥ 5) = 1 - P(X ≤ 4)
c <- 1 - ppois(4, lambda)
cat("(c) P(X ≥ 5) = 1 - P(X ≤ 4) =", round(c, 4))
```

## Example 5-29 {#example-5-29}

:::{.callout-warning}
## Poisson Approximation to Binomial

Approximately 2% of people in a room of 200 are left-handed. Find the probability that exactly 5 are left-handed, using the Poisson approximation.
:::
---

```{r e5-29-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
n <- 200; p <- 0.02
lambda <- n * p   # = 4
X <- 5

# Poisson approximation
prob_pois <- dpois(X, lambda)
cat("λ = n·p =", n, "×", p, "=", lambda, "\n")
cat("Poisson approx: P(X=5; λ=4) =", round(prob_pois, 4), "\n")

# Exact binomial
prob_binom <- dbinom(X, n, p)
cat("Exact binomial: P(X=5)      =", round(prob_binom, 4), "\n")
cat("Difference:", round(abs(prob_pois - prob_binom), 4))
```

:::{.callout-important}
The Poisson approximation is reliable when $\lambda = n \cdot p < 5$.
:::

## The Hypergeometric Distribution

:::{.callout-note}
## Definition & Formula

The **hypergeometric distribution** applies when sampling is done **without replacement** from a finite population containing two types of objects.

Given a population with $a$ items of type 1 and $b$ items of type 2 (total $a + b$), the probability of selecting exactly $X$ items of type 1 in a sample of size $n$ is:

$$P(X) = \frac{C^a_X \cdot  C^b_{n-X}}{C^{a+b}_n}$$
:::

**Key difference from binomial:** trials are **dependent** (no replacement).

## Example 5-30 {#example-5-30}

:::{.callout-warning}
## Assistant Manager Applicants

Ten people apply for a restaurant assistant manager position: 5 are college graduates and 5 are not. If the manager selects 3 at random, find the probability that all 3 are college graduates.
:::

## Example 5-30

:::{.callout-tip}
## Solution

Here $a = 5$ college graduates, $b = 5$ nongraduates, $n = 3$, and $X = 3$:

$$P(3) = \frac{C^5_3 \cdot C^5_0}{C^{10}_3} = \frac{10 \cdot 1}{120} = \frac{1}{12} \approx 0.0833$$

:::

```{r e5-30-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
a <- 5   # college graduates
b <- 5   # nongraduates
n <- 3; X <- 3

prob <- choose(a, X) * choose(b, n - X) / choose(a + b, n)
cat("a = 5 (graduates), b = 5 (nongraduates), n = 3, X = 3\n")
cat("P(X=3) = C(5,3)×C(5,0) / C(10,3)\n")
cat("       =", choose(a,X), "×", choose(b,n-X), "/", choose(a+b,n),
    "=", round(prob, 4), "(= 1/12)")
```

## Example 5-31 {#example-5-31}

:::{.callout-warning}
## House Insurance

A study found that 2 out of 10 houses in a neighborhood have no insurance. If 5 houses are selected from 10, find the probability that exactly 1 is uninsured.
:::

## Example 5-31

:::{.callout-tip}
## Solution

Here $a = 2$ uninsured houses, $b = 8$ insured houses, $n = 5$, and $X = 1$:

$$P(1) = \frac{C^2_1 \cdot C^8_4}{C^{10}_5} = \frac{2 \cdot 70}{252} = \frac{140}{252} = \frac{5}{9} \approx 0.5556$$

:::

```{r e5-31-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
a <- 2   # uninsured
b <- 8   # insured
n <- 5; X <- 1

prob <- choose(a, X) * choose(b, n - X) / choose(a + b, n)
cat("a = 2 (uninsured), b = 8 (insured), n = 5, X = 1\n")
cat("P(X=1) = C(2,1)×C(8,4) / C(10,5)\n")
cat("       =", choose(a,X), "×", choose(b,n-X), "/", choose(a+b,n),
    "=", round(prob, 4), "(= 5/9)")
```

## Example 5-32 {#example-5-32}
:::{.callout-warning}
A lot of 12 compressor tanks contains 3 defective ones. If 3 tanks are randomly selected for inspection and the lot is rejected when 1 or more defectives are found, what is the probability the lot will be rejected?
:::

---

```{r e5-32-1, echo=T, warning=FALSE, out.width="80%", fig.align='center'}
a <- 3   # defective
b <- 9   # non-defective
n <- 3; X <- 0   # find P(no defective)

p_none <- choose(a, X) * choose(b, n - X) / choose(a + b, n)
p_reject <- 1 - p_none

cat("a = 3 (defective), b = 9 (good), n = 3\n")
cat("P(X=0) = C(3,0)×C(9,3) / C(12,3) =", choose(a,0),"×",choose(b,3),"/",choose(a+b,n),
    "=", round(p_none, 4), "\n")
cat("P(at least 1 defective) = 1 - P(0) = 1 -", round(p_none, 4),
    "=", round(p_reject, 4))
```

There is a **62%** probability the lot will be rejected when 3 of the 12 tanks are defective.

## Summary of Discrete Distributions

```{r e5-32-2, echo=F, warning=FALSE, out.width="80%", fig.align='center'}
summary_df <- data.frame(
  Distribution = c("Binomial", "Multinomial", "Poisson", "Hypergeometric"),
  Formula = c(
    "$P(X) = C^n_X·p^x·q^{n-x}$",
    "$P(X) = n!/(X_1!·X_2!···X_k!)·p_1^{X_1}···p_k^{X_k}$",
    "$P(X;λ) = e^{-λ}·λ^X/X!$",
    "$P(X) = C^a_X·C^b_{n-X}/C^{a+b}_n$"
  ),
  `Use When` = c(
    "Fixed n trials; 2 outcomes; independent; constant p",
    "Fixed n trials; >2 outcomes; outcomes are independent",
    "n large, p small; events over time/area/volume",
    "Two outcomes; sampling WITHOUT replacement"
  ),
  check.names = FALSE
)
knitr::kable(summary_df, caption = "Table 5-1: Summary of Discrete Distributions")
```

## Key Takeaways


- A **discrete probability distribution** lists all possible values and their probabilities; requires $\sum P(X)=1$ and $0 \leq P(X) \leq 1$.
- **Mean:** $\mu = \sum X \cdot P(X)$; **Variance:** $\sigma^2 = \sum[X^2 \cdot P(X)] - \mu^2$
- **Expected value** $E(X) = \mu$ — the long-run average outcome.
- **Binomial:** $n$ fixed trials, 2 outcomes, independent, constant $p$.
  - $P(X) = C(n,X) \cdot p^X \cdot q^{n-X}$; $\mu = np$; $\sigma = \sqrt{npq}$
- **Multinomial:** extension of binomial for $k > 2$ outcomes.
- **Poisson:** events over time/area; $P(X;\lambda) = e^{-\lambda}\lambda^X / X!$
- **Hypergeometric:** two outcomes, sampling without replacement.

## Acknowledgement
-   **Copyright notice.** These teaching materials have been adapted from **Bluman, A. G. (2012). Elementary statistics: A step by step approach (8th ed.). McGraw-Hill**. All rights are reserved by the original authors and publishers.

-   **Non-commercial use only.** These materials are strictly intended for educational purposes and must not be used for commercial gain or profit.

-   **Proper attribution.** Any reproduction, distribution, or use of these materials must provide proper attribution to the original source.
