## Problem sets

### Statistics (1)

- [In-Class Exercise 1](https://yyliou.github.io/stat/stat1-ps/ex1.html) — **Chapter 1**: The Nature of Probability and Statistics

- [In-Class Exercise 2](https://yyliou.github.io/stat/stat1-ps/ex2.html) — **Chapter 2**: Frequency Distributions and Graphs

- [In-Class Exercise 3](https://yyliou.github.io/stat/stat1-ps/ex3.html) — **Chapter 3**: Measures of Central Tendency

- [In-Class Exercise 4](https://yyliou.github.io/stat/stat1-ps/ex4.html) — **Chapter 3**: Measures of Variation and Position

- [In-Class Exercise 5](https://yyliou.github.io/stat/stat1-ps/ex5.html) — **Chapter 4**: Sample Spaces and the Addition Rules

- [In-Class Exercise 6](https://yyliou.github.io/stat/stat1-ps/ex6.html) — **Chapter 4**: Multiplication Rules, Conditional Probability, and Counting Rules

- [In-Class Exercise 7](https://yyliou.github.io/stat/stat1-ps/ex7.html) — **Chapter 5**: Probability Distributions, Mean, Variance, and Expectation

- [In-Class Exercise 8](https://yyliou.github.io/stat/stat1-ps/ex8.html) — **Chapter 5**: The Binomial and Poisson Distributions

- [In-Class Exercise 9](https://yyliou.github.io/stat/stat1-ps/ex9.html) — **Chapter 6**: The Normal Distribution and Its Applications

- [In-Class Exercise 10](https://yyliou.github.io/stat/stat1-ps/ex10.html) — **Chapter 6**: The Central Limit Theorem and the Normal Approximation

- [In-Class Exercise 11](https://yyliou.github.io/stat/stat1-ps/ex11.html) — **Chapter 7**: Confidence Intervals for the Mean

- [In-Class Exercise 12](https://yyliou.github.io/stat/stat1-ps/ex12.html) — **Chapter 7**: Confidence Intervals for Proportions and Variances, and Sample Size

### Statistics (2)

- [In-Class Exercise 1](https://yyliou.github.io/stat/stat2-ps/ex1.html) — **Chapter 8**: Steps in Hypothesis Testing and the z Test for a Mean

- [In-Class Exercise 2](https://yyliou.github.io/stat/stat2-ps/ex2.html) — **Chapter 8**: The t Test for a Mean

- [In-Class Exercise 3](https://yyliou.github.io/stat/stat2-ps/ex3.html) — **Chapter 8**: Tests for a Proportion and for a Variance

- [In-Class Exercise 4](https://yyliou.github.io/stat/stat2-ps/ex4.html) — **Chapter 9**: Testing the Difference Between Two Means

- [In-Class Exercise 5](https://yyliou.github.io/stat/stat2-ps/ex5.html) — **Chapter 9**: Dependent Samples, Two Proportions, and Two Variances

- [In-Class Exercise 6](https://yyliou.github.io/stat/stat2-ps/ex6.html) — **Chapter 10**: Scatter Plots and Correlation

- [In-Class Exercise 7](https://yyliou.github.io/stat/stat2-ps/ex7.html) — **Chapter 10**: Regression

- [In-Class Exercise 8](https://yyliou.github.io/stat/stat2-ps/ex8.html) — **Chapter 10**: Coefficient of Determination and Standard Error of the Estimate

- [In-Class Exercise 9](https://yyliou.github.io/stat/stat2-ps/ex9.html) — **Chapter 11**: Test for Goodness of Fit

- [In-Class Exercise 10](https://yyliou.github.io/stat/stat2-ps/ex10.html) — **Chapter 11**: Tests Using Contingency Tables

- [In-Class Exercise 11](https://yyliou.github.io/stat/stat2-ps/ex11.html) — **Chapter 12**: One-Way Analysis of Variance

- [In-Class Exercise 12](https://yyliou.github.io/stat/stat2-ps/ex12.html) — **Chapter 12**: The Tukey Test and Two-Way Analysis of Variance

## Submission Requirements

> ⚠️ **All assignments must be submitted using the official template provided below.**
>
> Submissions that are missing any of the following elements will receive a grade of **zero**. This policy is non-negotiable.

### Template

Copy the following template as the starting point for every submission. Replace all placeholder values before submitting.

````{r,eval=F}
---
title: "Ex. X"
author: "Student ID"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
print(uuid::UUIDgenerate())
```

## Question 1

*(Write your response here.)*

## Question 2

*(Write your response here.)*
````

### Checklist

Before submitting, verify that your document satisfies all of the following criteria:

-   [ ] The `title` field in the YAML header is completed.
-   [ ] The `author` field includes your student ID *only*.
-   [ ] The final document must be submitted strictly in `.pdf` format.
-   [ ] The document's name should be `exN-[StudentID].pdf`, e.g. (`ex1-A113290001.pdf`).

## Submit Your Assignments

Upload your PDF and place it in the corresponding folder (e.g., `ex1` for Exercise 1).

- [Statistics (1)](https://u.pcloud.link/publink/show?code=kZNceI5ZF28nybQ9FvyKOJY4QTAHDyOSfKB7)

- [Statistics (2)](https://u.pcloud.link/publink/show?code=kZIknI5Z45y21se2lRmd41s9QJ7Xs4wNxBWX)

## Check your grade

- [Statistics (1)](https://u.pcloud.link/publink/show?code=kZbqdI5ZzY8KVlcj6JyhkIkWK9mRsm207ikX)

- [Statistics (2)](https://u.pcloud.link/publink/show?code=kZizqI5Z8lHjVGsYe1BrxIrQKLNg8hVeuSJX)


