## Installing R and RStudio

### Step 1: Install R

R is the underlying statistical computing environment and must be installed before RStudio.

1.  Navigate to the official CRAN repository: <https://cran.r-project.org>
2.  Select the installer appropriate for your operating system:
    -   **Windows**: Click *Download R for Windows* → *base* → download and execute the installer
    -   **macOS**: Click *Download R for macOS* → select the build matching your architecture (Apple Silicon users should select the `arm64` build)
    -   **Linux**: Follow the distribution-specific installation instructions provided on the page
3.  Run the installer using the default configuration settings.

### Step 2: Install RStudio Desktop

RStudio is an integrated development environment (IDE) for R. It requires a prior R installation to function.

1.  Navigate to: <https://posit.co/download/rstudio-desktop/>
2.  Download the free *RStudio Desktop* release for your operating system
3.  Execute the installer and follow the on-screen instructions
4.  Upon launching RStudio, the application should automatically detect the installed R environment

> **Important**: R and RStudio are distinct pieces of software. RStudio does not bundle R; it serves solely as an interface to an existing R installation.

------------------------------------------------------------------------

## Authoring Documents with R Markdown and Generating PDF Output

R Markdown is a document format that integrates prose, code, and computational output within a single reproducible file. The following steps outline the process of creating an R Markdown document and rendering it as a PDF.

### Step 1: Install Required Packages

Execute the following commands in the RStudio Console:

``` r
install.packages("rmarkdown")
install.packages("knitr")
```

### Step 2: Install a LaTeX Distribution

PDF rendering requires a LaTeX engine. The `tinytex` package provides a lightweight, self-contained LaTeX distribution suitable for most use cases:

``` r
install.packages("tinytex")
tinytex::install_tinytex()
```

**Restart RStudio** after installation to ensure the LaTeX binaries are correctly registered.

### Step 3: Create a New R Markdown File

1.  From the RStudio menu, select: *File* → *New File* → *R Markdown...*
2.  Specify a title and author in the dialogue box; set the default output format to **PDF**
3.  Click **OK** to generate the template file

### Step 4: Configure the YAML Header

Every R Markdown document begins with a YAML front matter block delimited by `---`. This block controls document metadata and output settings:

``` yaml
---
title: "Assignment Title"
author: "Full Name (Student ID)"
date: "2024-01-01"
output: pdf_document
---
```

### Step 5: Write Document Content

R Markdown uses standard Markdown syntax for formatting prose, supplemented by fenced code chunks for embedding R code:

| Syntax                       | Rendered Output             |
|------------------------------|-----------------------------|
| `# Heading 1`                | First-level heading         |
| `## Heading 2`               | Second-level heading        |
| `**bold**`                   | **bold**                    |
| `*italic*`                   | *italic*                    |
| Fenced code block with `{r}` | Executed R code with output |

**Example code chunk:**

```{r,eval=F}
summary(mtcars)
```

```{r,echo=FALSE}
summary(mtcars)
```

### Step 6: Render to PDF

Click the **Knit** button in the RStudio toolbar, or use the keyboard shortcut `Ctrl+Shift+K` (Windows/Linux) or `Cmd+Shift+K` (macOS), and select *Knit to PDF*.

> **Note on CJK character rendering**: If Chinese, Japanese, or Korean characters appear as missing glyphs in the PDF output, specify an alternative LaTeX engine in the YAML header:
>
> ``` yaml
> output:
>   pdf_document:
>     latex_engine: xelatex
> header-includes:
>   - \usepackage{fontspec}
>   - \setmainfont{Noto Sans CJK TC}
> ```

------------------------------------------------------------------------
